### 2024-06-04 doc/_paper/paper.tex
add paper

%%
%% Copyright 2022 OXFORD UNIVERSITY PRESS
%%
%% This file is part of the 'oup-authoring-template Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'oup-authoring-template Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
%% with bibliographic references
%%

%%%CONTEMPORARY%%%
%\documentclass[,webpdf,contemporary,large]{oup-authoring-template}%
%\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}

%%%MODERN%%%
%\documentclass[webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
\documentclass[unnumsec,webpdf,modern,large]{_oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}

%%%TRADITIONAL%%%
%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}

%\onecolumn % for one column layouts

%\usepackage{showframe}
\usepackage{cleveref}

\graphicspath{{Fig/}}

% line numbers
%\usepackage[mathlines, switch]{lineno}
%\usepackage[right]{lineno}

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

\DeclareMathSymbol{\shortminus}{\mathbin}{AMSa}{"39}
\newcommand{\Ex}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\var}{\mathrm{Var}}
\newcommand{\HW}{\mathrm{HW}}
\newcommand{\RI}{\mathrm{RI}}
\newcommand{\Bin}{\text{Binomial}}
\newcommand{\Uniform}{\text{Uniform}}
\newcommand{\Normal}{\mathcal{N}}
\newcommand{\Beta}{\text{Beta}}
\newcommand{\Exp}{\text{Exponential}}
\newcommand{\Gam}{\text{Gamma}}
\newcommand{\Bfun}{\mathrm{B}}
\newcommand{\eps}{\epsilon}
\newcommand{\all}{A}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\erfi}{\mathrm{erfi}}
\newcommand{\fix}{\mathrm{fix}}
\newcommand{\logit}{\mathop{\mathrm{logit}}}
\usepackage{caption}
\captionsetup[figure]{labelfont=bf,font=small}
\captionsetup[ table]{labelfont=bf,font=small}
\captionsetup[longtable]{labelfont=bf,font=small}
\usepackage{float,soul}
\usepackage[normalem]{ulem}
\makeatletter
\def\fps@figure{tb}
\makeatother
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{lipsum}
\usepackage{tikz}

\begin{document}

\journaltitle{Journal Title Here}
\DOI{HERE}
\copyrightyear{2024}
\pubyear{}
\access{}
\appnotes{}

\firstpage{1}

%\subtitle{Subject Section}

\title[Infinitesimal model for polyploids]{
    The infinitesimal model for autopolyploids and mixed-ploidy populations}

\author[1,$\ast$]{Arthur Zwaenepoel}

\authormark{Zwaenepoel}

\address[1]{
    \orgdiv{CNRS}, 
    \orgname{Univ. Lille, UMR 8198 -- Evo-Eco-Paleo}, 
    \orgaddress{\postcode{F-59000}, \state{Lille}, \country{France}}}

\corresp[$\ast$]{
\href{email:arthur.zwaenepoel@univ-lille.fr}{email:arthur.zwaenepoel@univ-lille.fr, }}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

%\editor{Associate Editor: Name}

%\abstract{
%\textbf{Motivation:} .\\
%\textbf{Results:} .\\
%\textbf{Availability:} .\\
%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
%online.}

\abstract{
    We define the infinitesimal model of quantitative genetics (\textit{sensu}
    \cite{barton2017}) for the inheritance of an additive quantitative trait in
    a mixed-ploidy population consisting of diploid, triploid and autotetraploid
    individuals producing haploid and diploid gametes.
    We implement efficient simulation methods and use these to study the
    quantitative genetics of mixed-ploidy populations and the establishment of
    tetraploids after an environmental challenge and in a new habitat.
}
\keywords{Polyploidy, quantitative genetics, infinitesimal model}

% \boxedtext{
% \begin{itemize}
% \item Key boxed text here.
% \item Key boxed text here.
% \item Key boxed text here.
% \end{itemize}}

\maketitle

\section*{Introduction}

\section*{Model \& Methods}

We first develop the infinitesimal model (in the sense of \cite{barton2017},
i.e.~the `Gaussian descendants' approximation \citep{turelli2017}) for an
autotetraploid population. We then consider mixed-ploidy populations in which
diploids, triploids and tetraploids coexist and interbreed through the
production of haploid and diploid gametes that combine freely.

{\subsection{The infinitesimal model}\label{the-infinitesimal-model}}

Consider a population which expresses a quantitative trait determined by a
large number of additive loci of small effect.
The infinitesimal model approximates the inheritance of such a trait by
assuming that the trait value \(Z_{ij}\) of a random offspring from parents
with trait values \(z_i\) and \(z_j\) follows a Gaussian distribution with mean
equal to the midparent value and variance which is independent of the mean:
\begin{align}
  Z_{ij} \sim \Normal\left(\frac{z_i + z_j}{2}, V_{ij}\right) \label{eq:inf}
\end{align} 
Here, \(V_{ij}\) is referred to as the \emph{segregation variance} in family
\((i,j)\).
This is the variation generated among offspring from the same parental pair due
to random Mendelian segregation in meiosis.
This approximation can be justified as arising from the limit where the number
of loci determining the trait tends to infinity \citep{barton2017}.

An equivalent, and for our purposes useful, way to characterize the model in
diploids and polyploids is to write \(Z_{ij} = Y_i + Y_j\), where \(Y_i\) and
\(Y_j\) are independent Gaussian random variables \(Y_i \sim
\Normal\left(\frac{z_i}{2}, V_i\right)\) (and similarly for \(Y_j\)).
We refer to \(Y_i\) as the (random) gametic value of individual \(i\), and to
\(V_i\) as the \emph{gametic segregation variance} of individual \(i\)
(clearly $V_{ij} = V_i + V_j$).
This formulation is helpful in that it highlights that Mendelian segregation
occurs independently in both parents when gametes are produced, which then
combine additively to determine the offspring trait value.

If we assume a population consisting of infinitely many unrelated
individuals, expressing a trait with genetic variance \(V_z\) and
segregation variance \(V_0\), we find that under random mating the
variance in the offspring generation is
\begin{align}
V_z' &= \Ex[\Var[Z_{ij}|Z_i, Z_j]] + \Var[\Ex[Z_{ij}|Z_i,Z_j]] \nonumber \\
   &= \Ex[V_{ij}] + \Var\left(\frac{Z_i + Z_j}{2}\right) \nonumber \\
   &= V_0 + \frac{V_z}{2}\ .
   \label{eq:eq}
\end{align}
So that at equilibrium (\(V_z'=V_z\)), we have \(V_z = 2V_0\) \citep{barton2017}.

Denoting by \(X\) the contribution to the trait value associated with a
randomly sampled haploid genome from the population, we also have that, at
equilibrium, \(V_z=2V_0=mV\), where \(V = \Var[X]\) and $m$ is the ploidy
level.
Note that no assumptions regarding $m$ have been made in the definition of the
model.
Indeed, as long as all individuals are of the same \emph{even} ploidy level,
\cref{eq:inf} and \cref{eq:eq} apply equally well to the inheritance of an
additive trait in polyploids as in diploids.
Two factors however do lead to differences with the standard diploid model due
to their effect on \(V_i\):
(1) the potential occurrence of double reduction in polyploids affects the
relationship between \(V\) and the segregation variance and 
(2) the evolution of the segregation variance over time in finite populations
is affected by the ploidy level.

\textbf{Perhaps we should organize the text around the table below. First
outline mixed-ploidy population, then indicate that we need the variances
associated with the different gametes that are produced. Gamete formation
occurs independently, so that is all we need. So start with the infinitesimal
model framed in terms of uniting gametes, and highlight that this works for any
ploidy level, and mixed-ploidy populations as long as we can work out the
segregation variances.}

\begin{table}[]
\begin{tabular}{l|rr}
cytotype   & haploid gamete variance & diploid gamete variance             \\ \cline{1-3}
diploid    & $\frac{1}{2}(1- F)V$    & $2\alpha_2(1-F)V$  \\
triploid   & $\frac{2}{3}(1-F)V$     & $\frac{2}{3}(1 + \alpha_3)(1-F)V$   \\
tetraploid & --                      & $(1+2\alpha_4)(1-F)V$             
\end{tabular}%
\end{table}

\section*{Results}

\section*{Discussion}

\section*{Competing interests}
No competing interest is declared.

\section*{Author contributions statement}
Not appliccable.

\section*{Data availability statement}

The authors affirm that all data necessary for confirming the conclusions of
the article are present within the article, figures, and supplementary
material. Software implementing the numerical methods and individual-based
simulations is available at \texttt{https://github.com/arzwa/InfGenetics}.

\section*{Acknowledgments}

This work was funded by the European Union (ERC BryoFit 101041201 granted to
CF). Views and opinions expressed are however those of the author(s) only and
do not necessarily reflect those of the European Union or the European Research
Council. Neither the European Union nor the granting authority can be held
responsible for them.

%%%%%%%%%%%%%%

\begin{appendices}

\end{appendices}

%\bibliographystyle{plain}
    %\bibliography{/home/arthur_z/vimwiki/bib.bib}

%USE THE BELOW OPTIONS IN CASE YOU NEED AUTHOR YEAR FORMAT.
\bibliographystyle{abbrvnat}
\bibliography{/home/arthur_z/vimwiki/bib.bib}

\clearpage
\setcounter{page}{1}

%SUPPLEMENT
\clearpage
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thealgorithm}{S\arabic{algorithm}}
\setcounter{figure}{0}
\setcounter{section}{0}
\setcounter{equation}{0}
\setcounter{algorithm}{0}

\onecolumn % for one column layouts

\section{Supplementary figures}

\section{Supplementary Information}

\subsection*{Double reduction in autotetraploids}

When an autopolyploid forms multivalents during prophase I, a form of `internal
inbreeding' may occur as a result of the phenomenon called \emph{double
reduction} \citep{huang2019}.
Double reduction happens when, as a result of recombination, replicated gene
copies on sister chromatids move to the same pole during anaphase I (see
\cite{stift2010} or \cite{huang2019} for helpful illustrations).
The frequency of double reduction at a locus in the presence of multivalent
formation is determined by the frequency at which that locus is involved in a
cross-over (which depends on the distance to the centromere), and, in
tetraploids, has an upper bound at \(1/6\) \citep{mather1935,stift2010}.

Consider a locus with genotype $ABCD$ in an autotetraploid.
In the absence of double reduction, six distinct gametes are produced with
equal frequencies.
When double reduction happens, four additional types of gametes carrying
replicated gene copies ($AA, BB, CC$ or $DD$ gametes) will be produced with
equal frequencies.
The segregation variance will hence be increased in the presence of double
reduction, as more distinct types of gametes are produced compared to disomic
inheritance.
For a random genotype \(X_1X_2X_3X_4\), we can find the gametic segregation
variance contributed by a locus when double reduction happens as
\begin{align*}
    \Ex[\var[Y&|X_1,X_2,X_3,X_4]] = \var[Y] - \Var[\Ex[Y|X_1,X_2,X_3,X_4]] \\
        &= \var[2X] - \var\left[\frac1 4 (2X_1 + 2X_2 + 2X_3 + 2X_4)\right] \\
        &= 4v - \frac{1}{4}4v
        = 3v
\end{align*} where \(X\) denotes the additive effect of a random allele
at the locus drawn from the reference population and \(v = \Var[X]\).
In the absence of double reduction we have
\begin{align*}
    \Ex[\var[Y&|X_1,X_2,X_3,X_4]] \\
        &= 2\Var[X] - \var\left[\frac1 6 \sum_{i=1}^3 \sum_{j=i+1}^4(X_i+X_j)\right]
        = v
\end{align*}
Assuming that the probability of double reduction at any locus is \(\alpha\)
(also \(\alpha_4\) below), and summing over independent loci, we find that the
gametic segregation variance in the presence of double reduction should be
\begin{equation}
\frac{V_0}{2} = (1-\alpha)V + 3\alpha V = V(1+2\alpha).
\label{eq:dr}
\end{equation}

\end{document}

### 2024-06-04 doc/draft.tex
add paper

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{model-and-methods}{%
\section{Model and Methods}\label{model-and-methods}}

We first develop the infinitesimal model (in the sense of
{[}@barton2017{]}, i.e.~the `Gaussian descendants' approximation
{[}@turelli2017{]}) for an autotetraploid population. We then consider
mixed-ploidy populations in which diploids, triploids and tetraploids
coexist and interbreed through the production of haploid and diploid
gametes that combine freely.

\hypertarget{the-infinitesimal-model}{%
\subsection{The infinitesimal model}\label{the-infinitesimal-model}}

Consider a population which expresses a quantitative trait determined by
a large number of additive loci of small effect. The infinitesimal model
approximates the inheritance of such a trait by assuming that the trait
value \(Z_{ij}\) of a random offspring from parents with trait values
\(z_i\) and \(z_j\) follows a Gaussian distribution with mean equal to
the midparent value and variance which is independent of the mean:
\begin{align}
  Z_{ij} \sim \Normal\left(\frac{z_i + z_j}{2}, V_{ij}\right)
  \label{eq:inf}
  \end{align} Here, \(V_{ij}\) is referred to as the \emph{segregation
variance} in family \((i,j)\). This is the variation generated among
offspring from the same parental pair due to random Mendelian
segregation in meiosis. This approximation can be justified as arising
from the limit where the number of loci determining the trait tends to
infinity {[}@barton2017{]}. An alternative, and for our purposes useful,
way to characterize the model in diploids and polyploids is to write
\(Z_{ij} = Y_i + Y_j\), where \(Y_i\) and \(Y_j\) are independent
Gaussian random variables
\(Y_i \sim \Normal\left(\frac{z_i}{2}, V_i\right)\) (and similarly for
\(Y_j\)). We refer to \(Y_i\) as the (random) gametic value of
individual \(i\), and to \(V_i\) as the \emph{gametic segregation
variance} of individual \(i\). This formulation is helpful in that it
highlights that Mendelian segregation occurs independently in both
parents when gametes are produced, which then combine additively to
determine the offspring trait value.

If we assume a population consisting of infinitely many unrelated
individuals, expressing a trait with genetic variance \(V\) and
segregation variance \(V_0\), we find that under random mating the
variance in the offspring generation is \begin{align}
V' &= \Ex[\Var[Z_{ij}|Z_i, Z_j]] + \Var[\Ex[Z_{ij}|Z_i,Z_j]] \nonumber \\
   &= \Ex[V_{ij}] + \Var\left(\frac{Z_i + Z_j}{2}\right) \nonumber \\
   &= V_0 + \frac{V}{2}\ .
   \label{eq:eq}
\end{align} So that at equilibrium (\(V'=V\)), we have \(V = 2V_0\)
{[}@barton2017{]}. If we denote by \(X\) the contribution to the trait
value associated with a randomly sampled haploid genome from the
population, we also have that, at equilibrium, \(V=2V_0=mV_x\), where
\(V_x = \Var[X]\). Note that no assumptions regarding ploidy level have
been made in the definition of the model. Indeed, as long as all
individuals are of the same \emph{even} ploidy level, eq.~\ref{eq:inf}
and eq.~\ref{eq:eq} apply equally well to the inheritance of an additive
trait in polyploids as in diploids. Two factors however do lead to
differences with the standard diploid model due to their effect on
\(V_i\): (1) the potential occurrence of double reduction in polyploids
affects the relationship between \(V_x\) and \(V_0\) and (2) the
evolution of the segregation variance over time in finite populations is
affected by the ploidy level.

\hypertarget{the-effect-of-double-reduction-on-the-segregation-variance}{%
\subsection{The effect of double reduction on the segregation
variance}\label{the-effect-of-double-reduction-on-the-segregation-variance}}

When an autotetraploid forms quadrivalents during prophase I, a form of
`internal inbreeding' may occur as a result of the phenomenon called
\emph{double reduction} {[}@lynch1998 p.~57{]}. Double reduction happens
when, as a result of recombination, replicated gene copies on sister
chromatids move to the same pole during anaphase I, as illustrated in
fig.~\ref{fig:dr}. In the example shown in fig.~\ref{fig:dr}, one of the
four generated gametes is \(AA\), which would not occur in the ordinary
bivalent meiosis, because in that case, paired chromosomes (involved in
cross-overs) are separated during anaphase I. The frequency of double
reduction at a locus in the presence of multivalent formation is hence
determined by the frequency at which that locus is involved in a
cross-over (which depends on the distance to the centromere), and has an
upper bound at \(1/6\) {[}@stift2008{]}.

When double reduction occurs, an \(ABCD\) genotype would generate 10
distinct gametes, as opposed to 6 in when chromosomes form bivalents. As
a result, the segregation variance is increased by double reduction. 
For
a random genotype \(X_1X_2X_3X_4\), we can find the gametic segregation
variance contributed by a locus when double reduction happens as
\begin{align*}
    \Ex[\var[Y|X_1,X_2,X_3,X_4]] &= \var[Y] - \Var[\Ex[Y|X_1,X_2,X_3,X_4]] \\
        &= \var[2X] - \var\left[\frac1 4 (2X_1 + 2X_2 + 2X_3 + 2X_4)\right] \\
        &= 4v_x - \frac{1}{4}4v_x 
        = 3v_x
\end{align*} where \(X\) denotes the additive effect of a random allele
at the locus drawn from the reference population and \(v_x = \Var[X]\).
In the absence of double reduction we have \begin{align*}
    \Ex[\var[Y|X_1,X_2,X_3,X_4]] 
        &= 2\Var[X] - \var\left[\frac1 6 \sum_{i=1}^3 \sum_{j=i+1}^4(X_i+X_j)\right]
        = v_x
\end{align*} Assuming that the probability of double reduction at any
locus is \(\alpha\) (also \(\alpha_4\) below), and summing over
independent loci, we find that the gametic segregation variance in the
presence of double reduction should be \begin{equation}
\frac{V_0}{2} = (1-\alpha)V_x + 3\alpha V_x = V_x(1+2\alpha).
\label{eq:dr}
\end{equation}

\begin{figure}
\begin{center}
\begin{tikzpicture}
\footnotesize
\node (n1) at (0, 5) [] {$A\ B\ C\ D$};
\node (n2) at (0, 4) [] {$A\underline{A}\ \underline{B}\overline{B}\ CC\ \overline{D}D$};
\node (n3) at (0, 3) [] {$AB\ AD | CC\ BD$};
\node (n4) at (-2, 2) [] {$AB\ AD$};
\node (n5) at ( 2, 2) [] {$CC\ BD$};
\node (n6) at (-3, 1) [] {$AA$};
\node (n7) at (-1, 1) [] {$BD$};
\node (n8) at ( 3, 1) [] {$CB$};
\node (n9) at ( 1, 1) [] {$CD$};
\draw[->] (n1) -- (n2) node [right,midway] {\small replication};
\draw[->] (n2) -- (n3) node [right,midway] {\small recombination};
\draw[->] (n3) -- (n4);
\draw[->] (n3) -- (n5) node [right,midway] {\small \ \ meiosis I};
\draw[->] (n5) -- (n8) node [right,midway] {\small \ \ meiosis II};
\draw[->] (n5) -- (n9);
\draw[->] (n4) -- (n6);
\draw[->] (n4) -- (n7);
\end{tikzpicture}
\end{center}
\begin{caption}{
Schematic illustration of a meiotic division in an autotetraploid leading to
double reduction at a locus with genotype $ABCD$.
Two recombination events are assumed to occur at the locus (denoted by the
bars).
\label{fig:dr}
}\end{caption}
\end{figure}

\hypertarget{inbreeding-in-autotetraploids}{%
\subsection{Inbreeding in
autotetraploids}\label{inbreeding-in-autotetraploids}}

In a finite population the segregation variance will decrease over time
due to inbreeding. Indeed, in a diploid individual for instance, when
the two gene copies at a locus are identical, segregation will not
generate any variation. The rate of inbreeding depends on the ploidy
level {[}@moody1993; @arnold2012{]}, and hence the rate at which the
segregation variance declines \emph{does} so too. Let \(F_i\) the
inbreeding coefficient of individual \(i\), defined as the probability
that two \emph{distinct} gene copies at a locus in individual \(i\) are
identical by descent (IBD) with respect to some reference population,
and let \(\Phi_{ij}\) be the coancestry coefficient for individuals
\(i\) and \(j\), defined as the probability that two genes sampled
independently from \(i\) and \(j\) at a given locus are IBD. In
diploids, one can show that the gametic segregation variance \(V_i\) is
decreased by a factor \(1-F_i\) relative to the segregation variance in
the reference population \(V_0/2\) {[}@barton2017{]}. In polyploids,
however, \(F_i\) does not suffice to describe the state of homozygosity
in individual \(i\). In tetraploids, for instance, we have five distinct
homozygosity states, which we can symbolically represent as
\(abcd, aabc, aabb, aaab\) and \(aaaa\) (in general, the number of
homozygosity states grows according to the partition function
\((1,2,3,5,7,11,15,22,\dots)\)). Representing the probability of being
in these five increasingly homozygous states as
\(\delta_1, \dots, \delta_5\), we find that the gametic segregation
variance is reduced by a factor
\[\phi = \delta_1 + \left(1 - \frac1 6\right) \delta_2 + \left(1 - \frac1
        3\right)\delta_3 + \left(1 - \frac1 2\right) \delta_4\] which is
precisely \(1-F_i\), as in diploids (see also {[}@moody1993{]}). This
shows that we do not need to track the array of homozygosity
coefficients in order to compute the segregation variance in a
tetraploid family, but only require the inbreeding coefficients of the
parents. This is a consequence of the fact that, in tetraploids, gametes
are diploid. For higher ploidy levels, one would need to track higher
order identity coefficients. We shall hence not consider higher ploidy
levels in this paper.

In order to efficiently conduct individual-based simulations under the
infinitesimal model, we hence need to be able to track the \(F_i\) and
\(\Phi_{ij}\). Denoting the parents of individual \(i\) by \(k\) and
\(l\), the recursion for the inbreeding coefficients is \begin{align}
    F_i &= \Phi_{kl}& \text{in diploids} \nonumber \\  
    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{in tetraploids} 
    \label{eq:recursion}
\end{align} where \(F_k^\ast = \alpha + (1-\alpha)F_k\). The recursion
for the tetraploid case follows from considering three cases: either (1)
the two genes sampled in individual \(i\) both came from the gamete
contributed by parent \(k\), which happens with probability \(1/6\), in
which case they are IBD with probability \(F_k^\ast\); or (2) as in (1)
but from parent \(l\); or (3) with probability \(2/3\) the two genes
came from different gametes, in which case they are IBD with probability
\(\Phi_{kl}\). The recursion for the coancestry coefficients in
\(m\)-ploids is given by \begin{align}
    \Phi_{ii} &= \frac1 m \left(1 + (m-1)F_i\right) \nonumber \\
    \Phi_{ij} &= \sum_k \sum_l P_{ik}P_{jl} \Phi_{kl} & i \ne j 
    \label{eq:coancestry}
\end{align} where the sums are over individuals in the parental
population, and where \(P_{ik} \in \{0, \frac1 2, 1\}\) is the
probability that a gene copy in \(i\) is derived from parent \(k\). When
dealing with discrete generations, \(P_{ik}\) values can be conveniently
represented in a \(N(t) \times N(t-1)\) matrix, where \(N(t)\) is the
population size in generation \(t\), so that
\(\Phi(t) = P \Phi(t-1) P^T\), where \(\Phi(t)\) is the matrix of
coancestry coefficients in generation \(t\) {[}@barton2017{]}.

\hypertarget{the-infinitesimal-model-in-mixed-ploidy-populations}{%
\subsection{The infinitesimal model in mixed-ploidy
populations}\label{the-infinitesimal-model-in-mixed-ploidy-populations}}

When considering mixed-ploidy populations, we need to consider how
equilibrium variances in the model scale with ploidy level, and how
offspring of different ploidy levels are derived from a parental pair.
Throughout the following paragraphs, we specialize to a
diploid-triploid-tetraploid complex. A glossary of the relevant notation
can be found in tbl.~\ref{tbl:glossary}.

\hypertarget{scaling-of-the-genetic-variance-across-ploidy-levels}{%
\subsubsection{Scaling of the genetic variance across ploidy
levels}\label{scaling-of-the-genetic-variance-across-ploidy-levels}}

The equilibrium phenotypic variance \(V\) under the infinitesimal model
was derived above as \(2V_0\), where \(V_0\) is the segregation variance
(eq.~\ref{eq:eq}). As noted below eq.~\ref{eq:eq}, this entails that for
\(m\)-ploids, \(2V_0 = mV_{x,m}\), where \(V_{x,m}\) is the variance of
the additive effect of a randomly sampled haploid genome from the
reference \(m\)-ploid population. Now, in a mixed-ploidy system, it is
not immediately clear how the \(V_{x,m}\) should scale across ploidy
levels, as this depends on the effects of polyploidization \emph{per se}
on additive allelic effects {[}@porturas2019; @clo2021; @clo2022{]}. If
we assume equal equilibrium variances (and concomitantly segregation
variances), the genetic variance for a haploid genome in tetraploids
will be half that of diploids, entailing a reduction in the additive
allelic effect per locus as ploidy level rises. On the other hand, if we
assume equal allelic effects, and hence that the genetic variance
associated with a hypothetical haploid genome copy is identical across
ploidy levels, then the segregation and equilibrium variances in
tetraploids will be twice those of diploids, and the associated
phenotypic range will be doubled as well.

To ease interpretation, let us consider an underlying Mendelian system,
consisting of \(n\) unlinked additive bi-allelic loci in Hardy-Weinberg
and linkage equilibrium (HWLE). Assuming that the additive effects in
tetraploids are homogeneously scaled by a factor \(\beta\) relative to
those in diploids, and writing \(V_{z,m}\), \(V_{0,m}\) and \(V_{x,m}\)
for respectively the equilibrium genetic variance, the segregation
variance and the genetic variance associated with a haploid genome in
\(m\)-ploids, the following relationships hold: \[
 \frac{V_{z,4}}{V_{z,2}} = \frac{2V_{0,4}}{2V_{0,2}} = \frac{4V_{x,4}}{2V_{x,2}} =
 \frac{2 \sum_i^n (\beta a_i)^2 p_i(1-p_i)}{\sum_i^n a_i^2 p_i(1-p_i)} =
 2\beta^2
\] So we see that assuming equal equilibrium phenotypic variances
\(V_{z,4}/V_{z,2} = 1\) entails that allelic effects are scaled by a
factor \(\beta=\sqrt{1/2}\). On the other hand, assuming equal allelic
effects (\(\beta=1\)) entails that the phenotypic variance at
equilibrium in tetraploids is twice that of diploids, as we noted above.
To study the effect of such assumptions, we introduce \(\beta_m\) as a
parameter so that \(V_{z,m} = \frac{m}{2}\beta_m^2 V_{z,2}\), keeping in
mind its interpretation as a scaler of allelic effects in the underlying
Mendelian system, and setting \(\beta_2=1\). The trait value
distribution for offspring of ploidy level \(m_{ij}\) from parental pair
\((i,j)\) with ploidy levels \(m_i\) and \(m_j\), formed through the
union of a \(g_i\)-ploid gamete from parent \(i\) and a \(g_j\)-ploid
gamete from parent \(j\) will then be \[Z_{ij} \sim \Normal\left(
    \beta_{m_{ij}} \left(
          \frac{g_i}{m_i}\frac{z_i}{\beta_{m_i}} 
        + \frac{g_j}{m_j}\frac{z_j}{\beta_{m_j}}\right), V_{i,g_i} + V_{j,g_j}\right)\]
Where \(V_{i,g_i}\) is the gametic segregation variance associated with
the formation of a \(g_i\)-ploid gamete by individual \(i\). We now
proceed to characterize the \(V_{i,g_i}\) and their evolution over time.

\hypertarget{unreduced-gamete-formation-in-diploids}{%
\subsubsection{Unreduced gamete formation in
diploids}\label{unreduced-gamete-formation-in-diploids}}

\begin{figure}
\begin{center}
\begin{tikzpicture}
\footnotesize
\node (rep) at (-6, 4.5) [] {\small Replication};
\node (rep) at (-6, 3.5) [] {\small Recombination};
\node (rep) at (-6, 2.5) [] {\small First division};
\node (rep) at (-6, 1.5) [] {\small Second division};
\node (rep) at (-6, 0)   [] {\small Gametes};
\node (Aa)  at (0, 5) [] {$AB$};
\node (AAaa)  at (0, 4) [] {$AA\ BB$};
\node (rAaAa) at (-2, 3) [] {$AB\ AB$};
\node (rAAaa) at ( 2, 3) [] {$AA\ BB$};
\path (Aa) edge (AAaa);
\path (AAaa) edge node[near end, above] {$c$} (rAaAa) ;
\path (AAaa) edge node[near end, above] {\qquad $1-c$} (rAAaa);

\node (fdr11) at (-3, 2) [] {$AB\ AB | \emptyset$};
\node (fdr12) at (-3, 1) [] {$A|B\  A|B$}; 
\node (AA) at (-3.8,0) [] {$AA$};
\node (p1) at (-3.8,-0.5) [] {$\frac{1}{4}$};
\node (Aa) at (-3,0) [] {$AB$};
\node (p2) at (-3,-0.5) [] {$\frac{1}{2}$};
\node (aa) at (-2.2,0) [] {$BB$};
\node (p2) at (-2.2,-0.5) [] {$\frac{1}{4}$};
\path (fdr12) edge (AA);
\path (fdr12) edge (Aa);
\path (fdr12) edge (aa);
\path (fdr11) edge (fdr12);
\path (rAaAa) edge node[left] {$f$} (fdr11);

\node (sdr11) at (-1, 2) [] {$AB | AB$}; 
\node (sdr12) at (-1, 1) [] {$AB | \emptyset$}; 
\node (Aa2) at (-1,0) [] {$AB$};
\path (sdr12) edge (Aa2);
\path (sdr11) edge (sdr12);
\path (rAaAa) edge node[right] {$1-f$} (sdr11);

\node (fdr21) at (1,2) [] {$AA\ BB|\emptyset$};
\node (fdr22) at (1,1) [] {$A|A\ B|B$};
\node (Aa3) at (1,0) [] {$AB$};
\path (rAAaa) edge node[left] {$f$} (fdr21);
\path (fdr21) edge (fdr22);
\path (fdr22) edge (Aa3);

\node (sdr21) at (3, 2) [] {$AA | BB$};
\node (sdr22) at (3, 1) [] {$AA| \emptyset, BB|\emptyset$};
\node (AA4) at (2.2,0) [] {$AA$};
\node (p1) at (2.2, -0.5) [] {$\frac{1}{2}$};
\node (aa4) at (3.8,0) [] {$BB$};
\node (p2) at (3.8, -0.5) [] {$\frac{1}{2}$};
\path (rAAaa) edge node[right] {$1-f$} (sdr21);
\path (sdr21) edge (sdr22);
\path (sdr22) edge (AA4);
\path (sdr22) edge (aa4);
\end{tikzpicture}
\begin{caption}{
Schematic representation of the different pathways for unreduced gamete
formation in diploids and their different outcomes.
\label{fig:udip}
}\end{caption}
\end{center}
\end{figure}

The mechanisms of unreduced gamete formation do not necessarily lead to
a faithful transmission of the complete diploid genome. Unreduced
gametes are formed in two ways, depending on the meiotic abberation that
leads to their origin: (1) first division restitution (FDR) of (2)
second division restitution (SDR) {[}@bretagnolle1995{]}. Consider a
locus in a diploid with two distinct genes \(A\) and \(a\). Assume
recombination happens with probability \(c\) and that conditional on
unreduced gamete formation, formation is due to FDR with probability
\(f\) while it is due to SDR with probability \(1-f\). The different
unreduced gametes that are formed are represented schematically in
fig.~\ref{fig:udip}. Writing the genotype at a locus in the diploid
parent as \(X_1X_2\), with allelic effects \(X_1\) and \(X_2\), the
genotypic value of an unreduced gamete at this locus will be
\[Y = \begin{cases}
    2X_1 & \text{w.p. } \alpha_2/2 = \frac1 4 f c + \frac1 2 (1-f)(1-c) \\
    2X_2 & \text{w.p. } \alpha_2/2 \\
    X_1 + X_2 & \text{w.p. } 1-\alpha_2 
    \end{cases}\] where \(\alpha_2 = 1 - f - c + \frac{3}{2}cf\) is the
probability that two copies of the same gene end up in a diploid gamete
produced by a diploid individual. Conditional on the latter event, we
get, by a similar reasoning as the one leading eq.~\ref{eq:dr}, a
segregation variance equal to \begin{align*}
\Ex[\Var[Y|X_1,X_2]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2]] \\
    &= \Var[2X] - \Var\left[\frac{1}{2}(2X_1 + 2X_2)\right] \\
    &= 2v_x
\end{align*} Conditioning on the complementary event, all gametes have
genetic value \(Y = X_1+X_2\), so that the segregation variance is 0.
Summing across independent loci, we have
\(V_{i,2} = 2\alpha_2(1-F_i)V_x\) for a diploid individual \(i\).

\hypertarget{meiosis-in-triploids}{%
\subsubsection{Meiosis in triploids}\label{meiosis-in-triploids}}

Triploids, when viable, may be important for the dynamics of
mixed-ploidy populations due to the formation of a so-called triploid
bridge. The formation of triploids presents no immediate issues, we
simply need to track the segregation variance contributions from both
donor gametes, and relate these to \(V_{0,3}\). Sexual reproduction in
triploids is however more complicated. There are no known mechanisms to
coordinate the assortment of chromosomes in for instance a haploid and
diploid gamete, and meiosis, if it happens, usually results in aneuploid
gametes {[}@ramsey1998{]}. Experimental results indicate that, at least
in yeast, triploids usually form trivalents and undergo recombination,
after which each trivalent is randomly assorted in the daughter cells,
some receiving one, others two copies of a given chromosome
{[}@charles2010{]}. In the absence of gametic nonreduction, the
probability of obtaining euploid gametes (two diploid and two haploid
gametes) from such a process is \((1/2)^n\), where \(n\) is the number
of chromosomes. If the number of chromosomes is small this is not
negligible, for instance in \emph{A. thaliana} we would have
\((1/2)^5 \approx 0.03\), which is of the same order as the unreduced
gamete formation rate. Unreduced (triploid) gametes may also be produced
and important for the dynamics of mixed-ploidy populations
{[}@ramsey1998{]}. However, they generate additional difficulty, since
in order to compute the contributed variance under inbreeding, we would
need an additional identity coefficient recording the probability that
three genes are IBD at a locus. We will hence ignore the possibility of
unreduced gamete production in triploids. Denoting by \(\alpha_3\) the
probability that a diploid gamete produced by a triploid individual
contains two copies of the same parental gene, we can use the same
approach as above to derive \(V_{i,1}\) and \(V_{i,2}\) for a triploid
individual \(i\). We note that, on the supposition that diploid gametes
are produced by random assortment of chromosomes in a haploid and
diploid gametes, \(\alpha_3 \le 1/4\). The formulae for the different
\(V_{i,g_i}\) are summarized in tbl.~\ref{tbl:segvar}.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2857}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3571}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3571}}@{}}
\caption{Gametic segregation variance for different cytotype \(\times\)
gametic cytotype combinations. \(V_x\) refers to genetic variance
associated with a haploid genome in the reference population of the
cytotype to which the gamete will contribute. Under the assumptions
outlined in the main text, we have the following constraints for the
\(\alpha_k\): \(\alpha_2 \le 1\), \(\alpha_3 \le 1/4\) and
\(\alpha_4 \le 1/6\). \label{tbl:segvar}}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{cytotype}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\(V_{i,1}\) (haploid)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(V_{i,2}\) (diploid)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{cytotype}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\(V_{i,1}\) (haploid)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(V_{i,2}\) (diploid)
\end{minipage} \\
\midrule()
\endhead
diploid \((m_i = 2)\) & \(\frac{1}{2}(1-F_i)V_x\) &
\(2\alpha_2 (1-F_i)V_x\) \\
triploid \((m_i = 3)\) & \(\frac{2}{3}(1-F_i)V_x\) &
\(\frac{2}{3}(1 + 3\alpha_3)(1-F_i)V_x\) \\
tetraploid \((m_i = 4)\) & -- & \((1+2\alpha_4)(1-F_i)V_x\) \\
\bottomrule()
\end{longtable}

\hypertarget{recursions-for-inbreeding-coefficients}{%
\subsubsection{Recursions for inbreeding
coefficients}\label{recursions-for-inbreeding-coefficients}}

In a mixed-ploidy system tracking inbreeding coefficients becomes only
slightly more complicated. Recall that \(\alpha_m\) is the probability
that a diploid gamete from an \(m\)-ploid individual transmits two
copies of the same gene at some locus. Let \(m_i\) denote the ploidy
level of individual \(i\). We still have the recursion shown in
eq.~\ref{eq:recursion}, where now
\[F_k^\ast = F_k(1-\alpha_{m_k}) + \alpha_{m_k}\] For a triploid
individual, where the parent contributing the \(2n\) gamete is \(k\), we
have \[ F_i = \frac1 3 \left(F_k^\ast + 2\Phi_{kl}\right) \] For
\(\Phi\) the recursion in eq.~\ref{eq:coancestry} remains valid, but
where diagonal elements are now given by
\[\Phi_{ii} = \frac{1}{m_i} \left(1 + (m_i-1)F_i\right)\] and where the
elements of the the pedigree matrix \(P\) now have entries in
\(\{0,\frac1 3,\frac1 2,\frac2 3,1\}\).

\hypertarget{population-dynamics}{%
\subsubsection{Population dynamics}\label{population-dynamics}}

\emph{What comes here depends on the results we will include.}

The dynamics of the mixed-ploidy system (in the absence of selection on
the trait) are determined by the rates at which different types of
gametes are formed. We assume an individual of ploidy level \(k\) forms
haploid and diploid gametes with proportions \(u_{k1}\) and \(u_{k2}\),
so that the (relative) fecundity of a \(k\)-ploid individual is
\(u_{k1} + u_{k2}\). Unless stated otherwise, we will assume \[U = 
    \begin{pmatrix} 
    u_{21} & u_{22} \\ 
    u_{31} & u_{32} \\ 
    u_{41} & u_{42} 
    \end{pmatrix} =
    \begin{pmatrix} 
    1-u & u \\
    v & v \\
    0 & 1-u
    \end{pmatrix}\] where \(u\) is referred to as the proportion of
unreduced gametes, and \(2v\) is the proportion of euploid gametes
produced by a triploid individual. An analysis of a deterministic model
for the cytotype dynamics {[}also analyzed in@felber1991; @felber1997{]}
is included in sec.~\ref{sec:det}.

\hypertarget{tbl:glossary}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.0870}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.9130}}@{}}
\caption{\label{tbl:glossary}Glossary for the notation used to define
the mixed-ploidy infinitesimal model.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Symbol
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Explanation
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Symbol
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Explanation
\end{minipage} \\
\midrule()
\endhead
\(z_i\) & Trait value of individual \(i\) \\
\(Z_{ij}\) & Trait value of a random offspring from parental pair
\((i,j)\) \\
\(V_{z,m}\) & Equilibrium variance in an \(m\)-ploid population \\
\(V_{0,m}\) & Segregation variance in the \(m\)-ploid reference
population \\
\(V_{x,m}\) & Equilibrium variance of a haploid genome in the
\(m\)-ploid reference population \\
\(V_{i,m}\) & Gametic segregation variance in individual \(i\) \\
\(F_i\) & Inbreeding coefficient of individual \(i\) \\
\(\Phi_{ij}\) & Coancestry coefficient of individuals \(i\) and \(j\) \\
\(m_i\) & Ploidy level of individual \(i\) \\
\(\alpha_m\) & Probability that a diploid gamete from an \(m\)-ploid
individual contains two copies of the same parental gene \\
\(\beta_m\) & Scaling factor for the allelic effects in ploidy level
\(m\) \\
\(u_{m,g}\) & Proportion of \(g\)-ploid gametes produced by an
\(m\)-ploid individual \\
\(u\) & Proportion of unreduced gametes in even-ploids \\
\(v\) & Proportion of haploid and diploid gametes in triploids \\
\bottomrule()
\end{longtable}

\hypertarget{results}{%
\section{Results}\label{results}}

We first verify that, in the absence of inbreeding, our implementation
of the mixed-ploidy infinitesimal model yields the predicted equilibrium
variance for each cytotype (fig.~\ref{fig:z}) and the expected
equilibrium proportions (fig.~\ref{fig:eqp}). We next evaluate the
accuracy of the infinitesimal model as an approximation to the evolution
of a quantitative trait determined by \(L\) additive loci. We find that
the infinitesimal model with inbreeding generally yields accurate
predictions for the evolution of the genetic variance, even as the
number of loci becomes fairly small (e.g.~\(L=50\), fig.~\ref{fig:vz}).
For \(L=20\), substantial deviations from the infinitesimal
approximation can be observed, although it may still be considered
reasonable (fig.~\ref{fig:vz}).

Although inbreeding is slower in autotetraploids than in diploids for
the same population size, the tetraploid fraction of a diploid-dominated
mixed-ploidy population will have an equal or higher average inbreeding
coefficient (fig.~\ref{fig:f1}). This is because in such a population,
triploid and tetraploid individuals mostly arise from gametes formed by
diploid individuals or polyploid individuals with very recent diploid
ancestry (about \(1+u+2v\) generations ago for tetraploids, and
\(1+\frac{2}{3}(u+2v)\) generations ago for triploids, see
sec.~\ref{sec:ttdip}), so that the polyploid subpopulations will show an
average relatedness similar to the diploid population and not evolve as
an isolated higher-ploidy population would. A nonzero probability of
producing IBD diploid gametes (\(\alpha\)) will then further increase
the inbreeding coefficient in the tetraploid and triploid fraction of
the population relative to their diploid progenitors
(fig.~\ref{fig:f1}). Therefore, as long as diploids dominate, the effect
of harboring some of the gene pool in triploid and tetraploid
individuals on the rate of inbreeding is negligible, and we find that
the evolution of the inbreeding coefficient over time is well predicted
by \(1-e^{-t/2N_e}\), where the inbreeding-effective population size is,
to first order in \(u\), given by \((1-2u)N\) (sec.~\ref{sec:effsize},
fig.~\ref{fig:fNe}).

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\begin{itemize}
\item
  It appears that many papers on diploid -- tetraploid contact zones or
  mixed-ploidy populations fail to find significant numbers of
  triploids. Note though that such studies focus on established
  tetraploids, where the cytotypes may have diverged to some extent,
  whereas we focus on the case where a tetraploid cytotype has not yet
  established. The model is therefore more relevant for populations that
  are likely to be referred to as diploid. It remains to be seen to what
  extent such populations actually exhibit a mixed-ploidy equilibrium.
  See {[}@suda2013{]} and {[}@herben2016{]}.
\item
  The issue with dominance. On the one hand, it is often said that
  polyploids accumulate a larger mutation load, and will hence show
  inbreeding depression, but this really only applies to already
  established polyploid populations. Similarly, the sort of effects of
  dominancestudied in Booker and Schrider during range expansions apply
  mostly to polyploids that have accumulated a large mutation load over
  a rather long time as a stable polyploid population. In our case,
  polyploids are virtually always derived very recently from diploid
  ancestors, so they will not have accumulated more deleterious
  recessives than their diploid counterparts. As inbreeding will be
  slower in tetraploids, we should hence have reduced inbreeding
  expression, as we have the same load as in diploids, but it is less
  expressed (at least when considering recessives). There should
  therefore be an immediate benefit to polyploidy (reduced inbreeding
  depression), which is however a long term disadvantagem because
  eventually the established polyploid will start to accumulate a higher
  load due to less efficient purging.
\end{itemize}

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\end{CSLReferences}

\clearpage
\setcounter{page}{1}

\clearpage
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thealgorithm}{S\arabic{algorithm}}
\setcounter{figure}{0}
\setcounter{section}{0}
\setcounter{equation}{0}
\setcounter{algorithm}{0}

\hypertarget{supplementary-figures}{%
\section{Supplementary figures}\label{supplementary-figures}}

\clearpage

\begin{figure}
\hypertarget{fig:z}{%
\centering
\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-10/mixedploidy-eq.svg}
\caption{Equilibrium trait value (\(z\)) distribution in a mixed-ploidy
population evolving according to the infinitesimal model without
inbreeding for three different scaling assumptions of allelic effects
across ploidy levels. Histograms (lines) show simulation results,
whereas the filled Gaussian densities show the expected distribution at
equilibrium \(\Normal\left(0, k\beta_k^2V_x\right)\). In the
\(\beta=(1,1,1)\) case (left), allelic effects remain constant across
ploidy levels, causing the equilibrium variance to be doubled in
tetraploids relative to diploids. In the \(\beta=(1, 3/4, 1/2)\) case
(middle), allelic effects are scaled so that the phenotypic range is
held constant (i.e.~the maximum and mimimum phenotypes are teh same
across ploidy levels). In the \(\beta=(1, \sqrt{2/3}, \sqrt{1/2})\) case
(right), allelic effects are scaled so that the variance in trait values
at equilibrium is equal across ploidy levels.}\label{fig:z}
}
\end{figure}

\begin{figure}
\hypertarget{fig:eqp}{%
\centering
\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-10/mixedploidy-eqf.svg}
\caption{Equilibrium proportions of the different cytotypes in a
mixed-ploidy population where \(u=v=0.05\), light lines show the
observed proportions in a simulation of a Wright-Fisher mixed-ploidy
infinitesimal model with \(N=500\). The horizontal lines show the
deterministic predictions
(\(\tilde{g}_1^2, 2\tilde{g}_1\tilde{g}_2, \tilde{g}_2^2)\).}\label{fig:eqp}
}
\end{figure}

\begin{figure}
\hypertarget{fig:vz}{%
\centering
\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-Vz.svg}
\caption{Comparison of the mixed-ploidy infinitesimal model with the
\(L\)-locus model, for \(L=500,100,50\) and \(20\). The decline in the
genetic variance \(V_z\) within each cytotype due to drift is shown. The
transparent lines show the complete simulation, whereas the solid line
shows the same data but smoothed in overlapping windows of 20
generations. We assume \(N=500, u=v=0.08\) and no selection. In the top
row where \(\beta_k^2=1, \alpha_k=0\), the equilibrium variance in the
absence of inbreeding in triploids is 2/3 that of diploids, and in
tetraploids it is twice that in diploids. In the middle row,
\(\beta_3^2=2/3\) and \(\beta_4^2=1/2\), so that the equilibrium
variance in the absence of inbreeding is equal across cytotypes. In the
bottom row, \(\alpha_2= \alpha_3 =1/2\) and \(\alpha_4=1/6\), causing an
immediate increase in the genetic variance in higher cytotypes, but also
accelerated inbreeding.}\label{fig:vz}
}
\end{figure}

\begin{figure}
\hypertarget{fig:vz2}{%
\centering
\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-Vz2.svg}
\caption{As in fig.~\ref{fig:vz} (\(L \rightarrow \infty\) simulations)
but highlighting the effect of \(\beta\) and \(\alpha\) within ploidy
levels. Only smoothed data is shown.}\label{fig:vz2}
}
\end{figure}

\begin{figure}
\hypertarget{fig:f1}{%
\centering
\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-F.svg}
\caption{Average inbreeding coefficient \(\bar{F}\) in each cytotype in
a mixed-ploidy population for different values of \(\alpha\) (we assume
\(\alpha_k=\alpha\), where \(\alpha_k\) is the probability that a
diploid gamete produced by a \(k\)-ploid cytotype contains two copies of
the same parental gene at a random locus). We assume \(N=500, u=v=0.08\)
and no selection.}\label{fig:f1}
}
\end{figure}

\clearpage

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

\hypertarget{sec:det}{%
\subsection{Deterministic mixed-ploidy model}\label{sec:det}}

\begin{figure}
\centering
\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-13/g1diff.svg}
\caption{Deterministic mixed-ploidy equilibrium. The left plot shows the
stable (solid lines) and unstable (dashed lines) equilibria for the
proportion of haploid gametes in the gamete pool \(g_1\) as a function
of \(u\) for different values of \(v\). The middle plot shows the
relationship between \(\Delta g_1 = g_1' - g_1\) and \(g_1\). The zeros
of this graph are the fixed points of the dynamical system and are
indicated by the hollow (unstable equilibrium) and solid (stable
equilibrium) dots. The rightmost plot shows the region of parameter
space where a stable mixed-ploidy equilibrium exist. \label{fig:g1diff}}
\end{figure}

Let \(g_k\) be the frequency of \(k\)-ploid gametes in the gamete pool,
and let us consider only haploid and diploid gametes, so that
\(g_2 = 1-g_1\). Diploids produce unreduced gametes with probability
\(u\) and reduced ones with probability \(1-u\), triploids produce
haploid and triploid gametes both with probability \(v\), and
tetraploids produce reduced diploid gametes with probability \((1-u)\)
(we assume they produce, just like diploids, a proportion \(u\) of
unreduced gametes, but these are assumed not to lead to viable offspring
and are ignored). We get after one generation of random mating
\[g_1' = \frac{(1-u)g_1^2 + 2vg_1g_2}{g_1^2 + 4vg_1g_2 + (1-u)g_2^2}.\]
We see that \(g_1=0\) is always an equilibrium (no haploid gametes,
tetraploids take over). Two more fixed points are obtained at
\begin{equation}
  \tilde{g_1}, \tilde{g_1}' =
    \frac{3 - 3u - 6v \pm \sqrt{(u + 2v - 1)(5u + 2v - 1)}}
         {2(2 - u - 4v)} \label{eq:gameteq}
  \end{equation} Of which the larger one, when it exists, corresponds to
a stable equilibrium, and the smaller one to an unstable equilibrium. As
there are no viability differences, the equilibrium cytotype frequencies
can be readily obtained from these through the relations \begin{align}
\pi_2 = \tilde{g_1}^2 & & \pi_3 = 2\tilde{g_1}\tilde{g_2} & & 
    \pi_4 = \tilde{g_2}^2
\label{eq:cyteq}
\end{align} Assuming \(v=O(u)\), we have to second order in \(u\)
\begin{align}
 \pi_2 &= 1 - 2u - 4uv - u^2 + O(u^3)  \nonumber \\
 \pi_3 &= 2u + 4uv + O(u^3) \nonumber \\
 \pi_4 &= u^2 + O(u^3) 
\end{align} At the critical point where the stable equilibrium
disappears, we have that \(\Delta g_1 = \frac{d\Delta g_1}{d g_1} = 0\)
(fig.~\ref{fig:g1diff}, middle). We find that, in the region of
parameter space that is biologically relevant (roughly
\(u < 0.1, v < 0.1\), say), the critical unreduced gamete formation rate
\(u_c\) beyond which tetraploids take over can be expressed as a linear
function of triploid fertility (\(2v\)): \[u_c = \frac{1}{5}(1-2v)\]
(fig.~\ref{fig:g1diff}, right). This shows that, for plausible parameter
values, we can safely assume that an initially diploid population will
evolve to a mixed-ploidy equilibrium.

\hypertarget{sec:ttdip}{%
\subsection{Expected time to diploid ancestry}\label{sec:ttdip}}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{/home/arthur_z/vimwiki/build/img/2023-10-22/timetodiploidancestord.svg}
\caption{Expected time to diploid ancestry. The solid lines show
\(\Ex[T_4]\), i.e.~the expected time since being inherited from a
diploid ancestor for a random gene in a tetraploid individual at
equilibrium, for different values of \(v\) (half the triploid
fertility). The dashed lines show \(\Ex[T_3]\), i.e.~the same quantity
for a gene sampled from a triploid. Note that \(\Ex[T]\) blows up
whenever \(u\) and \(v\) exceed their critical value for tetraploid
establishment. \label{fig:ttdip}}
\end{figure}

Consider a gene sampled from a tetraploid individual in a mixed-ploidy
population at equilibrium and not subjected to selection. Let \(T_4\)
denote the number of generations in the past until such a gene is found
in a diploid ancestor, and let \(T_3\) denote a similar random variable
for a randomly sampled gene from a triploid in the same population.
Assuming the different cytotypes are at their deterministic equilibrium
frequencies \(\pi_2, \pi_3\) and \(\pi_4\) (see sec.~\ref{sec:det},
eq.~\ref{eq:cyteq}), we have the recursive relations \begin{align}
  \Ex[T_4] &= \frac{1}{Z_2}\Big(\pi_2u + (1+\Ex[T_3])\pi_3 v +
    (1+\Ex[T_4])\pi_4(1-u)\Big) \nonumber \\
  \Ex[T_3] &= \frac{1}{3Z_1}\Big(\pi_2(1-u) + (1+\Ex[T_3])\pi_3 v\Big)
  \nonumber
    \\ &\qquad + \frac{2}{3Z_2}\Big(\pi_2 u + (1+\Ex[T_3])\pi_3 v +
      (1+\Ex[T_4])\pi_4(1-u)\Big) \label{eq:ttdip}
\end{align} where \begin{align*}
  Z_1 &= \pi_2(1-u) + \pi_3v \\
  Z_2 &= \pi_2u + \pi_3v + \pi_4(1-u)
\end{align*} (these expressions are straightforwardly modified when more
general \(u_{ij}\) are assumed, see e.g. sec.~\ref{sec:effsize},
eq.~\ref{eq:markovch}). The system in eq.~\ref{eq:ttdip} can be solved
to yield expressions for \(\Ex[T_4]\) and \(\Ex[T_3]\), which are
however rather unwieldy. Again assuming \(v=O(u)\), we obtain to first
order in \(u\) \begin{align*}
\Ex[T_4] &= 1 + u + 2v + O(u^2) \\
\Ex[T_3] &= 1 + \frac{2}{3}(u + 2v) + O(u^2) 
\end{align*} Numerical examples are shown in (fig.~\ref{fig:ttdip}).
Clearly, for plausible parameter values, \(\Ex[T]\) will be very close
to 1. For instance, for \(u=0.05\) and \(v=0.05\) (which are already
rather large values for these parameters), we would have
\(\Ex[T_3] \approx 1.13\) and \(\Ex[T_4] \approx 1.19\).

\hypertarget{sec:effsize}{%
\subsection{Effective population size of a mixed-ploidy
deme}\label{sec:effsize}}

\begin{figure}
\hypertarget{fig:fNe}{%
\centering
\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-13/mixedploidy-F-Ne.svg}
\caption{The evolution of \(\bar{F}\) in the mixed-ploidy population and
in the diploid subpopulation are shown for different values of \(u\) and
associated values of \(N\), keeping \(N_e = (1-2u)N\) constant at 200.
We assume \(u=v\). All lines coincide almost completely and are
indistinguishable from \(1-e^{-t/2N_e}\). Results are shown for
\(\alpha_k=1/6\). As \(\alpha_k\) decreases to 0, \(\bar{F}\) in the
mixed-ploidy population becomes completely indistinguishable from
\(\bar{F}\) in the diploid subpopulation.}\label{fig:fNe}
}
\end{figure}

We use the approach outlined in {[}@rousset2004{]} (pp.~153, 157) to
determine the effective size of a randomly mating mixed-ploidy
population. Denote by \(\nu_k(t)\) the probability that the ancestral
lineage of a given gene in the present is found in a individual of
ploidy level \(k\) \(t\) generations in the past, and let
\(\nu(t) = (\nu_2(t)\ \nu_3(t)\ \nu_4(t))\) be the corresponding row
vector. Assuming the population is at cytotype equilibrium
(eq.~\ref{eq:cyteq}), we have \begin{align}
      \nu(t+1) &= \nu(t) P \nonumber \\
               &= \nu(t) \begin{pmatrix}
      \frac{u_{21}}{Z_1}\pi_2 & \frac{u_{31}}{Z_1}\pi_3 & 0 \\
      \left(\frac{u_{21}}{3Z_1} + \frac{2u_{22}}{3Z_2}\right)\pi_2 &
          \left(\frac{u{31}}{3Z_1} + \frac{2u_{32}}{3Z_2}\right)\pi_3 & 
          \frac{2u_{42}}{3Z_2}\pi_4 \\
      \frac{u_{22}}{Z_2}\pi_2 & \frac{u_{32}}{Z_2}\pi_3 & \frac{u_{42}}{Z_2}\pi_4
      \label{eq:markovch}
       \end{pmatrix}
    \end{align} where we assume, as usual, that tetraploids do not
produce haploid gametes (\(u_{41} = 0\)), and where \begin{align*}
  Z_1 &= u_{21} \pi_2+ u_{31} \pi_3 \\
  Z_2 &= u_{22} \pi_2 + u_{32} \pi_3 + u_{42} \pi_4
\end{align*} At stationarity,
\(\lim_{t\rightarrow \infty} \nu(t) = \nu\), and we have
\(\nu = \nu P\). Hence, the probability that the ancestral lineage of a
given gene in the present is found in an individual of ploidy level
\(k\) in an indefinite past is given by \(\nu_k\), where \(\nu\) is the
left eigenvector of \(P\) associated with the unit eigenvalue. The
effective size of a mixed-ploidy population of size \(N\) can then be
obtained as \[N_e = N\left(\sum_k \frac{\nu_k^2}{\pi_k}\right)^{-1}\]
After plugging in \(\pi\) in accordance with eq.~\ref{eq:cyteq} and
solving the eigenvalue problem, this yields an unwieldy expression in
the \(u_{ij}\). For our usual parameterization where
\(u_{21} = u_{42} = 1-u, u_{22} = u\) and \(u_{31} = u_{32} = v\), and
\(v = O(u)\), we can find that
\[\begin{pmatrix}\nu_2 \\ \nu_3 \\ \nu_4\end{pmatrix} = \begin{pmatrix}
    1-2uv + O(u^3) \\ 2uv + O(u^3) \\ O(u^3)\end{pmatrix}\] and
\[\frac{N_e}{N} = 1 - 2u + O(u^2)\] which yields an excellent fit in
simulations for plausible parameter values (fig.~\ref{fig:fNe}). When
\(v=0\) and \(u < u_c\) (see sec.~\ref{sec:det}), \(N_e = \pi_2N\), as
in that case (i.e.~when triploids are infertile) there can be no gene
flow from tetraploids to diploids. Since we assume the cytotype
composition to be constant, and polyploids are continually formed from
diploids, no gene in a triploid or tetraploid will have any descendants
in the distant future in this case, so that the effective size is just
the diploid fraction of the population.

### 2024-06-04 doc/paper/paper.tex
add paper

\documentclass[11pt,a4paper]{article}
\input{preamble.tex}
\usepackage{geometry}
\usepackage{authblk}
\externaldocument[s-]{supp}

\begin{document}
\title{Autopolyploid establishment through polygenic adaptation}
\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@univ-lille.fr}}
\affil[1]{\footnotesize University of Lille, CNRS, UMR 8198 -- Evo-Eco-Paleo,
F-59000 Lille, France}
\date{\vspace{-5ex}}
\maketitle
\begin{abstract}
We define the infinitesimal model of quantitative genetics (\textit{sensu}
\cite{barton2017}) for the inheritance of an additive quantitative trait in a
mixed-ploidy population consisting of diploid, triploid and autotetraploid
individuals producing haploid and diploid gametes.
We implement efficient simulation methods and use these to study the
quantitative genetics of mixed-ploidy populations and the establishment of
autotetraploids in a new habitat.
\end{abstract}

\section*{Introduction}

Many plant species exhibit ploidy variation
\citep{levin2002,soltis2007,rice2015}, and many of these \textit{mixed-ploidy}
species have populations in which different cytotypes coexist or form contact
zones \citep{kolar2017}.
How such mixed-ploidy populations emerge and are maintained has proven somewhat
challenging to understand.

Consider for instance a randomly mating diploid population.
Under the commonly accepted view that polyploids mostly emerge through the
union of unreduced gametes \citep{herben2016,kreiner2017b}, a new tetraploid
individual originating by a chance encounter of two unreduced diploid gametes
(an event occurring at an appreciable rate, \citep{kreiner2017}) is highly
unlikely to establish a stable tetraploid subpopulation, as most of its gametes
will end up in unfit hybrids of odd ploidy level (triploids).
This negative frequency dependence effect is commonly referred to as
\textit{minority cytotype exclusion} (MCE) after \cite{levin1975}.
It is well-appreciated that, due to MCE, in a large random mating mixed-ploidy
population dominated by diploids, the rate of unreduced gamete formation needs
to be extraordinarily high for tetraploids to establish (\cite{felber1997}, see
also \cref{s-sec:det}).

Hence, to explain the widespread occurrence of mixed-ploidy populations,
additional factors besides the continuous formation of polyploids through the
union of unreduced gametes need to be considered.
On the one hand, chance establishment of tetraploids through drift could occur.
However, MCE is quite strong in randomly mating mixed-ploidy populations, and
the population size has to be very small for tetraploid establishment to occur
at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
On the other hand, any form of prezygotic isolation between cytotypes could
promote establishment of polyploid cytotypes by alleviating MCE.
Particularly relevant are assortative mating by cytotype  (for instance through
phenological differences across cytotypes, or differences in pollinators,
\citep{kolar2017}), self-fertilization \citep{rausch2005}, and localized
dispersal \citep{baack2005,kolar2017}.
Finally, selection may be invoked to explain the establishment of polyploids.
Tetraploids may have higher relative fitness than their diploid counterparts
due to reduced inbreeding depression \citep{ronfort1999}, or due to being
better adapted to (changing) environmental conditions \citep{vandepeer2021}.

None of these factors is likely to explain by itself the establishment of
polyploids, and the consensus in the field appears to be that some mix of the
above is required to explain the occurrence of mixed-ploidy populations in
nature \citep{kolar2017,mortier2024}.
In particular, polyploids are thought to establish mainly in novel, unoccupied
habitats where they evade MCE (for instance at range edges, or after local
extinction due to environmental change).
If they are able to colonize such habitats at an appreciable rate, they must
somehow be better adapted to local conditions, or more able to adapt to those
conditions despite inbreeding and migration from the source population while
the population is still small, compared to diploids.
Indeed, when the source population is dominated by diploids, the number of
tetraploid migrants arriving in the novel habitat will be almost negligible
($O(u^2)$ if $u$ is the probability that a diploid produces an unreduced
gamete in meiosis) compared to the number of diploid migrant individuals, so
that tetraploids need a substantial advantage during colonization if they are
to establish with reasonable probability.

While there have been substantial modeling efforts aimed at understanding
autotetraploid establishment within diploid populations (e.g. \cite{levin1975,
felber1991, felber1997, rausch2005, oswald2011, clo2022c, griswold2021}), the
problem of polyploid establishment in a novel habitat presenting some adaptive
challenge remains largely unaddressed, despite its centrality to verbal
arguments about the establishment of polyploids in natural populations
\citep{kolar2017, vandepeer2021, clo2022d}.

Here we develop a model for the establishment of a mixed-ploidy population in a
novel, unoccupied habitat based on \cite{barton2018}.
In order to establish in the novel habitat, the population has to adapt to
local environmental conditions.
We assume fitness is determined by directional selection on a single polygenic
trait which can be interpreted as log fitness at low density in the new
habitat.
As in \cite{barton2018}, we assume the trait follows the infinitesimal
model (\textit{sensu} \cite{barton2017}, i.e. the `Gaussian descendants'
infinitesimal model).
We extend the infinitesimal model, and the approach for exact simulation of
trait evolution under the infinitesimal model, to mixed-ploidy populations.
We then use simulations to study tetraploid establishment, both from single
migrants and under continuous migration from a predominantly diploid source
population, examining the effects of polyploid (quantitative) genetics,
maladaptive migration, selfing and assortative mating on the probability that
autotetraploids establish in the novel habitat before diploids do.

\section*{Model and Methods}

\begin{table}[t]
\caption{Glossary of the main notation used in the main text.
} \label{tbl:glossary}
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\cline{1-2}
\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
$N, N_k$ & total population size, population size of the $k$-ploid cytotype \\
$\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
$u$ & probability that an unreduced gamete is formed in diploids
($u=u_{22}=u_{32}=1-u_{42}$)\\
$v$ & probability that a triploid produces a haploid/diploid gamete
($v=u_{31}=u_{32}$)\\
$z_i$ & trait value of individual $i$ \\
$w_{ij}$ & fitness of parental pair $(i,j)$ \\
$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
$c_i$ & ploidy level of individual $i$ \\
$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
$\gamma$ & strength of directional selection \\
$F_i$ & inbreeding coefficient in individual $i$ \\
$V_{i,k}$ & gametic segregation variance associated with the production of a
  $k$-ploid gamete by individual $i$ \\
$V$ & segregation variance in the reference diploid population \\
  $\alpha_k$ & probability that the two genes at a locus in a diploid gamete
  formed by a $k$-ploid individual descend from the same parental gene copy\\
$\sigma_k$ & selfing rate in $k$-ploids \\
$\rho_k$ & probability of assortative mating in $k$-ploids \\
$m$ & expected number of migrants per generation arriving in the new habitat \\
\cline{1-2}
\end{tabularx}%
\end{table}

\subsection*{Mixed-ploidy population model}

We consider a mixed-ploidy population of size $N$ consisting of $N_2$ diploid,
$N_3$ triploid and $N_4=N-N_2-N_3$ tetraploid individuals.
We assume an individual of ploidy level $k$ forms haploid and diploid gametes
with proportions $u_{k1}$ and $u_{k2}$, as well as a proportion
$1-u_{k1}-u_{k2}$ inviable (e.g. aneuploid or polyploid) gametes.
The (relative) fecundity of a $k$-ploid individual is hence $u_{k1} + u_{k2}$.
Unless stated otherwise, we will assume 
\begin{equation}
    \begin{pmatrix} 
    u_{21} & u_{22} \\ 
    u_{31} & u_{32} \\ 
    u_{41} & u_{42} 
    \end{pmatrix} =
    \begin{pmatrix} 
    1-u & u \\
    v & v \\
    0 & 1-u
    \end{pmatrix} \label{eq:U}
\end{equation}
where $u$ is referred to as the proportion of unreduced gametes, and $2v$ is
the proportion of euploid gametes produced by a triploid individual.

When two individuals mate, we assume they produce gametes according to their
ploidy level (\cref{eq:U}), which randomly combine to produce offspring (which
may be inviable if one of the contributing gametes is inviable).
Intrinsic fitness disadvantages associated with particular zygotic ploidy
levels or cross types (e.g. modeling phenomena such as `triploid block',
\citep{ramsey1998,brown2024}) can be straighforwardly included at this level.
An analysis of a deterministic model (i.e. where $N \rightarrow \infty$) for
the cytotype dynamics and equilibrium cytotype composition under random mating
is included in \cref{s-sec:det} (see also \cite{felber1997,kauai2024}).
The stochastic version for finite and constant $N$ is analyzed briefly in
\cref{s-sec:mchain}.

\subsection*{Infinitesimal model}

\paragraph*{The basic infinitesimal model.}

Consider a population which expresses a quantitative trait determined by a
large number of additive loci of small effect.
The infinitesimal model approximates the inheritance of such a trait by
assuming that the trait value $Z_{ij}$ of a random offspring from parents with
trait values $z_i$ and $z_j$ follows a Gaussian distribution with mean equal to
the midparent value and variance which is independent of the mean:
  \begin{align}
  Z_{ij} \sim \Normal\left(\frac{z_i + z_j}{2}, V_{ij}\right)
  \label{eq:inf}
  \end{align}
Here, $V_{ij}$ is referred to as the \textit{segregation variance} in family
$(i,j)$.
This is the variation generated among offspring from the same parental pair due
to random Mendelian segregation in meiosis.
This approximation can be justified as arising from the limit where the number
of loci determining the trait tends to infinity \citep{barton2017}.

An alternative, and for our purposes useful, way to characterize the model
is to write $Z_{ij} = Y_i + Y_j$, where $Y_i$ and $Y_j$
are independent Gaussian random variables $Y_i \sim \Normal\left(\frac{z_i}{2},
V_i\right)$ (and similarly for $Y_j$).
We refer to $Y_i$ as the (random) \textit{gametic value} of individual $i$, and
to $V_i$ as the \textit{gametic segregation variance} of individual $i$.
This formulation is helpful in that it highlights that Mendelian segregation
occurs independently in both parents when gametes are produced, which then
combine additively to determine the offspring trait value.
This model applies readily to an autopolyploid population expressing a trait
with infinitesimal genetics.
For instance, when an autotetraploid produces a diploid gamete, it will pass on
half its trait value to the gamete on average, with a variance determined by
the details of tetraploid meiosis (which are not, in general, identical to
those of diploid meiosis, see below).

Note that in a finite population, the segregation variance will decay over time
as the population becomes more inbred.
Indeed, Mendelian segregation generates less variation among the gametes
produced by individual $i$ when that individual is more inbred, as segregation
at homozygous loci does not generate any variation.
When $F_i$ is the inbreeding coefficient relative to some ancestral reference
population with gametic segregation variance $V$ (i.e. the probability that two
genes at a locus in individual $i$ sampled without replacement are identical by
descent), the gametic segregation variance of individual $i$ will be
$V_i = (1-F_i)V$.
This holds for both diploids and tetraploids (\cref{s-sec:tetinbred}, also
\cite{moody1993}).

\paragraph{Mixed-ploidy infinitesimal model.}

\begin{table}[t]
\caption{Gametic segregation variance for haploid and diploid gametes produced
by the three cytotypes in the mixed-ploidy model. $F_i$ is the inbreeding
coefficient of individual $i$ (producing the gamete), whereas $\alpha_k$ is the
probability that a diploid gamete from a $k$-ploid individual receives two
copies of the same parental gene. Note that we have $\alpha_3 \le 1/4$ and
$\alpha_4 \le 1/6$.
} \label{tbl:segvar}
\centering
\small
\begin{tabularx}{\linewidth}{XXX}
\cline{1-3}
\textbf{cytotype}   & \textbf{haploid gamete variance} & \textbf{diploid gamete
variance}        \\ \cline{1-3} \\[-2.5ex]
diploid    & $\frac{1}{2}(1-F_i)V$    & $2\alpha_2(1-F_i)V$  \\ \\[-2.5ex]
triploid   & $\frac{2}{3}(1-F_i)V$     & $\frac{2}{3}(1 + 3\alpha_3)(1-F_i)V$ \\ \\[-2.5ex]
tetraploid & --                      & $(1+2\alpha_4)(1-F_i)V$             \\
\cline{1-3}
\end{tabularx}%
\end{table}

We can extend this model to the mixed-ploidy case, where we assume that the
gametic value associated with a $k$-ploid gamete ($k \in \{1,2\}$) from
individual $i$ of ploidy level $c_i \in \{2,3,4\}$ is a Gaussian random
variable with distribution
  $$Y_{i,k} \sim \Normal\left(\frac{k}{c_i}z_i, V_{i,k}\right)$$
where $V_{i,k}$ is the gametic segregation variance associated with the
production of a $k$-ploid gamete by individual $i$.
For diploids producing diploid unreduced gametes and triploid producing euploid
gametes, the latter depends not only on the segregation variance in the base
population ($V$) and the inbreeding coefficient ($F$), but also on the detailed
assumptions on how these aberrant meiotic processes occur.
Importantly however, when only haploid and diploid gametes are assumed to occur,
these details only affect the gametic segregation variance through the quantity
$\alpha_k$, which is the probability that a $k$-ploid transmits two copies of 
the same homolog to a diploid gamete.

Note that $\alpha_4$, the probability that a diploid gamete of a tetraploid
individual carries two copies of the same homolog, is the probability of
\textit{double reduction} (e.g. \cite{lynch1998} p.57), and is upper bounded by
1/6 \citep{stift2008}.
The value of $\alpha_2$ depends on the relative frequency of unreduced gamete
formation through so-called \textit{first} and \textit{second division
restitution} \citep{storme2013}.
We summarize the expressions for the gametic segregation variance in
\cref{tbl:segvar}.
Detailed derivations can be found in \cref{s-sec:segvar}.

If we would somewhat naively assume $Z_{ij} = Y_{i,k} + Y_{j,l}$ for a
$(k+l)$-ploid offspring from parental pair $(i,j)$, as we did in the standard
infinitesimal model (\cref{eq:inf}), 
a tetraploid offspring from a cross between two diploids will yield, on average,
a trait value which is the sum of the parental trait values (and hence twice
the midparent value).
This is not likely to reflect biological reality: tetraploids do not tend to
have, for instance, twice the size of their diploid progenitors on average.
Similarly, the genetic variance at Hardy-Weinberg and linkage equilibrium
(HWLE) in tetraploids will be twice that of their diploid counterparts under
such assumptions, which is similarly unrealistic \citep{clo2022}.

In order to account for this, we introduce a scaling factor $\beta_k$,
accounting for the effects of polyploidization \textit{per se} on trait expression.
Specifically, we assume that $\beta_k$ determines how the variance
$\tilde{V}_{z,k}$ at Hardy-Weinberg and linkage equilibrium (HWLE) in
$k$-ploids is related to that in their diploid counterparts:
\begin{equation}
  \frac{\tilde{V}_{z,k}}{\tilde{V}_{z,2}} = \frac{k}{2} \beta_{k}^2
\end{equation}
In the discrete locus model, $\beta_k$ is interpreted as a scaler of additive
effects in the cytotype of ploidy level $k$ (see \cref{s-sec:scaling} for
details).
Under these assumptions, and in the infinitesimal limit, the trait value of a
$k$-ploid descendant from parental pair $i$ and $j$, where parent $i$
contributes a gamete of ploidy level $g_i$ and likewise parent $j$ a gamete of
ploidy level $g_j$, is distributed according to a Gaussian:
\begin{equation}
  Z_{ij} \sim \Normal\left(
    \beta_{k} \left(
          \frac{g_i}{c_i}\frac{z_i}{\beta_{c_i}} 
        + \frac{g_j}{c_j}\frac{z_j}{\beta_{c_j}}\right), 
    \beta_{k}^2 (V_{i,g_i} + V_{j,g_j})\right)
   \label{eq:oneline}
\end{equation}
where the values of $V_{i,g_{i}}$ are as in \cref{tbl:segvar} (see
\cref{s-sec:scaling}).

\paragraph{Recursions for inbreeding coefficients}

We can simulate the mixed-ploidy infinitesimal model for a finite population
through a straightforward extension of the approach outlined in
\cite{barton2017}, provided we can efficiently track inbreeding and coancestry
coefficients across the different ploidy levels.
Denoting the parents of individual $i$ by $k$ and $l$, the recursion for the
inbreeding coefficients in the mixed-ploidy case becomes
\begin{align}
    F_i &= \Phi_{kl} & \text{if } & c_i = 2 \\ 
    F_i &= \frac{1}{3}\left(F_k^\ast + 2\Phi_{kl}\right) & \text{if } 
        & c_i = 3, g_k = 2, g_l = 1 \\ 
    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{if } & c_i = 4
\end{align}
where $F_k^\ast = \alpha_{c_k} + (1-\alpha_{c_k})F_k$ (\cref{s-sec:tetinbred}).
The recursion for the coancestry coefficients in is given by
\begin{align}
    \Phi_{ii} &= \frac{1}{c_{i}} \left(1 + (c_i-1)F_i\right) \nonumber \\
    \Phi_{ij} &= \sum_k \sum_l P_{ik}P_{jl} \Phi_{kl} & i \ne j 
    \label{eq:coancestry}
\end{align}
where the sums are over individuals in the parental population, and where
$P_{ik} \in \{0, \frac1 3, \frac1 2, \frac2 3, 1\}$ is the probability that a
gene copy in $i$ is derived from parent $k$.

\subsection*{Establishment model}

Our model for the establishment of a population in an initially unoccupied
habitat is based on \cite{barton2018}.
We assume a large non-inbred `mainland' population at HWLE and cytotype
equilibrium, with $\Ex[z] = 0$ irrespective of the cytotype.
In generation $t$, $M(t)$ migrant individuals arrive on an island (the new
habitat) joining $N^\ast(t)$ resident individuals, where $M(t)$ is Poisson
distributed with mean $m$. 
The migrant individuals are assumed to be unrelated to the resident
individuals.
After migration in generation $t$, the $N(t) = N^\ast(t) + M(t)$ individuals
reproduce sexually, and the offspring thus produced survives until the next
generation with a probability determined by their trait value.
In the basic model, random selfing is allowed (but see below for a model with
self-incompatibility).
We assume the trait is under directional selection, with fitness is $w(z) =
e^{\gamma(z - \theta)}$, where $\gamma$ is the intensity of directional
selection and $\theta$ is the trait value for which the growth rate of the
population becomes positive.

Again following \cite{barton2018}, we simulate the model by first calculating
the fitness of each parental pair $(i,j)$, which is the expected fitness of
offspring of this pair
\begin{equation}
  w_{ij}
    = \sum_{k=1}^2\sum_{l=1}^2 w_{ij}^{kl}
    = \sum_{k=1}^2\sum_{l=1}^2 u_{c_{i},k}u_{c_{j},l}
        \Ex\left[e^{\gamma(Z_{ij} - \theta)}|g_i=k,g_j=l\right]
\end{equation}
The expectation on the right hand side can be calculated from \cref{eq:oneline}
using the moment-generating function of the Gaussian.
Having calculated the $w_{ij}$, the number of offspring surviving into the next
generation is calculated as $N^\ast(t+1) = \sum_{i,j}w_{ij}/N(t)$.
Next, $N^\ast(t+1)$ offspring individuals are sampled by sampling parental
pairs and gametes proportional to $w_{ij}^{kl}$, and sampling a trait value
in accordance with \cref{eq:oneline}.

\subsection*{Self-fertilization and assortative mating}

We model partial self-fertilization by assuming that a proportion
$\sigma_{c_i}$ of the ovules of individual $i$ with ploidy level $c_i$ are
fertilized by self-pollen, while the remaining proportion $1-\sigma_{c_i}$ are
fertilized by randomly sampled pollen (which may be self-pollen with
probability $1/N$). 
That is, the expected number of offspring from individual $i$ as mother
surviving after selection is
\begin{equation}
\Ex[w_i] = \sigma_{c_i} \Ex[w_{ii}] +
  (1-\sigma_{c_i})\left[\frac{1}{N}\sum_{j=1}^N \Ex[w_{ij}]\right]
\end{equation}
Where $\Ex[w_{ij}] = e^{\gamma \bar{z}_{ij} + \gamma^2 V_{ij}/2}$ is the
expected fitness of offspring from the parental pair with $i$ as mother and $j$
as father.
We hence assume no pollen limitation (all outcrossing ovules are fertilized),
and no pollen discounting (the probability of being a father is unaffected by
an individual's selfing rate).
When modeling self-incompatibility, we assume there is no intrinsic
disadvantage to self-incompatibility, except when there is only a single
individual in the population i.e.
\begin{equation}
\Ex[w_i] = \begin{cases}
    \frac{1}{N-1}\sum_{j \ne i }\Ex[w_{ij}] & \text{if } N > 1\\ 
    0 & \text{if } N=1 \end{cases}
\end{equation}

We model assortative mating by ploidy level in a similar way, assuming that a
fraction $\rho_{c_i}$ of the ovules of individual $i$ are fertilized by pollen
sampled from the $c_i$-ploid portion of the population, while a fraction
$1-\rho_{c_i}$ is fertilized by pollen randomly sampled from the entire
population.
\begin{equation}
\Ex[w_i] = \sum_{j=1}^N \left(\frac{\delta_{c_i,c_j}}{N_{c_i}}\rho_{c_i} +
\frac{1}{N}(1-\rho_{c_i})\right)\Ex[w_{ij}]
\end{equation}

\subsection*{Implementation and availability}

Individual-based simulations for the mixed-ploidy infinitesimal model were
implemented in Julia \citep{julia}.
Documented code and simulation notebooks are available at
\url{https://github.com/arzwa/InfGenetics}.

\section*{Results}

\subsection*{Autotetraploid and mixed-ploidy infinitesimal model}

We evaluate the accuracy of the autotetraploid infinitesimal model as an
approximation to the evolution of a quantitative trait determined by $L$
additive loci.
We find that the infinitesimal model with inbreeding generally yields accurate
predictions for the evolution of the genetic variance when the number of loci
is sufficiently large ($L \ge 100$, say, \cref{fig:vztet}).
Furthermore, we confirm that, in the absence of double reducyion, the decay in
genetic variance due to inbreeding after a time $t$ is well-predicted by
$e^{-t/4N}$ (\cref{fig:vztet}A), as expected from the results of \cite{arnold2012}.
As predicted, double reduction leads to an immediate increase in genetic
variance, but leads to accelerated inbreeding, causing faster loss of genetic
variance in the long-term (\cref{fig:vztet}B).
As noted by \cite{arnold2012}, the loss of genetic variation in the presence of
double reduction cannot, however, be described by a single long-term effective
population size.

Simulations for the mixed-ploidy model further confirm the correctness of our
infinitesimal approximation and highlight the importance of assumptions
regarding the scaling of allelic effects across ploidy levels
($\beta$ parameters, \cref{s-fig:vz}).
Although inbreeding is slower in autotetraploids than in diploids for the same
population size, the tetraploid fraction of a diploid-dominated mixed-ploidy 
population will have an equal or higher average inbreeding coefficient
(\cref{s-fig:f1}).
This is because in such a population, triploid and tetraploid individuals
mostly arise from gametes formed by diploid individuals or polyploid
individuals with very recent diploid ancestry (about $1+u+2v$ generations ago
for tetraploids, and $1+\frac{2}{3}(u+2v)$ generations ago for triploids, see
\cref{s-sec:ttdip}), so that the polyploid subpopulations will show an average
relatedness similar to the diploid population and not evolve as an isolated
higher-ploidy population would.
A nonzero probability of producing IBD diploid gametes ($\alpha_k > 0$) will
then further increase the inbreeding coefficient in the tetraploid and triploid
fraction of the population relative to their diploid progenitors
(\cref{s-fig:f1}).
Therefore, as long as diploids dominate, the effect of harboring some of the
gene pool in triploid and tetraploid individuals on the rate of inbreeding is
negligible, and we find that the evolution of the inbreeding coefficient over
time is well predicted by $1-e^{-t/2N_e}$, where the inbreeding-effective
population size is, to first order in $u$, given by $(1-2u)N$
(\cref{s-sec:effsize}).

\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/tetfininf.pdf}
\caption{
The infinitesimal model in autotetraploids.
Comparisons are shown for the decay of the genetic variance ($V_z$) due to
inbreeding in exact simulations of the infinitesimal model in autotetraploids
against individual-based simulations of autotetraploid populations with $L$
unlinked additive loci determining the quantitative trait. 
(A) Simulations of a model without double reduction ($\alpha_4=0$).
(B) Simulations of a model with maximal double reduction ($\alpha_4=1/6$) (for
all loci in the finite $L$ simulations).
We show window-smoothed values for visual clarity, with observed genetic
variances averaged in windows of 20 generations every 10 generations.
The black line marks $e^{-t/4N}$. 
We assume $N=250$ and $V_z(0) = 1$.
\label{fig:vztet}}
\end{figure}

\subsection*{Establishment from a single individual}

Having established the validity of the mixed-ploidy infinitesimal model, we now
use it to study the establishment of polyploids in a marginal habitat to which
migrants from a mixed-ploidy source population are maladapted.

We first consider the establishment of a population from a single migrant
individual with trait value $z_0 < 0$.
We assume $u=0$ and compare the probability of establishment when the migrant
is diploid vs. tetraploid (\cref{fig:est1}).
As noted by \cite{barton2018}, the establishment probability depends
essentially on two dimensionless parameters, $\gamma\sqrt{2V}$ and
$z_0/\sqrt{2V}$, corresponding to the intensity of selection and the degree of
maladaptation, respectively.
We shall scale our results accordingly, assuming $2V=1$ throughout.

We find that reduced inbreeding in tetraploids substantially increases the
establishment probability.
Indeed, in the case where allelic effects are scaled so as to yield the same
equilibrium genetic variance ($\beta = \sqrt{1/2}$), the establishment probability 
for tetraploids can be almost five times as high as in diploids depending on
the selection gradient (\cref{fig:est1}A).
While the probability of establishment becomes lower as the strength of
selection becomes large, the establishment probability does not decrease
monotonically with $\gamma$, i.e. as selection becomes very weak, the
establishment probability also lowers.
This happens because, although the probability of surviving the first couple of
generations becomes higher for weaker selection, adaptation will be slower, and
the risk that genetic variation is exhausted due to inbreeding before the
population is able to reach a consistently positive growth rate is increased.

As expected, the scaling of the equilibrium genetic variance has a profound
effect on the establishment probability, but only when $\beta$ is close to 0.5
(i.e. individual alleles have almost half the effect size in tetraploids
compared to diploids) is the benefit of the slower rate of inbreeding in
tetraploids canceled (\cref{fig:est1}B).

\begin{figure}[t]
\centering
\includegraphics[width=0.8\textwidth]{/home/arthur_z/dev/InfGenetics/doc/img/est2.pdf}
\caption{
(B) Establishment probability from a single diploid or tetraploid individual
for increasing selection gradient $\gamma$.
(A) Probability of a tetraploid individual with trait value $z_0 = -2$
successfully founding a population ($P_4$), relative to the probability for a
diploid individual with the same trait value ($P_2$).
The vertical dashed line marks the $\beta$ value that yields equal equilibrium
variances across ploidy levels ($\beta = \sqrt{1/2}$).
All results are estimated from 500.000 replicate simulations.
\label{fig:est1}}
\end{figure}

\subsection*{Establishment under recurrent migration}

We next consider establishment in the new habitat when there is a continuous
influx of migrants coming from a large diploid-dominated mixed-ploidy source
population at cytotype equilibrium.
In this case, establishment is certain to happen eventually, and we are
interested in the probability that a tetraploid population established before a
diploid one does.

We hypothesized that two counteracting processes affect the probability of
autotetraploid establishment in this case.
On the one hand, increased migration will increase the probability that an
otherwise likely successful tetraploid migrant suffers from MCE in the early
generations while the population size is low, because migrants are likely to be
diploid.
On the other hand, tetraploids are strongly reproductively isolated from
the typical migrant, so that they are less prone to maladaptive swamping.
Hence, conditional on evading MCE, they should be able to adapt to the new
habitat at a rate which is not strongly affected by the migration rate.

\begin{figure}[t]
\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration1.pdf}
\caption{
Establishment with recurrent migration.
(A) Expected time until a population is established in the novel habitat for
increasing rates of migration and different degrees of maladaptation
($\bar{z}_s$, the mean trait value in the source population). 
(B) Proportion of simulation replicates in which tetraploids established. 
The shaded area corresponds to Jeffreys' 95\% interval.
The horizontal dashed line marks the proportion of tetraploid migrants (i.e.
the proportion of tetraploids at equilibrium in the source population).
All results are based on 10.000 replicate simulations.
We assume $\gamma=0.25$ and $u=v=0.05$.
\label{fig:estwmig}}
\end{figure}

\subsection*{Loss of self-incompatibility}

\begin{figure}
\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-si.pdf}
\caption{
Establishment with recurrent migration and loss of self-incompatibility in
polyploids.
(A) Expected time until a population is established in the novel habitat for
increasing rates of migration and different self-fertilization rates in
polyploids. Diploids are assumed to be self-incompatible.
Note that $\sigma=0.0$ refers to random self-fertilization (i.e.
self-fertilization with probability $1/N$). 
(B) Proportion of simulation replicates in which tetraploids established. 
The shaded area corresponds to Jeffreys' 95\% interval.
All results are based on 10.000 replicate simulations.
We assume $\gamma=0.25$ and $u=v=0.05$.
\label{fig:}}
\end{figure}

\subsection*{Assortative mating}

\begin{figure}
\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-am.pdf}
\caption{
\label{fig:}}
\end{figure}

\section*{Discussion}

\bibliographystyle{abbrvnat}
\bibliography{/home/arthur_z/vimwiki/bib.bib}
\end{document}

### 2024-06-04 doc/paper/supp.tex
add paper

--- 2024-06-04 doc/draft.tex
@@
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-\hypertarget{model-and-methods}{%
-\section{Model and Methods}\label{model-and-methods}}
-
-We first develop the infinitesimal model (in the sense of
-{[}@barton2017{]}, i.e.~the `Gaussian descendants' approximation
-{[}@turelli2017{]}) for an autotetraploid population. We then consider
-mixed-ploidy populations in which diploids, triploids and tetraploids
-coexist and interbreed through the production of haploid and diploid
-gametes that combine freely.
-
-\hypertarget{the-infinitesimal-model}{%
-\subsection{The infinitesimal model}\label{the-infinitesimal-model}}
-
-Consider a population which expresses a quantitative trait determined by
-a large number of additive loci of small effect. The infinitesimal model
-approximates the inheritance of such a trait by assuming that the trait
-value \(Z_{ij}\) of a random offspring from parents with trait values
-\(z_i\) and \(z_j\) follows a Gaussian distribution with mean equal to
-the midparent value and variance which is independent of the mean:
-\begin{align}
-  Z_{ij} \sim \Normal\left(\frac{z_i + z_j}{2}, V_{ij}\right)
-  \label{eq:inf}
-  \end{align} Here, \(V_{ij}\) is referred to as the \emph{segregation
-variance} in family \((i,j)\). This is the variation generated among
-offspring from the same parental pair due to random Mendelian
-segregation in meiosis. This approximation can be justified as arising
-from the limit where the number of loci determining the trait tends to
-infinity {[}@barton2017{]}. An alternative, and for our purposes useful,
-way to characterize the model in diploids and polyploids is to write
-\(Z_{ij} = Y_i + Y_j\), where \(Y_i\) and \(Y_j\) are independent
-Gaussian random variables
-\(Y_i \sim \Normal\left(\frac{z_i}{2}, V_i\right)\) (and similarly for
-\(Y_j\)). We refer to \(Y_i\) as the (random) gametic value of
-individual \(i\), and to \(V_i\) as the \emph{gametic segregation
-variance} of individual \(i\). This formulation is helpful in that it
-highlights that Mendelian segregation occurs independently in both
-parents when gametes are produced, which then combine additively to
-determine the offspring trait value.
-
-If we assume a population consisting of infinitely many unrelated
-individuals, expressing a trait with genetic variance \(V\) and
-segregation variance \(V_0\), we find that under random mating the
-variance in the offspring generation is \begin{align}
-V' &= \Ex[\Var[Z_{ij}|Z_i, Z_j]] + \Var[\Ex[Z_{ij}|Z_i,Z_j]] \nonumber \\
-   &= \Ex[V_{ij}] + \Var\left(\frac{Z_i + Z_j}{2}\right) \nonumber \\
-   &= V_0 + \frac{V}{2}\ .
-   \label{eq:eq}
-\end{align} So that at equilibrium (\(V'=V\)), we have \(V = 2V_0\)
-{[}@barton2017{]}. If we denote by \(X\) the contribution to the trait
-value associated with a randomly sampled haploid genome from the
-population, we also have that, at equilibrium, \(V=2V_0=mV_x\), where
-\(V_x = \Var[X]\). Note that no assumptions regarding ploidy level have
-been made in the definition of the model. Indeed, as long as all
-individuals are of the same \emph{even} ploidy level, eq.~\ref{eq:inf}
-and eq.~\ref{eq:eq} apply equally well to the inheritance of an additive
-trait in polyploids as in diploids. Two factors however do lead to
-differences with the standard diploid model due to their effect on
-\(V_i\): (1) the potential occurrence of double reduction in polyploids
-affects the relationship between \(V_x\) and \(V_0\) and (2) the
-evolution of the segregation variance over time in finite populations is
-affected by the ploidy level.
-
-\hypertarget{the-effect-of-double-reduction-on-the-segregation-variance}{%
-\subsection{The effect of double reduction on the segregation
-variance}\label{the-effect-of-double-reduction-on-the-segregation-variance}}
-
-When an autotetraploid forms quadrivalents during prophase I, a form of
-`internal inbreeding' may occur as a result of the phenomenon called
-\emph{double reduction} {[}@lynch1998 p.~57{]}. Double reduction happens
-when, as a result of recombination, replicated gene copies on sister
-chromatids move to the same pole during anaphase I, as illustrated in
-fig.~\ref{fig:dr}. In the example shown in fig.~\ref{fig:dr}, one of the
-four generated gametes is \(AA\), which would not occur in the ordinary
-bivalent meiosis, because in that case, paired chromosomes (involved in
-cross-overs) are separated during anaphase I. The frequency of double
-reduction at a locus in the presence of multivalent formation is hence
-determined by the frequency at which that locus is involved in a
-cross-over (which depends on the distance to the centromere), and has an
-upper bound at \(1/6\) {[}@stift2008{]}.
-
-When double reduction occurs, an \(ABCD\) genotype would generate 10
-distinct gametes, as opposed to 6 in when chromosomes form bivalents. As
-a result, the segregation variance is increased by double reduction. 
-For
-a random genotype \(X_1X_2X_3X_4\), we can find the gametic segregation
-variance contributed by a locus when double reduction happens as
-\begin{align*}
-    \Ex[\var[Y|X_1,X_2,X_3,X_4]] &= \var[Y] - \Var[\Ex[Y|X_1,X_2,X_3,X_4]] \\
-        &= \var[2X] - \var\left[\frac1 4 (2X_1 + 2X_2 + 2X_3 + 2X_4)\right] \\
-        &= 4v_x - \frac{1}{4}4v_x 
-        = 3v_x
-\end{align*} where \(X\) denotes the additive effect of a random allele
-at the locus drawn from the reference population and \(v_x = \Var[X]\).
-In the absence of double reduction we have \begin{align*}
-    \Ex[\var[Y|X_1,X_2,X_3,X_4]] 
-        &= 2\Var[X] - \var\left[\frac1 6 \sum_{i=1}^3 \sum_{j=i+1}^4(X_i+X_j)\right]
-        = v_x
-\end{align*} Assuming that the probability of double reduction at any
-locus is \(\alpha\) (also \(\alpha_4\) below), and summing over
-independent loci, we find that the gametic segregation variance in the
-presence of double reduction should be \begin{equation}
-\frac{V_0}{2} = (1-\alpha)V_x + 3\alpha V_x = V_x(1+2\alpha).
-\label{eq:dr}
-\end{equation}
-
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}
-\footnotesize
-\node (n1) at (0, 5) [] {$A\ B\ C\ D$};
-\node (n2) at (0, 4) [] {$A\underline{A}\ \underline{B}\overline{B}\ CC\ \overline{D}D$};
-\node (n3) at (0, 3) [] {$AB\ AD | CC\ BD$};
-\node (n4) at (-2, 2) [] {$AB\ AD$};
-\node (n5) at ( 2, 2) [] {$CC\ BD$};
-\node (n6) at (-3, 1) [] {$AA$};
-\node (n7) at (-1, 1) [] {$BD$};
-\node (n8) at ( 3, 1) [] {$CB$};
-\node (n9) at ( 1, 1) [] {$CD$};
-\draw[->] (n1) -- (n2) node [right,midway] {\small replication};
-\draw[->] (n2) -- (n3) node [right,midway] {\small recombination};
-\draw[->] (n3) -- (n4);
-\draw[->] (n3) -- (n5) node [right,midway] {\small \ \ meiosis I};
-\draw[->] (n5) -- (n8) node [right,midway] {\small \ \ meiosis II};
-\draw[->] (n5) -- (n9);
-\draw[->] (n4) -- (n6);
-\draw[->] (n4) -- (n7);
-\end{tikzpicture}
-\end{center}
-\begin{caption}{
-Schematic illustration of a meiotic division in an autotetraploid leading to
-double reduction at a locus with genotype $ABCD$.
-Two recombination events are assumed to occur at the locus (denoted by the
-bars).
-\label{fig:dr}
-}\end{caption}
-\end{figure}
-
-\hypertarget{inbreeding-in-autotetraploids}{%
-\subsection{Inbreeding in
-autotetraploids}\label{inbreeding-in-autotetraploids}}
-
-In a finite population the segregation variance will decrease over time
-due to inbreeding. Indeed, in a diploid individual for instance, when
-the two gene copies at a locus are identical, segregation will not
-generate any variation. The rate of inbreeding depends on the ploidy
-level {[}@moody1993; @arnold2012{]}, and hence the rate at which the
-segregation variance declines \emph{does} so too. Let \(F_i\) the
-inbreeding coefficient of individual \(i\), defined as the probability
-that two \emph{distinct} gene copies at a locus in individual \(i\) are
-identical by descent (IBD) with respect to some reference population,
-and let \(\Phi_{ij}\) be the coancestry coefficient for individuals
-\(i\) and \(j\), defined as the probability that two genes sampled
-independently from \(i\) and \(j\) at a given locus are IBD. In
-diploids, one can show that the gametic segregation variance \(V_i\) is
-decreased by a factor \(1-F_i\) relative to the segregation variance in
-the reference population \(V_0/2\) {[}@barton2017{]}. In polyploids,
-however, \(F_i\) does not suffice to describe the state of homozygosity
-in individual \(i\). In tetraploids, for instance, we have five distinct
-homozygosity states, which we can symbolically represent as
-\(abcd, aabc, aabb, aaab\) and \(aaaa\) (in general, the number of
-homozygosity states grows according to the partition function
-\((1,2,3,5,7,11,15,22,\dots)\)). Representing the probability of being
-in these five increasingly homozygous states as
-\(\delta_1, \dots, \delta_5\), we find that the gametic segregation
-variance is reduced by a factor
-\[\phi = \delta_1 + \left(1 - \frac1 6\right) \delta_2 + \left(1 - \frac1
-        3\right)\delta_3 + \left(1 - \frac1 2\right) \delta_4\] which is
-precisely \(1-F_i\), as in diploids (see also {[}@moody1993{]}). This
-shows that we do not need to track the array of homozygosity
-coefficients in order to compute the segregation variance in a
-tetraploid family, but only require the inbreeding coefficients of the
-parents. This is a consequence of the fact that, in tetraploids, gametes
-are diploid. For higher ploidy levels, one would need to track higher
-order identity coefficients. We shall hence not consider higher ploidy
-levels in this paper.
-
-In order to efficiently conduct individual-based simulations under the
-infinitesimal model, we hence need to be able to track the \(F_i\) and
-\(\Phi_{ij}\). Denoting the parents of individual \(i\) by \(k\) and
-\(l\), the recursion for the inbreeding coefficients is \begin{align}
-    F_i &= \Phi_{kl}& \text{in diploids} \nonumber \\  
-    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{in tetraploids} 
-    \label{eq:recursion}
-\end{align} where \(F_k^\ast = \alpha + (1-\alpha)F_k\). The recursion
-for the tetraploid case follows from considering three cases: either (1)
-the two genes sampled in individual \(i\) both came from the gamete
-contributed by parent \(k\), which happens with probability \(1/6\), in
-which case they are IBD with probability \(F_k^\ast\); or (2) as in (1)
-but from parent \(l\); or (3) with probability \(2/3\) the two genes
-came from different gametes, in which case they are IBD with probability
-\(\Phi_{kl}\). The recursion for the coancestry coefficients in
-\(m\)-ploids is given by \begin{align}
-    \Phi_{ii} &= \frac1 m \left(1 + (m-1)F_i\right) \nonumber \\
-    \Phi_{ij} &= \sum_k \sum_l P_{ik}P_{jl} \Phi_{kl} & i \ne j 
-    \label{eq:coancestry}
-\end{align} where the sums are over individuals in the parental
-population, and where \(P_{ik} \in \{0, \frac1 2, 1\}\) is the
-probability that a gene copy in \(i\) is derived from parent \(k\). When
-dealing with discrete generations, \(P_{ik}\) values can be conveniently
-represented in a \(N(t) \times N(t-1)\) matrix, where \(N(t)\) is the
-population size in generation \(t\), so that
-\(\Phi(t) = P \Phi(t-1) P^T\), where \(\Phi(t)\) is the matrix of
-coancestry coefficients in generation \(t\) {[}@barton2017{]}.
-
-\hypertarget{the-infinitesimal-model-in-mixed-ploidy-populations}{%
-\subsection{The infinitesimal model in mixed-ploidy
-populations}\label{the-infinitesimal-model-in-mixed-ploidy-populations}}
-
-When considering mixed-ploidy populations, we need to consider how
-equilibrium variances in the model scale with ploidy level, and how
-offspring of different ploidy levels are derived from a parental pair.
-Throughout the following paragraphs, we specialize to a
-diploid-triploid-tetraploid complex. A glossary of the relevant notation
-can be found in tbl.~\ref{tbl:glossary}.
-
-\hypertarget{scaling-of-the-genetic-variance-across-ploidy-levels}{%
-\subsubsection{Scaling of the genetic variance across ploidy
-levels}\label{scaling-of-the-genetic-variance-across-ploidy-levels}}
-
-The equilibrium phenotypic variance \(V\) under the infinitesimal model
-was derived above as \(2V_0\), where \(V_0\) is the segregation variance
-(eq.~\ref{eq:eq}). As noted below eq.~\ref{eq:eq}, this entails that for
-\(m\)-ploids, \(2V_0 = mV_{x,m}\), where \(V_{x,m}\) is the variance of
-the additive effect of a randomly sampled haploid genome from the
-reference \(m\)-ploid population. Now, in a mixed-ploidy system, it is
-not immediately clear how the \(V_{x,m}\) should scale across ploidy
-levels, as this depends on the effects of polyploidization \emph{per se}
-on additive allelic effects {[}@porturas2019; @clo2021; @clo2022{]}. If
-we assume equal equilibrium variances (and concomitantly segregation
-variances), the genetic variance for a haploid genome in tetraploids
-will be half that of diploids, entailing a reduction in the additive
-allelic effect per locus as ploidy level rises. On the other hand, if we
-assume equal allelic effects, and hence that the genetic variance
-associated with a hypothetical haploid genome copy is identical across
-ploidy levels, then the segregation and equilibrium variances in
-tetraploids will be twice those of diploids, and the associated
-phenotypic range will be doubled as well.
-
-To ease interpretation, let us consider an underlying Mendelian system,
-consisting of \(n\) unlinked additive bi-allelic loci in Hardy-Weinberg
-and linkage equilibrium (HWLE). Assuming that the additive effects in
-tetraploids are homogeneously scaled by a factor \(\beta\) relative to
-those in diploids, and writing \(V_{z,m}\), \(V_{0,m}\) and \(V_{x,m}\)
-for respectively the equilibrium genetic variance, the segregation
-variance and the genetic variance associated with a haploid genome in
-\(m\)-ploids, the following relationships hold: \[
- \frac{V_{z,4}}{V_{z,2}} = \frac{2V_{0,4}}{2V_{0,2}} = \frac{4V_{x,4}}{2V_{x,2}} =
- \frac{2 \sum_i^n (\beta a_i)^2 p_i(1-p_i)}{\sum_i^n a_i^2 p_i(1-p_i)} =
- 2\beta^2
-\] So we see that assuming equal equilibrium phenotypic variances
-\(V_{z,4}/V_{z,2} = 1\) entails that allelic effects are scaled by a
-factor \(\beta=\sqrt{1/2}\). On the other hand, assuming equal allelic
-effects (\(\beta=1\)) entails that the phenotypic variance at
-equilibrium in tetraploids is twice that of diploids, as we noted above.
-To study the effect of such assumptions, we introduce \(\beta_m\) as a
-parameter so that \(V_{z,m} = \frac{m}{2}\beta_m^2 V_{z,2}\), keeping in
-mind its interpretation as a scaler of allelic effects in the underlying
-Mendelian system, and setting \(\beta_2=1\). The trait value
-distribution for offspring of ploidy level \(m_{ij}\) from parental pair
-\((i,j)\) with ploidy levels \(m_i\) and \(m_j\), formed through the
-union of a \(g_i\)-ploid gamete from parent \(i\) and a \(g_j\)-ploid
-gamete from parent \(j\) will then be \[Z_{ij} \sim \Normal\left(
-    \beta_{m_{ij}} \left(
-          \frac{g_i}{m_i}\frac{z_i}{\beta_{m_i}} 
-        + \frac{g_j}{m_j}\frac{z_j}{\beta_{m_j}}\right), V_{i,g_i} + V_{j,g_j}\right)\]
-Where \(V_{i,g_i}\) is the gametic segregation variance associated with
-the formation of a \(g_i\)-ploid gamete by individual \(i\). We now
-proceed to characterize the \(V_{i,g_i}\) and their evolution over time.
-
-\hypertarget{unreduced-gamete-formation-in-diploids}{%
-\subsubsection{Unreduced gamete formation in
-diploids}\label{unreduced-gamete-formation-in-diploids}}
-
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}
-\footnotesize
-\node (rep) at (-6, 4.5) [] {\small Replication};
-\node (rep) at (-6, 3.5) [] {\small Recombination};
-\node (rep) at (-6, 2.5) [] {\small First division};
-\node (rep) at (-6, 1.5) [] {\small Second division};
-\node (rep) at (-6, 0)   [] {\small Gametes};
-\node (Aa)  at (0, 5) [] {$AB$};
-\node (AAaa)  at (0, 4) [] {$AA\ BB$};
-\node (rAaAa) at (-2, 3) [] {$AB\ AB$};
-\node (rAAaa) at ( 2, 3) [] {$AA\ BB$};
-\path (Aa) edge (AAaa);
-\path (AAaa) edge node[near end, above] {$c$} (rAaAa) ;
-\path (AAaa) edge node[near end, above] {\qquad $1-c$} (rAAaa);
-
-\node (fdr11) at (-3, 2) [] {$AB\ AB | \emptyset$};
-\node (fdr12) at (-3, 1) [] {$A|B\  A|B$}; 
-\node (AA) at (-3.8,0) [] {$AA$};
-\node (p1) at (-3.8,-0.5) [] {$\frac{1}{4}$};
-\node (Aa) at (-3,0) [] {$AB$};
-\node (p2) at (-3,-0.5) [] {$\frac{1}{2}$};
-\node (aa) at (-2.2,0) [] {$BB$};
-\node (p2) at (-2.2,-0.5) [] {$\frac{1}{4}$};
-\path (fdr12) edge (AA);
-\path (fdr12) edge (Aa);
-\path (fdr12) edge (aa);
-\path (fdr11) edge (fdr12);
-\path (rAaAa) edge node[left] {$f$} (fdr11);
-
-\node (sdr11) at (-1, 2) [] {$AB | AB$}; 
-\node (sdr12) at (-1, 1) [] {$AB | \emptyset$}; 
-\node (Aa2) at (-1,0) [] {$AB$};
-\path (sdr12) edge (Aa2);
-\path (sdr11) edge (sdr12);
-\path (rAaAa) edge node[right] {$1-f$} (sdr11);
-
-\node (fdr21) at (1,2) [] {$AA\ BB|\emptyset$};
-\node (fdr22) at (1,1) [] {$A|A\ B|B$};
-\node (Aa3) at (1,0) [] {$AB$};
-\path (rAAaa) edge node[left] {$f$} (fdr21);
-\path (fdr21) edge (fdr22);
-\path (fdr22) edge (Aa3);
-
-\node (sdr21) at (3, 2) [] {$AA | BB$};
-\node (sdr22) at (3, 1) [] {$AA| \emptyset, BB|\emptyset$};
-\node (AA4) at (2.2,0) [] {$AA$};
-\node (p1) at (2.2, -0.5) [] {$\frac{1}{2}$};
-\node (aa4) at (3.8,0) [] {$BB$};
-\node (p2) at (3.8, -0.5) [] {$\frac{1}{2}$};
-\path (rAAaa) edge node[right] {$1-f$} (sdr21);
-\path (sdr21) edge (sdr22);
-\path (sdr22) edge (AA4);
-\path (sdr22) edge (aa4);
-\end{tikzpicture}
-\begin{caption}{
-Schematic representation of the different pathways for unreduced gamete
-formation in diploids and their different outcomes.
-\label{fig:udip}
-}\end{caption}
-\end{center}
-\end{figure}
-
-The mechanisms of unreduced gamete formation do not necessarily lead to
-a faithful transmission of the complete diploid genome. Unreduced
-gametes are formed in two ways, depending on the meiotic abberation that
-leads to their origin: (1) first division restitution (FDR) of (2)
-second division restitution (SDR) {[}@bretagnolle1995{]}. Consider a
-locus in a diploid with two distinct genes \(A\) and \(a\). Assume
-recombination happens with probability \(c\) and that conditional on
-unreduced gamete formation, formation is due to FDR with probability
-\(f\) while it is due to SDR with probability \(1-f\). The different
-unreduced gametes that are formed are represented schematically in
-fig.~\ref{fig:udip}. Writing the genotype at a locus in the diploid
-parent as \(X_1X_2\), with allelic effects \(X_1\) and \(X_2\), the
-genotypic value of an unreduced gamete at this locus will be
-\[Y = \begin{cases}
-    2X_1 & \text{w.p. } \alpha_2/2 = \frac1 4 f c + \frac1 2 (1-f)(1-c) \\
-    2X_2 & \text{w.p. } \alpha_2/2 \\
-    X_1 + X_2 & \text{w.p. } 1-\alpha_2 
-    \end{cases}\] where \(\alpha_2 = 1 - f - c + \frac{3}{2}cf\) is the
-probability that two copies of the same gene end up in a diploid gamete
-produced by a diploid individual. Conditional on the latter event, we
-get, by a similar reasoning as the one leading eq.~\ref{eq:dr}, a
-segregation variance equal to \begin{align*}
-\Ex[\Var[Y|X_1,X_2]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2]] \\
-    &= \Var[2X] - \Var\left[\frac{1}{2}(2X_1 + 2X_2)\right] \\
-    &= 2v_x
-\end{align*} Conditioning on the complementary event, all gametes have
-genetic value \(Y = X_1+X_2\), so that the segregation variance is 0.
-Summing across independent loci, we have
-\(V_{i,2} = 2\alpha_2(1-F_i)V_x\) for a diploid individual \(i\).
-
-\hypertarget{meiosis-in-triploids}{%
-\subsubsection{Meiosis in triploids}\label{meiosis-in-triploids}}
-
-Triploids, when viable, may be important for the dynamics of
-mixed-ploidy populations due to the formation of a so-called triploid
-bridge. The formation of triploids presents no immediate issues, we
-simply need to track the segregation variance contributions from both
-donor gametes, and relate these to \(V_{0,3}\). Sexual reproduction in
-triploids is however more complicated. There are no known mechanisms to
-coordinate the assortment of chromosomes in for instance a haploid and
-diploid gamete, and meiosis, if it happens, usually results in aneuploid
-gametes {[}@ramsey1998{]}. Experimental results indicate that, at least
-in yeast, triploids usually form trivalents and undergo recombination,
-after which each trivalent is randomly assorted in the daughter cells,
-some receiving one, others two copies of a given chromosome
-{[}@charles2010{]}. In the absence of gametic nonreduction, the
-probability of obtaining euploid gametes (two diploid and two haploid
-gametes) from such a process is \((1/2)^n\), where \(n\) is the number
-of chromosomes. If the number of chromosomes is small this is not
-negligible, for instance in \emph{A. thaliana} we would have
-\((1/2)^5 \approx 0.03\), which is of the same order as the unreduced
-gamete formation rate. Unreduced (triploid) gametes may also be produced
-and important for the dynamics of mixed-ploidy populations
-{[}@ramsey1998{]}. However, they generate additional difficulty, since
-in order to compute the contributed variance under inbreeding, we would
-need an additional identity coefficient recording the probability that
-three genes are IBD at a locus. We will hence ignore the possibility of
-unreduced gamete production in triploids. Denoting by \(\alpha_3\) the
-probability that a diploid gamete produced by a triploid individual
-contains two copies of the same parental gene, we can use the same
-approach as above to derive \(V_{i,1}\) and \(V_{i,2}\) for a triploid
-individual \(i\). We note that, on the supposition that diploid gametes
-are produced by random assortment of chromosomes in a haploid and
-diploid gametes, \(\alpha_3 \le 1/4\). The formulae for the different
-\(V_{i,g_i}\) are summarized in tbl.~\ref{tbl:segvar}.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2857}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3571}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3571}}@{}}
-\caption{Gametic segregation variance for different cytotype \(\times\)
-gametic cytotype combinations. \(V_x\) refers to genetic variance
-associated with a haploid genome in the reference population of the
-cytotype to which the gamete will contribute. Under the assumptions
-outlined in the main text, we have the following constraints for the
-\(\alpha_k\): \(\alpha_2 \le 1\), \(\alpha_3 \le 1/4\) and
-\(\alpha_4 \le 1/6\). \label{tbl:segvar}}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-\textbf{cytotype}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-\(V_{i,1}\) (haploid)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-\(V_{i,2}\) (diploid)
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-\textbf{cytotype}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-\(V_{i,1}\) (haploid)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-\(V_{i,2}\) (diploid)
-\end{minipage} \\
-\midrule()
-\endhead
-diploid \((m_i = 2)\) & \(\frac{1}{2}(1-F_i)V_x\) &
-\(2\alpha_2 (1-F_i)V_x\) \\
-triploid \((m_i = 3)\) & \(\frac{2}{3}(1-F_i)V_x\) &
-\(\frac{2}{3}(1 + 3\alpha_3)(1-F_i)V_x\) \\
-tetraploid \((m_i = 4)\) & -- & \((1+2\alpha_4)(1-F_i)V_x\) \\
-\bottomrule()
-\end{longtable}
-
-\hypertarget{recursions-for-inbreeding-coefficients}{%
-\subsubsection{Recursions for inbreeding
-coefficients}\label{recursions-for-inbreeding-coefficients}}
-
-In a mixed-ploidy system tracking inbreeding coefficients becomes only
-slightly more complicated. Recall that \(\alpha_m\) is the probability
-that a diploid gamete from an \(m\)-ploid individual transmits two
-copies of the same gene at some locus. Let \(m_i\) denote the ploidy
-level of individual \(i\). We still have the recursion shown in
-eq.~\ref{eq:recursion}, where now
-\[F_k^\ast = F_k(1-\alpha_{m_k}) + \alpha_{m_k}\] For a triploid
-individual, where the parent contributing the \(2n\) gamete is \(k\), we
-have \[ F_i = \frac1 3 \left(F_k^\ast + 2\Phi_{kl}\right) \] For
-\(\Phi\) the recursion in eq.~\ref{eq:coancestry} remains valid, but
-where diagonal elements are now given by
-\[\Phi_{ii} = \frac{1}{m_i} \left(1 + (m_i-1)F_i\right)\] and where the
-elements of the the pedigree matrix \(P\) now have entries in
-\(\{0,\frac1 3,\frac1 2,\frac2 3,1\}\).
-
-\hypertarget{population-dynamics}{%
-\subsubsection{Population dynamics}\label{population-dynamics}}
-
-\emph{What comes here depends on the results we will include.}
-
-The dynamics of the mixed-ploidy system (in the absence of selection on
-the trait) are determined by the rates at which different types of
-gametes are formed. We assume an individual of ploidy level \(k\) forms
-haploid and diploid gametes with proportions \(u_{k1}\) and \(u_{k2}\),
-so that the (relative) fecundity of a \(k\)-ploid individual is
-\(u_{k1} + u_{k2}\). Unless stated otherwise, we will assume \[U = 
-    \begin{pmatrix} 
-    u_{21} & u_{22} \\ 
-    u_{31} & u_{32} \\ 
-    u_{41} & u_{42} 
-    \end{pmatrix} =
-    \begin{pmatrix} 
-    1-u & u \\
-    v & v \\
-    0 & 1-u
-    \end{pmatrix}\] where \(u\) is referred to as the proportion of
-unreduced gametes, and \(2v\) is the proportion of euploid gametes
-produced by a triploid individual. An analysis of a deterministic model
-for the cytotype dynamics {[}also analyzed in@felber1991; @felber1997{]}
-is included in sec.~\ref{sec:det}.
-
-\hypertarget{tbl:glossary}{}
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.0870}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.9130}}@{}}
-\caption{\label{tbl:glossary}Glossary for the notation used to define
-the mixed-ploidy infinitesimal model.}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Symbol
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Explanation
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Symbol
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Explanation
-\end{minipage} \\
-\midrule()
-\endhead
-\(z_i\) & Trait value of individual \(i\) \\
-\(Z_{ij}\) & Trait value of a random offspring from parental pair
-\((i,j)\) \\
-\(V_{z,m}\) & Equilibrium variance in an \(m\)-ploid population \\
-\(V_{0,m}\) & Segregation variance in the \(m\)-ploid reference
-population \\
-\(V_{x,m}\) & Equilibrium variance of a haploid genome in the
-\(m\)-ploid reference population \\
-\(V_{i,m}\) & Gametic segregation variance in individual \(i\) \\
-\(F_i\) & Inbreeding coefficient of individual \(i\) \\
-\(\Phi_{ij}\) & Coancestry coefficient of individuals \(i\) and \(j\) \\
-\(m_i\) & Ploidy level of individual \(i\) \\
-\(\alpha_m\) & Probability that a diploid gamete from an \(m\)-ploid
-individual contains two copies of the same parental gene \\
-\(\beta_m\) & Scaling factor for the allelic effects in ploidy level
-\(m\) \\
-\(u_{m,g}\) & Proportion of \(g\)-ploid gametes produced by an
-\(m\)-ploid individual \\
-\(u\) & Proportion of unreduced gametes in even-ploids \\
-\(v\) & Proportion of haploid and diploid gametes in triploids \\
-\bottomrule()
-\end{longtable}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-We first verify that, in the absence of inbreeding, our implementation
-of the mixed-ploidy infinitesimal model yields the predicted equilibrium
-variance for each cytotype (fig.~\ref{fig:z}) and the expected
-equilibrium proportions (fig.~\ref{fig:eqp}). We next evaluate the
-accuracy of the infinitesimal model as an approximation to the evolution
-of a quantitative trait determined by \(L\) additive loci. We find that
-the infinitesimal model with inbreeding generally yields accurate
-predictions for the evolution of the genetic variance, even as the
-number of loci becomes fairly small (e.g.~\(L=50\), fig.~\ref{fig:vz}).
-For \(L=20\), substantial deviations from the infinitesimal
-approximation can be observed, although it may still be considered
-reasonable (fig.~\ref{fig:vz}).
-
-Although inbreeding is slower in autotetraploids than in diploids for
-the same population size, the tetraploid fraction of a diploid-dominated
-mixed-ploidy population will have an equal or higher average inbreeding
-coefficient (fig.~\ref{fig:f1}). This is because in such a population,
-triploid and tetraploid individuals mostly arise from gametes formed by
-diploid individuals or polyploid individuals with very recent diploid
-ancestry (about \(1+u+2v\) generations ago for tetraploids, and
-\(1+\frac{2}{3}(u+2v)\) generations ago for triploids, see
-sec.~\ref{sec:ttdip}), so that the polyploid subpopulations will show an
-average relatedness similar to the diploid population and not evolve as
-an isolated higher-ploidy population would. A nonzero probability of
-producing IBD diploid gametes (\(\alpha\)) will then further increase
-the inbreeding coefficient in the tetraploid and triploid fraction of
-the population relative to their diploid progenitors
-(fig.~\ref{fig:f1}). Therefore, as long as diploids dominate, the effect
-of harboring some of the gene pool in triploid and tetraploid
-individuals on the rate of inbreeding is negligible, and we find that
-the evolution of the inbreeding coefficient over time is well predicted
-by \(1-e^{-t/2N_e}\), where the inbreeding-effective population size is,
-to first order in \(u\), given by \((1-2u)N\) (sec.~\ref{sec:effsize},
-fig.~\ref{fig:fNe}).
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\begin{itemize}
-\item
-  It appears that many papers on diploid -- tetraploid contact zones or
-  mixed-ploidy populations fail to find significant numbers of
-  triploids. Note though that such studies focus on established
-  tetraploids, where the cytotypes may have diverged to some extent,
-  whereas we focus on the case where a tetraploid cytotype has not yet
-  established. The model is therefore more relevant for populations that
-  are likely to be referred to as diploid. It remains to be seen to what
-  extent such populations actually exhibit a mixed-ploidy equilibrium.
-  See {[}@suda2013{]} and {[}@herben2016{]}.
-\item
-  The issue with dominance. On the one hand, it is often said that
-  polyploids accumulate a larger mutation load, and will hence show
-  inbreeding depression, but this really only applies to already
-  established polyploid populations. Similarly, the sort of effects of
-  dominancestudied in Booker and Schrider during range expansions apply
-  mostly to polyploids that have accumulated a large mutation load over
-  a rather long time as a stable polyploid population. In our case,
-  polyploids are virtually always derived very recently from diploid
-  ancestors, so they will not have accumulated more deleterious
-  recessives than their diploid counterparts. As inbreeding will be
-  slower in tetraploids, we should hence have reduced inbreeding
-  expression, as we have the same load as in diploids, but it is less
-  expressed (at least when considering recessives). There should
-  therefore be an immediate benefit to polyploidy (reduced inbreeding
-  depression), which is however a long term disadvantagem because
-  eventually the established polyploid will start to accumulate a higher
-  load due to less efficient purging.
-\end{itemize}
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\end{CSLReferences}
-
-\clearpage
-\setcounter{page}{1}
-
-\clearpage
+\documentclass[11pt,a4paper]{article}
+\input{preamble.tex}
+\usepackage{geometry}
+\usepackage{authblk}
+\externaldocument[m-]{main}
 \renewcommand{\thefigure}{S\arabic{figure}}
 \renewcommand{\thesection}{S\arabic{section}}
-\renewcommand{\theequation}{S\arabic{equation}}
 \renewcommand{\thealgorithm}{S\arabic{algorithm}}
-\setcounter{figure}{0}
-\setcounter{section}{0}
-\setcounter{equation}{0}
-\setcounter{algorithm}{0}
 
-\hypertarget{supplementary-figures}{%
-\section{Supplementary figures}\label{supplementary-figures}}
+\begin{document}
+\title{Supplement: Autopolyploid establishment through polygenic adaptation}
+\author{Arthur Zwaenepoel}
+\date{\vspace{-5ex}}
+\maketitle
+\tableofcontents
 
-\clearpage
+\section{Supplementary Figures}
 
-\begin{figure}
-\hypertarget{fig:z}{%
+\begin{figure}[H]
 \centering
-\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-10/mixedploidy-eq.svg}
-\caption{Equilibrium trait value (\(z\)) distribution in a mixed-ploidy
-population evolving according to the infinitesimal model without
-inbreeding for three different scaling assumptions of allelic effects
-across ploidy levels. Histograms (lines) show simulation results,
-whereas the filled Gaussian densities show the expected distribution at
-equilibrium \(\Normal\left(0, k\beta_k^2V_x\right)\). In the
-\(\beta=(1,1,1)\) case (left), allelic effects remain constant across
-ploidy levels, causing the equilibrium variance to be doubled in
-tetraploids relative to diploids. In the \(\beta=(1, 3/4, 1/2)\) case
-(middle), allelic effects are scaled so that the phenotypic range is
-held constant (i.e.~the maximum and mimimum phenotypes are teh same
-across ploidy levels). In the \(\beta=(1, \sqrt{2/3}, \sqrt{1/2})\) case
-(right), allelic effects are scaled so that the variance in trait values
-at equilibrium is equal across ploidy levels.}\label{fig:z}
-}
-\end{figure}
-
-\begin{figure}
-\hypertarget{fig:eqp}{%
-\centering
-\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-10/mixedploidy-eqf.svg}
-\caption{Equilibrium proportions of the different cytotypes in a
-mixed-ploidy population where \(u=v=0.05\), light lines show the
-observed proportions in a simulation of a Wright-Fisher mixed-ploidy
-infinitesimal model with \(N=500\). The horizontal lines show the
-deterministic predictions
-(\(\tilde{g}_1^2, 2\tilde{g}_1\tilde{g}_2, \tilde{g}_2^2)\).}\label{fig:eqp}
-}
-\end{figure}
-
-\begin{figure}
-\hypertarget{fig:vz}{%
-\centering
-\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-Vz.svg}
+\includegraphics[width=\textwidth]{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-Vz.pdf}
 \caption{Comparison of the mixed-ploidy infinitesimal model with the
@@
 immediate increase in the genetic variance in higher cytotypes, but also
-accelerated inbreeding.}\label{fig:vz}
-}
+accelerated inbreeding. \label{fig:vz}}
 \end{figure}
 
-\begin{figure}
-\hypertarget{fig:vz2}{%
+\begin{figure}[H]
 \centering
-\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-Vz2.svg}
-\caption{As in fig.~\ref{fig:vz} (\(L \rightarrow \infty\) simulations)
-but highlighting the effect of \(\beta\) and \(\alpha\) within ploidy
-levels. Only smoothed data is shown.}\label{fig:vz2}
-}
-\end{figure}
-
-\begin{figure}
-\hypertarget{fig:f1}{%
-\centering
-\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-F.svg}
+\includegraphics[width=\linewidth]{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-F.pdf}
 \caption{Average inbreeding coefficient \(\bar{F}\) in each cytotype in
@@
 and no selection.}\label{fig:f1}
-}
 \end{figure}
@@
 
-\hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
+\section{Supplementary Information}
 
-\hypertarget{sec:det}{%
-\subsection{Deterministic mixed-ploidy model}\label{sec:det}}
+\subsection{Inbreeding in autotetraploids \label{sec:tetinbred}}
 
-\begin{figure}
-\centering
-\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-13/g1diff.svg}
-\caption{Deterministic mixed-ploidy equilibrium. The left plot shows the
-stable (solid lines) and unstable (dashed lines) equilibria for the
-proportion of haploid gametes in the gamete pool \(g_1\) as a function
-of \(u\) for different values of \(v\). The middle plot shows the
-relationship between \(\Delta g_1 = g_1' - g_1\) and \(g_1\). The zeros
-of this graph are the fixed points of the dynamical system and are
-indicated by the hollow (unstable equilibrium) and solid (stable
-equilibrium) dots. The rightmost plot shows the region of parameter
-space where a stable mixed-ploidy equilibrium exist. \label{fig:g1diff}}
-\end{figure}
+In polyploids, the inbreeding coefficient $F_i$ does not suffice to describe
+the state of homozygosity in individual $i$.
+In tetraploids, for instance, we have five distinct homozygosity states, which
+we can symbolically represent as $abcd, aabc, aabb, aaab$ and $aaaa$ (in
+general, the number of homozygosity states grows according to the partition
+function $(1,2,3,5,7,11,15,22,\dots)$).
+Representing the probability of being in these five increasingly homozygous
+states as $\delta_1, \dots, \delta_5$, we find that the gametic segregation
+variance is reduced by a factor
+  $$\phi = \delta_1 + \left(1 - \frac1 6\right) \delta_2 + \left(1 - \frac1
+        3\right)\delta_3 + \left(1 - \frac1 2\right) \delta_4$$
+which is precisely $1-F_i$, as in diploids (see also \cite{moody1993}).
+This shows that we do not need to track the array of homozygosity coefficients
+in order to compute the segregation variance in a tetraploid family, but only
+require the inbreeding coefficients of the parents.
+This is a consequence of the fact that, in tetraploids, gametes are diploid.
+For higher ploidy levels, one would need to track higher order identity
+coefficients.
 
-Let \(g_k\) be the frequency of \(k\)-ploid gametes in the gamete pool,
-and let us consider only haploid and diploid gametes, so that
-\(g_2 = 1-g_1\). Diploids produce unreduced gametes with probability
-\(u\) and reduced ones with probability \(1-u\), triploids produce
-haploid and triploid gametes both with probability \(v\), and
-tetraploids produce reduced diploid gametes with probability \((1-u)\)
-(we assume they produce, just like diploids, a proportion \(u\) of
-unreduced gametes, but these are assumed not to lead to viable offspring
-and are ignored). We get after one generation of random mating
-\[g_1' = \frac{(1-u)g_1^2 + 2vg_1g_2}{g_1^2 + 4vg_1g_2 + (1-u)g_2^2}.\]
-We see that \(g_1=0\) is always an equilibrium (no haploid gametes,
-tetraploids take over). Two more fixed points are obtained at
+Denoting the parents of individual $i$ by $k$ and $l$, the recursion for the
+inbreeding coefficients in an autotetraploid population is
+\begin{align}
+    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) \label{eq:rectet}
+\end{align}
+where $F_k^\ast = \alpha_4 + (1-\alpha_4)F_k$.
+The recursion follows from considering three cases:
+either (1) the two genes sampled in individual $i$ both came from the gamete
+contributed by parent $k$, which happens with probability $1/6$, in which case
+they are IBD with probability $F_k^\ast$; or (2) as in (1) but from parent $l$; or
+(3) with probability $2/3$ the two genes came from different gametes, in which
+case they are IBD with probability $\Phi_{kl}$ (the coancestry coefficient for
+individuals $k$ and $l$).
+
+\subsection{Mixed-ploidy infinitesimal model}
+
+\subsubsection{Discrete locus model and variance scaling \label{sec:scaling}}
+
+Consider an $L$-locus additive model, with two alleles (0 and 1) at each locus.
+For a $k$-ploid individual, let $X_{i,j}$ be the allele at homolog $j$ of locus
+$i$.
+We assume the trait value is determined by
+\begin{equation}
+  z = \sum_{i=1}^L\sum_{j=1}^k a_{i,k} X_{i,j}
+\end{equation}
+Where $a_{i,k}$ is the allelic effect of the 1 allele at locus $i$ in
+$k$-ploids.
+The genetic variance at HWLE in $k$-ploids ($\tilde{V}_{z,k}$) will then be
+\begin{equation}
+  \tilde{V}_{z,k} = k\sum_{i=1} a_{i,k}^2 p_iq_i = kV_{x,k}
+\end{equation}
+where we refer to $V_{x,k}$ as the variance associated with a haploid genome in
+$k$-ploids at HWLE.
+Note that we also have $\tilde{V}_{z,2} = 2V_{x,k} = 2V$, where $V$ is the
+segregation variance in the diploid population.
+
+We now assume $a_{i,k} = \beta_k a_{i,2}$, i.e. allelic effects in $k$-ploids
+are as in diploids, but scaled by a factor $\beta_k$.
+This implies that
+\begin{equation}
+  \frac{\tilde{V}_{z,k}}{\tilde{V}_{z,2}} 
+  = \frac{kV_{x,k}}{2V_{x,2}} = \frac{k}{2} \beta_{k}^2
+\end{equation}
+and hence also that $V_{x,k} = \beta_k^2 V_{x,2} = \beta_k^2 V$.
+These relations will also hold in the infinitesimal limit.
+Below, we derive the segregation variance expressions for the different meiotic
+processes in the mixed-ploidy models in terms of the $V_{x,k}$.
+Using the relationships outlined here, we will then be able to express all the
+required variances in terms of $V$.
+
+Note that under the above assumptions, not only the variance is rescaled, but
+also the expected value of offspring within a family.
+Specifically, for a $k$-ploid offspring of parental pair $i$ and $j$, where the
+gamete coming from $i$ is $g_i$-ploid and the other gamete $g_j$-ploid, we have
+\begin{equation}
+  \Ex[Z_{ij}] = \beta_k \left(\frac{g_i}{c_i}\frac{z_i}{\beta_{c_i}} + 
+    \frac{g_j}{c_j}\frac{z_j}{\beta_{c_j}}\right)
+\end{equation}
+where $c_i$ is the ploidy level of individual $i$.
+Here we rescale the parental trait values to the diploid scale by dividing by
+$\beta_{c_i}$, sum the expected rescaled gametic values, and rescale to the
+offspring ploidy level.
+
+\subsubsection{Segregation variance \label{sec:segvar}}
+
+\subsection{Deterministic mixed-ploidy model \label{sec:det}}
+
+Let $g_k$ be the frequency of $k$-ploid gametes in the gamete pool, and let us
+consider only haploid and diploid gametes, so that $g_2 = 1-g_1$.
+Diploids produce unreduced gametes with probability $u$ and reduced ones with
+probability $1-u$, triploids produce haploid and triploid gametes both with
+probability $v$, and tetraploids produce reduced diploid gametes with
+probability $(1-u)$ (we assume they produce, just like diploids, a proportion
+$u$ of unreduced gametes, but these are assumed not to lead to viable offspring
+and are ignored).
+We get after one generation of random mating
+  $$g_1' = \frac{(1-u)g_1^2 + 2vg_1g_2}{g_1^2 + 4vg_1g_2 + (1-u)g_2^2}.$$
+We see that $g_1=0$ is always an equilibrium (no haploid gametes, tetraploids
+take over).
+Two more fixed points are obtained at
   \begin{equation}
@@
          {2(2 - u - 4v)} \label{eq:gameteq}
-  \end{equation} Of which the larger one, when it exists, corresponds to
-a stable equilibrium, and the smaller one to an unstable equilibrium. As
-there are no viability differences, the equilibrium cytotype frequencies
-can be readily obtained from these through the relations \begin{align}
+  \end{equation}
+Of which the larger one, when it exists, corresponds to a stable equilibrium,
+and the smaller one to an unstable equilibrium.
+As there are no viability differences, the equilibrium cytotype frequencies can
+be readily obtained from these through the relations 
+\begin{align}
 \pi_2 = \tilde{g_1}^2 & & \pi_3 = 2\tilde{g_1}\tilde{g_2} & & 
@@
 \label{eq:cyteq}
-\end{align} Assuming \(v=O(u)\), we have to second order in \(u\)
+\end{align}
+Assuming $v=O(u)$, we have to second order in $u$
 \begin{align}
@@
  \pi_4 &= u^2 + O(u^3) 
-\end{align} At the critical point where the stable equilibrium
-disappears, we have that \(\Delta g_1 = \frac{d\Delta g_1}{d g_1} = 0\)
-(fig.~\ref{fig:g1diff}, middle). We find that, in the region of
-parameter space that is biologically relevant (roughly
-\(u < 0.1, v < 0.1\), say), the critical unreduced gamete formation rate
-\(u_c\) beyond which tetraploids take over can be expressed as a linear
-function of triploid fertility (\(2v\)): \[u_c = \frac{1}{5}(1-2v)\]
-(fig.~\ref{fig:g1diff}, right). This shows that, for plausible parameter
-values, we can safely assume that an initially diploid population will
-evolve to a mixed-ploidy equilibrium.
-
-\hypertarget{sec:ttdip}{%
-\subsection{Expected time to diploid ancestry}\label{sec:ttdip}}
+\end{align}
+At the critical point where the stable equilibrium disappears, we have that
+$\Delta g_1 = \frac{d\Delta g_1}{d g_1} = 0$ (\cref{fig:g1diff}, middle).
+We find that, in the region of parameter space that is biologically relevant
+(roughly $u < 0.1, v < 0.1$, say), the critical unreduced gamete formation rate
+$u_c$ beyond which tetraploids take over can be expressed as a linear function
+of triploid fertility ($2v$):
+  $$u_c = \frac{1}{5}(1-2v)$$
+(\cref{fig:g1diff}, right).
+This shows that, for plausible parameter values, we can safely assume that an
+initially diploid population will evolve to a mixed-ploidy equilibrium.
+A similar model was first analyzed in \cite{felber1997}.
 
@@
 \centering
-\includegraphics[width=0.5\textwidth,height=\textheight]{/home/arthur_z/vimwiki/build/img/2023-10-22/timetodiploidancestord.svg}
-\caption{Expected time to diploid ancestry. The solid lines show
-\(\Ex[T_4]\), i.e.~the expected time since being inherited from a
-diploid ancestor for a random gene in a tetraploid individual at
-equilibrium, for different values of \(v\) (half the triploid
-fertility). The dashed lines show \(\Ex[T_3]\), i.e.~the same quantity
-for a gene sampled from a triploid. Note that \(\Ex[T]\) blows up
-whenever \(u\) and \(v\) exceed their critical value for tetraploid
-establishment. \label{fig:ttdip}}
+\includegraphics[width=\linewidth]{/home/arthur_z/vimwiki/build/img/2023-10-13/g1diff.pdf}
+\caption{
+Deterministic mixed-ploidy equilibrium. The left plot shows the stable (solid
+lines) and unstable (dashed lines) equilibria for the proportion of haploid
+gametes in the gamete pool $g_1$ as a function of $u$ for different values of
+$v$. The middle plot shows the relationship between $\Delta g_1 = g_1' - g_1$
+and $g_1$. The zeros of this graph are the fixed points of the dynamical system
+and are indicated by the hollow (unstable equilibrium) and solid (stable
+equilibrium) dots. The rightmost plot shows the region of parameter space where
+a stable mixed-ploidy equilibrium exist.
+\label{fig:g1diff}}
 \end{figure}
+
+\subsection{Stochastic mixed-ploidy model \label{sec:mchain}}
+
+For finite $N$, the basic mixed-ploidy model defines a Markov chain on the
+state space $[0..N] \times [0..N]$.
+\begin{align}
+  p_{ij,kl} = \Pr\{N_2(t+1)=k,& N_3(t+1)=l | N_2(t)=i, N_3(t)=j\} \nonumber \\
+    &= \frac{N!}{k!l!(N-k-l)!} p_2^k p_3^l (1-p_2-p_3)^{N-k-l}
+\end{align}
+where 
+\begin{align}
+  p_2 &= \left(\frac{i(1-u) + jv}{N(1-u) + (i+j)u + j(2v-1)}\right)^2 \\
+  p_3 &= \left(\frac{2(i(1-u) + jv)(N(1-u) + i(2u - 1) + j(u + v - 1))}
+    {N(1-u) + (i+j)u + j(2v-1)}\right)^2 
+\end{align}
+Associating a unique index with each pair $(i,j)$ with $0 \le i,j \le N$, we
+can define a transition probability matrix $P$ of dimensions $(N+1)^2 \times
+(N+1)^2$ for this Markov chain.
+
+For nonzero $u$ and $v$, the only absorbing state is the one where $N_2 = N_3 =
+0$, i.e. the tetraploid cytotype fixes.
+All other states are transient, and hence tetraploid fixation occurs with
+probability one.
+The expected time until fixation may however be extremely long.
+Using standard theory for absorbing Markov chains, we can numerically compute
+the expected time until fixation $\Ex[T_{\text{fix}}]$ from the transition
+probability matrix.
+Calculations for the case where $u=v=0.05$ (which are large parameter values
+conducive for tetraploid fixation) are shown in \cref{tbl:mchain-ex}.  Clearly,
+tetraploid establishment by drift alone requires very small population sizes to
+occur at an appreciable rate.
+A similar model without triploids has been analyzed by \cite{rausch2005}.
+
+\begin{table}[t]
+\caption{
+Expected number of generations until fixation of the tetraploid cytotype for
+different population sizes, assuming $u=v=0.05$ and an initially diploid
+population. \label{tbl:mchain-ex}
+}
+\begin{tabularx}{\linewidth}{XXXXXX}
+\cline{1-6}
+$N$ & 10 & 20 & 30 & 40 & 50 \\
+\cline{1-6}
+$\Ex[T_{\text{fix}}]$ & $5.4 \times 10^3$ & $6.4 \times 10^5$ & 
+    $7.9 \times 10^7$ & $9.8 \times 10^9$ & $1.2 \times 10^{12}$ 
+\end{tabularx}
+\end{table}
+
+\subsection{Expected time to diploid ancestry \label{sec:ttdip}}
 
@@
 
-\hypertarget{sec:effsize}{%
-\subsection{Effective population size of a mixed-ploidy
-deme}\label{sec:effsize}}
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{/home/arthur_z/vimwiki/build/img/2023-10-22/timetodiploidancestord.pdf}
+\caption{Expected time to diploid ancestry. The solid lines show
+\(\Ex[T_4]\), i.e.~the expected time since being inherited from a
+diploid ancestor for a random gene in a tetraploid individual at
+equilibrium, for different values of \(v\) (half the triploid
+fertility). The dashed lines show \(\Ex[T_3]\), i.e.~the same quantity
+for a gene sampled from a triploid. Note that \(\Ex[T]\) blows up
+whenever \(u\) and \(v\) exceed their critical value for tetraploid
+establishment. \label{fig:ttdip}}
+\end{figure}
+
+\subsection{Effective population size of a mixed-ploidy deme \label{sec:effsize}}
 
 \begin{figure}
-\hypertarget{fig:fNe}{%
 \centering
-\includegraphics{/home/arthur_z/vimwiki/build/img/2023-10-13/mixedploidy-F-Ne.svg}
+\includegraphics[width=\textwidth]{/home/arthur_z/vimwiki/build/img/2023-10-13/mixedploidy-F-Ne.pdf}
 \caption{The evolution of \(\bar{F}\) in the mixed-ploidy population and
@@
 \(\bar{F}\) in the diploid subpopulation.}\label{fig:fNe}
-}
 \end{figure}
 
-We use the approach outlined in {[}@rousset2004{]} (pp.~153, 157) to
+We use the approach outlined in \citep{rousset2004} (pp.~153, 157) to
 determine the effective size of a randomly mating mixed-ploidy
@@
 vector. Assuming the population is at cytotype equilibrium
-(eq.~\ref{eq:cyteq}), we have \begin{align}
+(eq.~\ref{eq:cyteq}), we have 
+  \begin{align}
       \nu(t+1) &= \nu(t) P \nonumber \\
@@
        \end{pmatrix}
-    \end{align} where we assume, as usual, that tetraploids do not
+  \end{align}
+where we assume, as usual, that tetraploids do not
 produce haploid gametes (\(u_{41} = 0\)), and where \begin{align*}
@@
 the diploid fraction of the population.
+
+\bibliographystyle{abbrvnat}
+\bibliography{/home/arthur_z/vimwiki/bib.bib}
+\end{document}

### 2024-06-05 doc/paper/paper.tex
up

@@
 $\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
-$u$ & probability that an unreduced gamete is formed in diploids
-($u=u_{22}=u_{32}=1-u_{42}$)\\
+$u$ & probability of unreduced gamete formation ($u=u_{22}=u_{32}=1-u_{42}$)\\
 $v$ & probability that a triploid produces a haploid/diploid gamete
   ($v=u_{31}=u_{32}$)\\
+$m$ & expected number of migrants per generation arriving in the new habitat \\
 $z_i$ & trait value of individual $i$ \\
+$c_i$ & ploidy level of individual $i$ \\
+$g_i$ & ploidy level of gamete produced by individual $i$ in a particular cross\\
+$V$ & segregation variance in the reference diploid population \\
+$V_{i,k}$ & gametic segregation variance associated with the production of a
+  $k$-ploid gamete by individual $i$ \\
+$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
+$F_i$ & inbreeding coefficient in individual $i$ \\
+$\Phi_{ij}$ & coancestry coefficient for individuals $i$ and $j$ \\
+$\alpha_k$ & probability that the two genes at a locus in a diploid gamete
+  formed by a $k$-ploid individual descend from the same parental gene copy\\
+$\gamma$ & strength of directional selection in the new habitat\\
+$\theta$ & trait value beyond which the growth rate becomes positive in the new habitat \\
 $w_{ij}$ & fitness of parental pair $(i,j)$ \\
@@
 $i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
-$c_i$ & ploidy level of individual $i$ \\
-$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
-$\gamma$ & strength of directional selection \\
-$F_i$ & inbreeding coefficient in individual $i$ \\
-$V_{i,k}$ & gametic segregation variance associated with the production of a
-  $k$-ploid gamete by individual $i$ \\
-$V$ & segregation variance in the reference diploid population \\
-  $\alpha_k$ & probability that the two genes at a locus in a diploid gamete
-  formed by a $k$-ploid individual descend from the same parental gene copy\\
 $\sigma_k$ & selfing rate in $k$-ploids \\
 $\rho_k$ & probability of assortative mating in $k$-ploids \\
-$m$ & expected number of migrants per generation arriving in the new habitat \\
 \cline{1-2}
@@
 formation through so-called \textit{first} and \textit{second division
-restitution} \citep{storme2013}.
+restitution} \citep{bretagnolle1995,storme2013}.
 We summarize the expressions for the gametic segregation variance in
@@
 \begin{align}
-    F_i &= \Phi_{kl} & \text{if } & c_i = 2 \\ 
+    F_i &= \Phi_{kl} & \text{if } & c_i = 2 \nonumber \\ 
     F_i &= \frac{1}{3}\left(F_k^\ast + 2\Phi_{kl}\right) & \text{if } 
-        & c_i = 3, g_k = 2, g_l = 1 \\ 
+        & c_i = 3, g_k = 2, g_l = 1 \nonumber \\ 
+    F_i &= \frac{1}{3}\left(F_l^\ast + 2\Phi_{kl}\right) & \text{if } 
+        & c_i = 3, g_k = 1, g_l = 2 \nonumber \\ 
     F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{if } & c_i = 4
@@
 predictions for the evolution of the genetic variance when the number of loci
-is sufficiently large ($L \ge 100$, say, \cref{fig:vztet}).
-Furthermore, we confirm that, in the absence of double reducyion, the decay in
+is sufficiently large ($L \ge 100$, say, \cref{fig:vztet,s-fig:fininf}).
+Furthermore, we confirm that, in the absence of double reduction, the decay in
 genetic variance due to inbreeding after a time $t$ is well-predicted by
@@
 variance, but leads to accelerated inbreeding, causing faster loss of genetic
-variance in the long-term (\cref{fig:vztet}B).
+variance in the long-term (\cref{fig:vztet,s-fig:fininf}).
 As noted by \cite{arnold2012}, the loss of genetic variation in the presence of
@@
 \centering
-\includegraphics[width=0.8\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/tetfininf.pdf}
+\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/tetfininf.pdf}
 \caption{
@@
 We first consider the establishment of a population from a single migrant
-individual with trait value $z_0 < 0$.
+individual with trait value $z_0 = 0$.
 We assume $u=0$ and compare the probability of establishment when the migrant
@@
 essentially on two dimensionless parameters, $\gamma\sqrt{2V}$ and
-$z_0/\sqrt{2V}$, corresponding to the intensity of selection and the degree of
+$\theta/\sqrt{2V}$, corresponding to the intensity of selection and the degree of
 maladaptation, respectively.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{/home/arthur_z/dev/InfGenetics/doc/img/est2.pdf}
+\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/est1.pdf}
 \caption{
-(B) Establishment probability from a single diploid or tetraploid individual
-for increasing selection gradient $\gamma$.
-(A) Probability of a tetraploid individual with trait value $z_0 = -2$
+(A) Probability of establishment from a single diploid or tetraploid individual
+with trait value $z=0$ for increasing selection intensity $\gamma$.
+We assume $m=0$ and $u=0$, i.e. there is no migration, and no unreduced
+gametes are produced. The trait is scaled in tetraploids so as to yield the
+same genetic variance at HWLE ($\beta_4 = \sqrt{1/2}$)
+(B) Probability of a tetraploid individual with trait value $z = 0$
 successfully founding a population ($P_4$), relative to the probability for a
 diploid individual with the same trait value ($P_2$).
-The vertical dashed line marks the $\beta$ value that yields equal equilibrium
-variances across ploidy levels ($\beta = \sqrt{1/2}$).
+The vertical dashed line marks $\beta_4=\sqrt{1/2}$, for which the variance at
+HWLE is identical between diploids and autotetraploids.
 All results are estimated from 500.000 replicate simulations.
@@
 \begin{figure}[t]
-\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration1.pdf}
+\centering
+\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration1.pdf}
 \caption{
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-si.pdf}
+\centering
+\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-si.pdf}
 \caption{

### 2024-06-05 doc/paper/supp.tex
up

@@
 \usepackage{authblk}
-\externaldocument[m-]{main}
+\externaldocument[m-]{paper}
 \renewcommand{\thefigure}{S\arabic{figure}}
@@
 \tableofcontents
+\clearpage
 
 \section{Supplementary Figures}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=\textwidth]{/home/arthur_z/dev/InfGenetics/doc/img/finsvsinf-polyploids-validation.pdf}
+\caption{
+Validation of the autotetraploid and triploid infinitesimal model by a
+comparison against the discrete locus model with $L=500$ additive loci, showing
+the average phenotypic variance in each generation averaged over 10 replicate
+simulations for both models.
+We assume the initial phenotypic variance to be one for all simulations, and
+all replicates are initialized randomly in accordance with this initial
+phenotypic variance.
+Allelic effects for the discrete locus model are sampled from a Gaussian with
+mean $0$ and variance $1/2L$.
+\label{fig:fininf}}
+\end{figure}
 
@@
 \section{Supplementary Information}
-
-\subsection{Inbreeding in autotetraploids \label{sec:tetinbred}}
-
-In polyploids, the inbreeding coefficient $F_i$ does not suffice to describe
-the state of homozygosity in individual $i$.
-In tetraploids, for instance, we have five distinct homozygosity states, which
-we can symbolically represent as $abcd, aabc, aabb, aaab$ and $aaaa$ (in
-general, the number of homozygosity states grows according to the partition
-function $(1,2,3,5,7,11,15,22,\dots)$).
-Representing the probability of being in these five increasingly homozygous
-states as $\delta_1, \dots, \delta_5$, we find that the gametic segregation
-variance is reduced by a factor
-  $$\phi = \delta_1 + \left(1 - \frac1 6\right) \delta_2 + \left(1 - \frac1
-        3\right)\delta_3 + \left(1 - \frac1 2\right) \delta_4$$
-which is precisely $1-F_i$, as in diploids (see also \cite{moody1993}).
-This shows that we do not need to track the array of homozygosity coefficients
-in order to compute the segregation variance in a tetraploid family, but only
-require the inbreeding coefficients of the parents.
-This is a consequence of the fact that, in tetraploids, gametes are diploid.
-For higher ploidy levels, one would need to track higher order identity
-coefficients.
-
-Denoting the parents of individual $i$ by $k$ and $l$, the recursion for the
-inbreeding coefficients in an autotetraploid population is
-\begin{align}
-    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) \label{eq:rectet}
-\end{align}
-where $F_k^\ast = \alpha_4 + (1-\alpha_4)F_k$.
-The recursion follows from considering three cases:
-either (1) the two genes sampled in individual $i$ both came from the gamete
-contributed by parent $k$, which happens with probability $1/6$, in which case
-they are IBD with probability $F_k^\ast$; or (2) as in (1) but from parent $l$; or
-(3) with probability $2/3$ the two genes came from different gametes, in which
-case they are IBD with probability $\Phi_{kl}$ (the coancestry coefficient for
-individuals $k$ and $l$).
-
-\subsection{Mixed-ploidy infinitesimal model}
-
-\subsubsection{Discrete locus model and variance scaling \label{sec:scaling}}
-
-Consider an $L$-locus additive model, with two alleles (0 and 1) at each locus.
-For a $k$-ploid individual, let $X_{i,j}$ be the allele at homolog $j$ of locus
-$i$.
-We assume the trait value is determined by
-\begin{equation}
-  z = \sum_{i=1}^L\sum_{j=1}^k a_{i,k} X_{i,j}
-\end{equation}
-Where $a_{i,k}$ is the allelic effect of the 1 allele at locus $i$ in
-$k$-ploids.
-The genetic variance at HWLE in $k$-ploids ($\tilde{V}_{z,k}$) will then be
-\begin{equation}
-  \tilde{V}_{z,k} = k\sum_{i=1} a_{i,k}^2 p_iq_i = kV_{x,k}
-\end{equation}
-where we refer to $V_{x,k}$ as the variance associated with a haploid genome in
-$k$-ploids at HWLE.
-Note that we also have $\tilde{V}_{z,2} = 2V_{x,k} = 2V$, where $V$ is the
-segregation variance in the diploid population.
-
-We now assume $a_{i,k} = \beta_k a_{i,2}$, i.e. allelic effects in $k$-ploids
-are as in diploids, but scaled by a factor $\beta_k$.
-This implies that
-\begin{equation}
-  \frac{\tilde{V}_{z,k}}{\tilde{V}_{z,2}} 
-  = \frac{kV_{x,k}}{2V_{x,2}} = \frac{k}{2} \beta_{k}^2
-\end{equation}
-and hence also that $V_{x,k} = \beta_k^2 V_{x,2} = \beta_k^2 V$.
-These relations will also hold in the infinitesimal limit.
-Below, we derive the segregation variance expressions for the different meiotic
-processes in the mixed-ploidy models in terms of the $V_{x,k}$.
-Using the relationships outlined here, we will then be able to express all the
-required variances in terms of $V$.
-
-Note that under the above assumptions, not only the variance is rescaled, but
-also the expected value of offspring within a family.
-Specifically, for a $k$-ploid offspring of parental pair $i$ and $j$, where the
-gamete coming from $i$ is $g_i$-ploid and the other gamete $g_j$-ploid, we have
-\begin{equation}
-  \Ex[Z_{ij}] = \beta_k \left(\frac{g_i}{c_i}\frac{z_i}{\beta_{c_i}} + 
-    \frac{g_j}{c_j}\frac{z_j}{\beta_{c_j}}\right)
-\end{equation}
-where $c_i$ is the ploidy level of individual $i$.
-Here we rescale the parental trait values to the diploid scale by dividing by
-$\beta_{c_i}$, sum the expected rescaled gametic values, and rescale to the
-offspring ploidy level.
-
-\subsubsection{Segregation variance \label{sec:segvar}}
 
@@
 
+\subsection{Inbreeding in the mixed-ploidy model}
+
+\subsubsection{Effect of inbreeding on segregation variance in autotetraploids \label{sec:tetinbred}}
+
+In polyploids, the inbreeding coefficient $F_i$ does not suffice to describe
+the state of homozygosity in individual $i$.
+In tetraploids, for instance, we have five distinct homozygosity states, which
+we can symbolically represent as $abcd, aabc, aabb, aaab$ and $aaaa$ (in
+general, the number of homozygosity states grows according to the partition
+function $(1,2,3,5,7,11,15,22,\dots)$).
+Representing the probability of being in these five increasingly homozygous
+states as $\delta_1, \dots, \delta_5$, we find that the gametic segregation
+variance is reduced by a factor
+  $$\phi = \delta_1 + \left(1 - \frac1 6\right) \delta_2 + \left(1 - \frac1
+        3\right)\delta_3 + \left(1 - \frac1 2\right) \delta_4$$
+which is precisely $1-F_i$, as in diploids (see also \cite{moody1993}).
+This shows that we do not need to track the array of homozygosity coefficients
+in order to compute the segregation variance in a tetraploid family, but only
+require the inbreeding coefficients of the parents.
+This is a consequence of the fact that, in tetraploids, gametes are diploid.
+Similar considerations apply to triploids if we only model haploid and diploid
+gametes.
+For higher gametic ploidy levels, one would need to track higher order identity
+coefficients in polyploids, which is intractable in general \citep{barton2023}.
+
+\subsubsection{Recursions for inbreeding coefficients in the mixed-ploidy model}
+
+Denoting the parents of individual $i$ by $k$ and $l$, the recursion for the
+inbreeding coefficients in an autotetraploid population is
+\begin{align}
+    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) \label{eq:rectet}
+\end{align}
+where $F_k^\ast = \alpha_4 + (1-\alpha_4)F_k$.
+The recursion follows from considering three cases:
+either (1) the two genes sampled in individual $i$ both came from the gamete
+contributed by parent $k$, which happens with probability $1/6$, in which case
+they are IBD with probability $F_k^\ast$; or (2) as in (1) but from parent $l$; or
+(3) with probability $2/3$ the two genes came from different gametes, in which
+case they are IBD with probability $\Phi_{kl}$ (the coancestry coefficient for
+individuals $k$ and $l$).
+
+There is little difficulty in extending the recursions for diploids
+\citep{barton2017} and autotetraploids (\cref{eq:rectet}) to the mixed-ploidy
+case. 
+Denoting the parents of individual $i$ by $k$ and $l$, the recursion for the
+inbreeding coefficients in the mixed-ploidy case becomes
+\begin{align}
+    F_i &= \Phi_{kl} & \text{if } & c_i = 2 \nonumber \\ 
+    F_i &= \frac{1}{3}\left(F_k^\ast + 2\Phi_{kl}\right) & \text{if } 
+        & c_i = 3, g_k = 2, g_l = 1 \nonumber \\ 
+    F_i &= \frac{1}{3}\left(F_l^\ast + 2\Phi_{kl}\right) & \text{if } 
+        & c_i = 3, g_k = 1, g_l = 2 \nonumber \\ 
+    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{if } & c_i = 4
+\end{align}
+where $F_k^\ast = \alpha_{c_k} + (1-\alpha_{c_k})F_k$, as in the
+autotetraploid model.
+The recursion for the coancestry coefficients in is given by
+\begin{align}
+    \Phi_{ii} &= \frac{1}{c_{i}} \left(1 + (c_i-1)F_i\right) \nonumber \\
+    \Phi_{ij} &= \sum_k \sum_l P_{ik}P_{jl} \Phi_{kl} & i \ne j 
+    \label{eq:coancestry}
+\end{align}
+where the sums are over individuals in the parental population, and where
+$P_{ik} \in \{0, \frac1 3, \frac1 2, \frac2 3, 1\}$ is the probability that a
+gene copy in $i$ is derived from parent $k$.
+
+\subsection{Mixed-ploidy infinitesimal model}
+
+\subsubsection{Discrete locus model and variance scaling \label{sec:scaling}}
+
+Consider an $L$-locus additive model, with two alleles (0 and 1) at each locus.
+For a $k$-ploid individual, let $X_{i,j}$ be the allele at homolog $j$ of locus
+$i$.
+We assume the trait value is determined by
+\begin{equation}
+  z = \sum_{i=1}^L\sum_{j=1}^k a_{i,k} X_{i,j}
+\end{equation}
+Where $a_{i,k}$ is the allelic effect of the 1 allele at locus $i$ in
+$k$-ploids.
+The genetic variance at HWLE in $k$-ploids ($\tilde{V}_{z,k}$) will then be
+\begin{equation}
+  \tilde{V}_{z,k} = k\sum_{i=1} a_{i,k}^2 p_iq_i = kV_{x,k}
+\end{equation}
+where we refer to $V_{x,k}$ as the variance associated with a haploid genome in
+$k$-ploids at HWLE.
+Note that we also have $\tilde{V}_{z,2} = 2V_{x,k} = 2V$, where $V$ is the
+segregation variance in the diploid population.
+
+We now assume $a_{i,k} = \beta_k a_{i,2}$, i.e. allelic effects in $k$-ploids
+are as in diploids, but scaled by a factor $\beta_k$.
+This implies that
+\begin{equation}
+  \frac{\tilde{V}_{z,k}}{\tilde{V}_{z,2}} 
+  = \frac{kV_{x,k}}{2V_{x,2}} = \frac{k}{2} \beta_{k}^2
+\end{equation}
+and hence also that $V_{x,k} = \beta_k^2 V_{x,2} = \beta_k^2 V$.
+These relations will also hold in the infinitesimal limit.
+Below, we derive the segregation variance expressions for the different meiotic
+processes in the mixed-ploidy models in terms of the $V_{x,k}$.
+Using the relationships outlined here, we will then be able to express all the
+required variances in terms of $V$.
+
+Note that under the above assumptions, not only the variance is rescaled, but
+also the expected value of offspring within a family.
+Specifically, for a $k$-ploid offspring of parental pair $i$ and $j$, where the
+gamete coming from $i$ is $g_i$-ploid and the other gamete $g_j$-ploid, we have
+\begin{equation}
+  \Ex[Z_{ij}] = \beta_k \left(\frac{g_i}{c_i}\frac{z_i}{\beta_{c_i}} + 
+    \frac{g_j}{c_j}\frac{z_j}{\beta_{c_j}}\right)
+\end{equation}
+where $c_i$ is the ploidy level of individual $i$.
+Here we rescale the parental trait values to the diploid scale by dividing by
+$\beta_{c_i}$, sum the expected rescaled gametic values, and rescale to the
+offspring ploidy level.
+
+\subsubsection{Segregation variance \label{sec:segvar}}
+
+Consider a single locus in a population of $k$-ploids, and let $X_1,\dots,X_k$
+denote a random genotype at this locus, where $X_i$ is the additive genetic
+value of the allele on homolog $i$.
+The variance in gametic values $Y$ produced by some meiotic process can be
+decomposed as 
+\begin{equation}
+  \Var[Y] = \underbrace{\Ex\left[\Var[Y|X_1, \dots, X_k]\right]}_{\text{segregation variance}} 
+    + \underbrace{\Var\left[\Ex[Y|X_1, \dots, X_k]\right]}_{\text{population
+variation}}
+\label{eq:eve}
+\end{equation}
+We use this relationship to derive the segregation variance for a given meiotic
+process.
+
+As an example, consider ordinary diploid meiosis at a single locus, where a
+diploid produces a haploid gamete. $Y$ is a random haploid gamete sampled from the
+population, and $\Var[Y] = \Var[X] = v_x$, where $X$ is a a randomly sampled
+gene from the population and $v_x$ is the variance in additive genetic values
+at the locus.
+We have
+  $$\Var[\Ex[Y|X_1,X_2]] = \Var\left[\frac{X_1 + X_2}{2}\right] = \frac{v_x}{2}$$
+Using \cref{eq:eve}, we can then find the segregation variance
+  $$\Ex[\Var[Y|X_1,X_2]] = v_x - \frac{v_x}{2} = \frac{v_x}{2}$$
+Summing over $L$ independent loci, we find the gametic segregation variance in
+diploids (which is halve the zygotic segregation variance $V$, by definition) as
+  $$V_{S(2,1)} = \frac{V}{2} = \sum_{i=1}^L \frac{v_{x,i}}{2} = \frac{V_{x,2}}{2}$$
+and hence $V=V_{x,2}$, where $V_{x,2}$ is as defined in \cref{sec:scaling}.
+This will hold in the infinitsimal limit, where $L$ becomes very large and
+$v_x$ smaller and smaller.
+
+Below, we derive the segregation variance associated with haploid and diploid
+gamete production in the three cytotypes of the mixed-ploidy population.
+We use \cref{eq:eve} to do this, and express the segregation variance for a
+$k$-ploid individual producing an $l$-ploid $V_{S(k,l)}$ in terms of the
+variance $V_{x,k}$ associated with a haploid genome in a (hypothetical)
+$k$-ploid reference population at HWLE. Using the scaling assumptions above,
+i.e. $V_{x,k} = \beta_k^2 V$, we can then express all segregation variances in
+terms of the diploid segregation variance in the reference population $V$.
+
+\paragraph{Meiosis in autotetraploids.}
+
+When an autotetraploid forms quadrivalents during prophase I, a form of
+'internal inbreeding' may occur as a result of the phenomenon called
+\textit{double reduction} (see e.g. \cite{lynch1998} p. 57).
+Double reduction happens when, as a result of recombination, replicated gene
+copies on sister chromatids move to the same pole during anaphase I, as
+illustrated in \cref{fig:dr}.
+In the example shown in \cref{fig:dr}, one of the four generated gametes is $AA$,
+which would not occur in the ordinary bivalent meiosis, because in that case,
+paired chromosomes (involved in cross-overs) are separated during anaphase I.
+The frequency of double reduction at a locus in the presence of multivalent
+formation is hence determined by the frequency at which that locus is involved
+in a cross-over (which depends on the distance to the centromere), and has an
+upper bound at $1/6$ \citep{stift2008}.
+
+When double reduction occurs, an $ABCD$ genotype would generate 10 distinct
+gametes, as opposed to 6 in when chromosomes form bivalents. As a result, the
+segregation variance is increased by double reduction.
+For a random genotype $X_1X_2X_3X_4$, we can find the gametic segregation
+variance contributed by a locus when double reduction happens as
+\begin{align*}
+    \Ex[\Var[Y|X_1,X_2,X_3,X_4]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3,X_4]] \\
+        &= \Var[2X] - \Var\left[\frac1 4 (2X_1 + 2X_2 + 2X_3 + 2X_4)\right] \\
+        &= 4v_x - \frac{1}{4}4v_x 
+        = 3v_x
+\end{align*}
+where $X$ denotes the additive effect of a random allele at the locus drawn from the reference
+population and $v_x = \Var[X]$.
+In the absence of double reduction we have
+\begin{align*}
+    \Ex[\Var[Y|X_1,X_2,X_3,X_4]] 
+        &= 2\Var[X] - \Var\left[\frac1 6 \sum_{i=1}^3 \sum_{j=i+1}^4(X_i+X_j)\right]
+        = v_x
+\end{align*}
+Assuming that the probability of double reduction at any locus is $\alpha_4$,
+and summing over independent loci, we find that the gametic segregation
+variance in the presence of double reduction should be
+\begin{equation}
+V_{S(4,2)} = (1-\alpha_4)V_{x,4} + 3\alpha_4 V_{x,4} = V_{x,4}(1+2\alpha_4).
+\label{eq:dr}
+\end{equation}
+where, again, $V_{x,4}$ is as defined in \cref{sec:scaling}.
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}
+\footnotesize
+\node (n1) at (0, 5) [] {$A\ B\ C\ D$};
+\node (n2) at (0, 4) [] {$A\underline{A}\ \underline{B}\overline{B}\ CC\ \overline{D}D$};
+\node (n3) at (0, 3) [] {$AB\ AD | CC\ BD$};
+\node (n4) at (-2, 2) [] {$AB\ AD$};
+\node (n5) at ( 2, 2) [] {$CC\ BD$};
+\node (n6) at (-3, 1) [] {$AA$};
+\node (n7) at (-1, 1) [] {$BD$};
+\node (n8) at ( 3, 1) [] {$CB$};
+\node (n9) at ( 1, 1) [] {$CD$};
+\draw[->] (n1) -- (n2) node [right,midway] {\small replication};
+\draw[->] (n2) -- (n3) node [right,midway] {\small recombination};
+\draw[->] (n3) -- (n4);
+\draw[->] (n3) -- (n5) node [right,midway] {\small \ \ meiosis I};
+\draw[->] (n5) -- (n8) node [right,midway] {\small \ \ meiosis II};
+\draw[->] (n5) -- (n9);
+\draw[->] (n4) -- (n6);
+\draw[->] (n4) -- (n7);
+\end{tikzpicture}
+\end{center}
+\begin{caption}{
+Schematic illustration of a meiotic division in an autotetraploid leading to
+double reduction at a locus with genotype $ABCD$.
+Two recombination events are assumed to occur at the locus (denoted by the
+bars).
+\label{fig:dr}
+}\end{caption}
+\end{figure}
+
+\paragraph{Unreduced gamete formation in diploids.}
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}
+\footnotesize
+\node (rep) at (-6, 4.5) [] {\small Replication};
+\node (rep) at (-6, 3.5) [] {\small Recombination};
+\node (rep) at (-6, 2.5) [] {\small First division};
+\node (rep) at (-6, 1.5) [] {\small Second division};
+\node (rep) at (-6, 0)   [] {\small Gametes};
+\node (Aa)  at (0, 5) [] {$AB$};
+\node (AAaa)  at (0, 4) [] {$AA\ BB$};
+\node (rAaAa) at (-2, 3) [] {$AB\ AB$};
+\node (rAAaa) at ( 2, 3) [] {$AA\ BB$};
+\path (Aa) edge (AAaa);
+\path (AAaa) edge node[near end, above] {$c$} (rAaAa) ;
+\path (AAaa) edge node[near end, above] {\qquad $1-c$} (rAAaa);
+
+\node (fdr11) at (-3, 2) [] {$AB\ AB | \emptyset$};
+\node (fdr12) at (-3, 1) [] {$A|B\  A|B$}; 
+\node (AA) at (-3.8,0) [] {$AA$};
+\node (p1) at (-3.8,-0.5) [] {$\frac{1}{4}$};
+\node (Aa) at (-3,0) [] {$AB$};
+\node (p2) at (-3,-0.5) [] {$\frac{1}{2}$};
+\node (aa) at (-2.2,0) [] {$BB$};
+\node (p2) at (-2.2,-0.5) [] {$\frac{1}{4}$};
+\path (fdr12) edge (AA);
+\path (fdr12) edge (Aa);
+\path (fdr12) edge (aa);
+\path (fdr11) edge (fdr12);
+\path (rAaAa) edge node[left] {$f$} (fdr11);
+
+\node (sdr11) at (-1, 2) [] {$AB | AB$}; 
+\node (sdr12) at (-1, 1) [] {$AB | \emptyset$}; 
+\node (Aa2) at (-1,0) [] {$AB$};
+\path (sdr12) edge (Aa2);
+\path (sdr11) edge (sdr12);
+\path (rAaAa) edge node[right] {$1-f$} (sdr11);
+
+\node (fdr21) at (1,2) [] {$AA\ BB|\emptyset$};
+\node (fdr22) at (1,1) [] {$A|A\ B|B$};
+\node (Aa3) at (1,0) [] {$AB$};
+\path (rAAaa) edge node[left] {$f$} (fdr21);
+\path (fdr21) edge (fdr22);
+\path (fdr22) edge (Aa3);
+
+\node (sdr21) at (3, 2) [] {$AA | BB$};
+\node (sdr22) at (3, 1) [] {$AA| \emptyset, BB|\emptyset$};
+\node (AA4) at (2.2,0) [] {$AA$};
+\node (p1) at (2.2, -0.5) [] {$\frac{1}{2}$};
+\node (aa4) at (3.8,0) [] {$BB$};
+\node (p2) at (3.8, -0.5) [] {$\frac{1}{2}$};
+\path (rAAaa) edge node[right] {$1-f$} (sdr21);
+\path (sdr21) edge (sdr22);
+\path (sdr22) edge (AA4);
+\path (sdr22) edge (aa4);
+\end{tikzpicture}
+\begin{caption}{
+Schematic representation of the different pathways for unreduced gamete
+formation in diploids and their different outcomes.
+\label{fig:udip}
+}\end{caption}
+\end{center}
+\end{figure}
+
+The mechanisms of unreduced gamete formation do not necessarily lead to a
+faithful transmission of the complete diploid genome.
+Unreduced gametes are formed in two ways, depending on the meiotic abberation
+that leads to their origin: (1) first division restitution (FDR) of (2) second
+division restitution (SDR) \citep{bretagnolle1995,storme2013}.
+Consider a locus in a diploid with two distinct genes $A$ and $a$. Assume
+recombination happens with probability $c$ and that conditional on unreduced
+gamete formation, formation is due to FDR with probability $f$ while it is due
+to SDR with probability $1-f$.
+The different unreduced gametes that are formed are represented schematically
+in \cref{fig:udip}.
+Writing the genotype at a locus in the diploid parent as $X_1X_2$, with allelic
+effects $X_1$ and $X_2$, the genotypic value of an unreduced gamete at this
+locus will be
+$$Y = \begin{cases}
+    2X_1 & \text{w.pr.  } \alpha_2/2 \\
+    2X_2 & \text{w.pr.  } \alpha_2/2 \\
+    X_1 + X_2 & \text{w.pr.  } 1-\alpha_2 
+    \end{cases}$$
+where $\alpha_2 = 1 - f - c + \frac{3}{2}cf$ is the probability that two copies
+of the same gene end up in a diploid gamete produced by a diploid individual
+(see the diagram in \cref{fig:udip}). 
+Conditional on the latter event, we get
+\begin{align*}
+\Ex[\Var[Y|X_1,X_2]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2]] \\
+    &= \Var[2X] - \Var\left[\frac{1}{2}(2X_1 + 2X_2)\right] \\
+    &= 2v_x
+\end{align*}
+Conditioning on the complementary event, all gametes have genetic value
+$Y = X_1+X_2$, so that the segregation variance is 0.
+Summing across independent loci, we have $V_{S(2,2)} = 2\alpha_2 V_{x,2}$.
+
+\paragraph{Meiosis in triploids.}
+
+Triploids, when viable, may be important for the dynamics of mixed-ploidy
+populations due to the formation of a so-called triploid bridge.
+The formation of triploids presents no immediate issues, we simply need to
+track the segregation variance contributions from both donor gametes, and
+relate these to $V_{0,3}$.
+Sexual reproduction in triploids is however more complicated.
+There are no known mechanisms to coordinate the assortment of chromosomes in
+for instance a haploid and diploid gamete, and meiosis, if it happens, usually
+results in aneuploid gametes \citep{ramsey1998}.
+
+Experimental results indicate that, at least in yeast, triploids usually form
+trivalents and undergo recombination, after which each trivalent is randomly
+assorted in the daughter cells, some receiving one, others two copies of a
+given chromosome \citep{charles2010}.
+In the absence of gametic nonreduction, the probability
+of obtaining euploid gametes (two diploid and two haploid gametes) from such 
+a process is $(1/2)^n$, where $n$ is the number of chromosomes. If the number
+of chromosomes is small this is not negligible, for instance in
+\textit{Arabidopsis thaliana} we would have $(1/2)^5 \approx 0.03$, which is of
+the same order as the unreduced gamete formation rate.
+Unreduced (triploid) gametes may also be produced and important for the
+dynamics of mixed-ploidy populations \citep{ramsey1998}.
+However, they generate additional difficulty, since in order to compute the
+contributed variance under inbreeding, we would need an additional identity
+coefficient recording the probability that three genes are IBD at a locus.
+We will hence ignore the possibility of unreduced gamete production in
+triploids.
+We note that, on the supposition that diploid gametes are produced by random
+assortment of chromosomes in a haploid and diploid gametes, $\alpha_3 \le 1/4$.
+
+When a triploid produces a haploid gamete, we get as segregation variance at a
+single locus
+\begin{align*}
+  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
+    &= v_x - \Var\left[\frac{1}{3}\left(X_1 + X_2 + X_3\right)\right] \\ 
+    &= v_x - \frac{1}{9}3v_x = \frac{2}{3}v_x
+\end{align*}
+Summing over many loci, we have $V_{S(3,1)} = \frac{2}{3}V_{x,3}$.
+When a triploid produces a diploid gamete, we assume there is, as in diploids
+and tetraploids, a probability $\alpha_3$ that the same gene copy ends up twice
+in the gamete.
+Conditional on this happening, we have
+\begin{align*}
+  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
+    &= \Var[2X] - \Var\left[\frac{1}{3}\left(2X_1 + 2X_2 + 2X_3\right)\right] \\
+    &= 4V_x - \frac{4}{9}3V_x = \frac{8}{3}V_x
+\end{align*}
+Conditional on this \textit{not} happening, 
+\begin{align*}
+  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
+    &= 2V_x - \Var\left[\frac{1}{3}\left((X_1+X_2) + (X_1+X_3) +
+      (X_2+X_3)\right)\right] \\
+    &= 2V_x - \frac{4}{9}3V_x = \frac{2}{3}V_x
+\end{align*} 
+Putting this together and summing over many loci, the segregation variance for
+a diploid gamete from a triploid individual would be
+  $$V_{S(3,2)} = \frac{8}{3}V_{x,3}\alpha_3 + \frac{2}{3}V_{x,3} (1-\alpha_3) 
+            = \frac{2}{3}V_{x,3}(1 + 3\alpha_3)$$
+
+The different expressions for the gametic segregation variance in the
+mixed-ploidy model, adjusted for inbreeding, are summarized in
+\cref{m-tbl:segvar} in the main text.
+
 \bibliographystyle{abbrvnat}

### 2024-06-06 doc/paper/paper.tex
update

@@
 autotetraploids in a new habitat.
+We show that in the absence of migration from the source population,
+autotetraploids are more likely to found a successful population than diploids
+under a very broad range of conditions, but that this is unlikely to
+sufficiently counter the scarcity of tetraploid founders when the source is
+predominantly diploid.
+[...]
 \end{abstract}
@@
 This negative frequency dependence effect is commonly referred to as
-\textit{minority cytotype exclusion} (MCE) after \cite{levin1975}.
+\textit{minority cytotype exclusion} (MCE), after \cite{levin1975}.
 It is well-appreciated that, due to MCE, in a large random mating mixed-ploidy
@@
   $k$-ploid gamete by individual $i$ \\
+$\Vx_k$ & genetic variance associated with a haploid genome in the $k$-ploid
+  reference population (i.e. a $k$-ploid non-inbred population at HWLE) \\
 $\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
@@
 $\gamma$ & strength of directional selection in the new habitat\\
-$\theta$ & trait value beyond which the growth rate becomes positive in the new habitat \\
-$w_{ij}$ & fitness of parental pair $(i,j)$ \\
+$\theta$ & trait value beyond which the growth rate becomes positive in the new
+habitat \\ $w_{ij}$ & fitness of parental pair $(i,j)$ \\
 $w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
@@
 
+\paragraph{Scaling of traits across ploidy levels.}
+
+If we would somewhat naively assume that the allelic effects underlying an
+additive quantitative trait are identical across ploidy levels, 
+a tetraploid offspring from a cross between two diploids would yield, on average,
+a trait value which is the sum of the parental trait values, with a
+distribution that depends on the variance generated by the process producing
+unreduced gametes.
+This is not likely to reflect biological reality: tetraploids do not tend to
+have, for instance, twice the size of their diploid progenitors on average.
+Similarly, the genetic variance at Hardy-Weinberg and linkage equilibrium
+(HWLE) in tetraploids will be twice that of their diploid counterparts under
+such assumptions, which is similarly unrealistic \citep{clo2022}.
+
+In order to account for this, we introduce a scaling factor $\beta_k$,
+accounting for the effects of polyploidization \textit{per se} on trait
+expression in $k$-ploids.
+To introduce and interpret this parameter, we consider an $L$-locus additive
+model, with two alleles (0 and 1) at each locus.
+For a $k$-ploid individual, let $X_{i,j}$ be the allele at homolog $j$ of locus
+$i$.
+We assume the trait value is determined by
+\begin{equation}
+  z = \sum_{i=1}^L\sum_{j=1}^k a_{i,k} X_{i,j}
+\end{equation}
+Where $a_{i,k}$ is the allelic effect of the 1 allele at locus $i$ in
+$k$-ploids.
+The genetic variance at HWLE in $k$-ploids ($\tilde{V}_{z,k}$) will then be
+\begin{equation}
+  \tilde{V}_{z,k} = k\sum_{i=1}^L a_{i,k}^2 p_iq_i = k\Vx_{k}
+\end{equation}
+where we refer to $\Vx_{k}$ as the variance associated with a haploid genome in
+$k$-ploids at HWLE.
+Note that we also have $\tilde{V}_{z,2} = 2\Vx_{2} = 2V$ \citep{barton2017},
+where $V$ is the segregation variance in the diploid population.
+If we now assume $a_{i,k} = \beta_k a_{i,2}$, i.e. allelic effects in $k$-ploids
+are as in diploids, but scaled by a factor $\beta_k$.
+This implies that
+\begin{equation}
+  \frac{\tilde{V}_{z,k}}{\tilde{V}_{z,2}} 
+  = \frac{k\Vx_{k}}{2\Vx_{2}} = \frac{k}{2} \beta_{k}^2
+\end{equation}
+and hence also that $\Vx_{k} = \beta_k^2 \Vx_{2} = \beta_k^2 V$.
+These relations will also hold in the infinitesimal limit where $a_{i,2}
+\rightarrow 0$ as $L\rightarrow \infty$.
+
 \paragraph{Mixed-ploidy infinitesimal model.}
+
+We can extend the infinitesimal model to the mixed-ploidy case, assuming that the
+gametic value, on the diploid trait scale, associated with a $k$-ploid gamete ($k
+\in \{1,2\}$) from individual $i$ of ploidy level $c_i \in \{2,3,4\}$ is a
+Gaussian random variable $Y_{i,k}$ with distribution
+  $$Y_{i,k} \sim \Normal\left(\frac{k}{c_i}\frac{z_i}{\beta_{c_i}}, V_{i,k}\right)$$
+where $V_{i,k}$ is the gametic segregation variance associated with the
+production of a $k$-ploid gamete by individual $i$.
+The trait value of an individual originating from the union of a $k$-ploid
+gamete of individual $i$ and an $l$-ploid gamete from individual $j$ is then
+  $$Z_{ij}^{kl} = \beta_{k+l}\left(Y_{i,k} + Y_{j,l}\right)$$
+i.e., $Z_{ij}^{kl}$ is a Gaussian random variate with distribution
+\begin{equation}
+  Z_{ij}^{kl} \sim \Normal\left(
+    \beta_{k+l} \left(
+          \frac{k}{c_i}\frac{z_i}{\beta_{c_i}} 
+        + \frac{l}{c_j}\frac{z_j}{\beta_{c_j}}\right), 
+    \beta_{k+l}^2 (V_{i,k} + V_{j,l})\right)
+   \label{eq:oneline}
+\end{equation}
+
+For the production of diploid gametes by a $k$-ploid individual, the
+segregation variance depends not only on the segregation variance in the base
+population ($V$) and the inbreeding coefficient ($F$), but also on the detailed
+assumptions on how these aberrant meiotic processes occur.
+Importantly however, these details only affect the gametic segregation variance
+through the quantity $\alpha_k$, which is the probability that a $k$-ploid
+transmits two copies of the same homolog to a diploid gamete.
+Note that $\alpha_4$, the probability that a diploid gamete of a tetraploid
+individual carries two copies of the same homolog, is the probability of
+\textit{double reduction} (e.g. \cite{lynch1998} p.57), and is upper bounded by
+1/6 \citep{stift2008}.
+The value of $\alpha_2$ depends on the relative frequency of unreduced gamete
+formation through so-called \textit{first} and \textit{second division
+restitution} \citep{bretagnolle1995,storme2013}.
+We summarize the expressions for the gametic segregation variance in
+\cref{tbl:segvar}.
+Detailed derivations can be found in \cref{s-sec:segvar}.
 
@@
 \end{table}
-
-We can extend this model to the mixed-ploidy case, where we assume that the
-gametic value associated with a $k$-ploid gamete ($k \in \{1,2\}$) from
-individual $i$ of ploidy level $c_i \in \{2,3,4\}$ is a Gaussian random
-variable with distribution
-  $$Y_{i,k} \sim \Normal\left(\frac{k}{c_i}z_i, V_{i,k}\right)$$
-where $V_{i,k}$ is the gametic segregation variance associated with the
-production of a $k$-ploid gamete by individual $i$.
-For diploids producing diploid unreduced gametes and triploid producing euploid
-gametes, the latter depends not only on the segregation variance in the base
-population ($V$) and the inbreeding coefficient ($F$), but also on the detailed
-assumptions on how these aberrant meiotic processes occur.
-Importantly however, when only haploid and diploid gametes are assumed to occur,
-these details only affect the gametic segregation variance through the quantity
-$\alpha_k$, which is the probability that a $k$-ploid transmits two copies of 
-the same homolog to a diploid gamete.
-
-Note that $\alpha_4$, the probability that a diploid gamete of a tetraploid
-individual carries two copies of the same homolog, is the probability of
-\textit{double reduction} (e.g. \cite{lynch1998} p.57), and is upper bounded by
-1/6 \citep{stift2008}.
-The value of $\alpha_2$ depends on the relative frequency of unreduced gamete
-formation through so-called \textit{first} and \textit{second division
-restitution} \citep{bretagnolle1995,storme2013}.
-We summarize the expressions for the gametic segregation variance in
-\cref{tbl:segvar}.
-Detailed derivations can be found in \cref{s-sec:segvar}.
-
-If we would somewhat naively assume $Z_{ij} = Y_{i,k} + Y_{j,l}$ for a
-$(k+l)$-ploid offspring from parental pair $(i,j)$, as we did in the standard
-infinitesimal model (\cref{eq:inf}), 
-a tetraploid offspring from a cross between two diploids will yield, on average,
-a trait value which is the sum of the parental trait values (and hence twice
-the midparent value).
-This is not likely to reflect biological reality: tetraploids do not tend to
-have, for instance, twice the size of their diploid progenitors on average.
-Similarly, the genetic variance at Hardy-Weinberg and linkage equilibrium
-(HWLE) in tetraploids will be twice that of their diploid counterparts under
-such assumptions, which is similarly unrealistic \citep{clo2022}.
-
-In order to account for this, we introduce a scaling factor $\beta_k$,
-accounting for the effects of polyploidization \textit{per se} on trait expression.
-Specifically, we assume that $\beta_k$ determines how the variance
-$\tilde{V}_{z,k}$ at Hardy-Weinberg and linkage equilibrium (HWLE) in
-$k$-ploids is related to that in their diploid counterparts:
-\begin{equation}
-  \frac{\tilde{V}_{z,k}}{\tilde{V}_{z,2}} = \frac{k}{2} \beta_{k}^2
-\end{equation}
-In the discrete locus model, $\beta_k$ is interpreted as a scaler of additive
-effects in the cytotype of ploidy level $k$ (see \cref{s-sec:scaling} for
-details).
-Under these assumptions, and in the infinitesimal limit, the trait value of a
-$k$-ploid descendant from parental pair $i$ and $j$, where parent $i$
-contributes a gamete of ploidy level $g_i$ and likewise parent $j$ a gamete of
-ploidy level $g_j$, is distributed according to a Gaussian:
-\begin{equation}
-  Z_{ij} \sim \Normal\left(
-    \beta_{k} \left(
-          \frac{g_i}{c_i}\frac{z_i}{\beta_{c_i}} 
-        + \frac{g_j}{c_j}\frac{z_j}{\beta_{c_j}}\right), 
-    \beta_{k}^2 (V_{i,g_i} + V_{j,g_j})\right)
-   \label{eq:oneline}
-\end{equation}
-where the values of $V_{i,g_{i}}$ are as in \cref{tbl:segvar} (see
-\cref{s-sec:scaling}).
 
@@
     = \sum_{k=1}^2\sum_{l=1}^2 u_{c_{i},k}u_{c_{j},l}
-        \Ex\left[e^{\gamma(Z_{ij} - \theta)}|g_i=k,g_j=l\right]
+        \Ex\left[e^{\gamma(Z_{ij}^{kl} - \theta)}\right]
 \end{equation}
@@
 \centering
-\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/est1.pdf}
+\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/est1b.pdf}
 \caption{

### 2024-06-06 doc/paper/supp.tex
update

@@
 
-\subsection{Mixed-ploidy infinitesimal model}
-
-\subsubsection{Discrete locus model and variance scaling \label{sec:scaling}}
-
-Consider an $L$-locus additive model, with two alleles (0 and 1) at each locus.
-For a $k$-ploid individual, let $X_{i,j}$ be the allele at homolog $j$ of locus
-$i$.
-We assume the trait value is determined by
-\begin{equation}
-  z = \sum_{i=1}^L\sum_{j=1}^k a_{i,k} X_{i,j}
-\end{equation}
-Where $a_{i,k}$ is the allelic effect of the 1 allele at locus $i$ in
-$k$-ploids.
-The genetic variance at HWLE in $k$-ploids ($\tilde{V}_{z,k}$) will then be
-\begin{equation}
-  \tilde{V}_{z,k} = k\sum_{i=1} a_{i,k}^2 p_iq_i = kV_{x,k}
-\end{equation}
-where we refer to $V_{x,k}$ as the variance associated with a haploid genome in
-$k$-ploids at HWLE.
-Note that we also have $\tilde{V}_{z,2} = 2V_{x,k} = 2V$, where $V$ is the
-segregation variance in the diploid population.
-
-We now assume $a_{i,k} = \beta_k a_{i,2}$, i.e. allelic effects in $k$-ploids
-are as in diploids, but scaled by a factor $\beta_k$.
-This implies that
-\begin{equation}
-  \frac{\tilde{V}_{z,k}}{\tilde{V}_{z,2}} 
-  = \frac{kV_{x,k}}{2V_{x,2}} = \frac{k}{2} \beta_{k}^2
-\end{equation}
-and hence also that $V_{x,k} = \beta_k^2 V_{x,2} = \beta_k^2 V$.
-These relations will also hold in the infinitesimal limit.
-Below, we derive the segregation variance expressions for the different meiotic
-processes in the mixed-ploidy models in terms of the $V_{x,k}$.
-Using the relationships outlined here, we will then be able to express all the
-required variances in terms of $V$.
-
-Note that under the above assumptions, not only the variance is rescaled, but
-also the expected value of offspring within a family.
-Specifically, for a $k$-ploid offspring of parental pair $i$ and $j$, where the
-gamete coming from $i$ is $g_i$-ploid and the other gamete $g_j$-ploid, we have
-\begin{equation}
-  \Ex[Z_{ij}] = \beta_k \left(\frac{g_i}{c_i}\frac{z_i}{\beta_{c_i}} + 
-    \frac{g_j}{c_j}\frac{z_j}{\beta_{c_j}}\right)
-\end{equation}
-where $c_i$ is the ploidy level of individual $i$.
-Here we rescale the parental trait values to the diploid scale by dividing by
-$\beta_{c_i}$, sum the expected rescaled gametic values, and rescale to the
-offspring ploidy level.
-
-\subsubsection{Segregation variance \label{sec:segvar}}
+\subsection{Segregation variance expressions for the mixed-ploidy infinitesimal model \label{sec:segvar}}
 
@@
 diploids (which is halve the zygotic segregation variance $V$, by definition) as
-  $$V_{S(2,1)} = \frac{V}{2} = \sum_{i=1}^L \frac{v_{x,i}}{2} = \frac{V_{x,2}}{2}$$
-and hence $V=V_{x,2}$, where $V_{x,2}$ is as defined in \cref{sec:scaling}.
+  $$V_{S(2,1)} = \frac{V}{2} = \sum_{i=1}^L \frac{v_{x,i}}{2} = \frac{\Vx_{2}}{2}$$
+and hence $V=\Vx_{2}$, where $\Vx_{2}$ is the genetic variance associated with
+a haploid genome in the diploid cytotype, as defined in the main text.
 This will hold in the infinitsimal limit, where $L$ becomes very large and
@@
 $k$-ploid individual producing an $l$-ploid $V_{S(k,l)}$ in terms of the
-variance $V_{x,k}$ associated with a haploid genome in a (hypothetical)
-$k$-ploid reference population at HWLE. Using the scaling assumptions above,
-i.e. $V_{x,k} = \beta_k^2 V$, we can then express all segregation variances in
-terms of the diploid segregation variance in the reference population $V$.
+variance $\Vx_{k}$ associated with a haploid genome in a (hypothetical)
+$k$-ploid reference population at HWLE. Using the scaling assumptions outlined
+in the main text, i.e. $\Vx_{k} = \beta_k^2 V$, we can then express all
+segregation variances in terms of the diploid segregation variance in the
+reference population $V$.
 
@@
 \begin{equation}
-V_{S(4,2)} = (1-\alpha_4)V_{x,4} + 3\alpha_4 V_{x,4} = V_{x,4}(1+2\alpha_4).
+V_{S(4,2)} = (1-\alpha_4)\Vx_{4} + 3\alpha_4 \Vx_{4} = \Vx_{4}(1+2\alpha_4).
 \label{eq:dr}
 \end{equation}
-where, again, $V_{x,4}$ is as defined in \cref{sec:scaling}.
+where, again, $\Vx_{4}$ is the genetic variance associated with a haploid
+genome in the (hypothetical) tetraploid reference population, as defined in the
+main text.
 
@@
 $Y = X_1+X_2$, so that the segregation variance is 0.
-Summing across independent loci, we have $V_{S(2,2)} = 2\alpha_2 V_{x,2}$.
+Summing across independent loci, we have $V_{S(2,2)} = 2\alpha_2 \Vx_{2}$.
 
@@
 \end{align*}
-Summing over many loci, we have $V_{S(3,1)} = \frac{2}{3}V_{x,3}$.
+Summing over many loci, we have $V_{S(3,1)} = \frac{2}{3}\Vx_{3}$.
 When a triploid produces a diploid gamete, we assume there is, as in diploids
@@
 a diploid gamete from a triploid individual would be
-  $$V_{S(3,2)} = \frac{8}{3}V_{x,3}\alpha_3 + \frac{2}{3}V_{x,3} (1-\alpha_3) 
-            = \frac{2}{3}V_{x,3}(1 + 3\alpha_3)$$
+  $$V_{S(3,2)} = \frac{8}{3}\Vx_{3}\alpha_3 + \frac{2}{3}\Vx_{3} (1-\alpha_3) 
+            = \frac{2}{3}\Vx_{3}(1 + 3\alpha_3)$$
 

### 2024-12-17 doc/_paper/paper.tex
update

@@
 
-We first develop the infinitesimal model (in the sense of \cite{barton2017},
-i.e.~the `Gaussian descendants' approximation \citep{turelli2017}) for an
-autotetraploid population. We then consider mixed-ploidy populations in which
-diploids, triploids and tetraploids coexist and interbreed through the
-production of haploid and diploid gametes that combine freely.
+% Should connect more with literature and concepts of isolating mechanisms pre
+% vs. postzygotic isolation, and reinforcement etc. (see e.g. Parvathy &
+% Himani's paper introduction)
 
-{\subsection{The infinitesimal model}\label{the-infinitesimal-model}}
+Many plant species exhibit ploidy variation
+\citep{levin2002,soltis2007,rice2015}, and many of these \textit{mixed-ploidy}
+species have populations in which different cytotypes coexist or form contact
+zones \citep{kolar2017}.
+How such mixed-ploidy populations emerge and are maintained has proven somewhat
+challenging to understand.
+
+Consider for instance a randomly mating diploid population.
+Under the commonly accepted view that polyploids mostly emerge through the
+union of unreduced gametes \citep{bretagnolle1995,herben2016,kreiner2017b}, a
+new tetraploid individual originating by a chance encounter of two unreduced
+diploid gametes (an event occurring at an appreciable rate; \cite{kreiner2017})
+is highly unlikely to establish a stable tetraploid subpopulation, as most of
+its gametes will end up in unfit hybrids of odd ploidy level (triploid block,
+\cite{ramsey1998,kohler2010,brown2024}).
+This negative frequency dependence effect is commonly referred to as
+\textit{minority cytotype exclusion} (MCE), after \cite{levin1975}.
+It is well-appreciated that, due to MCE, in a large random mating mixed-ploidy
+population dominated by diploids, the rate of unreduced gamete formation needs
+to be extraordinarily high for tetraploids to establish (\cite{felber1997}, see
+also \cref{s-sec:det}).
+
+Hence, to explain the widespread occurrence of mixed-ploidy populations,
+additional factors besides the continuous formation of polyploids through the
+union of unreduced gametes need to be considered.
+Firstly, chance establishment of tetraploids through drift could occur.
+Indeed, the problem is somewhat analogous to the spread of underdominant
+chromosomal rearrangements, where local establishment through genetic drift and
+subsequent spreading in a subdivided population through local extinction and 
+recolonization dynamics has been suggested as a plausible model
+\citep{lande1985}.
+However, MCE is quite strong in randomly mating mixed-ploidy populations, and
+the population size has to be very small for local tetraploid establishment to
+occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
+Secondly, any form of prezygotic isolation between cytotypes could
+promote establishment of polyploid cytotypes by alleviating MCE.
+Particularly relevant are assortative mating by cytotype  (for instance through
+phenological differences across cytotypes, or differences in pollinators;
+\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, and
+localized dispersal \citep{baack2005,kolar2017}.
+Finally, selection may be invoked to explain the establishment of polyploids.
+Tetraploids may have higher relative fitness than their diploid counterparts
+due to reduced inbreeding depression \citep{ronfort1999}, or due to being
+better adapted to (changing) environmental conditions \citep{vandepeer2021}.
+
+None of these factors is likely to explain by itself the establishment of
+polyploids, and the consensus in the field appears to be that some mix of the
+above is required to explain the occurrence of mixed-ploidy populations in
+nature \citep{kolar2017,mortier2024}.
+In particular, polyploids are thought to establish mainly in novel, unoccupied
+habitats where they evade MCE (for instance at range edges, or after local
+extinction due to environmental change).
+However, such peripheral habitats are likely to present adaptive challenges to
+establishment \citep{kawecki2008}, and if polyploids are able to colonize such
+habitats at an appreciable rate, they must somehow be better adapted to local
+conditions, or more able to adapt to those conditions despite inbreeding and
+migration from the source population while the population is still small,
+compared to diploids.
+Indeed, when the source population is dominated by diploids, the probability
+that a migrant individual is tetraploid will be very small ($O(u^2)$ if $u$ is
+the probability that a diploid in the source population produces an unreduced
+gamete in meiosis), so that tetraploids need a substantial advantage during
+colonization if they are to establish before diploids do.
+
+While there have been substantial modeling efforts aimed at understanding
+autotetraploid establishment within diploid populations (e.g. \cite{levin1975,
+felber1991, felber1997, rausch2005, oswald2011, clo2022c, griswold2021}), the
+problem of polyploid establishment in a novel habitat which presents some adaptive
+challenge remains largely unaddressed, despite its centrality to verbal
+arguments about the establishment of polyploids in natural populations
+\citep{kolar2017, vandepeer2021, clo2022d}.
+
+Here we develop a model for the establishment of a mixed-ploidy population in a
+novel, unoccupied habitat based on \cite{barton2018}.
+In order to establish in the novel habitat, the population has to adapt to
+local environmental conditions.
+We assume fitness is determined by directional selection on a single polygenic
+trait which can be interpreted as log fitness at low density in the new
+habitat.
+As in \cite{barton2018}, we assume the trait follows the infinitesimal
+model (\textit{sensu} \cite{barton2017}, i.e. the `Gaussian descendants'
+infinitesimal model \citep{turelli2017}).
+We extend the infinitesimal model, and the approach for exact simulation of
+trait evolution under the infinitesimal model, to mixed-ploidy populations.
+We then use simulations to study tetraploid establishment, both from single
+migrants and under continuous migration from a predominantly diploid source
+population, examining the effects of autopolyploid genetics,
+maladaptive migration, selfing and assortative mating on the probability that
+autotetraploids establish in the novel habitat.
+
+\section*{Model and Methods}
+
+\begin{table}[t]
+\caption{Glossary of the main notation used in the main text.
+} \label{tbl:glossary}
+\centering
+\small
+\begin{tabularx}{\linewidth}{lX}
+\cline{1-2}
+\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
+$N$ & total population size\\
+$N_k$ & population size of the $k$-ploid cytotype \\
+$\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
+$u$ & probability of unreduced gamete formation ($u=u_{22}=u_{32}=1-u_{42}$)\\
+$v$ & probability that a triploid produces a haploid/diploid gamete
+  ($v=u_{31}=u_{32}$)\\
+$m$ & expected number of migrants per generation arriving in the new habitat \\
+$z_i$ & trait value of individual $i$ \\
+$c_i$ & ploidy level of individual $i$ \\
+$g_i$ & ploidy level of gamete produced by individual $i$ in a particular cross\\
+$V$ & segregation variance in the reference diploid population \\
+$V_{i,k}$ & gametic segregation variance associated with the production of a
+  $k$-ploid gamete by individual $i$ \\
+$\Vx_k$ & genetic variance associated with a haploid genome in the $k$-ploid
+  reference population (i.e. a $k$-ploid non-inbred population at HWLE) \\
+$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
+$F_i$ & inbreeding coefficient in individual $i$ \\
+$\Phi_{ij}$ & coancestry coefficient for individuals $i$ and $j$ \\
+$\alpha_k$ & probability that the two genes at a locus in a diploid gamete
+  formed by a $k$-ploid individual descend from the same parental gene copy\\
+$\gamma$ & strength of directional selection in the new habitat\\
+$\theta$ & trait value beyond which the growth rate becomes positive in the new
+habitat \\ $w_{ij}$ & fitness of parental pair $(i,j)$ \\
+$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
+$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
+$\sigma_k$ & selfing rate in $k$-ploids \\
+$\rho_k$ & probability of assortative mating in $k$-ploids \\
+\cline{1-2}
+\end{tabularx}%
+\end{table}
+
+\subsection*{Mixed-ploidy population model}
+
+We consider a mixed-ploidy population of size $N$ consisting of $N_2$ diploid,
+$N_3$ triploid and $N_4=N-N_2-N_3$ tetraploid individuals.
+We assume an individual of ploidy level $k$ forms haploid and diploid gametes
+with proportions $u_{k1}$ and $u_{k2}$, as well as a proportion
+$1-u_{k1}-u_{k2}$ inviable (e.g. aneuploid or polyploid) gametes.
+The (relative) fecundity of a $k$-ploid individual is hence $u_{k1} + u_{k2}$.
+Unless stated otherwise, we will assume 
+\begin{equation}
+    \begin{pmatrix} 
+    u_{21} & u_{22} \\ 
+    u_{31} & u_{32} \\ 
+    u_{41} & u_{42} 
+    \end{pmatrix} =
+    \begin{pmatrix} 
+    1-u & u \\
+    v & v \\
+    0 & 1-u
+    \end{pmatrix} \label{eq:U}
+\end{equation}
+where $u$ is referred to as the proportion of unreduced gametes, and $2v$ is
+the proportion of euploid gametes produced by a triploid individual.
+
+When two individuals mate, we assume they produce gametes according to their
+ploidy level (\cref{eq:U}), which randomly combine to produce offspring (which
+may be inviable if one of the contributing gametes is inviable).
+Intrinsic fitness disadvantages associated with particular zygotic ploidy
+levels or cross types (e.g. modeling phenomena such as `triploid block') can be
+straighforwardly included at this level.
+An analysis of a deterministic model (i.e. where $N \rightarrow \infty$) for
+the cytotype dynamics and equilibrium cytotype composition under random mating
+is included in \cref{s-sec:det} (see also \cite{felber1997,kauai2024}).
+The stochastic version for finite and constant $N$ is analyzed briefly in
+\cref{s-sec:mchain}.
+
+\subsection*{Infinitesimal model}
+
+\paragraph*{The basic infinitesimal model.}
 
@@
 The infinitesimal model approximates the inheritance of such a trait by
-assuming that the trait value \(Z_{ij}\) of a random offspring from parents
-with trait values \(z_i\) and \(z_j\) follows a Gaussian distribution with mean
-equal to the midparent value and variance which is independent of the mean:
+assuming that the trait value $Z_{ij}$ of a random offspring from parents with
+trait values $z_i$ and $z_j$ follows a Gaussian distribution with mean equal to
+the midparent value and variance which is independent of the mean:
   \begin{align}
-  Z_{ij} \sim \Normal\left(\frac{z_i + z_j}{2}, V_{ij}\right) \label{eq:inf}
+  Z_{ij} \sim \Normal\left(\frac{z_i + z_j}{2}, V_{ij}\right)
+  \label{eq:inf}
   \end{align}
-Here, \(V_{ij}\) is referred to as the \emph{segregation variance} in family
-\((i,j)\).
+Here, $V_{ij}$ is referred to as the \textit{segregation variance} in family
+$(i,j)$.
 This is the variation generated among offspring from the same parental pair due
@@
 
-An equivalent, and for our purposes useful, way to characterize the model in
-diploids and polyploids is to write \(Z_{ij} = Y_i + Y_j\), where \(Y_i\) and
-\(Y_j\) are independent Gaussian random variables \(Y_i \sim
-\Normal\left(\frac{z_i}{2}, V_i\right)\) (and similarly for \(Y_j\)).
-We refer to \(Y_i\) as the (random) gametic value of individual \(i\), and to
-\(V_i\) as the \emph{gametic segregation variance} of individual \(i\)
-(clearly $V_{ij} = V_i + V_j$).
+An alternative, and for our purposes useful, way to characterize the model
+is to write $Z_{ij} = Y_i + Y_j$, where $Y_i$ and $Y_j$
+are independent Gaussian random variables $Y_i \sim \Normal\left(\frac{z_i}{2},
+V_i\right)$ (and similarly for $Y_j$).
+We refer to $Y_i$ as the (random) \textit{gametic value} of individual $i$, and
+to $V_i$ as the \textit{gametic segregation variance} of individual $i$.
 This formulation is helpful in that it highlights that Mendelian segregation
@@
 combine additively to determine the offspring trait value.
+This model applies readily to an autopolyploid population expressing a trait
+with infinitesimal genetics.
+For instance, when an autotetraploid produces a diploid gamete, it will pass on
+half its trait value to the gamete on average, with a variance determined by
+the details of tetraploid meiosis (which are not, in general, identical to
+those of diploid meiosis, see below).
 
-If we assume a population consisting of infinitely many unrelated
-individuals, expressing a trait with genetic variance \(V_z\) and
-segregation variance \(V_0\), we find that under random mating the
-variance in the offspring generation is
+Note that in a finite population, the segregation variance will decay over time
+as the population becomes more inbred.
+Indeed, Mendelian segregation generates less variation among the gametes
+produced by individual $i$ when that individual is more inbred, as segregation
+at homozygous loci does not generate any variation.
+When $F_i$ is the inbreeding coefficient relative to some ancestral reference
+population with gametic segregation variance $V$ (i.e. the probability that two
+genes at a locus in individual $i$ sampled without replacement are identical by
+descent), the gametic segregation variance of individual $i$ will be
+$V_i = (1-F_i)V$.
+This holds for both diploids and tetraploids (\cref{s-sec:tetinbred}, also
+\cite{moody1993}).
+
+\paragraph{Scaling of traits across ploidy levels.}
+
+If we would somewhat naively assume that the allelic effects underlying an
+additive quantitative trait are identical across ploidy levels, 
+a tetraploid offspring from a cross between two diploids would yield, on average,
+a trait value which is the sum of the parental trait values, with a
+distribution that depends on the variance generated by the process producing
+unreduced gametes.
+This is not likely to reflect biological reality: tetraploids do not tend to
+have, for instance, twice the size of their diploid progenitors on average.
+Similarly, the genetic variance at Hardy-Weinberg and linkage equilibrium
+(HWLE) in tetraploids will be twice that of their diploid counterparts under
+such assumptions, which is similarly unrealistic \citep{clo2022}.
+
+In order to account for this, we introduce a scaling factor $\beta_k$,
+accounting for the effects of polyploidization \textit{per se} on trait
+expression in $k$-ploids.
+To introduce and interpret this parameter, we consider an $L$-locus additive
+model, with two alleles (0 and 1) at each locus.
+For a $k$-ploid individual, let $X_{i,j}$ be the allele at homolog $j$ of locus
+$i$.
+We assume the trait value is determined by
+\begin{equation}
+  z = \sum_{i=1}^L\sum_{j=1}^k a_{i,k} X_{i,j}
+\end{equation}
+Where $a_{i,k}$ is the allelic effect of the 1 allele at locus $i$ in
+$k$-ploids.
+The genetic variance at HWLE in $k$-ploids ($\tilde{V}_{z,k}$) will then be
+\begin{equation}
+  \tilde{V}_{z,k} = k\sum_{i=1}^L a_{i,k}^2 p_iq_i = k\Vx_{k}
+\end{equation}
+where we refer to $\Vx_{k}$ as the variance associated with a haploid genome in
+$k$-ploids at HWLE.
+Note that we also have $\tilde{V}_{z,2} = 2\Vx_{2} = 2V$ \citep{barton2017},
+where $V$ is the segregation variance in the diploid population.
+If we now assume $a_{i,k} = \beta_k a_{i,2}$, i.e. allelic effects in $k$-ploids
+are as in diploids, but scaled by a factor $\beta_k$.
+This implies that
+\begin{equation}
+  \frac{\tilde{V}_{z,k}}{\tilde{V}_{z,2}} 
+  = \frac{k\Vx_{k}}{2\Vx_{2}} = \frac{k}{2} \beta_{k}^2
+\end{equation}
+and hence also that $\Vx_{k} = \beta_k^2 \Vx_{2} = \beta_k^2 V$.
+These relations will also hold in the infinitesimal limit where $a_{i,2}
+\rightarrow 0$ as $L\rightarrow \infty$.
+
+\paragraph{Mixed-ploidy infinitesimal model.}
+
+We can extend the infinitesimal model to the mixed-ploidy case, assuming that the
+gametic value, on the diploid trait scale, associated with a $k$-ploid gamete ($k
+\in \{1,2\}$) from individual $i$ of ploidy level $c_i \in \{2,3,4\}$ is a
+Gaussian random variable $Y_{i,k}$ with distribution
+  $$Y_{i,k} \sim \Normal\left(\frac{k}{c_i}\frac{z_i}{\beta_{c_i}}, V_{i,k}\right)$$
+where $V_{i,k}$ is the gametic segregation variance associated with the
+production of a $k$-ploid gamete by individual $i$.
+The trait value of an individual originating from the union of a $k$-ploid
+gamete of individual $i$ and an $l$-ploid gamete from individual $j$ is then
+  $$Z_{ij}^{kl} = \beta_{k+l}\left(Y_{i,k} + Y_{j,l}\right)$$
+i.e., $Z_{ij}^{kl}$ is a Gaussian random variate with distribution
+\begin{equation}
+  Z_{ij}^{kl} \sim \Normal\left(
+    \beta_{k+l} \left(
+          \frac{k}{c_i}\frac{z_i}{\beta_{c_i}} 
+        + \frac{l}{c_j}\frac{z_j}{\beta_{c_j}}\right), 
+    \beta_{k+l}^2 (V_{i,k} + V_{j,l})\right)
+   \label{eq:oneline}
+\end{equation}
+
+For the production of diploid gametes by a $k$-ploid individual, the
+segregation variance depends not only on the segregation variance in the base
+population ($V$) and the inbreeding coefficient ($F$), but also on the detailed
+assumptions on how these aberrant meiotic processes occur.
+Importantly however, these details only affect the gametic segregation variance
+through the quantity $\alpha_k$, which is the probability that a $k$-ploid
+transmits two copies of the same homolog to a diploid gamete.
+Note that $\alpha_4$, the probability that a diploid gamete of a tetraploid
+individual carries two copies of the same homolog, is the probability of
+\textit{double reduction} (e.g. \cite{lynch1998} p.57), and is upper bounded by
+1/6 \citep{stift2008}.
+The value of $\alpha_2$ depends on the relative frequency of unreduced gamete
+formation through so-called \textit{first} and \textit{second division
+restitution} \citep{bretagnolle1995,storme2013}.
+We summarize the expressions for the gametic segregation variance in
+\cref{tbl:segvar}.
+Detailed derivations can be found in \cref{s-sec:segvar}.
+
+\begin{table}[t]
+\caption{Gametic segregation variance for haploid and diploid gametes produced
+by the three cytotypes in the mixed-ploidy model. $F_i$ is the inbreeding
+coefficient of individual $i$ (producing the gamete), whereas $\alpha_k$ is the
+probability that a diploid gamete from a $k$-ploid individual receives two
+copies of the same parental gene. Note that we have $\alpha_3 \le 1/4$ and
+$\alpha_4 \le 1/6$.
+} \label{tbl:segvar}
+\centering
+\small
+\begin{tabularx}{\linewidth}{XXX}
+\cline{1-3}
+\textbf{cytotype}   & \textbf{haploid gamete variance} & \textbf{diploid gamete
+variance}        \\ \cline{1-3} \\[-2.5ex]
+diploid    & $\frac{1}{2}(1-F_i)V$    & $2\alpha_2(1-F_i)V$  \\ \\[-2.5ex]
+triploid   & $\frac{2}{3}(1-F_i)V$     & $\frac{2}{3}(1 + 3\alpha_3)(1-F_i)V$ \\ \\[-2.5ex]
+tetraploid & --                      & $(1+2\alpha_4)(1-F_i)V$             \\
+\cline{1-3}
+\end{tabularx}%
+\end{table}
+
+\paragraph{Recursions for inbreeding coefficients}
+
+We can simulate the mixed-ploidy infinitesimal model for a finite population
+through a straightforward extension of the approach outlined in
+\cite{barton2017}, provided we can efficiently track inbreeding and coancestry
+coefficients across the different ploidy levels.
+Denoting the parents of individual $i$ by $k$ and $l$, the recursion for the
+inbreeding coefficients in the mixed-ploidy case becomes
 \begin{align}
-V_z' &= \Ex[\Var[Z_{ij}|Z_i, Z_j]] + \Var[\Ex[Z_{ij}|Z_i,Z_j]] \nonumber \\
-   &= \Ex[V_{ij}] + \Var\left(\frac{Z_i + Z_j}{2}\right) \nonumber \\
-   &= V_0 + \frac{V_z}{2}\ .
-   \label{eq:eq}
+    F_i &= \Phi_{kl} & \text{if } & c_i = 2 \nonumber \\ 
+    F_i &= \frac{1}{3}\left(F_k^\ast + 2\Phi_{kl}\right) & \text{if } 
+        & c_i = 3, g_k = 2, g_l = 1 \nonumber \\ 
+    F_i &= \frac{1}{3}\left(F_l^\ast + 2\Phi_{kl}\right) & \text{if } 
+        & c_i = 3, g_k = 1, g_l = 2 \nonumber \\ 
+    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{if } & c_i = 4
 \end{align}
-So that at equilibrium (\(V_z'=V_z\)), we have \(V_z = 2V_0\) \citep{barton2017}.
+where $F_k^\ast = \alpha_{c_k} + (1-\alpha_{c_k})F_k$ (\cref{s-sec:tetinbred}).
+The recursion for the coancestry coefficients in is given by
+\begin{align}
+    \Phi_{ii} &= \frac{1}{c_{i}} \left(1 + (c_i-1)F_i\right) \nonumber \\
+    \Phi_{ij} &= \sum_k \sum_l P_{ik}P_{jl} \Phi_{kl} & i \ne j 
+    \label{eq:coancestry}
+\end{align}
+where the sums are over individuals in the parental population, and where
+$P_{ik} \in \{0, \frac1 3, \frac1 2, \frac2 3, 1\}$ is the probability that a
+gene copy in $i$ is derived from parent $k$.
 
-Denoting by \(X\) the contribution to the trait value associated with a
-randomly sampled haploid genome from the population, we also have that, at
-equilibrium, \(V_z=2V_0=mV\), where \(V = \Var[X]\) and $m$ is the ploidy
-level.
-Note that no assumptions regarding $m$ have been made in the definition of the
-model.
-Indeed, as long as all individuals are of the same \emph{even} ploidy level,
-\cref{eq:inf} and \cref{eq:eq} apply equally well to the inheritance of an
-additive trait in polyploids as in diploids.
-Two factors however do lead to differences with the standard diploid model due
-to their effect on \(V_i\):
-(1) the potential occurrence of double reduction in polyploids affects the
-relationship between \(V\) and the segregation variance and 
-(2) the evolution of the segregation variance over time in finite populations
-is affected by the ploidy level.
+\subsection*{Establishment model}
 
-\textbf{Perhaps we should organize the text around the table below. First
-outline mixed-ploidy population, then indicate that we need the variances
-associated with the different gametes that are produced. Gamete formation
-occurs independently, so that is all we need. So start with the infinitesimal
-model framed in terms of uniting gametes, and highlight that this works for any
-ploidy level, and mixed-ploidy populations as long as we can work out the
-segregation variances.}
+Our model for the establishment of a population in an initially unoccupied
+habitat is based on \cite{barton2018}.
+We assume a large non-inbred `mainland' population at HWLE and cytotype
+equilibrium, with $\Ex[z] = 0$ irrespective of the cytotype.
+In generation $t$, $M(t)$ migrant individuals arrive on an island (the new
+habitat) joining $N^\ast(t)$ resident individuals, where $M(t)$ is Poisson
+distributed with mean $m$. 
+The migrant individuals are assumed to be unrelated to the resident
+individuals.
+After migration in generation $t$, the $N(t) = N^\ast(t) + M(t)$ individuals
+reproduce sexually, and the offspring thus produced survives until the next
+generation with a probability determined by their trait value.
+In the basic model, random selfing is allowed (but see below for a model with
+self-incompatibility).
+We assume the trait is under directional selection, with fitness is $w(z) =
+e^{\gamma(z - \theta)}$, where $\gamma$ is the intensity of directional
+selection and $\theta$ is the trait value for which the growth rate of the
+population becomes positive.
 
-\begin{table}[]
-\begin{tabular}{l|rr}
-cytotype   & haploid gamete variance & diploid gamete variance             \\ \cline{1-3}
-diploid    & $\frac{1}{2}(1- F)V$    & $2\alpha_2(1-F)V$  \\
-triploid   & $\frac{2}{3}(1-F)V$     & $\frac{2}{3}(1 + \alpha_3)(1-F)V$   \\
-tetraploid & --                      & $(1+2\alpha_4)(1-F)V$             
-\end{tabular}%
-\end{table}
+Again following \cite{barton2018}, we simulate the model by first calculating
+the fitness of each parental pair $(i,j)$, which is the expected fitness of
+offspring of this pair
+\begin{equation}
+  w_{ij}
+    = \sum_{k=1}^2\sum_{l=1}^2 w_{ij}^{kl}
+    = \sum_{k=1}^2\sum_{l=1}^2 u_{c_{i},k}u_{c_{j},l}
+        \Ex\left[e^{\gamma(Z_{ij}^{kl} - \theta)}\right]
+\end{equation}
+The expectation on the right hand side can be calculated from \cref{eq:oneline}
+using the moment-generating function of the Gaussian.
+Having calculated the $w_{ij}$, the number of offspring surviving into the next
+generation is calculated as $N^\ast(t+1) = \sum_{i,j}w_{ij}/N(t)$.
+Next, $N^\ast(t+1)$ offspring individuals are sampled by sampling parental
+pairs and gametes proportional to $w_{ij}^{kl}$, and sampling a trait value
+in accordance with \cref{eq:oneline}.
+
+\subsection*{Self-fertilization and assortative mating}
+
+We model partial self-fertilization by assuming that a proportion
+$\sigma_{c_i}$ of the ovules of individual $i$ with ploidy level $c_i$ are
+fertilized by self-pollen, while the remaining proportion $1-\sigma_{c_i}$ are
+fertilized by randomly sampled pollen (which may be self-pollen with
+probability $1/N$). 
+That is, the expected number of offspring from individual $i$ as mother
+surviving after selection is
+\begin{equation}
+\Ex[w_i] = \sigma_{c_i} w_{ii} +
+  (1-\sigma_{c_i})\left[\frac{1}{N}\sum_{j=1}^N w_{ij}\right]
+\end{equation}
+We hence assume no pollen limitation (all outcrossing ovules are fertilized),
+and no pollen discounting (the probability of being a father is unaffected by
+an individual's selfing rate).
+When modeling self-incompatibility, we assume there is no intrinsic
+disadvantage to self-incompatibility, except when there is only a single
+individual in the population, i.e.
+\begin{equation}
+\Ex[w_i] = \begin{cases}
+    \frac{1}{N-1}\sum_{j \ne i }w_{ij} & \text{if } N > 1\\ 
+    0 & \text{if } N=1 \end{cases}
+\end{equation}
+
+We model assortative mating by ploidy level in a similar way, assuming that a
+fraction $\rho_{c_i}$ of the ovules of individual $i$ are fertilized by pollen
+sampled from the $c_i$-ploid portion of the population, while a fraction
+$1-\rho_{c_i}$ is fertilized by pollen randomly sampled from the entire
+population.
+\begin{equation}
+\Ex[w_i] = \rho_{c_i} \frac{1}{N_{c_{i}}} \sum_{j=1}^N \delta_{c_i,c_j}w_{ij}
+ + (1-\rho_{c_i})
+\left[\frac{1}{N}\sum_{j=1}^N w_{ij}\right]
+\end{equation}
+
+\subsection*{Implementation and availability}
+
+Individual-based simulations for the mixed-ploidy infinitesimal model were
+implemented in Julia \citep{julia}.
+Documented code and simulation notebooks are available at
+\url{https://github.com/arzwa/InfGenetics}.
 

### 2024-12-17 doc/paper/paper.tex
update

@@
-\documentclass[11pt,a4paper]{article}
+\documentclass[12pt,a4paper]{article}
 \input{preamble.tex}
-\usepackage{geometry}
+\usepackage[margin=1in]{geometry}
 \usepackage{authblk}
@@
 \title{Autopolyploid establishment through polygenic adaptation}
-\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@univ-lille.fr}}
-\affil[1]{\footnotesize University of Lille, CNRS, UMR 8198 -- Evo-Eco-Paleo,
-F-59000 Lille, France}
+\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
+\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
 \date{\vspace{-5ex}}
@@
 
+% Should connect more with literature and concepts of isolating mechanisms pre
+% vs. postzygotic isolation, and reinforcement etc. (see e.g. Parvathy &
+% Himani's paper introduction)
+
 Many plant species exhibit ploidy variation
@@
 Under the commonly accepted view that polyploids mostly emerge through the
-union of unreduced gametes \citep{herben2016,kreiner2017b}, a new tetraploid
-individual originating by a chance encounter of two unreduced diploid gametes
-(an event occurring at an appreciable rate, \citep{kreiner2017}) is highly
-unlikely to establish a stable tetraploid subpopulation, as most of its gametes
-will end up in unfit hybrids of odd ploidy level (triploids).
+union of unreduced gametes \citep{bretagnolle1995,herben2016,kreiner2017b}, a
+new tetraploid individual originating by a chance encounter of two unreduced
+diploid gametes (an event occurring at an appreciable rate; \cite{kreiner2017})
+is highly unlikely to establish a stable tetraploid subpopulation, as most of
+its gametes will end up in unfit hybrids of odd ploidy level (triploid block,
+\cite{ramsey1998,kohler2010,brown2024}).
 This negative frequency dependence effect is commonly referred to as
@@
 union of unreduced gametes need to be considered.
-On the one hand, chance establishment of tetraploids through drift could occur.
+Firstly, chance establishment of tetraploids through drift could occur.
+Indeed, the problem is somewhat analogous to the spread of underdominant
+chromosomal rearrangements, where local establishment through genetic drift and
+subsequent spreading in a subdivided population through local extinction and 
+recolonization dynamics has been suggested as a plausible model
+\citep{lande1985}.
 However, MCE is quite strong in randomly mating mixed-ploidy populations, and
-the population size has to be very small for tetraploid establishment to occur
-at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
-On the other hand, any form of prezygotic isolation between cytotypes could
+the population size has to be very small for local tetraploid establishment to
+occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
+Secondly, any form of prezygotic isolation between cytotypes could
 promote establishment of polyploid cytotypes by alleviating MCE.
 Particularly relevant are assortative mating by cytotype  (for instance through
-phenological differences across cytotypes, or differences in pollinators,
-\citep{kolar2017}), self-fertilization \citep{rausch2005}, and localized
-dispersal \citep{baack2005,kolar2017}.
+phenological differences across cytotypes, or differences in pollinators;
+\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, and
+localized dispersal \citep{baack2005,kolar2017}.
 Finally, selection may be invoked to explain the establishment of polyploids.
@@
 extinction due to environmental change).
-If they are able to colonize such habitats at an appreciable rate, they must
-somehow be better adapted to local conditions, or more able to adapt to those
-conditions despite inbreeding and migration from the source population while
-the population is still small, compared to diploids.
-Indeed, when the source population is dominated by diploids, the number of
-tetraploid migrants arriving in the novel habitat will be almost negligible
-($O(u^2)$ if $u$ is the probability that a diploid produces an unreduced
-gamete in meiosis) compared to the number of diploid migrant individuals, so
-that tetraploids need a substantial advantage during colonization if they are
-to establish with reasonable probability.
+However, such peripheral habitats are likely to present adaptive challenges to
+establishment \citep{kawecki2008}, and if polyploids are able to colonize such
+habitats at an appreciable rate, they must somehow be better adapted to local
+conditions, or more able to adapt to those conditions despite inbreeding and
+migration from the source population while the population is still small,
+compared to diploids.
+Indeed, when the source population is dominated by diploids, the probability
+that a migrant individual is tetraploid will be very small ($O(u^2)$ if $u$ is
+the probability that a diploid in the source population produces an unreduced
+gamete in meiosis), so that tetraploids need a substantial advantage during
+colonization if they are to establish before diploids do.
 
@@
 felber1991, felber1997, rausch2005, oswald2011, clo2022c, griswold2021}), the
-problem of polyploid establishment in a novel habitat presenting some adaptive
+problem of polyploid establishment in a novel habitat which presents some adaptive
 challenge remains largely unaddressed, despite its centrality to verbal
@@
 model (\textit{sensu} \cite{barton2017}, i.e. the `Gaussian descendants'
-infinitesimal model).
+infinitesimal model \citep{turelli2017}).
 We extend the infinitesimal model, and the approach for exact simulation of
@@
 migrants and under continuous migration from a predominantly diploid source
-population, examining the effects of polyploid (quantitative) genetics,
+population, examining the effects of autopolyploid genetics,
 maladaptive migration, selfing and assortative mating on the probability that
-autotetraploids establish in the novel habitat before diploids do.
+autotetraploids establish in the novel habitat.
 
@@
 \textbf{notation}   & \textbf{description}   \\ \cline{1-2}
-$N, N_k$ & total population size, population size of the $k$-ploid cytotype \\
+$N$ & total population size\\
+$N_k$ & population size of the $k$-ploid cytotype \\
 $\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
@@
 Intrinsic fitness disadvantages associated with particular zygotic ploidy
-levels or cross types (e.g. modeling phenomena such as `triploid block',
-\citep{ramsey1998,brown2024}) can be straighforwardly included at this level.
+levels or cross types (e.g. modeling phenomena such as `triploid block') can be
+straighforwardly included at this level.
 An analysis of a deterministic model (i.e. where $N \rightarrow \infty$) for
@@
 \begin{equation}
-\Ex[w_i] = \sigma_{c_i} \Ex[w_{ii}] +
-  (1-\sigma_{c_i})\left[\frac{1}{N}\sum_{j=1}^N \Ex[w_{ij}]\right]
+\Ex[w_i] = \sigma_{c_i} w_{ii} +
+  (1-\sigma_{c_i})\left[\frac{1}{N}\sum_{j=1}^N w_{ij}\right]
 \end{equation}
-Where $\Ex[w_{ij}] = e^{\gamma \bar{z}_{ij} + \gamma^2 V_{ij}/2}$ is the
-expected fitness of offspring from the parental pair with $i$ as mother and $j$
-as father.
 We hence assume no pollen limitation (all outcrossing ovules are fertilized),
@@
 disadvantage to self-incompatibility, except when there is only a single
-individual in the population i.e.
+individual in the population, i.e.
 \begin{equation}
 \Ex[w_i] = \begin{cases}
-    \frac{1}{N-1}\sum_{j \ne i }\Ex[w_{ij}] & \text{if } N > 1\\ 
+    \frac{1}{N-1}\sum_{j \ne i }w_{ij} & \text{if } N > 1\\ 
     0 & \text{if } N=1 \end{cases}
@@
 \begin{equation}
-\Ex[w_i] = \sum_{j=1}^N \left(\frac{\delta_{c_i,c_j}}{N_{c_i}}\rho_{c_i} +
-\frac{1}{N}(1-\rho_{c_i})\right)\Ex[w_{ij}]
+\Ex[w_i] = \rho_{c_i} \frac{1}{N_{c_{i}}} \sum_{j=1}^N \delta_{c_i,c_j}w_{ij}
+ + (1-\rho_{c_i})
+\left[\frac{1}{N}\sum_{j=1}^N w_{ij}\right]
 \end{equation}
@@
 \subsection*{Autotetraploid and mixed-ploidy infinitesimal model}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/tetfininf.pdf}
+\caption{
+The infinitesimal model in autotetraploids.
+Comparisons are shown for the decay of the genetic variance ($V_z$) due to
+inbreeding in exact simulations of the infinitesimal model in autotetraploids
+against individual-based simulations of autotetraploid populations with $L$
+unlinked additive loci determining the quantitative trait. 
+(A) Simulations of a model without double reduction ($\alpha_4=0$).
+(B) Simulations of a model with maximal double reduction ($\alpha_4=1/6$) (for
+all loci in the finite $L$ simulations).
+We show window-smoothed values for visual clarity, with observed genetic
+variances averaged in windows of 20 generations every 10 generations.
+The black line marks $e^{-t/4N}$. 
+We assume $N=250$ and $V_z(0) = 1$.
+\label{fig:vztet}}
+\end{figure}
 
@@
 \cref{s-sec:ttdip}), so that the polyploid subpopulations will show an average
-relatedness similar to the diploid population and not evolve as an isolated
-higher-ploidy population would.
+relatedness similar to that of the diploid population and not evolve as an
+isolated higher-ploidy population would.
 A nonzero probability of producing IBD diploid gametes ($\alpha_k > 0$) will
@@
 (\cref{s-fig:f1}).
-Therefore, as long as diploids dominate, the effect of harboring some of the
-gene pool in triploid and tetraploid individuals on the rate of inbreeding is
+Therefore, as long as diploids dominate, the effect of harboring some fraction
+of the gene pool in polyploid individuals on the rate of inbreeding is
 negligible, and we find that the evolution of the inbreeding coefficient over
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/tetfininf.pdf}
-\caption{
-The infinitesimal model in autotetraploids.
-Comparisons are shown for the decay of the genetic variance ($V_z$) due to
-inbreeding in exact simulations of the infinitesimal model in autotetraploids
-against individual-based simulations of autotetraploid populations with $L$
-unlinked additive loci determining the quantitative trait. 
-(A) Simulations of a model without double reduction ($\alpha_4=0$).
-(B) Simulations of a model with maximal double reduction ($\alpha_4=1/6$) (for
-all loci in the finite $L$ simulations).
-We show window-smoothed values for visual clarity, with observed genetic
-variances averaged in windows of 20 generations every 10 generations.
-The black line marks $e^{-t/4N}$. 
-We assume $N=250$ and $V_z(0) = 1$.
-\label{fig:vztet}}
-\end{figure}
-
 \subsection*{Establishment from a single individual}
-
-Having established the validity of the mixed-ploidy infinitesimal model, we now
-use it to study the establishment of polyploids in a marginal habitat to which
-migrants from a mixed-ploidy source population are maladapted.
-
-We first consider the establishment of a population from a single migrant
-individual with trait value $z_0 = 0$.
-We assume $u=0$ and compare the probability of establishment when the migrant
-is diploid vs. tetraploid (\cref{fig:est1}).
-As noted by \cite{barton2018}, the establishment probability depends
-essentially on two dimensionless parameters, $\gamma\sqrt{2V}$ and
-$\theta/\sqrt{2V}$, corresponding to the intensity of selection and the degree of
-maladaptation, respectively.
-We shall scale our results accordingly, assuming $2V=1$ throughout.
-
-We find that reduced inbreeding in tetraploids substantially increases the
-establishment probability.
-Indeed, in the case where allelic effects are scaled so as to yield the same
-equilibrium genetic variance ($\beta = \sqrt{1/2}$), the establishment probability 
-for tetraploids can be almost five times as high as in diploids depending on
-the selection gradient (\cref{fig:est1}A).
-While the probability of establishment becomes lower as the strength of
-selection becomes large, the establishment probability does not decrease
-monotonically with $\gamma$, i.e. as selection becomes very weak, the
-establishment probability also lowers.
-This happens because, although the probability of surviving the first couple of
-generations becomes higher for weaker selection, adaptation will be slower, and
-the risk that genetic variation is exhausted due to inbreeding before the
-population is able to reach a consistently positive growth rate is increased.
-
-As expected, the scaling of the equilibrium genetic variance has a profound
-effect on the establishment probability, but only when $\beta$ is close to 0.5
-(i.e. individual alleles have almost half the effect size in tetraploids
-compared to diploids) is the benefit of the slower rate of inbreeding in
-tetraploids canceled (\cref{fig:est1}B).
 
@@
 \centering
-\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/est1b.pdf}
+\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/est1b.pdf}
 \caption{
@@
 HWLE is identical between diploids and autotetraploids.
-All results are estimated from 500.000 replicate simulations.
+(C) Probability of tetraploid establishment relative to the probability of
+diploid establishment (on a $\log_{10}$ scale) across a range of values for
+$\gamma$ and $\beta_4$. The vertical line again marks $\beta_4=\sqrt{1/2}$.
+All results are estimated from 500.000 replicate simulations and assume
+$\theta=2$.
 \label{fig:est1}}
@@
 
+Having established the validity of the mixed-ploidy infinitesimal model, we now
+use it to study the establishment of polyploids in a marginal habitat to which
+migrants from a mixed-ploidy source population are maladapted.
+
+We first consider the establishment of a population from a single migrant
+individual with trait value $z_0 = 0$.
+We assume $u=0$ and compare the probability of establishment when the migrant
+is diploid vs. tetraploid (\cref{fig:est1}).
+As noted by \cite{barton2018}, the establishment probability (for a given
+cytotype) depends essentially on two dimensionless parameters,
+$\gamma\sqrt{2V}$ and $\theta/\sqrt{2V}$, corresponding to the intensity of
+selection and the degree of maladaptation, respectively.
+We shall scale our results accordingly, assuming $2V=1$ throughout, and shall
+assume $\theta=2$, i.e. adaptation has to increase the trait value by two
+standard deviations to achieve a positive growth rate.
+
+We find that reduced inbreeding in tetraploids substantially increases the
+establishment probability.
+Indeed, in the case where allelic effects are scaled across ploidy levels
+to yield the same equilibrium genetic variance ($\beta = \sqrt{1/2}$, so that
+polyploid migrants are not more likely to have high fitness than diploid
+migrants), the establishment probability for tetraploids can be almost five
+times as high as in diploids depending on the selection gradient
+(\cref{fig:est1}A).
+As the segregation variance and initial trait value are the same across these
+simulations, this is a consequence of the reduced rate of inbreeding, which
+slows down the exhaustion of the genetic variance carried by the initial
+migrant individual.
+
+While the probability of establishment becomes lower as the strength of
+selection becomes large, the establishment probability does not decrease
+monotonically with $\gamma$, i.e. as selection becomes very weak, the
+establishment probability also decreases.
+This happens because, although the probability of surviving the first couple of
+generations becomes higher when selection is weaker, adaptation (i.e. increase
+in the trait mean) will be slower,
+and the risk that genetic variation is exhausted due to inbreeding before the
+population is able to reach a consistently positive growth rate is increased as
+a result.
+
+Evidently, the scaling of the genetic variance across ploidy levels has a
+profound effect on the establishment probability, but only when $\beta$ is
+close to 0.5 (i.e. individual alleles have almost half the effect size in
+tetraploids compared to diploids) is the benefit of the slower rate of
+inbreeding in tetraploids canceled (\cref{fig:est1}B,C).
+
 \subsection*{Establishment under recurrent migration}
-
-We next consider establishment in the new habitat when there is a continuous
-influx of migrants coming from a large diploid-dominated mixed-ploidy source
-population at cytotype equilibrium.
-In this case, establishment is certain to happen eventually, and we are
-interested in the probability that a tetraploid population established before a
-diploid one does.
-
-We hypothesized that two counteracting processes affect the probability of
-autotetraploid establishment in this case.
-On the one hand, increased migration will increase the probability that an
-otherwise likely successful tetraploid migrant suffers from MCE in the early
-generations while the population size is low, because migrants are likely to be
-diploid.
-On the other hand, tetraploids are strongly reproductively isolated from
-the typical migrant, so that they are less prone to maladaptive swamping.
-Hence, conditional on evading MCE, they should be able to adapt to the new
-habitat at a rate which is not strongly affected by the migration rate.
 
@@
 \centering
-\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration1.pdf}
+\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration1b.pdf}
 \caption{
@@
 (B) Proportion of simulation replicates in which tetraploids established. 
-The shaded area corresponds to Jeffreys' 95\% interval.
-The horizontal dashed line marks the proportion of tetraploid migrants (i.e.
+The dashed horizontal line marks the proportion of tetraploid migrants (i.e.
 the proportion of tetraploids at equilibrium in the source population).
+The solid horizontal lines mark $\frac{p_4\pi_4}{(p_4\pi_4 + p_2\pi_2)}$, where $p_2$ and
+$p_4$ are the establishment probabilities for a single diploid and a single
+tetraploid individual in the absence of migration (as in \cref{fig:est1}, but
+with unreduced gamete formation included).
 All results are based on 10.000 replicate simulations.
-We assume $\gamma=0.25$ and $u=v=0.05$.
+We assume $\gamma=0.25, \theta=1.5$ and $u=v=0.05$.
 \label{fig:estwmig}}
@@
 
-\subsection*{Loss of self-incompatibility}
+We next consider establishment in the new habitat when there is a continuous
+influx of migrants (at rate $m$) coming from a large, noninbred and
+predominantly diploid source population at cytotype equilibrium.
+In this setting, establishment is certain to happen eventually, and we are
+interested in the probability that a tetraploid population establishes before a
+diploid one does, as well as the effects of maladaptive gene flow and MCE
+(which both increase with increasing migration rate) on the time until
+establishment.
+
+We hypothesized that two counteracting processes affect the probability of
+autotetraploid establishment in this scenario.
+On the one hand, increased migration will increase the probability that an
+otherwise likely successful tetraploid migrant suffers from MCE in the early
+generations while the population size is low, because migrants are likely to be
+diploid.
+On the other hand, tetraploids are strongly reproductively isolated from
+a typical migrant, so that they are less prone to maladaptive swamping.
+Hence, conditional on evading MCE, they should be able to adapt to the new
+habitat at a rate which is not strongly affected by the migration rate.
+This contrasts with diploids, which interbreed freely with maladapted
+migrants, resulting in a pulling back of the trait mean towards that of the
+source population.
+
+As expected, we find that the time to establishment (of a population of either
+ploidy level) first decreases with increasing migration as a result of a larger
+influx of potentially succesful migrants, but later increases with increasing
+migration as a result of swamping by gene flow (\cref{fig:estwmig}A).
+The probability that a tetraploid population establishes decreases
+monotonically with increasing migration and is considerably larger than the
+probability that a migrant individual is tetraploid over a large part of the
+parameter range (\cref{fig:estwmig}B).
+Together, this suggests that the negative effects of MCE generally outweigh the
+benefits that come from reproductive isolation as migration increases.
+
+\subsection*{Loss of self-incompatibility and assortative mating}
+
+When polyploidization disrupts an existing SI system, we expect that
+tetraploids suffer less from MCE as some portion of their ovules are now
+assured to be fertilized by diploid gametes.
+At the same time, we expect that accelerated inbreeding in selfing tetraploids
+diminishes the adaptive advantage of tetraploids that we found above.
+As shown in \cref{fig:selfing}, we find that when polyploidization is
+associated with the loss of a SI system, tetraploids have a strongly increases
+establishment probability.
+This is the case even when the selfing rate $\sigma$ in tetraploids is zero (in
+which case, under our modeling assumptions, there is only random selfing, i.e.
+the \textit{realized} selfing rate in tetraploids is $1/N$).
+Besides reducing MCE, selfing is also expected to simultaneously reduce the
+effects of maladaptive gene flow, and we indeed find that when the selfing rate
+is sufficiently high ($\ge 0.4$ in \cref{fig:selfing}), the relative
+establishment probability of tetraploids increases with increasing migration
+rate.
+In this regime, the increased strength of MCE due to increased migration is
+compensated by the stronger maladaptive gene flow experienced by diploids.
 
@@
 \centering
-\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-si.pdf}
+\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-si2.pdf}
 \caption{
 Establishment with recurrent migration and loss of self-incompatibility in
-polyploids.
+polyploids (diploids are assumed to be self-incompatible).
 (A) Expected time until a population is established in the novel habitat for
 increasing rates of migration and different self-fertilization rates in
-polyploids. Diploids are assumed to be self-incompatible.
-Note that $\sigma=0.0$ refers to random self-fertilization (i.e.
-self-fertilization with probability $1/N$). 
+polyploids. Triploids and tetraploids have the same selfing rate. 
+Diploids are assumed to be self-incompatible.  Note that $\sigma=0.0$ refers to
+random self-fertilization (i.e.  self-fertilization with probability $1/N$). 
 (B) Proportion of simulation replicates in which tetraploids established. 
-The shaded area corresponds to Jeffreys' 95\% interval.
-All results are based on 10.000 replicate simulations.
-We assume $\gamma=0.25$ and $u=v=0.05$.
-\label{fig:}}
+All results are based on 25.000 replicate simulations.
+We assume $\gamma=0.25, \theta=1.5$ and $u=v=0.05$.
+\label{fig:selfing}}
 \end{figure}
 
-\subsection*{Assortative mating}
-
 \begin{figure}
-\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-am.pdf}
+\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-am2.pdf}
 \caption{
-\label{fig:}}
+Establishment with recurrent migration and assortative mating by cytotype.
+(A) Expected time until a population is established in the novel habitat for
+increasing rates of migration and different probabilities of assortative mating
+($\rho_k = \rho$ for $k=2,3,4$, where $\rho_k$ is the probability that an ovule
+from a $k$-ploid mother is pollinated by a $k$-ploid father).
+(B) Proportion of simulation replicates in which tetraploids established. 
+All results are based on 25.000 replicate simulations.
+We assume $\gamma=0.25$ and $u=v=0.05$.
+\label{fig:am}}
 \end{figure}
 
+Another prezygotic isolating mechanism that has often been considered relevant
+for explaining tetraploid establishment is assortative mating by ploidy level,
+where ovules from a tetraploid are more likely to be fertilized by
+pollen coming from a tetraploid -- irrespective of the trait values of these
+individuals.
+Clearly, assortative mating increases the probability of tetraploid
+establishment (\cref{fig:am}), although not as strongly as the loss of an SI
+system does.
+Again, we find that for some parameter values (roughly $\rho \ge 0.4$),
+assortative mating may be strong enough so that tetraploid establishment
+increases with increasing migration rates, suggesting that tetraploids evade
+maladapive gene flow sufficiently to overcome MCE.
+
 \section*{Discussion}
+
+Despite a sizeable literature on the topic, the question whether the
+establishment of autopolyploids requires particular fitness advantages
+associated with polyploidy or not has remained rather unclear and a topic of
+considerable speculation.
+The observation that polyploid populations tend to inhabit more extreme
+habitats or occur at range margins of their diploid progenitors has been
+interpreted as supporting the hypothesis that polyploids may be more tolerant
+to extreme environmental conditions.
+
+In a recent study, \cite{griswold2021} also investigated the effects of
+maladaptive gene flow on polyploid establishment in a peripheral habitat,
+focusing however on the case where fitness is determined by a single biallelic
+locus rather than a polygenic trait.
+The author found that autotetraploids can establish when adaptation in the
+peripheral habitat is conferred by recessive alleles, when inbreeding
+depression in tetraploids is lower than in diploids, and when maladaptive gene
+flow is due to pollen dispersal, rather than seed dispersal.
+
+We have ignored the effects of dominance throughout. Polyploids are expected to
+accumulate a larger mutation load when deleterious variation is recessive due
+to less efficient purging, and this may yield increased inbreeding depression.
+These effects have been studied in the context of range expansions
+\citep{booker2024}.
+However, this applies only to polyploids that have been established for a long
+time.
+In our case, polyploids are always recently descended from diploid ancestors,
+so they will not have accumulated more deleterious recessives than their
+diploid counterparts.
+As inbreeding is slower in tetraploids, we should hence have reduced inbreeding
+depression in tetraploids, since they have the same deleterious load as
+diploids but it is less expressed (when recessive).  
+There should therefore be an immediate benefit to polyploidy (reduced
+inbreeding depression), which is however associated with a long term
+disadvantage because eventually the established polyploid will start to
+accumulate a higher load due to less efficient purging.
+
+\citep{novikova2023}
+
+\section*{Acknowledgements}
+
+I thank Christelle Fraïsse for employment, encouragement and feedback. I
+acknowledge funding from the Research Foundation -- Flanders (FWO, Junior
+Postdoctoral Fellowship 1272625N).
 

### 2024-12-17 notebooks/equilibrium.jl

@@ -1,0 +1,24 @@
+
+using InfGenetics, StatsBase, Distributions, Parameters, Plots
+
+MM = InfDemeMix(U=Umat(0.0, 0.0), V=0.5)
+N  = 1000
+map([2,4]) do c
+    pop = InfPop(z=zeros(N), c=fill(c,N))
+    for i=1:20
+        pop = generation(pop, MM)
+        pop.F .= 0.0
+    end
+    mean(pop.z), var(pop.z)
+end
+
+MM = InfDemeMix(U=Umat(0.1, 0.1), V=0.5)
+N  = 1000
+pop = InfPop(z=zeros(N), c=fill(2,N))
+for i=1:100
+    pop = generation(pop, MM)
+    pop.F .= 0.0
+    i % 10 != 0 && continue
+    zs = pop.z[pop.c .== 4]
+    @info var(zs), 4*MM.β[3]^2*MM.V
+end

### 2024-12-17 notebooks/est1.jl

@@ -131,9 +131,9 @@ md"""
 """
 
 # ╔═╡ 0ea00635-fade-48d3-b971-f910a200b0b9
-ps = let n=100000
-	reps1 = sim(()->simest1(reconstruct(M1, V=0.5), InfPop(z=[-2.0], c=[2])), n)
-	reps2 = sim(()->simest1(M2, InfPop(z=[-2.0], c=[4])), n)
+ps = let n=500_000
+	reps1 = sim(()->simest1(reconstruct(M1, V=0.5), InfPop(z=[0.0], c=[2])), n)
+    reps2 = sim(()->simest1(M2, InfPop(z=[0.0], c=[4], θ=[2.0,2.0,2.0])), n)
 	a = sum(first.(reps1))/n
 	b = sum(first.(reps2))/n
 	a, b

### 2024-12-17 notebooks/establishment.jl

@@ -1,2 +1,106 @@
 # Consider first, as in Barton & Etheridge 2018, for a given source population
 # the probability of a single diploid/tetraploid migrant establishing.
+using Distributed, Plots
+#addprocs(6)
+@everywhere using Parameters, InfGenetics
+@everywhere M = InfDemeMixEst(U=Umat(0.0,0.0), γ=0.25, m=0.0, θ=[2.0,2.0,2.0], V=0.5)
+gs = [0.0 ; 10 .^ range(log10(0.02), stop=log10(2), length=10)]
+
+res = pmap(gs) do g
+    nr = 500_000
+    x2 = map(_->simest2(reconstruct(M, γ=g), x0=InfPop(z=[0.0], c=[2])), 1:nr)
+    x4 = map(_->simest2(reconstruct(M, γ=g), x0=InfPop(z=[0.0], c=[4])), 1:nr)
+    (sum(first.(x2)) / nr, sum(first.(x4)) / nr)
+end
+
+using Serialization
+#serialize("data/est1.jls", res)
+res = deserialize("data/est1.jls")
+
+p2 = first.(res)
+p4 = last.(res)
+x2 = floor.(Int, p2 .* 500000)
+x4 = floor.(Int, p4 .* 500000)
+e2 = jeffreys_interval.(x2, 500000, 0.01)
+e4 = jeffreys_interval.(x4, 500000, 0.01)
+
+P1 = plot(gs, p2, marker=true, ms=2, label="diploid",
+    ribbon=(p2 .- first.(e2), last.(e2) .- p2), fillalpha=0.4)
+plot!(gs, p4 , marker=true, ms=2, label="tetraploid",
+    ribbon=(p4 .- first.(e4), last.(e4) .- p4), fillalpha=0.4)
+plot!(title="(A) \$\\theta=2, \\beta_4 = \\sqrt{1/2}\$", 
+    ylabel="\$P\$", xlabel="\$\\gamma\$", ylim=(0,0.01))
+
+bs = 0.25:0.05:1.0
+resb = pmap(bs) do b
+    nr = 500_000
+    x4 = map(_->simest2(
+        reconstruct(M, β=[1.0, 1.0, √b]), 
+        x0=InfPop(z=[0.0], c=[4])), 1:nr)
+    sum(first.(x4)) / nr
+end
+
+#serialize("data/est1b.jls", resb)
[…]

### 2024-12-17 notebooks/estwmigration.jl

@@ -1,5 +1,6 @@
+using Pkg; Pkg.activate("/home/arthur_z/dev/InfGenetics/")
 using Distributed, StatsBase, Plots, Serialization, Distributions, Optim
-addprocs(6)
+addprocs(4)
 @everywhere using InfGenetics
 
 # estimate parameter of a censored geometric distribution by ML
@@ -12,26 +13,51 @@ function mle_censored_geometric(c, xs)
     optimize(fun, 0.0, 1.0).minimizer
 end
 
-# example        
-sim = simest(InfDemeMixEst(m=1.0, θ=fill(2.0, 3), U=U), nmax=nmax, Nest=Nest)
+# Example        
+# ========
+function dosim(M=InfDemeMixEst(m=1.0, θ=fill(2.0, 3), U=Umat(0.05,0.05)))
+    while true
+        xs = simest(M)
+        cs = counts(xs[2][end][2].c, 2:4)
+        @info cs; cs[3] > cs[2] && return xs
+    end
+end
 
-z_ = map(x->x.z, last.(sim[2]))
-P1 = plot(); for i=1:length(z_)
-    scatter!(fill(i, length(z_[i])), z_[i], color=:black, ms=2)
-end; plot!(legend=false)
-P2 = plot(mapreduce(x->log10.(counts(x.c, 2:4)), hcat, last.(sim[2]))',
-    legend=false, marker=true, ms=2)
-plot(P1, P2, size=(600,200))
+ex = dosim()
+
+z_ = map(x->x.z, last.(ex[2]))
+P1 = plot()
+for i=1:length(z_)
+    scatter!(fill(i, length(z_[i])), z_[i], color=:black, ms=2, label="")
+end
+plot!(legend=false, xlabel="generation", ylabel="\$z\$")
+hline!([2])
+P2 = plot(mapreduce(x->log10.(counts(x.c, 2:4)), hcat, last.(ex[2]))',
+    legend=:topleft, marker=true, ms=2, ylabel="\$\\log_{10}N_k\$",
+    xlabel="generation", label=["diploid" "triploid" "tetraploid"])
+plot(P1, P2, size=(260*2,210), margin=3Plots.mm)
+
 
 # Basic model
 # ===========
-nrep = 5000
+nrep = 50_000
 ms   = 10 .^ range(log10(0.1), stop=log10(3), length=10)
 zs   = [2.0, 1.5, 1.0]
 γ    = 0.25
[…]

### 2024-12-17 notebooks/finvsinf3.jl

@@ -4,7 +4,7 @@ addprocs(6)
 
 nrep = 10
 ngen = 2000
-L    = 500
+L    = 50
 N    = 250
 M    = InfDemeMix(U=[1.0 0.0; 0.0 0.0; 0.0 1.0])
 
@@ -84,22 +84,3 @@ plot!(P3, title="triploids, \$\\alpha=1/4\$")
 plot(P1, P2, P3, layout=(1,3), size=(600,160), margin=3Plots.mm, guidefont=9, titlefont=9)
 
 
-nrep = 1000
-M   = InfDemeMix(U=Umat(0.05,0.05), α=[1/4, 1/4, 1/6])
-Xid = pmap(1:nrep) do _
-    P = InfPop(z=rand(Normal(0, √(2M.V)), N), c=fill(2,N))
-    X = InfGenetics.fiterate((x,i)->generation(x, M, z->1), deepcopy(P), ngen, callback=cb)
-end    
-Xfd = pmap(1:nrep) do _
-    G = [randn(L,3)/√(2L) for i=1:N]
-    P = FinPop(sum.(G), G, fill(2, N))
-    X = fiterate((x,i)->generation(x, M, z->1), deepcopy(P), ngen, callback=cb)
-end
-
-res = map(1:3) do i
-    map([Xid, Xfd]) do X 
-        xs = map(x->getindex.(x, i), X)
-        Z = mapreduce(i->map(x->isempty(x) ? NaN : var(x), xs[i]), hcat, 1:nrep)
-        map(x->filter(!isnan, x) |> y->isempty(y) ? NaN : mean(y), eachrow(Z))
-    end
-end

### 2024-12-17 notebooks/markovchain.jl

@@ -1,5 +1,5 @@
 ### A Pluto.jl notebook ###
-# v0.19.42
+# v0.19.40
 
 using Markdown
 using InteractiveUtils

### 2024-12-17 src/infmixest.jl

@@ -146,7 +146,8 @@ function migration(M::InfDemeMixEst, pop::InfPop)
     nm = rand(Poisson(m))
     cm = rand(Multinomial(nm, M.c))
     cm = vcat([fill(i+1, k) for (i, k) in enumerate(cm)]...)
-    zm = [rand(Normal(0.0, β[m-1]*√2V)) for m in cm]
+    # Vzₖ/Vz₂ = (k/2)βₖ² => Vzₖ = kβₖ²V
+    zm = [rand(Normal(0.0, √(m*β[m-1]^2*V))) for m in cm]
     z  = [pop.z ; zm]
     c  = [pop.c ; cm]
     F  = [pop.F ; zeros(nm)]

### 2024-12-17 test/selfing.jl

@@ -1,0 +1,7 @@
+
+M = InfDemeMixEst(m=1.0, γ=0.0, σ=ones(3), θ=fill(0.0, 3), U=Umat(0.0,0.0))
+x = InfPop(z=zeros(4))
+W, Z, V = InfGenetics.families(M, x)
+InfGenetics.selfing!(W, fill(Inf, 3), x.c)
+sum(W)
+

### 2025-01-14 doc/paper/paper.tex
paper update

@@
 autotetraploids in a new habitat.
-We show that in the absence of migration from the source population,
-autotetraploids are more likely to found a successful population than diploids
-under a very broad range of conditions, but that this is unlikely to
-sufficiently counter the scarcity of tetraploid founders when the source is
+We show that for rare migration from the source population, autotetraploids are
+more likely to found a successful population than diploids under a very
+broad range of conditions, but that this is unlikely to sufficiently
+counter the scarcity of tetraploid founders when the source is
 predominantly diploid.
-[...]
+
 \end{abstract}
@@
 Consider for instance a randomly mating diploid population.
-Under the commonly accepted view that polyploids mostly emerge through the
-union of unreduced gametes \citep{bretagnolle1995,herben2016,kreiner2017b}, a
-new tetraploid individual originating by a chance encounter of two unreduced
-diploid gametes (an event occurring at an appreciable rate; \cite{kreiner2017})
-is highly unlikely to establish a stable tetraploid subpopulation, as most of
-its gametes will end up in unfit hybrids of odd ploidy level (triploid block,
-\cite{ramsey1998,kohler2010,brown2024}).
+Under the commonly accepted view that new polyploid plants are mostly formed
+through the union of unreduced gametes
+\citep{bretagnolle1995,herben2016,kreiner2017b}, a new tetraploid individual
+originating by a chance encounter of two unreduced diploid gametes (an event
+occurring at an appreciable rate; \cite{kreiner2017}) is highly unlikely to
+establish a stable tetraploid subpopulation, as most of its gametes will end up
+in unfit hybrids of odd ploidy level (a phenomenon referred to as `triploid
+block', see \cite{ramsey1998,kohler2010,brown2024}).
 This negative frequency dependence effect is commonly referred to as
 \textit{minority cytotype exclusion} (MCE), after \cite{levin1975}.
-It is well-appreciated that, due to MCE, in a large random mating mixed-ploidy
-population dominated by diploids, the rate of unreduced gamete formation needs
-to be extraordinarily high for tetraploids to establish (\cite{felber1997}, see
+It is well-appreciated that, as a consequence of MCE, the rate of unreduced
+gamete formation needs to be extraordinarily high for tetraploids to establish
+in a large random mating initially diploid population (\cite{felber1997}, see
 also \cref{s-sec:det}).
@@
 occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
-Secondly, any form of prezygotic isolation between cytotypes could
+Secondly, any form of \textit{prezygotic isolation} between cytotypes could
 promote establishment of polyploid cytotypes by alleviating MCE.
@@
 habitats where they evade MCE (for instance at range edges, or after local
-extinction due to environmental change).
+extinction due to environmental change) \citep{griswold2021}.
 However, such peripheral habitats are likely to present adaptive challenges to
-establishment \citep{kawecki2008}, and if polyploids are able to colonize such
-habitats at an appreciable rate, they must somehow be better adapted to local
-conditions, or more able to adapt to those conditions despite inbreeding and
-migration from the source population while the population is still small,
-compared to diploids.
+establishment \citep{kawecki2008adaptation}, and if polyploids are able to
+colonize such habitats at an appreciable rate, they must somehow be better
+adapted to local conditions, or more able to adapt to those conditions despite
+inbreeding and maladaptive migration, compared to diploids.
 Indeed, when the source population is dominated by diploids, the probability
@@
 While there have been substantial modeling efforts aimed at understanding
-autotetraploid establishment within diploid populations (e.g. \cite{levin1975,
-felber1991, felber1997, rausch2005, oswald2011, clo2022c, griswold2021}), the
-problem of polyploid establishment in a novel habitat which presents some adaptive
-challenge remains largely unaddressed, despite its centrality to verbal
-arguments about the establishment of polyploids in natural populations
-\citep{kolar2017, vandepeer2021, clo2022d}.
+autotetraploid establishment within diploid populations \citep{levin1975,
+felber1991, felber1997, rausch2005, oswald2011, clo2022c}, the problem of
+polyploid establishment in peripheral habitats remains largely unaddressed
+(but see \cite{griswold2021}),
+despite its centrality to verbal arguments about the establishment of
+polyploids in natural populations \citep{kolar2017, vandepeer2021, clo2022d}.
 
-Here we develop a model for the establishment of a mixed-ploidy population in a
-novel, unoccupied habitat based on \cite{barton2018}.
+Here we develop a rather general model for the establishment of a mixed-ploidy
+population in a novel, unoccupied habitat based on \cite{barton2018}.
 In order to establish in the novel habitat, the population has to adapt to
@@
 trait evolution under the infinitesimal model, to mixed-ploidy populations.
+This is a first contribution of the present paper.
 We then use simulations to study tetraploid establishment, both from single
@@
 \begin{table}[t]
-\caption{Glossary of the main notation used in the main text.
+\caption{Glossary of the notation used in the main text.
 } \label{tbl:glossary}
@@
 $i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
-$\sigma_k$ & selfing rate in $k$-ploids \\
+$\sigma_k$ & rate of self-fertilization in $k$-ploids \\
 $\rho_k$ & probability of assortative mating in $k$-ploids \\
@@
 
-Note that in a finite population, the segregation variance will decay over time
+In a finite population, the segregation variance will decay over time
 as the population becomes more inbred.
@@
 Similarly, the genetic variance at Hardy-Weinberg and linkage equilibrium
-(HWLE) in tetraploids will be twice that of their diploid counterparts under
-such assumptions, which is similarly unrealistic \citep{clo2022}.
+(HWLE) in a large non-inbred tetraploid population will be twice that of their
+diploid counterparts under such assumptions, which is similarly unrealistic
+\citep{clo2022}.
 
@@
 $k$-ploids at HWLE.
-Note that we also have $\tilde{V}_{z,2} = 2\Vx_{2} = 2V$ \citep{barton2017},
-where $V$ is the segregation variance in the diploid population.
 If we now assume $a_{i,k} = \beta_k a_{i,2}$, i.e. allelic effects in $k$-ploids
-are as in diploids, but scaled by a factor $\beta_k$.
-This implies that
+are as in diploids, but scaled by a factor $\beta_k$, and assume equal allele
+frequencies in the different cytotypes, 
+we find that
 \begin{equation}
@@
 \end{equation}
-and hence also that $\Vx_{k} = \beta_k^2 \Vx_{2} = \beta_k^2 V$.
-These relations will also hold in the infinitesimal limit where $a_{i,2}
-\rightarrow 0$ as $L\rightarrow \infty$.
+Note that by definition $\beta_2=1$.
+Under the infinitesimal model (where $a_{i,2} \rightarrow 0$ as $L\rightarrow
+\infty$), we have $\tilde{V}_{z,2} = 2\Vx_{2} = 2V$ \citep{barton2017}, where
+$V$ is the segregation variance in the diploid population.
+Hence, in the infinitesimal limit we have $\Vx_{k} = \beta_k^2 \Vx_{2} =
+\beta_k^2 V$.
 
@@
 Gaussian random variable $Y_{i,k}$ with distribution
-  $$Y_{i,k} \sim \Normal\left(\frac{k}{c_i}\frac{z_i}{\beta_{c_i}}, V_{i,k}\right)$$
+\begin{equation}
+    Y_{i,k} \sim \Normal\left(\frac{k}{c_i}\frac{z_i}{\beta_{c_i}}, V_{i,k}\right)
+    \label{eq:gamete}
+\end{equation}
 where $V_{i,k}$ is the gametic segregation variance associated with the
-production of a $k$-ploid gamete by individual $i$.
+production of a $k$-ploid gamete by individual $i$ (see below).
 The trait value of an individual originating from the union of a $k$-ploid
@@
 probability that a diploid gamete from a $k$-ploid individual receives two
-copies of the same parental gene. Note that we have $\alpha_3 \le 1/4$ and
-$\alpha_4 \le 1/6$.
+copies of the same parental gene. We assume $\alpha_3 \le 1/4$ and
+    $\alpha_4 \le 1/6$ (see \cref{s-sec:segvar}).
 } \label{tbl:segvar}
@@
 \end{table}
+
+Writing $\bar{z}_2$ for the mean trait value in the diploid subpopulation,
+\cref{eq:oneline} implies that a tetraploid offspring from a random diploid
+parental pair has an expected trait value equal to $\bar{z}_4 =
+2\beta_4\bar{z}_2$.
+This hence implies that $|\bar{z}_4| \ge |\bar{z}_2|$, with equality only when
+$\beta_4=1/2$ or $\bar{z}_2 = \bar{z}_4 = 0$. 
+In other words, when $\beta_4 > 1/2$, we would have for all but $\bar{z}_2 = 0$
+that newly formed tetraploids have more extreme phenotypes on average than
+their diploid parents.
+In our establishment model (see below), we shall therefore always consider 
+the case where $\bar{z}_2=0$ in the source population, and think of the trait
+value modeled as the deviation from the mean phenotype in the source
+population.
+
+The property that polyploid phenotypes are more extreme on average makes sense
+if we consider the underlying genetic model: if we assume the source population
+is at an equilibrium between mutation, drift and stabilizing selection,
+segregating genetic variants that affect the trait will be equally likely to
+have positive or negative allelic effects \citep{hayward2022}.
+At equilibrium, any diploid individual with trait value $z - \bar{z}_2 > 0$
+then carries an excess of variants with positive allelic effects on the trait,
+and this excess should be exaggerated in its polyploid offspring, which carry
+\textit{twice} the excess of their parent on average, unless allelic effects are
+scaled exactly by one half (i.e. $\beta_4 = 1/2$). 
+In other words, the mixed-ploidy model is not `coordinate free' as the basic
+infinitesimal model is: \cref{eq:oneline} only makes sense when the trait
+values that are modeled correspond to deviations from the mean values
+associated with an underlying equilibrium state.
 
@@
 habitat is based on \cite{barton2018}.
-We assume a large non-inbred `mainland' population at HWLE and cytotype
-equilibrium, with $\Ex[z] = 0$ irrespective of the cytotype.
+We assume a large non-inbred `mainland' mixed-ploidy population at HWLE and
+cytotype equilibrium, with $\Ex[z] = 0$ irrespective of the cytotype.
+The equilibrium trait value distribution for the different cytotypes on the
+mainland is complicated in general, but a very accurate approximation (which we
+use throughout) is readily obtained for the case where $u$ is small (see
+\cref{s-sec:mixeq}).
+
 In generation $t$, $M(t)$ migrant individuals arrive on an island (the new
@@
 self-incompatibility).
-We assume the trait is under directional selection, with fitness is $w(z) =
+We assume the trait is under directional selection, with fitness $w(z) =
 e^{\gamma(z - \theta)}$, where $\gamma$ is the intensity of directional
@@
 population becomes positive.
+Throughout the paper, we define establishment as reaching a population of size
+100.
 
@@
 
+The results section is organized as follows: first we verify the correctness
+and accuracy of the infinitesimal model for mixed-ploidy populations by
+comparison against simulations of a model with finitely many loci.
+Next we assess how autotetraploid genetics and different assumptions on the
+genetic variance of neotetraploids affect the probability of tetraploid
+establishment in a marginal habitat relative to diploid establishment, starting
+from a single maladapted migrant individual.
+We then study how continuous migration from a predominantly diploid maladapted
+source population affects the relative establishment probability, assessing the
+impact of migration load and MCE.
+Finally, we consider the impact of prezygotic isolation mechanisms on
+tetraploid establishment in the marginal habitat.
+Throughout, we assume $\beta_3 = \sqrt{2/3}, \beta_4=\sqrt{1/2}$
+and $\alpha_2 = \alpha_3 = \alpha_4 = 0$, unless stated otherwise.
+
 \subsection*{Autotetraploid and mixed-ploidy infinitesimal model}
@@
 \centering
-\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/tetfininf.pdf}
+\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/tetfininf.pdf}
 \caption{
@@
 As predicted, double reduction leads to an immediate increase in genetic
-variance, but leads to accelerated inbreeding, causing faster loss of genetic
-variance in the long-term (\cref{fig:vztet,s-fig:fininf}).
+variance (as it increases the segregation variance), but leads to accelerated
+inbreeding, causing faster loss of genetic variance in the long-term
+(\cref{fig:vztet,s-fig:fininf}).
 As noted by \cite{arnold2012}, the loss of genetic variation in the presence of
@@
 Simulations for the mixed-ploidy model further confirm the correctness of our
-infinitesimal approximation and highlight the importance of assumptions
-regarding the scaling of allelic effects across ploidy levels
-($\beta$ parameters, \cref{s-fig:vz}).
-Although inbreeding is slower in autotetraploids than in diploids for the same
-population size, the tetraploid fraction of a diploid-dominated mixed-ploidy 
-population will have an equal or higher average inbreeding coefficient
-(\cref{s-fig:f1}).
+infinitesimal approximation (\cref{s-fig:vz}), yielding accurate predictions
+for the evolution of the phenotypic variance across ploidy levels and for
+different assumptions regarding the scaling of allelic effects ($\beta$
+parameters) and unreduced gamete formation ($\alpha$ parameters).
+
+It is worth noting that, although inbreeding is slower in autotetraploids than
+in diploids for the same population size, the tetraploid fraction of a
+diploid-dominated mixed-ploidy population will have an equal or higher average
+inbreeding coefficient (\cref{s-fig:f1}).
 This is because in such a population, triploid and tetraploid individuals
-mostly arise from gametes formed by diploid individuals or polyploid
-individuals with very recent diploid ancestry (about $1+u+2v$ generations ago
+mostly arise from gametes formed by diploid individuals, or by polyploid
+individuals with very recent diploid ancestry (on average $1+u+2v$ generations ago
 for tetraploids, and $1+\frac{2}{3}(u+2v)$ generations ago for triploids, see
-\cref{s-sec:ttdip}), so that the polyploid subpopulations will show an average
-relatedness similar to that of the diploid population and not evolve as an
-isolated higher-ploidy population would.
+\cref{s-sec:ttdip}).
+As a result, the polyploid subpopulations will show an average relatedness
+similar to that of the diploid population and not evolve as an isolated
+higher-ploidy population would.
 A nonzero probability of producing IBD diploid gametes ($\alpha_k > 0$) will
@@
 (\cref{s-fig:f1}).
-Therefore, as long as diploids dominate, the effect of harboring some fraction
-of the gene pool in polyploid individuals on the rate of inbreeding is
-negligible, and we find that the evolution of the inbreeding coefficient over
-time is well predicted by $1-e^{-t/2N_e}$, where the inbreeding-effective
-population size is, to first order in $u$, given by $(1-2u)N$
-(\cref{s-sec:effsize}).
+Therefore, as long as diploids dominate, harboring some fraction of the gene
+pool in polyploid individuals has a negligible effect on the rate of inbreeding
+in the mixed-ploidy population as a whole, and we find that the evolution of
+the inbreeding coefficient over time is well predicted by $1-e^{-t/2N_e}$,
+where the inbreeding-effective population size is, to first order in $u$, given
+by $(1-2u)N$ (\cref{s-sec:effsize}). This is just the expected number of
+diploid individuals (to first order in $u$), highlighting that when diploids
+dominate, polyploids have low fitness and do not contribute to the effective
+population size.
 
@@
 \centering
-\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/est1b.pdf}
+\includegraphics[width=\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/est1b.pdf}
 \caption{
@@
 All results are estimated from 500.000 replicate simulations and assume
-$\theta=2$.
+$\theta=2$. Establishment is defined as reaching $N=100$.
 \label{fig:est1}}
@@
 individual with trait value $z_0 = 0$.
-We assume $u=0$ and compare the probability of establishment when the migrant
+We assume $u=0$ (i.e. there are no unreduced gametes, and hence no newly formed
+polyploids) and compare the probability of establishment when the migrant
 is diploid vs. tetraploid (\cref{fig:est1}).
@@
 We shall scale our results accordingly, assuming $2V=1$ throughout, and shall
-assume $\theta=2$, i.e. adaptation has to increase the trait value by two
+assume $\theta=2$, so that the average trait value has to increase by two
 standard deviations to achieve a positive growth rate.
+The \textit{relative} establishment probability for different cytotypes will of
+course also depend on how allelic effects scale across ploidy levels ($\beta$).
 
 We find that reduced inbreeding in tetraploids substantially increases the
-establishment probability.
+establishment probability of tetraploids relative to diploids across a large
+part of the parameter space (\cref{fig:est1}).
 Indeed, in the case where allelic effects are scaled across ploidy levels
-to yield the same equilibrium genetic variance ($\beta = \sqrt{1/2}$, so that
+to yield the same equilibrium genetic variance ($\beta_4 = \sqrt{1/2}$, so that
 polyploid migrants are not more likely to have high fitness than diploid
@@
 As the segregation variance and initial trait value are the same across these
-simulations, this is a consequence of the reduced rate of inbreeding, which
-slows down the exhaustion of the genetic variance carried by the initial
+simulations, this is a consequence only of the reduced rate of inbreeding,
+which slows down the exhaustion of the genetic variance carried by the initial
 migrant individual.
@@
 in the trait mean) will be slower,
-and the risk that genetic variation is exhausted due to inbreeding before the
-population is able to reach a consistently positive growth rate is increased as
-a result.
+increasing the risk that genetic variation is exhausted due to inbreeding before the
+population is able to reach a consistently positive growth rate.
+Note that when $\gamma = 0$, each individual has fitness 1 and we have a
+critical branching process with a Poisson offspring distribution, so that the
+probability to reach $N=100$ is $1/100$ \citep{barton2018}.
 
 Evidently, the scaling of the genetic variance across ploidy levels has a
-profound effect on the establishment probability, but only when $\beta$ is
+profound effect on the establishment probability, but only when $\beta_4$ is
 close to 0.5 (i.e. individual alleles have almost half the effect size in
@@
 \centering
-\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration1b.pdf}
+\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration-dr.pdf}
 \caption{
     Establishment with recurrent migration.
-(A) Expected time until a population is established in the novel habitat for
-increasing rates of migration and different degrees of maladaptation
-($\bar{z}_s$, the mean trait value in the source population). 
+    (A) Expected time until a population is established in the novel habitat
+    for increasing rates of migration and different degrees of maladaptation
+    ($\bar{z}_s$, the mean trait value in the source population). Results are
+    shown for the case with $\alpha_k = 0$ for $k=2,3,4$. 
     (B) Proportion of simulation replicates in which tetraploids established.
-The dashed horizontal line marks the proportion of tetraploid migrants (i.e.
-the proportion of tetraploids at equilibrium in the source population).
-The solid horizontal lines mark $\frac{p_4\pi_4}{(p_4\pi_4 + p_2\pi_2)}$, where $p_2$ and
-$p_4$ are the establishment probabilities for a single diploid and a single
-tetraploid individual in the absence of migration (as in \cref{fig:est1}, but
-with unreduced gamete formation included).
-All results are based on 10.000 replicate simulations.
-We assume $\gamma=0.25, \theta=1.5$ and $u=v=0.05$.
+    The dots connected by solid lines show simulation results with
+    $\alpha_k=0$, whereas the dots connected by dashed lines
+    show simulation results with $\alpha_2=1/2, \alpha_3=1/4$ and
+    $\alpha_4=1/6$ (i.e. maximum $\alpha$).
+    The solid horizontal lines mark $\frac{p_4\pi_4}{(p_4\pi_4 + p_2\pi_2)}$,
+    where $p_2$ and $p_4$ are the establishment probabilities for a single
+    diploid and a single tetraploid individual in the absence of migration (as
+    in \cref{fig:est1}, but with unreduced gamete formation included). 
+    The dashed horizontal line marks the proportion of tetraploid migrants
+    (i.e. the proportion of tetraploids at equilibrium in the source
+    population, $\approx 0.3$\%).
+    The baseline predictions (horizontal lines) are based on $5\times 10^6$
+    simulation replicates.
+    All other results are based on 50.000 replicate simulations.
+    We assume $\gamma=0.25$ and $u=v=0.05$.
     \label{fig:estwmig}}
@@
 We next consider establishment in the new habitat when there is a continuous
-influx of migrants (at rate $m$) coming from a large, noninbred and
-predominantly diploid source population at cytotype equilibrium.
+influx of migrants ($m$ migrants per generation on average) coming from a large,
+noninbred and predominantly diploid source population at cytotype equilibrium.
 In this setting, establishment is certain to happen eventually, and we are
 interested in the probability that a tetraploid population establishes before a
-diploid one does, as well as the effects of maladaptive gene flow and MCE
-(which both increase with increasing migration rate) on the time until
-establishment.
+diploid one does.
+More specifically, we are interested in how maladaptive gene flow and MCE
+(which both increase with $m$) affect the relative establishment probability
+and the time until establishment.
 
@@
 diploid.
-On the other hand, tetraploids are strongly reproductively isolated from
-a typical migrant, so that they are less prone to maladaptive swamping.
+On the other hand, tetraploids are more strongly reproductively isolated from
+a typical migrant, so that a tetraploid subpopulation should be less prone to
+maladaptive gene flow.
 Hence, conditional on evading MCE, they should be able to adapt to the new
@@
 source population.
+However, MCE is expected to be stronger than the latter effect,
+precluding the establishment of tetraploids as $m$ increases.
+Lastly, as the mean trait value on the island increases in diploids
+during adaptation, newly formed tetraploids on the island will, on average,
+have more extreme phenotypes than diploid offspring when $\beta_4 > 1/2$, which
+may also aid their establishment (irrespective of $m$).
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration1d.pdf}
+\caption{
+    Evolution of the mean inbreeding coefficient and trait value across
+    establishment replicates, averaged by population size, for increasing rates
+    of migration ($m$) from the predominantly diploid source population. 
+    All results are based on 1000 succesful establishment replicates.
+    We assume equal equilibrium variance across ploidy levels and $\gamma=0.25,
+    \theta=2, 2V=1$ and $u=v=0.05$.
+    The $m=0$ simulations are seeded with a single diploid or tetraploid
+    individual with trait value 0, and assume $u=v=0$.
+\label{fig:estwmig2}}
+\end{figure}
 
@@
 migration as a result of swamping by gene flow (\cref{fig:estwmig}A).
-The probability that a tetraploid population establishes decreases
-monotonically with increasing migration and is considerably larger than the
-probability that a migrant individual is tetraploid over a large part of the
-parameter range (\cref{fig:estwmig}B).
-Together, this suggests that the negative effects of MCE generally outweigh the
-benefits that come from reproductive isolation as migration increases.
+Importantly, the tetraploid establishment probability is considerably larger
+than the expected proportion of tetraploid migrants over a large part of the
+parameter range (\cref{fig:estwmig}B, black dashed line).
+Furthermore, we find that for small migration rates, the probability that
+tetraploids establish rather than diploids can be larger than expected based on
+the establishment probabilities in the absence of migration (colored lines in
+\cref{fig:estwmig}), depending on how maladapted migrant individuals are.
+This suggests that maladaptive gene flow in diploids can increase the
+probability of tetraploid establishment.
+However, at the same time we find that the tetraploid establishment probability
+declines monotonically with $m$, which highlights the negative effects of MCE.
+Together, these results confirm the view that, while cytotype differences are an
+effective barrier to maladaptive gene flow, this advantage to establishment is
+outweighed by the negative effects of MCE.
+
+Our simulations further show that the mechanism of unreduced gamete formation
+(as determined by the $\alpha_2$ parameter) can affect the establishment
+probability (\cref{fig:estwmig}B, dashed lines).
+This is mainly because the phenotypic variance of a newly formed tetraploid is
+increased by a factor $(1+\alpha_2)$, thereby increasing the chance that a
+newly formed tetraploid is well-adapted to the marginal habitat.
+The rate of double reduction ($\alpha_4$) has a more limited effect
+(\cref{s-fig:suppdr}).
+
+Established diploid populations are more inbred on average than established
+tetraploids when migration is weak, but the difference is slight except when
+there is no migration at all (\cref{fig:estwmig2}, top row).
+For stronger migration ($m > 0.1$), the opposite holds.
+This is a result of two interacting processes.
+On the one hand, inbreeding is slower in tetraploids, so that during adaptation
+and establishment from a single or limited number of barely inbred individuals, 
+the inbreeding coefficient is expected to increase less rapidly conditional on
+the event of tetraploid establishment.
+On the other hand, migration mostly introduces unrelated diploids, which cross
+more readily with diploids than tetraploids, so that migration reduces the
+average relatedness in established diploid populations much more than in
+established tetraploid populations.
+
+Conditional on establishment, tetraploids have a higher trait mean both in the
+absence of migration and for moderate to strong migration, but not when
+migration is weak (\cref{fig:estwmig2}, bottom row).
+In the absence of migration, this is a consequence of the reduced rate of
+inbreeding and the resulting increased adaptive potential of tetraploids.
+Another consequence of this is that, in the absence of migration, the average
+trait value in tetraploids at low population size (conditional on eventual
+establishment) is lower than that in diploids.
+This is because, in order to adapt before genetic variation is exhausted,
+tetraploids need not increase the trait mean as much as diploids do in the
+first couple of generation.
+
+It is clear from \cref{fig:estwmig2} that migration is not only maladaptive,
+but also introduces genetic variation upon which selection can act, thereby
+increasing the mean trait value at the time of establishment for both
+tetraploids and diploids.
+Interestingly, when migration is strong, tetraploids have markedly larger
+phenotypic values than diploids ($m=1$ in \cref{fig:estwmig2}).
+Diploids suffer strongly from maladaptive gene flow when the population size is
+low, while tetraploids enjoy some reproductive isolation.
+Furthermore, in these replicates, tetraploids tend to emerge and rise in
+frequency at larger population sizes on the island, and hence tend to derive
+from diploids that already experienced several generations of selection.
+Neotetraploids deriving from parents with $z>0$ will have more extreme
+phenotypes on average (see methods) and hence be better adapted.
 
@@
 
-When polyploidization disrupts an existing SI system, we expect that
-tetraploids suffer less from MCE as some portion of their ovules are now
-assured to be fertilized by diploid gametes.
+We now briefly examine the effects of prezygotic isolation mechanisms that may
+be associated with polyploidization and that could alleviate MCE.
+Plausible sources of prezygotic isolation include assortative mating by
+cytotype and self-fertilization.
+
+When polyploidization disrupts an existing SI system (see e.g.
+\cite{robertson2011comparative,zenil2019,novikova2023}), we expect
+that tetraploids suffer less from MCE, as an appreciable portion of their
+ovules are now assured to be fertilized by diploid gametes, irrespective of the
+composition of the population.
 At the same time, we expect that accelerated inbreeding in selfing tetraploids
@@
 As shown in \cref{fig:selfing}, we find that when polyploidization is
-associated with the loss of a SI system, tetraploids have a strongly increases
-establishment probability.
+associated with the loss of a SI system (i.e. when diploids are
+self-incompatible, but tetraploids are not), tetraploids have a strongly
+increases establishment probability.
 This is the case even when the selfing rate $\sigma$ in tetraploids is zero (in
@@
 the \textit{realized} selfing rate in tetraploids is $1/N$).
+
 Besides reducing MCE, selfing is also expected to simultaneously reduce the
-effects of maladaptive gene flow, and we indeed find that when the selfing rate
-is sufficiently high ($\ge 0.4$ in \cref{fig:selfing}), the relative
-establishment probability of tetraploids increases with increasing migration
-rate.
-In this regime, the increased strength of MCE due to increased migration is
-compensated by the stronger maladaptive gene flow experienced by diploids.
+effects of maladaptive gene flow, so we would expect that tetraploid
+establishment is promoted with increasing migration rates. 
+However, increased migration would at the same time alleviate the issue of
+reproductive assurance in diploids, as a diploid migrant is more likely to find
+a compatible mate.
+We find that when the selfing rate is sufficiently high ($\ge 0.4$ in
+\cref{fig:selfing}A), the relative establishment probability of tetraploids
+increases with increasing migration rate.
+In this regime, the increased strength of MCE due to increased migration and
+increased reproductive assurance in diploids is compensated by the stronger
+maladaptive gene flow experienced by diploids.
+
+Self-incompatibility is clearly a strong disadvantage when colonizing a novel
+habitat, as a self-incompatible population of size one can never reproduce.
+However, even when diploids are self-compatible, polyploids may still have
+increased rates of self-fertilization (for instance due to altered flower
+morphology). As an illustration, \cref{fig:selfing}A also shows results where
+diploids are assumed to be self-compatible with $\sigma=0$ in diploids (i.e.
+random selfing, dashed transparent lines).
+The tetraploid establishment probability is still markedly increased when
+$\sigma \ge 0.4$, and as for the simulations with self-incompatibility,
+migration still promotes the probability of tetraploid establishment when the
+selfing rate in polyploids is sufficiently large.
 
@@
 \centering
-\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-si2.pdf}
+\includegraphics[width=0.9\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/prez.pdf} 
     \caption{
-Establishment with recurrent migration and loss of self-incompatibility in
-polyploids (diploids are assumed to be self-incompatible).
-(A) Expected time until a population is established in the novel habitat for
-increasing rates of migration and different self-fertilization rates in
-polyploids. Triploids and tetraploids have the same selfing rate. 
-Diploids are assumed to be self-incompatible.  Note that $\sigma=0.0$ refers to
-random self-fertilization (i.e.  self-fertilization with probability $1/N$). 
-(B) Proportion of simulation replicates in which tetraploids established. 
-All results are based on 25.000 replicate simulations.
-We assume $\gamma=0.25, \theta=1.5$ and $u=v=0.05$.
-\label{fig:selfing}}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-am2.pdf}
-\caption{
-Establishment with recurrent migration and assortative mating by cytotype.
-(A) Expected time until a population is established in the novel habitat for
-increasing rates of migration and different probabilities of assortative mating
-($\rho_k = \rho$ for $k=2,3,4$, where $\rho_k$ is the probability that an ovule
-from a $k$-ploid mother is pollinated by a $k$-ploid father).
-(B) Proportion of simulation replicates in which tetraploids established. 
-All results are based on 25.000 replicate simulations.
-We assume $\gamma=0.25$ and $u=v=0.05$.
-\label{fig:am}}
+    (A) Establishment with recurrent migration selfing in polyploids. 
+    The solid lines show the case where diploids are self-incompatible 
+    The dashed transparent lines show the case where diploids do random
+    self-fertilization (i.e. self-fertilization occurs with probability $1/N$),
+    Triploids and tetraploids have the same selfing rate. 
+    $\sigma=0.0$ refers to random self-fertilization.  
+    (B) Establishment with recurrent migration and assortative mating by
+    cytotype. The rate of assortative mating is determined by $\rho_k = \rho$
+    for $k=2,3,4$, where $\rho_k$ is the probability that an ovule from a
+    $k$-ploid mother is pollinated by a $k$-ploid father.
+    All results are based on 25.000 replicate simulations.  We assume
+    $\gamma=0.25, \theta=1.5$ and $u=v=0.05$.  \label{fig:selfing}}
 \end{figure}
@@
 Clearly, assortative mating increases the probability of tetraploid
-establishment (\cref{fig:am}), although not as strongly as the loss of an SI
+establishment (\cref{fig:selfing}B), although not as strongly as the loss of an SI
 system does.
@@
 
-Despite a sizeable literature on the topic, the question whether the
-establishment of autopolyploids requires particular fitness advantages
-associated with polyploidy or not has remained rather unclear and a topic of
-considerable speculation.
 The observation that polyploid populations tend to inhabit more extreme
-habitats or occur at range margins of their diploid progenitors has been
-interpreted as supporting the hypothesis that polyploids may be more tolerant
-to extreme environmental conditions.
+habitats or occur at the edge of the range of their conspecific diploids
+has spurred considerable interest among botanists and evolutionary biologists
+\citep{rice2015,kolar2017,vandepeer2021,griswold2021,mortier2024}.
+An important question is whether such patterns emerge because polyploids are
+somehow more tolerant to extreme environmental conditions (i.e. they somehow
+are intrinsically more fit than diploids at the range margin), or whether other
+aspects of the population dynamics of mixed-ploidy populations may favor the
+establishment of polyploid subpopulations in peripheral habitats.
 
-In a recent study, \cite{griswold2021} also investigated the effects of
-maladaptive gene flow on polyploid establishment in a peripheral habitat,
-focusing however on the case where fitness is determined by a single biallelic
-locus rather than a polygenic trait.
-The author found that autotetraploids can establish when adaptation in the
-peripheral habitat is conferred by recessive alleles, when inbreeding
-depression in tetraploids is lower than in diploids, and when maladaptive gene
-flow is due to pollen dispersal, rather than seed dispersal.
+In this study, we worked out the infinitesimal model for an additive polygenic
+trait in autotetraploids and mixed-ploidy populations.
+We used the mixed-ploidy infinitesimal model to study the establishment of
+tetraploids in a marginal habitat by means of individual-based simulations. 
+Assuming the trait to be under directional selection in the marginal habitat,
+and migration of maladapted individuals from a predominantly diploid source,
+we sought to determine the conditions under which tetraploids are more likely
+to establish a stable population.
+Importantly, we assume no intrinsic advantage of polyploids in the marginal
+habitat, i.e. the expected fitness of a migrant individual is the same
+regardless of the ploidy level.
+Differences in establishment probabilities are hence caused solely by aspects
+of autopolyploid genetics and the barrier to gene flow between subpopulations
+of different ploidy levels.
 
-We have ignored the effects of dominance throughout. Polyploids are expected to
-accumulate a larger mutation load when deleterious variation is recessive due
-to less efficient purging, and this may yield increased inbreeding depression.
+When migration is weak, succesful establishment is not affected by
+maladaptive gene flow and we can treat establishment in the marginal habitat as
+independent trials of founding a population from a single individual.
+In order to avoid extinction, the population has to increase the trait mean
+by a sufficient amount before the genetic variation carried by the initial
+migrant individual is exhausted.
+The probability that the population manages to do so depends on the degree of
+maladaptation, the intensity of selection and the rate of inbreeding.
+We find that the decreased rate of inbreeding in autotetraploids gives a
+rare tetraploid migrant a larger adaptive potential than a diploid migrant,
+even if the genetic variance carried by the founding individual is the same.
+
+In the presence of maladaptive gene flow, a nascent tetraploid subpopulation
+suffers from MCE, and although polyploids are more reproductively isolated from
+a typical migrant (and hence suffer less maladaptive gene flow), MCE will
+preclude polyploid establishment as the rate of migration grows.
+Nevertheless, it is important to remark that, the probability of tetraploid
+establishment in the marginal habitat can be an order of magnitude higher than
+expected based on the frequency of tetraploid migrants (i.e. is roughly $O(u)$
+instead of $O(u^2)$) when migration is sufficiently weak and maladaptation
+sufficiently high.
+
+Additional sources of prezygotic isolation such as selfing and assortative
+mating by cytotype may further boost the probability of tetraploid
+establishment.
+These processes interact with the rate of migration, so that when
+selfing/assortative mating occurs above some threshold rate, the tetraploid
+establishment probability increases with increasing migration rates, whereas
+below the threshold it decreases with increasing migration pressure.
+In the latter case, the advantage that tetraploids have when it comes to
+avoiding maladaptive gene flow is not strong enough to overcome the effects of
+MCE, whereas in the former case it is.
+
+A major weakness of the present work is that we have ignored inbreeding
+depression and dominance throughout.
+Including dominance in the infinitesimal framework is already challenging for
+diploids (requiring the tracking of 4-way identity coefficients;
+\cite{barton2023}), and appears intractable for higher ploidy levels.
+Nevertheless, autopolyploidy has important consequences whenever dominance is
+relevant, as in the case of inbreeding depression.
+Indeed, when inbreeding depression is due to recessive deleterious variation,
+it should be less expressed in polyploids simply because, for the same allele
+frequency, homozygous genotypes are much rarer in polyploids than in diploids
+at Hardy-Weinberg equilibrium.
+When there is indeed deleterious recessive variation, inbreeding during the
+establishment process should therefore incur a higher fitness cost in diploids
+relative to tetraploids, and hence further increase the probability of
+tetraploid establishment.
+In addition, inbreeding itself occurs at a slower rate in polyploids, further
+decreasing inbreeding depression during establishment.
+
+Dominance and inbreeding depression may strongly affect the complicated
+relationship between selfing, migration load and tetraploid establishment
+probability.
+\cite{griswold2021} performed a study somewhat similar to ours, but did include
+a model of inbreeding depression.
+The author focused on the case where local fitness is determined by a single
+biallelic locus, and investigated the interaction between inbreeding depression
+and migration load (where inbreeding depression is modeled as an immediate
+fitness reduction in offspring produced by selfing).
+In his model, inbreeding depression is different between cytotypes, so that
+tetraploids are able to produce more offspring through selfing relative to
+diploids, who have to rely more on outcrossing.
+However, outcrossing incurs maladaptive gene flow, and thereby puts the
+diploids at a disadvantage. 
+He found that autotetraploids can establish when adaptation in the
+peripheral habitat is conferred by recessive alleles (so that migration load is
+expressed when migrant alleles are rare) and when inbreeding depression in
+tetraploids is lower than in diploids.
+It would be very interesting to combine the infinitesimal framework with a
+simple model of inbreeding depression to investigate in a somewhat more
+realistic model whether the combination of maladaptive migration and
+differential inbreeding depression could explain the prevalence of polyploid
+subpopulations at range edges.
+
+In the long term, polyploids are expected to accumulate a larger
+mutation load when deleterious variation is recessive due to less efficient
+purging, and this may yield \textit{increased} inbreeding depression.
 These effects have been studied in the context of range expansions
@@
 so they will not have accumulated more deleterious recessives than their
-diploid counterparts.
-As inbreeding is slower in tetraploids, we should hence have reduced inbreeding
-depression in tetraploids, since they have the same deleterious load as
-diploids but it is less expressed (when recessive).  
-There should therefore be an immediate benefit to polyploidy (reduced
-inbreeding depression), which is however associated with a long term
-disadvantage because eventually the established polyploid will start to
-accumulate a higher load due to less efficient purging.
+diploid counterparts, in which case polyploidy would lead to reduced rather
+than increased inbreeding depression (as discussed above).
 
-\citep{novikova2023}
+It would be very interesting to experimentally assess whether typical
+quantitative traits in mixed-ploidy populations follow approximately the
+mixed-ploidy infinitesimal model we outlined in this paper, and to estimate the
+relevant model parameters (in particular the $\beta_3$ and $\beta_4$ values).
 
@@
 \bibliographystyle{abbrvnat}
-\bibliography{/home/arthur_z/vimwiki/bib.bib}
+\bibliography{/home/arzwa/vimwiki/bib.bib}
 \end{document}

### 2025-01-14 doc/paper/supp.tex
update
paper update

@@
 \renewcommand{\thesection}{S\arabic{section}}
-\renewcommand{\thealgorithm}{S\arabic{algorithm}}
 
@@
 \centering
-\includegraphics[width=\textwidth]{/home/arthur_z/dev/InfGenetics/doc/img/finsvsinf-polyploids-validation.pdf}
+\includegraphics[width=\textwidth]{/home/arzwa/dev/InfGenetics/doc/img/finsvsinf-polyploids-validation.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=\textwidth]{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-Vz.pdf}
+\includegraphics[width=\textwidth]{/home/arzwa/vimwiki/build/img/2023-10-23/mixedploidy-Vz.pdf}
 \caption{Comparison of the mixed-ploidy infinitesimal model with the
@@
 \centering
-\includegraphics[width=\linewidth]{/home/arthur_z/vimwiki/build/img/2023-10-23/mixedploidy-F.pdf}
+\includegraphics[width=\linewidth]{/home/arzwa/vimwiki/build/img/2023-10-23/mixedploidy-F.pdf}
 \caption{Average inbreeding coefficient \(\bar{F}\) in each cytotype in
@@
 and no selection.}\label{fig:f1}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.9\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration-dr-check.pdf}
+\caption{
+    Nonzero $\alpha_k$ parameters increase the tetraploid establishment
+    probability, but mainly due to $\alpha_2$, which increases the segregation
+    variance associated with the formatoin of diploid gametes by diploids.
+    All other parameters and simulation details are as in \cref{m-fig:estwmig}.
+\label{fig:suppdr}}
 \end{figure}
@@
 \centering
-\includegraphics[width=\linewidth]{/home/arthur_z/vimwiki/build/img/2023-10-13/g1diff.pdf}
+\includegraphics[width=\linewidth]{/home/arzwa/vimwiki/build/img/2023-10-13/g1diff.pdf}
 \caption{
@@
 \begin{align}
-  p_{ij,kl} = \Pr\{N_2(t+1)=k,& N_3(t+1)=l | N_2(t)=i, N_3(t)=j\} \nonumber \\
+  p_{ij,kl} &= \Pr\{N_2(t+1)=k, N_3(t+1)=l | N_2(t)=i, N_3(t)=j\} \nonumber \\
     &= \frac{N!}{k!l!(N-k-l)!} p_2^k p_3^l (1-p_2-p_3)^{N-k-l}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{/home/arthur_z/vimwiki/build/img/2023-10-22/timetodiploidancestord.pdf}
+\includegraphics[width=0.5\textwidth]{/home/arzwa/vimwiki/build/img/2023-10-22/timetodiploidancestord.pdf}
 \caption{Expected time to diploid ancestry. The solid lines show
@@
 \centering
-\includegraphics[width=\textwidth]{/home/arthur_z/vimwiki/build/img/2023-10-13/mixedploidy-F-Ne.pdf}
+\includegraphics[width=\textwidth]{/home/arzwa/vimwiki/build/img/2023-10-13/mixedploidy-F-Ne.pdf}
 \caption{The evolution of \(\bar{F}\) in the mixed-ploidy population and
@@
 
-\subsection{Segregation variance expressions for the mixed-ploidy infinitesimal model \label{sec:segvar}}
+\subsection{Segregation variance expressions for the mixed-ploidy infinitesimal 
+model \label{sec:segvar}}
 
@@
 We note that, on the supposition that diploid gametes are produced by random
-assortment of chromosomes in a haploid and diploid gametes, $\alpha_3 \le 1/4$.
+assortment of chromosomes in a haploid and diploid gamete, $\alpha_3 \le 1/4$.
 
@@
 
+\subsection{Equilibrium trait value distribution for a large mixed-ploidy
+population \label{sec:mixeq}}
+
+In our establishment model, we assume that migrant individuals are drawn
+randomly from a large (effectively infinite) non-inbred mixed-ploidy population
+at HWLE and cytotype equilibrium, with mean trait value zero.
+The equilibrium trait value distribution in a monocytotypic population
+under such conditions is straightforwardly obtained under the infinitesimal
+model as a Gaussian with variance $k \beta_k^2 V$ (see main text).
+However, the equilibrium trait value distributions for the different cytotypes
+in the mixed-ploidy model are more challenging to obtain.
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{/home/arzwa/dev/InfGenetics/doc/img/mixeq.pdf}
+\caption{
+    Equilibrium trait value distribution for diploids, triploids and
+    tetraploids.
+    The simulated equilibrium distribution, the small $u$ approximation and the
+    monocytotypic equilibrium distribution are shown for all three cytotypes.
+    The diploid monocytotypic distribution $\Normal(0,2V)$ is shown everywhere
+    as a reference. Here we assumed $\beta_3 = \beta_4 = 1$ and $\alpha_2 =
+    1/2, \alpha_3 = 1/4$ and $\alpha_4 = 1/6$.
+\label{fig:mixeq}}
+\end{figure}
+
+Indeed, each cytotype produces haploid and diploid gametes with Gaussian
+gametic values, distributed according to \cref{m-eq:gamete} in the main text.
+Conditional on the parental trait values, offspring trait values will be sums
+of independent Gaussians, and hence again Gaussian, distributed according to
+\cref{m-eq:oneline} in the main text.
+Consider now, for instance, a triploid individual, there are six different
+parental cytotype combinations that may yield a triploid offspring, with
+relative frequencies proportional to the terms in the expansion of the
+following polynomial in $\pi$
+  $$2((1-u)\pi_2 + v\pi_3)(u\pi_2 + v\pi_3 + (1-u)\pi_4)$$
+(where the term in $\pi_2^2$ is proportional to the frequency of triploids
+derived from two diploid parents, the term in $\pi_2\pi_3$ proportional to the
+frequency of triploids derived from a diploid and a triploid parent,
+\textit{etc.}).
+Each of these combinations of parental cytotypes potentially yields a different
+Gaussian distribution over offspring trait values.
+As a result, the triploid trait distribution in the next generation will be a
+\textit{mixture} of Gaussians, with mixture proportions given by the relative
+frequencies in the expansion of the above polynomial.
+A similar reasoning applies to the diploid and tetraploid subpopulation.
+Proceeding to the next generation, the trait distributions will be mixtures
+mixtures of Gaussians, and so on for further generations.
+
+We have not identified whether this process yields a tractable equilibrium
+distribution. However, for a predominantly diploid population with $u$
+sufficiently small, an accurate approximation is readily obtained by assuming
+that effectively all individuals in the population have diploid parents.
+In this case, still assuming the mean trait value to be zero, diploid trait
+values will be Gaussian with variance $2V$.
+Using \cref{m-eq:oneline}, the trait value distribution of newly formed triploids
+will be Gaussian with mean zero and variance
+\begin{align*}
+  \Var[Z_{ij}^{12}] 
+    &= \Ex[\Var[Z_{ij}^{12}|Z_i,Z_j]] + \Var[\Ex[Z_{ij}^{12}|Z_i,Z_j]] \\
+    &= \beta_3^2\left(\frac{V}{2} + 2\alpha_2 V\right) +
+    \Var\left[\beta_3\left(\frac{Z_i}{2} + Z_j\right)\right] \\
+    &= 3\beta_3^2V\left(1 + \frac{2}{3}\alpha_2\right)
+\end{align*}
+(where $Z_{ij}^{12}$ denotes the trait value of a random offspring from
+individuals $i$ and $j$, and where the superscript indicates that $i$
+contributes a haploid gamete, and $j$ a diploid gamete -- as in
+\cref{m-eq:oneline}).
+Under the small $u$ assumption, $Z_i$ and $Z_j$ are random draws from a
+Gaussian with mean zero and variance $2V$.
+The trait value distribution for tetraploids will be Gaussian with mean zero
+and variance
+  $$\Var[Z_{ij}^{22}] = 4\beta_4^2V\left(1 + \alpha_2\right)$$
+This approximation is illustrated in \cref{fig:mixeq}, where, the fit to
+simulations is shown to be very accurate.
+Note that when the $\alpha_k$ parameters are set to zero (as in most of our
+results), the equilibrium variance for each cytotype in the mixed-ploidy
+population is just the equilibrium variance in an isolated monocytotypic
+population.
+
 \bibliographystyle{abbrvnat}
-\bibliography{/home/arthur_z/vimwiki/bib.bib}
+\bibliography{/home/arzwa/vimwiki/bib.bib}
 \end{document}

### 2025-01-14 doc/paper/paper.tex
update

@@
 We define the infinitesimal model of quantitative genetics (\textit{sensu}
-\cite{barton2017}) for the inheritance of an additive quantitative trait in a
-mixed-ploidy population consisting of diploid, triploid and autotetraploid
-individuals producing haploid and diploid gametes.
-We implement efficient simulation methods and use these to study the
+    \cite{barton2017}) for the inheritance of an additive quantitative trait in
+    a mixed-ploidy population consisting of diploid, triploid and
+    autotetraploid individuals producing haploid and diploid gametes.  We
+    implement efficient simulation methods and use these to study the
     quantitative genetics of mixed-ploidy populations and the establishment of
-autotetraploids in a new habitat.
-We show that for rare migration from the source population, autotetraploids are
-more likely to found a successful population than diploids under a very
-broad range of conditions, but that this is unlikely to sufficiently
-counter the scarcity of tetraploid founders when the source is
-predominantly diploid.
-
+    autotetraploids in a new habitat.  We show that for rare migration from the
+    source population, autotetraploids are more likely to found a successful
+    population than diploids under a very broad range of conditions, but that
+    this is unlikely to sufficiently counter the scarcity of tetraploid
+    founders when the source is predominantly diploid.  We assess in more
+    detail how minority cytotype exclusion interacts with migration load in the
+    establishment process and evaluate the impact of additional sources of
+    prezygotic isolation, specifically selfing and assortative mating, on the
+    relative establishment probabilities of the different cytotypes.
+    In the discussion we consider how inbreeding depression may impact our
+    findings.
 \end{abstract}
@@
 trait in autotetraploids and mixed-ploidy populations.
+It would be very interesting to experimentally assess whether typical
+quantitative traits in mixed-ploidy populations follow approximately the
+mixed-ploidy infinitesimal model we outlined in this paper, and to estimate the
+relevant model parameters (in particular the $\beta_3$ and $\beta_4$ values).
+
 We used the mixed-ploidy infinitesimal model to study the establishment of
@@
 decreasing inbreeding depression during establishment.
+
+In the long term, polyploids are expected to accumulate a larger
+mutation load when deleterious variation is recessive due to less efficient
+purging, and this may yield \textit{increased} inbreeding depression.
+These effects have been studied in the context of range expansions
+\citep{booker2024}.
+However, this applies only to polyploids that have been established for a long
+time.
+In our case, polyploids are always recently descended from diploid ancestors,
+so they will not have accumulated more deleterious recessives than their
+diploid counterparts, in which case polyploidy would lead to reduced rather
+than increased inbreeding depression (as discussed above).
 
@@
 
-In the long term, polyploids are expected to accumulate a larger
-mutation load when deleterious variation is recessive due to less efficient
-purging, and this may yield \textit{increased} inbreeding depression.
-These effects have been studied in the context of range expansions
-\citep{booker2024}.
-However, this applies only to polyploids that have been established for a long
-time.
-In our case, polyploids are always recently descended from diploid ancestors,
-so they will not have accumulated more deleterious recessives than their
-diploid counterparts, in which case polyploidy would lead to reduced rather
-than increased inbreeding depression (as discussed above).
-
-It would be very interesting to experimentally assess whether typical
-quantitative traits in mixed-ploidy populations follow approximately the
-mixed-ploidy infinitesimal model we outlined in this paper, and to estimate the
-relevant model parameters (in particular the $\beta_3$ and $\beta_4$ values).
-
 \section*{Acknowledgements}

### 2025-01-15 doc/paper/paper.tex
first draft finalized

@@
     findings.
+
+    \textbf{Keywords:} polyploidy, adaptation, establishment, inbreeding,
+    infinitesimal model
 \end{abstract}
@@
 $\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
-$u$ & probability of unreduced gamete formation ($u=u_{22}=u_{32}=1-u_{42}$)\\
+$u$ & probability of unreduced gamete formation ($u=u_{22}=1-u_{42}$)\\
 $v$ & probability that a triploid produces a haploid/diploid gamete
@@
 
+Our notation is summarized in \cref{tbl:glossary}.
 We consider a mixed-ploidy population of size $N$ consisting of $N_2$ diploid,
@@
 have, for instance, twice the size of their diploid progenitors on average.
-Similarly, the genetic variance at Hardy-Weinberg and linkage equilibrium
+Furthermore, the genetic variance at Hardy-Weinberg and linkage equilibrium
 (HWLE) in a large non-inbred tetraploid population will be twice that of their
@@
 If we now assume $a_{i,k} = \beta_k a_{i,2}$, i.e. allelic effects in $k$-ploids
-are as in diploids, but scaled by a factor $\beta_k$, and assume equal allele
-frequencies in the different cytotypes, 
-we find that
+are as in diploids, but scaled homogeneously by a factor $\beta_k$, and assume
+equal allele frequencies in the different cytotypes, we will have
 \begin{equation}
@@
 and this excess should be exaggerated in its polyploid offspring, which carry
-\textit{twice} the excess of their parent on average, unless allelic effects are
-scaled exactly by one half (i.e. $\beta_4 = 1/2$). 
+\textit{twice} the excess of their parent on average.
 In other words, the mixed-ploidy model is not `coordinate free' as the basic
@@
 where $F_k^\ast = \alpha_{c_k} + (1-\alpha_{c_k})F_k$ (\cref{s-sec:tetinbred}).
-The recursion for the coancestry coefficients in is given by
+The recursion for the coancestry coefficients is given by
 \begin{align}
@@
 selection and $\theta$ is the trait value for which the growth rate of the
-population becomes positive.
+island population becomes positive.
 Throughout the paper, we define establishment as reaching a population of size
@@
 \end{equation}
+Where $\delta_{x,y}$ is the Kronecker delta function.
 
@@
 and $\alpha_2 = \alpha_3 = \alpha_4 = 0$, unless stated otherwise.
+Note that while the former assumption is a natural default, the latter is not
+easy to motivate, and we do investigate the impact of the $\alpha$ parameters
+in some detail below.
 
@@
 $e^{-t/4N}$ (\cref{fig:vztet}A), as expected from the results of \cite{arnold2012}.
-As predicted, double reduction leads to an immediate increase in genetic
-variance (as it increases the segregation variance), but leads to accelerated
-inbreeding, causing faster loss of genetic variance in the long-term
-(\cref{fig:vztet,s-fig:fininf}).
-As noted by \cite{arnold2012}, the loss of genetic variation in the presence of
-double reduction cannot, however, be described by a single long-term effective
-population size.
+As predicted, double reduction (i.e. $\alpha_4 > 0$) leads to an immediate
+increase in genetic variance (as it increases the segregation variance), but
+leads to accelerated inbreeding, causing faster loss of genetic variance in the
+long term (\cref{fig:vztet,s-fig:fininf}).
 
@@
 The \textit{relative} establishment probability for different cytotypes will of
-course also depend on how allelic effects scale across ploidy levels ($\beta$).
+course also depend on how allelic effects scale across ploidy levels ($\beta_4$).
 
@@
 tetraploids need not increase the trait mean as much as diploids do in the
-first couple of generation.
+first couple of generations.
 
@@
 morphology). As an illustration, \cref{fig:selfing}A also shows results where
-diploids are assumed to be self-compatible with $\sigma=0$ in diploids (i.e.
-random selfing, dashed transparent lines).
+diploids are assumed to be self-compatible with $\sigma_2=0$ (i.e. random
+selfing, dashed transparent lines).
 The tetraploid establishment probability is still markedly increased when
-$\sigma \ge 0.4$, and as for the simulations with self-incompatibility,
+$\sigma_4 \ge 0.4$, and as for the simulations with self-incompatibility,
 migration still promotes the probability of tetraploid establishment when the
@@
 maladapive gene flow sufficiently to overcome MCE.
+Note that the case $\rho = 1$ amounts to complete prezygotic isolation.
 
@@
 
-In the long term, polyploids are expected to accumulate a larger
-mutation load when deleterious variation is recessive due to less efficient
-purging, and this may yield \textit{increased} inbreeding depression.
-These effects have been studied in the context of range expansions
-\citep{booker2024}.
-However, this applies only to polyploids that have been established for a long
-time.
-In our case, polyploids are always recently descended from diploid ancestors,
-so they will not have accumulated more deleterious recessives than their
-diploid counterparts, in which case polyploidy would lead to reduced rather
-than increased inbreeding depression (as discussed above).
-
 Dominance and inbreeding depression may strongly affect the complicated
@@
 \cite{griswold2021} performed a study somewhat similar to ours, but did include
-a model of inbreeding depression.
+a simple model of inbreeding depression.
 The author focused on the case where local fitness is determined by a single
@@
 tetraploids is lower than in diploids.
-It would be very interesting to combine the infinitesimal framework with a
-simple model of inbreeding depression to investigate in a somewhat more
-realistic model whether the combination of maladaptive migration and
-differential inbreeding depression could explain the prevalence of polyploid
-subpopulations at range edges.
+It would be very interesting to combine the infinitesimal framework with some
+form of inbreeding depression to investigate in a somewhat more realistic model
+whether the combination of maladaptive migration and differential inbreeding
+depression could explain the prevalence of polyploid subpopulations at range
+edges.
+
+In the long term, polyploids are expected to accumulate a larger
+mutation load when deleterious variation is recessive due to less efficient
+purging, and this may yield \textit{increased} inbreeding depression
+\citep{vlcek2025biorxiv}.
+These effects have been studied in the context of range expansions
+\citep{booker2024}.
+However, this applies only to polyploids that have been established for a long
+time.
+In our case, polyploids are always recently descended from diploid ancestors,
+so they will not have accumulated more deleterious recessives than their
+diploid counterparts, so that polyploidy leads to reduced rather than increased
+inbreeding depression (as discussed above).
+An interesting picture emerges here however, where the different effects of
+polyploidy in the short and long term lead to some sort of equilibrium
+situation that may characterize mixed-ploidy populations in nature: although
+polyploids may enjoy enhanced establishment probabilities in peripheral
+habitats, the accumulation of mutational load may in the long-term preclude
+further range expansion or competitive exclusion of their diploid progenitors
+in the core habitat.
+Further modeling efforts could provide more insights into the plausibility of
+such a model.
 

### 2025-01-22 doc/paper/paper.tex
first draft, editing

@@
 Indeed, when the source population is dominated by diploids, the probability
-that a migrant individual is tetraploid will be very small ($O(u^2)$ if $u$ is
-the probability that a diploid in the source population produces an unreduced
-gamete in meiosis), so that tetraploids need a substantial advantage during
-colonization if they are to establish before diploids do.
+that a migrant individual is tetraploid will be very small ($\sim u^2$, where
+$u$ is the probability that a diploid produces an unreduced gamete in meiosis),
+so that tetraploids need a substantial advantage during colonization if they
+are to establish before diploids do.
 
@@
 
-Here we develop a rather general model for the establishment of a mixed-ploidy
-population in a novel, unoccupied habitat based on \cite{barton2018}.
+Here we develop a model for the establishment of a mixed-ploidy population in a
+novel, unoccupied habitat based on \cite{barton2018}.
 In order to establish in the novel habitat, the population has to adapt to
@@
 We assume fitness is determined by directional selection on a single polygenic
-trait which can be interpreted as log fitness at low density in the new
+trait, which can be interpreted as log fitness at low density in the new
 habitat.
@@
 $N_k$ & population size of the $k$-ploid cytotype \\
-$\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
+$\pi_k$ & deterministic equilibrium frequency of the $k$-ploid cytotype \\
 $u$ & probability of unreduced gamete formation ($u=u_{22}=1-u_{42}$)\\
@@
 This formulation is helpful in that it highlights that Mendelian segregation
-occurs independently in both parents when gametes are produced, which then
-combine additively to determine the offspring trait value.
+occurs independently in both parents to produce gametes, which then combine
+additively to determine the offspring trait value.
 This model applies readily to an autopolyploid population expressing a trait
 with infinitesimal genetics.
-For instance, when an autotetraploid produces a diploid gamete, it will pass on
-half its trait value to the gamete on average, with a variance determined by
-the details of tetraploid meiosis (which are not, in general, identical to
-those of diploid meiosis, see below).
-
-In a finite population, the segregation variance will decay over time
-as the population becomes more inbred.
-Indeed, Mendelian segregation generates less variation among the gametes
-produced by individual $i$ when that individual is more inbred, as segregation
-at homozygous loci does not generate any variation.
+However, the segregation variance will be determined by the details of
+tetraploid meiosis, which differ from those of diploid meiosis (see below).
+
+In a finite population, the segregation variance will decay over time as the
+population becomes more inbred (Mendelian segregation at homozygous loci does
+not generate any variation).
 When $F_i$ is the inbreeding coefficient relative to some ancestral reference
@@
 
-If we would somewhat naively assume that the allelic effects underlying an
-additive quantitative trait are identical across ploidy levels, 
-a tetraploid offspring from a cross between two diploids would yield, on average,
-a trait value which is the sum of the parental trait values, with a
-distribution that depends on the variance generated by the process producing
-unreduced gametes.
+If we would naively assume that the allelic effects underlying an additive
+trait are identical across ploidy levels, 
+a tetraploid offspring from a cross between two diploids would have, on average,
+a trait value which is the sum of the parental trait values.
 This is not likely to reflect biological reality: tetraploids do not tend to
@@
 
-For the production of diploid gametes by a $k$-ploid individual, the
-segregation variance depends not only on the segregation variance in the base
-population ($V$) and the inbreeding coefficient ($F$), but also on the detailed
-assumptions on how these aberrant meiotic processes occur.
-Importantly however, these details only affect the gametic segregation variance
+The gametic segregation variance associated with the production of diploid
+gametes depends not only on the segregation variance in the base population
+($V$) and the inbreeding coefficient ($F$), but also on the detailed
+assumptions of how the meiotic process takes place.
+Importantly however, the latter only affect the gametic segregation variance
 through the quantity $\alpha_k$, which is the probability that a $k$-ploid
@@
 biallelic locus, and investigated the interaction between inbreeding depression
-and migration load (where inbreeding depression is modeled as an immediate
-fitness reduction in offspring produced by selfing).
+and migration load (where inbreeding depression is modeled as a fixed fitness
+reduction in offspring produced by selfing).
 In his model, inbreeding depression is different between cytotypes, so that

### 2025-01-23 doc/paper/paper.tex
revisions

@@
     \textbf{Keywords:} polyploidy, adaptation, establishment, inbreeding,
-    infinitesimal model
+    quantitative genetics 
 \end{abstract}
@@
 Tetraploids may have higher relative fitness than their diploid counterparts
-due to reduced inbreeding depression \citep{ronfort1999}, or due to being
+due to reduced inbreeding depression \citep{ronfort1999,otto2000}, or due to being
 better adapted to (changing) environmental conditions \citep{vandepeer2021}.
@@
 island population becomes positive.
-Throughout the paper, we define establishment as reaching a population of size
-100.
+As the population becomes better adapted, eventually, some form of density
+regulation must limit its growth.
+We ignore density regulation throughout and focus on the initial establishment
+phase, defining establishment as reaching a population of size 100.
 
@@
 long term (\cref{fig:vztet,s-fig:fininf}).
-
 Simulations for the mixed-ploidy model further confirm the correctness of our
-infinitesimal approximation (\cref{s-fig:vz}), yielding accurate predictions
-for the evolution of the phenotypic variance across ploidy levels and for
-different assumptions regarding the scaling of allelic effects ($\beta$
-parameters) and unreduced gamete formation ($\alpha$ parameters).
+infinitesimal approximation (\cref{s-fig:vz}).
 
@@
 \cref{s-sec:ttdip}).
-As a result, the polyploid subpopulations will show an average relatedness
-similar to that of the diploid population and not evolve as an isolated
-higher-ploidy population would.
 A nonzero probability of producing IBD diploid gametes ($\alpha_k > 0$) will
@@
 
-\subsection*{Establishment under recurrent migration}
+\subsection*{Establishment with recurrent migration}
 
@@
     Establishment with recurrent migration.
-    (A) Expected time until a population is established in the novel habitat
+    (A) Expected time until a population is established in the marginal habitat
     for increasing rates of migration and different degrees of maladaptation
-    ($\bar{z}_s$, the mean trait value in the source population). Results are
-    shown for the case with $\alpha_k = 0$ for $k=2,3,4$. 
+    ($\theta$). Results are shown for the case with $\alpha_k = 0$ for
+    $k=2,3,4$. 
     (B) Proportion of simulation replicates in which tetraploids established.
@@
 diploid one does.
-More specifically, we are interested in how maladaptive gene flow and MCE
-(which both increase with $m$) affect the relative establishment probability
-and the time until establishment.
 
@@
 source population.
-However, MCE is expected to be stronger than the latter effect,
-precluding the establishment of tetraploids as $m$ increases.
 Lastly, as the mean trait value on the island increases in diploids
-during adaptation, newly formed tetraploids on the island will, on average,
-have more extreme phenotypes than diploid offspring when $\beta_4 > 1/2$, which
-may also aid their establishment (irrespective of $m$).
+during adaptation, tetraploid offspring will have more extreme phenotypes on
+average than diploid offspring when $\beta_4 > 1/2$, which may also aid their
+establishment (irrespective of $m$).
 
@@
     Evolution of the mean inbreeding coefficient and trait value across
-    establishment replicates, averaged by population size, for increasing rates
+    simulation replicates where diploids (blue) or tetraploids (orange)
+    established eventually.
+    Average $F$ and $z$ by population size are shown for increasing rates
     of migration ($m$) from the predominantly diploid source population. 
@@
 influx of potentially succesful migrants, but later increases with increasing
-migration as a result of swamping by gene flow (\cref{fig:estwmig}A).
+migration due to swamping by gene flow (\cref{fig:estwmig}A).
 Importantly, the tetraploid establishment probability is considerably larger
@@
 increased by a factor $(1+\alpha_2)$, thereby increasing the chance that a
-newly formed tetraploid is well-adapted to the marginal habitat.
+tetraploid migrant is well-adapted to the marginal habitat.
 The rate of double reduction ($\alpha_4$) has a more limited effect
@@
 On the one hand, inbreeding is slower in tetraploids, so that during adaptation
-and establishment from a single or limited number of barely inbred individuals, 
+and establishment from a single or limited number of outbred individuals, 
 the inbreeding coefficient is expected to increase less rapidly conditional on
@@
 On the other hand, migration mostly introduces unrelated diploids, which cross
-more readily with diploids than tetraploids, so that migration reduces the
-average relatedness in established diploid populations much more than in
-established tetraploid populations.
+more readily with diploids than tetraploids, reducing the average relatedness
+more strongly in established diploid than in tetraploid populations.
 
@@
 inbreeding and the resulting increased adaptive potential of tetraploids.
-Another consequence of this is that, in the absence of migration, the average
-trait value in tetraploids at low population size (conditional on eventual
-establishment) is lower than that in diploids.
+Another consequence of this is that the average trait value in tetraploids at
+low population size (conditional on eventual establishment) is lower than that
+in diploids.
 This is because, in order to adapt before genetic variation is exhausted,
@@
 first couple of generations.
-
-It is clear from \cref{fig:estwmig2} that migration is not only maladaptive,
-but also introduces genetic variation upon which selection can act, thereby
-increasing the mean trait value at the time of establishment for both
-tetraploids and diploids.
-Interestingly, when migration is strong, tetraploids have markedly larger
-phenotypic values than diploids ($m=1$ in \cref{fig:estwmig2}).
-Diploids suffer strongly from maladaptive gene flow when the population size is
-low, while tetraploids enjoy some reproductive isolation.
+The similar trait means for weak migration suggest that establishment depends
+mostly on the chance pick of a well-adapted migrant (note that the $m=0$
+simulations in \cref{fig:estwmig2} assume the initial migrant has $z=0$,
+whereas the $m=0.01$ simulations assume the initial trait value to be
+Gaussian).
+When migration is strong, tetraploids have markedly larger trait values
+than diploids ($m=1$ in \cref{fig:estwmig2}), showing that diploids suffer
+strongly from maladaptive gene flow when the population size is low, while
+tetraploids enjoy some reproductive isolation.
 Furthermore, in these replicates, tetraploids tend to emerge and rise in
@@
 from diploids that already experienced several generations of selection.
-Neotetraploids deriving from parents with $z>0$ will have more extreme
-phenotypes on average (see methods) and hence be better adapted.
-
-\subsection*{Loss of self-incompatibility and assortative mating}
-
-We now briefly examine the effects of prezygotic isolation mechanisms that may
-be associated with polyploidization and that could alleviate MCE.
-Plausible sources of prezygotic isolation include assortative mating by
-cytotype and self-fertilization.
+These neotetraploids, deriving from diploid parents with $z>0$, will have more
+extreme phenotypes on average (see methods) and hence be better adapted.
+
+\subsection*{Loss of self-incompatibility, selfing and assortative mating}
 
 When polyploidization disrupts an existing SI system (see e.g.
-\cite{robertson2011comparative,zenil2019,novikova2023}), we expect
-that tetraploids suffer less from MCE, as an appreciable portion of their
-ovules are now assured to be fertilized by diploid gametes, irrespective of the
-composition of the population.
+\cite{robertson2011comparative,zenil2019,novikova2023}), we expect that
+tetraploids suffer less from MCE, as some portion of their ovules are now
+assured to be fertilized by diploid gametes, irrespective of the composition of
+the population.
 At the same time, we expect that accelerated inbreeding in selfing tetraploids
-diminishes the adaptive advantage of tetraploids that we found above.
-As shown in \cref{fig:selfing}, we find that when polyploidization is
-associated with the loss of a SI system (i.e. when diploids are
-self-incompatible, but tetraploids are not), tetraploids have a strongly
-increases establishment probability.
+diminishes the adaptive advantage of tetraploids.
+We find that when polyploidization is associated with the loss of a SI system
+(i.e. when diploids are self-incompatible, but tetraploids are not),
+tetraploids have a strongly increased establishment probability
+(\cref{fig:selfing}).
 This is the case even when the selfing rate $\sigma$ in tetraploids is zero (in
@@
 the \textit{realized} selfing rate in tetraploids is $1/N$).
-
-Besides reducing MCE, selfing is also expected to simultaneously reduce the
-effects of maladaptive gene flow, so we would expect that tetraploid
-establishment is promoted with increasing migration rates. 
-However, increased migration would at the same time alleviate the issue of
-reproductive assurance in diploids, as a diploid migrant is more likely to find
-a compatible mate.
-We find that when the selfing rate is sufficiently high ($\ge 0.4$ in
-\cref{fig:selfing}A), the relative establishment probability of tetraploids
+Furthermore, we find that when the selfing rate is sufficiently high ($\ge 0.4$
+in \cref{fig:selfing}A), the relative establishment probability of tetraploids
 increases with increasing migration rate.
@@
 increased rates of self-fertilization (for instance due to altered flower
-morphology). As an illustration, \cref{fig:selfing}A also shows results where
+morphology).
+For the sake of comparison, \cref{fig:selfing}A also shows results where
 diploids are assumed to be self-compatible with $\sigma_2=0$ (i.e. random
@@
 migration still promotes the probability of tetraploid establishment when the
-selfing rate in polyploids is sufficiently large.
+selfing rate in polyploids is sufficiently large compared to the diploid
+selfing rate.
 
@@
 somehow more tolerant to extreme environmental conditions (i.e. they somehow
-are intrinsically more fit than diploids at the range margin), or whether other
+are intrinsically more fit than diploids in marginal habitats), or whether other
 aspects of the population dynamics of mixed-ploidy populations may favor the
-establishment of polyploid subpopulations in peripheral habitats.
+establishment of polyploid subpopulations.
 
 In this study, we worked out the infinitesimal model for an additive polygenic
-trait in autotetraploids and mixed-ploidy populations.
-It would be very interesting to experimentally assess whether typical
-quantitative traits in mixed-ploidy populations follow approximately the
-mixed-ploidy infinitesimal model we outlined in this paper, and to estimate the
-relevant model parameters (in particular the $\beta_3$ and $\beta_4$ values).
-
-We used the mixed-ploidy infinitesimal model to study the establishment of
-tetraploids in a marginal habitat by means of individual-based simulations. 
+trait in autotetraploids and mixed-ploidy populations and used it to study the
+establishment of tetraploids in a marginal habitat by means of individual-based
+simulations.
 Assuming the trait to be under directional selection in the marginal habitat,
@@
 to establish a stable population.
-Importantly, we assume no intrinsic advantage of polyploids in the marginal
-habitat, i.e. the expected fitness of a migrant individual is the same
-regardless of the ploidy level.
+
+Importantly, we assume no intrinsic advantage or disadvantage of polyploids in
+the marginal habitat, i.e. the expected fitness of a migrant individual is the
+same regardless of the ploidy level.
 Differences in establishment probabilities are hence caused solely by aspects
@@
 of different ploidy levels.
+This is undoubtedly unrealistic.
+For instance, neopolyploids are likely to suffer intrinsic fertility issues due
+to meiotic irregularities associated with multivalent formation
+\citep{bomblies2016,novikova2023}.
+Furthermore, in our simulations, we have focused on the case where the
+equilibrium genetic variance is constant across cytotypes ($\beta_4=\sqrt{1/2}$
+in our model).
+It would be very interesting to experimentally assess whether typical
+quantitative traits in mixed-ploidy populations follow approximately the
+mixed-ploidy infinitesimal model we outlined in this paper, and to actually
+estimate how the variance scales across ploidy levels.
+In addition, we have assumed a relatively high and constant rate of unreduced
+gamete formation $u$ and triploid fertility $v$ in all our simulations (5\%),
+whereas these are known to be variable across the population, and at least in
+part genetically determined \citep{kreiner2017,clo2022c}.
+We ignore such complications, and hence do not take the actual establishment
+probabilities very serious, focusing instead on how migration load and
+prezygotic isolation affect the tetraploid establishment probability.
 
@@
 a typical migrant (and hence suffer less maladaptive gene flow), MCE will
-preclude polyploid establishment as the rate of migration grows.
-Nevertheless, it is important to remark that, the probability of tetraploid
-establishment in the marginal habitat can be an order of magnitude higher than
-expected based on the frequency of tetraploid migrants (i.e. is roughly $O(u)$
-instead of $O(u^2)$) when migration is sufficiently weak and maladaptation
-sufficiently high.
+increasingly hamper the establishment of tetraploids as the rate of migration
+grows.
+Nevertheless, it is important to remark that despite MCE, the probability of
+tetraploid establishment in the marginal habitat can be an order of magnitude
+higher than expected based on the frequency of tetraploid migrants (i.e. is
+roughly $O(u)$ instead of $O(u^2)$) when migration is sufficiently weak and
+maladaptation sufficiently high.
 
@@
 
-A major weakness of the present work is that we have ignored inbreeding
-depression and dominance throughout.
+A major weakness of the present work, and an important caveat, is that we have
+ignored inbreeding depression and dominance throughout.
 Including dominance in the infinitesimal framework is already challenging for
-diploids (requiring the tracking of 4-way identity coefficients;
+diploids (requiring the tracking of four-way identity coefficients;
 \cite{barton2023}), and appears intractable for higher ploidy levels.
-Nevertheless, autopolyploidy has important consequences whenever dominance is
-relevant, as in the case of inbreeding depression.
+However, autopolyploidy has important consequences whenever dominance is
+relevant, as in the case of inbreeding depression \citep{ronfort1999}.
 Indeed, when inbreeding depression is due to recessive deleterious variation,
-it should be less expressed in polyploids simply because, for the same allele
-frequency, homozygous genotypes are much rarer in polyploids than in diploids
-at Hardy-Weinberg equilibrium.
+it is expected to be less expressed in neotetraploids because homozygous genotypes
+should be much rarer than in their diploid parents.
 When there is indeed deleterious recessive variation, inbreeding during the
@@
 tetraploid establishment.
-In addition, inbreeding itself occurs at a slower rate in polyploids, further
-decreasing inbreeding depression during establishment.
+How this plays out depends on the \textit{rate} of inbreeding, which will
+differ between cytotypes. 
+In outcrossing populations, inbreeding occurs at a slower rate in tetraploids,
+further decreasing inbreeding depression and aiding tetraploid establishment.
+However, when polyploidization is associated with increased selfing (as when it
+disrupts an existing SI system), increased inbreeding depression in
+autotetraploids may prevent their establishment.
 
@@
 probability.
-\cite{griswold2021} performed a study somewhat similar to ours, but did include
-a simple model of inbreeding depression.
-The author focused on the case where local fitness is determined by a single
-biallelic locus, and investigated the interaction between inbreeding depression
-and migration load (where inbreeding depression is modeled as a fixed fitness
-reduction in offspring produced by selfing).
-In his model, inbreeding depression is different between cytotypes, so that
-tetraploids are able to produce more offspring through selfing relative to
-diploids, who have to rely more on outcrossing.
+\cite{griswold2021} studied the case where local fitness is determined by a
+single biallelic locus, and investigated the interaction between inbreeding
+depression and migration load (where inbreeding depression is modeled as a
+fixed fitness reduction in offspring produced by selfing).
+In his model, inbreeding depression is different between cytotypes (assuming
+stronger inbreeding depression in diploids), so that tetraploids are able to
+produce more offspring through selfing relative to diploids, who have to rely
+more on outcrossing.
 However, outcrossing incurs maladaptive gene flow, and thereby puts the
@@
 It would be very interesting to combine the infinitesimal framework with some
-form of inbreeding depression to investigate in a somewhat more realistic model
-whether the combination of maladaptive migration and differential inbreeding
-depression could explain the prevalence of polyploid subpopulations at range
-edges.
+form of inbreeding depression to investigate in a more realistic model whether
+the combination of maladaptive migration and differential inbreeding depression
+could explain the prevalence of polyploid subpopulations at range edges.
 
@@
 In our case, polyploids are always recently descended from diploid ancestors,
-so they will not have accumulated more deleterious recessives than their
+and they will not have accumulated more deleterious mutations than their
 diploid counterparts, so that polyploidy leads to reduced rather than increased
-inbreeding depression (as discussed above).
-An interesting picture emerges here however, where the different effects of
-polyploidy in the short and long term lead to some sort of equilibrium
-situation that may characterize mixed-ploidy populations in nature: although
-polyploids may enjoy enhanced establishment probabilities in peripheral
-habitats, the accumulation of mutational load may in the long-term preclude
-further range expansion or competitive exclusion of their diploid progenitors
-in the core habitat.
+inbreeding depression when selfing rates are similar (as discussed above).
+Interestingly, the interplay between the effects of polyploidy on different
+timescales could yield an equilibrium situation that may characterize
+mixed-ploidy populations in nature: although sometimes polyploids could enjoy
+enhanced establishment probabilities in peripheral habitats, the accumulation
+of mutational load may in the long-term limit further range expansion or
+even lead to competitive exclusion by diploids.
 Further modeling efforts could provide more insights into the plausibility of
@@
 
-I thank Christelle Fraïsse for employment, encouragement and feedback. I
-acknowledge funding from the Research Foundation -- Flanders (FWO, Junior
+I thank Christelle Fraïsse for encouragement and feedback. Thanks to Quinten
+Bafort, Felipe Kauai and Frederik Mortier for comments on an early draft
+version.
+I acknowledge funding from the Research Foundation -- Flanders (FWO, Junior
 Postdoctoral Fellowship 1272625N).

### 2025-03-13 doc/paper/paper.tex
starting revisions

@@
 \usepackage{authblk}
+\usepackage{lineno}
+\linenumbers
 \externaldocument[s-]{supp}
@@
 \title{Autopolyploid establishment through polygenic adaptation}
-\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
-\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
+%\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
+%\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
+\author{}
 \date{\vspace{-5ex}}
@@
 \begin{abstract}
-We define the infinitesimal model of quantitative genetics (\textit{sensu}
-    \cite{barton2017}) for the inheritance of an additive quantitative trait in
-    a mixed-ploidy population consisting of diploid, triploid and
-    autotetraploid individuals producing haploid and diploid gametes.  We
-    implement efficient simulation methods and use these to study the
+We define the infinitesimal model of quantitative genetics for the inheritance
+    of an additive quantitative trait in a mixed-ploidy population consisting
+    of diploid, triploid and autotetraploid individuals producing haploid and
+    diploid gametes.
+    We implement efficient simulation methods and use these to study the
     quantitative genetics of mixed-ploidy populations and the establishment of
-    autotetraploids in a new habitat.  We show that for rare migration from the
-    source population, autotetraploids are more likely to found a successful
-    population than diploids under a very broad range of conditions, but that
-    this is unlikely to sufficiently counter the scarcity of tetraploid
-    founders when the source is predominantly diploid.  We assess in more
-    detail how minority cytotype exclusion interacts with migration load in the
-    establishment process and evaluate the impact of additional sources of
-    prezygotic isolation, specifically selfing and assortative mating, on the
-    relative establishment probabilities of the different cytotypes.
+    autotetraploids in a new habitat.
+    We show that for rare migration from the source population, autotetraploids
+    are more likely to found a successful population than diploids under a very
+    broad range of conditions, but that this is unlikely to sufficiently
+    counter the scarcity of tetraploid founders when the source is
+    predominantly diploid.
+    We assess in more detail how minority cytotype exclusion interacts with
+    migration load in the establishment process and evaluate the impact of
+    additional sources of prezygotic isolation, specifically selfing and
+    assortative mating, on the relative establishment probabilities of the
+    different cytotypes.
     In the discussion we consider how inbreeding depression may impact our
@@
 \section*{Introduction}
-
-% Should connect more with literature and concepts of isolating mechanisms pre
-% vs. postzygotic isolation, and reinforcement etc. (see e.g. Parvathy &
-% Himani's paper introduction)
 
@@
 phenological differences across cytotypes, or differences in pollinators;
-\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, and
-localized dispersal \citep{baack2005,kolar2017}.
+\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, 
+localized dispersal \citep{baack2005,kolar2017} and asexual reproduction
+\citep{vandrunen2022}.
 Finally, selection may be invoked to explain the establishment of polyploids.
 Tetraploids may have higher relative fitness than their diploid counterparts
-due to reduced inbreeding depression \citep{ronfort1999,otto2000}, or due to being
+due to reduced inbreeding depression
+\citep{ronfort1999,otto2000,husband2008,clo2022b}, or due to being
 better adapted to (changing) environmental conditions \citep{vandepeer2021}.
@@
 the proportion of euploid gametes produced by a triploid individual.
+The rate of unreduced gamete production is hence fixed across individuals, and
+is assumed to be the same in diploids and tetraploids.
 
@@
 This is not likely to reflect biological reality: tetraploids do not tend to
-have, for instance, twice the size of their diploid progenitors on average.
+have, for instance, twice the size of their diploid progenitors on average
+(e.g. \cite{porturas2019}).
 Furthermore, the genetic variance at Hardy-Weinberg and linkage equilibrium
@@
 The expectation on the right hand side can be calculated from \cref{eq:oneline}
-using the moment-generating function of the Gaussian.
+using the moment-generating function of the Gaussian, i.e.
+\begin{equation}
+    \Ex\left[e^{\gamma(Z_{ij}^{kl} - \theta)}\right] 
+    = e^{\gamma \left(\Ex \left[Z_{ij}^{kl}\right] -\theta\right) +
+    \frac{\gamma^2}{2}
+    \Var\left[Z_{ij}^{kl}\right]} 
+\end{equation}
 Having calculated the $w_{ij}$, the number of offspring surviving into the next
@@
 implemented in Julia \citep{julia}.
-Documented code and simulation notebooks are available at
-\url{https://github.com/arzwa/InfGenetics}.
+Documented code and simulation notebooks are available at \url{[anonymized]}.
+%\url{https://github.com/arzwa/InfGenetics}.
 
@@
 critical branching process with a Poisson offspring distribution, so that the
-probability to reach $N=100$ is $1/100$ \citep{barton2018}.
+probability to reach $N=100$ is $\sim 1/100$ \citep{barton2018}.
 
@@
 to meiotic irregularities associated with multivalent formation
-\citep{bomblies2016,novikova2023}.
-Furthermore, in our simulations, we have focused on the case where the
-equilibrium genetic variance is constant across cytotypes ($\beta_4=\sqrt{1/2}$
-in our model).
+\citep{bomblies2016,novikova2023}, and triploids may be inviable due to
+issues with endosperm development \citep{bretagnolle1995}.
+Similarly implausible is the assumption of a constant equilibrium genetic
+variance across cytotypes ($\beta_4=\sqrt{1/2}$ in our model), which we used in
+most of our results.
 It would be very interesting to experimentally assess whether typical
@@
 higher than expected based on the frequency of tetraploid migrants (i.e. is
-roughly $O(u)$ instead of $O(u^2)$) when migration is sufficiently weak and
+roughly of order $u$ instead of $u^2$) when migration is sufficiently weak and
 maladaptation sufficiently high.
@@
 However, autopolyploidy has important consequences whenever dominance is
-relevant, as in the case of inbreeding depression \citep{ronfort1999}.
+relevant, as in the case of inbreeding depression
+\citep{ronfort1999,husband2008,clo2022b}.
 Indeed, when inbreeding depression is due to recessive deleterious variation,
 it is expected to be less expressed in neotetraploids because homozygous genotypes
-should be much rarer than in their diploid parents.
-When there is indeed deleterious recessive variation, inbreeding during the
+should be much rarer than in their diploid parents (the `masking' effect;
+\cite{otto2000}).
+In the presence of deleterious recessive variation, inbreeding during the
 establishment process should therefore incur a higher fitness cost in diploids
@@
 tetraploid establishment.
-How this plays out depends on the \textit{rate} of inbreeding, which will
-differ between cytotypes. 
+How this plays out depends however on the \textit{rate} at which populations
+become inbred, which will differ between cytotypes and will depend strongly on
+the mating system.
 In outcrossing populations, inbreeding occurs at a slower rate in tetraploids,
@@
 and they will not have accumulated more deleterious mutations than their
-diploid counterparts, so that polyploidy leads to reduced rather than increased
-inbreeding depression when selfing rates are similar (as discussed above).
+diploid counterparts, so that polyploidy should lead to reduced rather than
+increased inbreeding depression when selfing rates are similar (as discussed
+above).
 Interestingly, the interplay between the effects of polyploidy on different
-timescales could yield an equilibrium situation that may characterize
+timescales could yield an equilibrium situation that may characterize many
 mixed-ploidy populations in nature: although sometimes polyploids could enjoy
@@
 
-\section*{Acknowledgements}
-
-I thank Christelle Fraïsse for encouragement and feedback. Thanks to Quinten
-Bafort, Felipe Kauai and Frederik Mortier for comments on an early draft
-version.
-I acknowledge funding from the Research Foundation -- Flanders (FWO, Junior
-Postdoctoral Fellowship 1272625N).
+%\section*{Acknowledgements}
+%
+%I thank Christelle Fraïsse for encouragement and feedback. Thanks to Quinten
+%Bafort, Felipe Kauai and Frederik Mortier for comments on an early draft
+%version.
+%I acknowledge funding from the Research Foundation -- Flanders (FWO, Junior
+%Postdoctoral Fellowship 1272625N).
 

### 2025-03-13 doc/paper/titlepage.tex
starting revisions

\documentclass[12pt,a4paper]{article}
\input{preamble.tex}
\usepackage[margin=1in]{geometry}
\usepackage{authblk}
\externaldocument[s-]{supp}

\begin{document}
\title{Autopolyploid establishment through polygenic adaptation}
\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
\date{\vspace{-5ex}}
\maketitle

\subsection*{Data availability}

Individual-based simulations for the mixed-ploidy infinitesimal model were
implemented in Julia \citep{julia}.
Documented code and simulation notebooks are available at 
\url{https://github.com/arzwa/InfGenetics}.

\subsection*{Funding}

I acknowledge funding from the Research Foundation -- Flanders (FWO, Junior
Postdoctoral Fellowship 1272625N).

\subsection*{Conflict of interest}

I have no conflict of interest to declare.

\subsection*{Acknowledgements}

I thank Christelle Fraïsse for encouragement and feedback. Thanks to Quinten
Bafort, Felipe Kauai and Frederik Mortier for comments on an early draft
version.

\bibliographystyle{abbrvnat}
\bibliography{/home/arzwa/vimwiki/bib.bib}

\end{document}

### 2025-03-24 doc/paper/response.md
update

## Associate Editor

>Thank you for submitting your article to Evolution. It has now received two
reviews, both of which recognize its valuable contributions and essentially
recommend publication after relatively minor revision. I share these views.
While both reviews are positive, they make a number of worthwhile suggestions
for you to consider for revising the manuscript. One reviewer has suggested
combining the Results and the Discussion into a single section to avoid
unnecessary discussion of the results in the current Results section.
Personally, I would favor retaining both sections, **moving more discussive text
from the Results to the Discussion**. I leave the decision to you, but I do agree
with the reviewers that some resorting of the Results text is needed. I also
agree that the **Introduction could be better motivated by referring more to
empirical work**, and the **Discussion section could relate the results more to
other theoretical findings** in the literature. Overall, think that revision will
be relatively straightforward, but I would ask that you consider all the
suggestions made by the reviewers and, if you disagree with them, that you
justify and explain your reasons in a response letter.

I have now revised the Introduction to include more discussion of empirical
work, although I also, for the sake of brevity, refer the reader to the article
by Griswold (2021) to this end, given that he provides a nice overview relevant
studies. **TODO!**

Regarding the Results and the Discussion sections: I do not want to do away
with Discussion section altogether (as suggested by Reviewer 1), given that I
wish to discuss factors relevant to the problem which are not studied in my
article (dominance, inbreeding depression, *etc.*) and want to relate my work
to other recent theoretical work.
My aim is to describe and interpret the results in the Results section, and to
discuss the limitations and links with other work in the Discussion section.
I have sought to move 'more discussive text' to the Discussion section,
although I found there was fairly limited scope to do so when I seek to keep
the interpretation of the results more or less self-contained. **TODO!**

I have expanded the discussion slightly to include more of the previous
theoretical literature on the subject, in particular following the suggestions
of Reviewer 2. **TODO!**

## Reviewer 1

>This interesting manuscript investigated how the infinitesimal model can be
expanded to polyploidy populations (triploids and tetraploids here). It
proposes a mathematical formulation of a polyploid population's expected
genetic diversity for a given amount of inbreeding. The model is then used to
study how polyploidy can affect the adaptation to a new habitat (based on a
sink-source model).
I found the manuscript to be interesting and only have a few things to say
about it. I will for sure motivate further expansion of the model to understand
the evolution of autopolyploidy.
Having no line number does not help to comment precisely on the manuscript; I
did my best to be as precise as possible.

Thank you for these encouraging words, and my apologies for the oversight
when it comes to line numbers!

>Results are already well discussed when presented. I would either merge the
results and discussion sections (as in brief communications for example),
because it reads well as presented, or remove any explanations/discussion from
the results to put them in the discussion section. The first option sounds
better to me.

I appreciate the comment and have reworked the text with this in mind.
However, following the suggestion of the associate editor, I went with the
second option. **TODO**

>A small concluding section where the authors mentioned other questions to
study with their model will be useful in my opinion. The author proposed to
complexify the current model of adaptation with dominance or to relax some
hypotheses, but I think other biological scenarios could be studied too.

**TODO**

>Defining establishment when N=100: Is it a classic way to define establishment
in such models?  Or does it come from your experience with the model and you
see that Under N=100 extinction remains likely? More details would be needed
here.

$N=100$ is indeed an arbitrary choice. This threshold was also adopted in
@barton2018, so it makes comparisons of establishment probabilities *etc.* with
their paper somewhat more straightforward.
$N=100$ seems to be a reasonable threshold in that extinction becomes
sufficiently unlikely, while we do not have to simulate large populations (our
approach requires tracking the $N \times N$ matrix of identity coefficients). 
I did a check by putting the threshold at $N=150$, and the baseline predictions
(as in fig. 1) do not change significantly.

>Abstract: I am not sure the reference to Barton et al. (2017) is appropriate
here. Maybe a more general definition would fit better.

The reason for the reference is that their appears to be some confusion about
different models that bear the name 'infinitesimal model' (as outlined in for
instance @turelli2018 and @walsh2018). I wanted to make it clear that we use
the "Gaussian descendants" infinitesimal model, which is very carefully
outlined in @barton2017. However, I guess this is sufficiently clear from the
introduction and the methods sections, so the reference in the abstract can
indeed be omitted, as it is in the revised manuscript.

>Introduction: P2 (lower inbreeding depression (ronfort 1999, Otto & Whitton
2000): Some empirical data?  Husband et al. (2008), Clo & Kolar (2022).

I have now added the suggested references.

>Model and methods: P3-P4 (mixed ploidy population model): maybe write that the
rate of reduced gametes is fixed and genetically inherited from diploids to
tetraploids. It is clear from the equations but I suspect it could be clearer
for non-theoretical people who will not necessarily focus that much on
equations or matrices.

I added a sentence to emphasize this. I note that this is also emphasized in
the discussion "In addition, we have assumed a relatively high and constant
rate of unreduced gamete formation $u$ and triploid fertility $v$ in all our
simulations (5%), whereas these are known to be variable across the population,
and at least in part genetically determined (Kreiner et al., 2017a; Clo et al.,
2022)"

>P4 (tetraploids are not twice as big as diploids): Some references would be
needed to justify (Porturas et al. 2019 for example).

I have added the reference to Porturas et al. (2019).

>Discussion about inbreeding depression: again, some empirical data are
available to support some statements (Husband et al. (2008), Clo & Kolar
(2022)).

**TODO**

## Reviewer 2

>This paper is an important foundational work in the area of theoretical polyploid
evolutionary genetics. It presents an infinitesimal model of trait evolution for
triploids and tetraploids, and then applies the model to assess conditions under
which a tetraploid population will establish in a peripheral region connected to
a central core by dispersal. In terms of the infinitesimal model, several aspects
are very helpful
- Demonstrating how genetic effects are scaled across triploids and tetraploids
such that segregating variance is equivalent across ploidies, including with diploids
- How to combine the genetic effects in an individual with mixed-ploidy ancestry
- Calculating inbreeding and co-ancestry values for mixed-ploidy individuals
- Calculating the effect of unreduced gamete formation on segregation variance
across diploids, triploids and tetraploids
Other informative analyses include recursive or Markov models of null expecta-
tions for frequencies of ploidies, times to diploid ancestry and effective popula-
tion size across a population of mixed-ploidy.
Besides this set of results, the paper then examines conditions under which a
tetraploid population is established in a peripheral population connected to a
central core population that is predominantly composed of diploids. Given the
additive nature of gene-action in the infinitesimal model, this analysis will likely
be of interest to empiricists because additive effects are thought to be the least
constrained basis for something to evolve and there is empirical evidence that
the genetic basis of polyploid establishment in some species is additive. This is
in contrast to the theoretical results of Griswold (2021), which highlighted the
potential for a recessive-basis for autotetraploid establishment in a peripheral
region.
Main comments focus on mutation and adding background/context.

I thank the reviewer for this appraisal.

>Mutation:
For low migration rates, the possibility that autotetraploid relative to diploid es-
tablishment is driven by differences in how inbreeding accumulates is interesting
(figure 4), but the model does not include mutation. New mutation dissociates
identity by descent from identity in state and likely slows the rate of loss of
segregation variance. It would be helpful to address how scaled mutation may
affect dynamics of segregating variance between diploids and autotetraploids
and whether mutation may affect results. Is there a relationship between μVm
(Barton et al. 2017) and m and V on establishment?
Recognizing the point seems to be important, but its analysis can be for later.

I have not considered mutation at all so far because I assumed it would not be
relevant for the population sizes and timescales considered in my study.
Specifically, any individual at the time of establishment derives from a
completely outbred migrant individual a relatively short time in the past (10
to 100 generations, say), so the contribution of new mutation to differences in
establishment probability between diploids and tetraploids should be
negligible.
This is definitely the case when there is ongoing migration: as long as $m$ is
sufficiently greater than $\mu$ and $V_m$ is sufficiently smaller than the
genetic variance in the mainland population, we will have that for every
mutation contributing a very slight decrease in the loss of segregation
variance on the island, there will be many more migrant arrivals, which each
introduce a completely unrelated genotype.

Of course, when considering the mixed-ploidy infinitesimal model independently
of the establishment question, mutation does become a relevant topic. I mention
this in the revised discussion but I think it is out of scope to discuss this
in any detail in the paper. I think that my approach for scaling across ploidy
levels that is explicitly based on underlying allelic effects in a finite-locus
model extends readily to mutation. The main difference between ploidy levels is
of course that the mutational target size is proportional to the ploidy level,
so that, unless the mutational variance is exactly halved in tetraploids, the
contribution of mutation to genetic variance ($\mu V_m$) in tetraploids is
larger than in diploids.

>In the intro providing empirical examples of polyploid establishment in
peripheral habitats would broaden readership and connections to the
literature. Also, reviewing empirical work on the genetics of adaptation to
peripheral environments in polyploids would broaden connections to the
literature.

I agree with the relevance of empirical work on polyploid establishment and
adaptation to marginal habitats for the introduction of the paper. However, I
think that to present an actual review of empirical studies here is a bit out
of scope. Griswold presents a brief review in his 2021 paper, and I prefer to
refer to his paper instead of paraphrasing his paragraph.

Similarly, I think that reviewing empirical work on adaptation to peripheral
habitats is somewhat out of scope (note that the paper is already at the word
limit for Evolution research articles). I refer to an important (if somewhat
old) review article of Kawecki (2008) and recent theoretical work by Sachdeva
et al. I think this gives a good sample of where to look for more references.

>The discussion could relate findings to other theoretical works more. For ex-
ample, it does not go back and place the results in the context of Oswald and
Nuismer (2011), which assumed a two-locus model and varied dosage effects, as
well as competition between diploids and tetraploids, including no competition.
In addition it allowed for assortative mating.

**TODO** (Discussion)

>It would be informative to expand on the the consequences of different scalings
of genetic effects. This paper made the sensible choice to scale such that segre-
gation variance is equal between ploidies as a baseline. Griswold (2021) scaled
genetic effects such that autotetraploid and diploid individuals have equal fit-
ness as a function of proportional allele count in an individual. A consequence
of Griswold’s scaling is autotetraploids have lower additive genetic variance at
HWE (Gallais 2003, p. 185), so are expected to respond more slowly in the
absence of another process. Gallais (2003, p. 186) mentions some studies that
compared additive genetic variance between diploids and autotetraploids.

To some extent, the consequences of different scalings is taken up in the
results and discussion throughout the paper where relevant. In the methods
section the relationship between allelic effects and additive genetic variance
is stated explicitly. In the results shown in Fig. 2, the effect of scaling
assumptions on establishment probability is explicitly shown and described. In
the section on establishment with recurrent migration, the phenomenon that
polyploid offspring is more extreme on average when allelic effects are not
exactly scaled by one half (as in Griswold) is taken up. 
**TODO** (Discussion) **I now added some further discussion, referring
explicitly to Griswold and Gallais in the Discussion section. I now also refer
to Porturas et al.**

>Neither the intro nor the discussion revisit what seems to be an important point
in Barton & Etheridge (2018, p. 111) about local adaptation via infinitely small
effects.

I assume the reviewer is referring to the following issue: classical single
locus population genetics predicts that one needs $m < s$ for selection to
maintain a locally beneficial allele when migrants introduce the deleterious
allele, and that when adaptation is due to alleles of very small effect this
may seem to imply that there is little scope for local adaptation.

There are multiple reasons why this is not the case. The argument outlined in
@barton2018 is simply that the classical theory assumes complete divergence
between mainland and island, but that when adaptation is from standing
variation instead this doe snot hold, since allele frequency differences
between mainland and island are only slight. In the latter case, many alleles
of small effect can be divergently maintained by selection. Another reason why
this does not hold is linkage disequilibrium between locally beneficial
alleles, this is explored in detail in @sachdeva2022 and @zwaenepoel2024.

I do not think, however, that it is pertinent to discuss these issues in the
present paper -- they are somewhat technical and require quite some space to
explain and do not have much to do with the problem studied, i.e. polyploid
establishment.

### Other comments:

>p.1, Abstract: Use the word “found”, but is “establish” more accurate? Overall,
the corresponding sentence is difficult to understand/parse. Suggest revision.

This sentence has been revised.

>p. 1: “negative frequency dependence” -> Isn’t MCE a positive-frequency de-
pendent process?

Indeed, it is! This has been corrected.

>p. 2: Suggest citing Bulmer (1971).

I am unsure whether this is a good idea, as it may lead to some confusion as to
which version of the infinitesimal model we are working with.
The model described in Bulmer (1971) is not exactly equivalent to the Gaussian
descendants approximation that I use (as outlined in Turelli (2017), which I
cite).
If the reference is suggested by the reviewer in order to cite a more
'original' reference, I don't think there is a reason to single out Bulmer
(1971), one could refer to Fisher (1918) or work by Alan Robertson in the
1960's.
I think the reference to Turelli (2017) which reviews the history of the
infinitesimal model suffices, and prevents confusion as to what we mean when we
refer to 'the infinitesimal model'.

>p.7: Suggest adding “, across gene copy arrangements and ploidies” at the end
of the last sentence before the “Establishment” section.

I do not fully understand this suggestion. I personally find this makes the
sentence more confusing, as I never talk of gene copy arrangements.

>p. 7, Establishment section: The organization of paragraphs two and three in
this section is a bit confusing. N is used before it is formally defined. In my
first reading I thought there was migration and then selection, whereas there is
selection and then migration. It would also help support a self-contained paper
to give the expression for E[eγ(...) ].

As far as I can tell, $N$ is defined when first mentioned. I slightly altered
the sentence to make this more clear.

I added the expression for the expected fitness (eq. 11).

The life cycle is indeed as follows: migrants arrive, they mate randomly and
offspring survives with a probability that is determined by its fitness. This
is indeed 'selection before migration'. I tried to make this more clear.

>p. 11, figure 3 (and later figure 5): The smallest value of m is 0.1 in these
figures, but based on figure 4, it looks like m = 0.01 is a transition point from
processes shaped by the accumulation of inbreeding versus migration becoming
more of a factor. Can m be extended to 0.01?

I extended the relevant figures as suggested, as $m=0.01$ is perhaps more
representative of weak migration than $m=0.1$
Weak migration means essentially that $m^2$ is negligible, so migrants and
early generation descendants thereof do not mate with migrants and the whole
process amounts more or less to independent establishment trials (there is no
swamping).

I do not think, however, that fig 4 suggests such a transition point though. As
noted in the main text (**line numbers**), the difference between the $m=0$ and
$m=0.01$ results in fig. 4 are mostly due to the fact that the latter assumes a
Gaussian distribution of migrant trait values, whereas the former assumes a
fixed initial migrant trait value.

>p. 12: Add qualifier “as m increases” at end of last sentence in last full para-
graph.

Done.

>p. 14, Figure 5 caption: “migration and selfing”

Fixed.

>p. 4, suppl., figure S4: “formation”

Fixed.

>p. 6, suppl., equation 6: Should the square only be in the denominator?

That's right, thank you!

>p. 10, suppl.: “coancestry coefficients in (10) is “ - missing the “(10)”

Fixed, thank you!

>p. 12, suppl.: add “between centromere and locus” after “. . . recombination
happens with probability c”

Done.

>p. 13-15, suppl.: Some notation seemed to not be defined, e.g. VS(2,2), V0,3, ...

The notation is introduced on p.11 in the beginning of S2.6. I have slightly
reorganized the first paragraphs of S2.6 to make the notation and general
approach more clear.

### 2025-03-24 doc/paper/paper.tex
update

@@
     autotetraploids in a new habitat.
-    We show that for rare migration from the source population, autotetraploids
-    are more likely to found a successful population than diploids under a very
-    broad range of conditions, but that this is unlikely to sufficiently
-    counter the scarcity of tetraploid founders when the source is
-    predominantly diploid.
+    We show that, when migration from the source population is rare,
+    autotetraploids are more likely to establish in the new habitat 
+    than diploids under a very broad range of conditions, but that this
+    is unlikely to sufficiently counter the scarcity of tetraploid founders
+    when the source is predominantly diploid.
     We assess in more detail how minority cytotype exclusion interacts with
@@
 block', see \cite{ramsey1998,kohler2010,brown2024}).
-This negative frequency dependence effect is commonly referred to as
+This positive frequency dependence effect is commonly referred to as
 \textit{minority cytotype exclusion} (MCE), after \cite{levin1975}.
@@
 nature \citep{kolar2017,mortier2024}.
-In particular, polyploids are thought to establish mainly in novel, unoccupied
-habitats where they evade MCE (for instance at range edges, or after local
-extinction due to environmental change) \citep{griswold2021}.
-However, such peripheral habitats are likely to present adaptive challenges to
-establishment \citep{kawecki2008adaptation}, and if polyploids are able to
+Many empirical studies show a pattern where polyploids established in
+peripheral habitats at the edge of a species' range (reviewed in
+\cite{griswold2021}), in accordance with large-scale biogeographical patterns
+\citep{rice2019}.
+In a peripheral habitat, a new polyploid population may be more likely to reach
+an appreciable size as it may evade to some extent the negative effects of MCE.
+However, at the same time, such peripheral habitats are likely to present
+adaptive challenges to establishment
+\citep{kawecki2008adaptation,sachdeva2022b}, and if polyploids are able to
 colonize such habitats at an appreciable rate, they must somehow be better
@@
 $P_{ik} \in \{0, \frac1 3, \frac1 2, \frac2 3, 1\}$ is the probability that a
-gene copy in $i$ is derived from parent $k$.
+gene copy in individual $i$ is derived from parent $k$.
 
@@
 habitat) joining $N^\ast(t)$ resident individuals, where $M(t)$ is Poisson
-distributed with mean $m$. 
+distributed with mean $m$, and $N^\ast(0)=0$ unless stated otherwise. 
 The migrant individuals are assumed to be unrelated to the resident
 individuals.
-After migration in generation $t$, the $N(t) = N^\ast(t) + M(t)$ individuals
-reproduce sexually, and the offspring thus produced survives until the next
-generation with a probability determined by their trait value.
+After migration there are $N(t) = N^\ast(t) + M(t)$ individuals on the island
+which reproduce sexually, and the offspring thus produced survives until the
+next generation with a probability determined by their trait value.
+Note that this corresponds to a life cycle where selection occurs before
+migration within a generation.
 In the basic model, random selfing is allowed (but see below for a model with
@@
 \centering
-\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration-dr.pdf}
+\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/paper/figs/estwmig.pdf}
 \caption{
@@
 effective barrier to maladaptive gene flow, this advantage to establishment is
-outweighed by the negative effects of MCE.
+outweighed by the negative effects of MCE as $m$ increases.
 
@@
 \centering
-\includegraphics[width=0.9\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/prez.pdf} 
+\includegraphics[width=0.9\linewidth]{/home/arzwa/dev/InfGenetics/doc/paper/figs/prez.pdf} 
     \caption{
-    (A) Establishment with recurrent migration selfing in polyploids. 
+    (A) Establishment with recurrent migration and selfing in polyploids. 
     The solid lines show the case where diploids are self-incompatible 

### 2025-03-24 doc/paper/supp.tex
starting revisions
update

@@
 \title{Supplement: Autopolyploid establishment through polygenic adaptation}
-\author{Arthur Zwaenepoel}
+%\author{Arthur Zwaenepoel}
+\author{}
 \date{\vspace{-5ex}}
@@
     probability, but mainly due to $\alpha_2$, which increases the segregation
-    variance associated with the formatoin of diploid gametes by diploids.
+    variance associated with the formation of diploid gametes by diploids.
     All other parameters and simulation details are as in \cref{m-fig:estwmig}.
@@
   p_2 &= \left(\frac{i(1-u) + jv}{N(1-u) + (i+j)u + j(2v-1)}\right)^2 \\
-  p_3 &= \left(\frac{2(i(1-u) + jv)(N(1-u) + i(2u - 1) + j(u + v - 1))}
-    {N(1-u) + (i+j)u + j(2v-1)}\right)^2 
+  p_3 &= \frac{2(i(1-u) + jv)(N(1-u) + i(2u - 1) + j(u + v - 1))}
+    {\left(N(1-u) + (i+j)u + j(2v-1)\right)^2} 
 \end{align}
@@
     F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{if } & c_i = 4
+    \label{eq:coanc}
 \end{align}
@@
 autotetraploid model.
-The recursion for the coancestry coefficients in is given by
+The recursion for the coancestry coefficients in \cref{eq:coanc} is given by
 \begin{align}
@@
 \end{equation}
-We use this relationship to derive the segregation variance for a given meiotic
-process.
+We use this relationship to derive the segregation variance associated with
+haploid and diploid gamete production in the three cytotypes of the
+mixed-ploidy population.
+
+We write $V_{S(k,l)}$ for the segregation variance associated with a
+$k$-ploid individual producing an $l$-ploid gamete, and we express this in
+terms of the variance $\Vx_{k}$ associated with a haploid genome in a
+(hypothetical) $k$-ploid reference population at HWLE.
+Using the scaling assumptions outlined in the main text, i.e. $\Vx_{k} =
+\beta_k^2 V$, we can then express all segregation variances in terms of the
+diploid segregation variance in the reference population $V$.
 
@@
 Summing over $L$ independent loci, we find the gametic segregation variance in
-diploids (which is halve the zygotic segregation variance $V$, by definition) as
+diploids producing haploid gametes (which is halve the zygotic segregation
+variance $V$, by definition) as
   $$V_{S(2,1)} = \frac{V}{2} = \sum_{i=1}^L \frac{v_{x,i}}{2} = \frac{\Vx_{2}}{2}$$
@@
 $v_x$ smaller and smaller.
-
-Below, we derive the segregation variance associated with haploid and diploid
-gamete production in the three cytotypes of the mixed-ploidy population.
-We use \cref{eq:eve} to do this, and express the segregation variance for a
-$k$-ploid individual producing an $l$-ploid $V_{S(k,l)}$ in terms of the
-variance $\Vx_{k}$ associated with a haploid genome in a (hypothetical)
-$k$-ploid reference population at HWLE. Using the scaling assumptions outlined
-in the main text, i.e. $\Vx_{k} = \beta_k^2 V$, we can then express all
-segregation variances in terms of the diploid segregation variance in the
-reference population $V$.
 
@@
 Consider a locus in a diploid with two distinct genes $A$ and $a$. Assume
-recombination happens with probability $c$ and that conditional on unreduced
-gamete formation, formation is due to FDR with probability $f$ while it is due
-to SDR with probability $1-f$.
+recombination between the centromere and the locus happens with probability $c$
+and that conditional on unreduced gamete formation, formation is due to FDR
+with probability $f$ while it is due to SDR with probability $1-f$.
 The different unreduced gametes that are formed are represented schematically

### 2025-03-26 doc/paper/response.md
revising

@@
 
-I have now revised the Introduction to include more discussion of empirical
-work, although I also, for the sake of brevity, refer the reader to the article
-by Griswold (2021) to this end, given that he provides a nice overview relevant
-studies. **TODO!**
+I have now revised the Introduction slightly to include more discussion of
+empirical work, although, in the interest of brevity (the manuscript is already
+near the Evolution word limit), I mostly resorted to explicitly referring the
+reader to the article by @griswold2021. In the latter, the author provides a
+nice overview of relevant recent studies.
 
@@
 to other recent theoretical work.
-My aim is to describe and interpret the results in the Results section, and to
-discuss the limitations and links with other work in the Discussion section.
+I believe it is better to separate the latter topics from the description *and*
+interpretation of the actual results. 
 I have sought to move 'more discussive text' to the Discussion section,
@@
 However, following the suggestion of the associate editor, I went with the
-second option. **TODO**
+second option. I refer to my answer to the associated editor above.
 
@@
 approach requires tracking the $N \times N$ matrix of identity coefficients). 
-I did a check by putting the threshold at $N=150$, and the baseline predictions
-(as in fig. 1) do not change significantly.
+I did a check by putting the threshold at $N=150$, and the establishment
+probabilities do not change appreciably.
 
@@
 
-The reason for the reference is that their appears to be some confusion about
+The reason for the reference is that there appears to be some confusion about
 different models that bear the name 'infinitesimal model' (as outlined in for
-instance @turelli2018 and @walsh2018). I wanted to make it clear that we use
+instance @turelli2017 and @walsh2018). I wanted to make it clear that we use
 the "Gaussian descendants" infinitesimal model, which is very carefully
@@
 
-I have now added the suggested references.
+I have now added the suggested references, thank you.
 
@@
 
-I added a sentence to emphasize this. I note that this is also emphasized in
-the discussion "In addition, we have assumed a relatively high and constant
-rate of unreduced gamete formation $u$ and triploid fertility $v$ in all our
-simulations (5%), whereas these are known to be variable across the population,
-and at least in part genetically determined (Kreiner et al., 2017a; Clo et al.,
-2022)"
+I added a sentence below eq. 1 to emphasize this. I note that this is also
+emphasized in the discussion "In addition, we have assumed a relatively high
+and constant rate of unreduced gamete formation $u$ and triploid fertility $v$
+in all our simulations (5%), whereas these are known to be variable across the
+population, and at least in part genetically determined (Kreiner et al., 2017a;
+Clo et al., 2022)"
 
@@
 
-**TODO**
+I have added the suggested references.
 
@@
 are very helpful
+
 - Demonstrating how genetic effects are scaled across triploids and tetraploids
@@
 across diploids, triploids and tetraploids
+
 Other informative analyses include recursive or Markov models of null expecta-
@@
 affect dynamics of segregating variance between diploids and autotetraploids
-and whether mutation may affect results. Is there a relationship between μVm
+and whether mutation may affect results. Is there a relationship between $\mu V_m$
 (Barton et al. 2017) and m and V on establishment?
@@
 of scope. Griswold presents a brief review in his 2021 paper, and I prefer to
-refer to his paper instead of paraphrasing his paragraph.
+refer to his paper instead of paraphrasing his paragraph. Specifically, I now
+write:
+"Many empirical studies of mixed-ploidy populations show a pattern where
+polyploids established in peripheral habitats at the edge of a species’ range
+(reviewed in Griswold (2021)), and these observations for specific study
+systems are in accord with large scale biogeographical patterns (Rice et al.,
+2019)."
 
-Similarly, I think that reviewing empirical work on adaptation to peripheral
-habitats is somewhat out of scope (note that the paper is already at the word
-limit for Evolution research articles). I refer to an important (if somewhat
-old) review article of Kawecki (2008) and recent theoretical work by Sachdeva
-et al. I think this gives a good sample of where to look for more references.
+With regard to adaptation to peripheral habitats: I refer to an important (if
+somewhat old) review article of Kawecki (2008) and recent theoretical work by
+Sachdeva et al. I think this gives a good sample of where to look for more
+references.
+When it comes to empirical work on the *genetics of adaptation to peripheral
+environments in polyploids*, I am quite frankly not aware of relevant studies...
+Reviewing empirical work on adaptation to peripheral habitats in general is
+somewhat out of scope, as this is a vast field on its own with which I myself
+am not that well acquainted, and the paper is already at the word limit for
+Evolution research articles.
 
@@
 
-To some extent, the consequences of different scalings is taken up in the
-results and discussion throughout the paper where relevant. In the methods
-section the relationship between allelic effects and additive genetic variance
-is stated explicitly. In the results shown in Fig. 2, the effect of scaling
-assumptions on establishment probability is explicitly shown and described. In
-the section on establishment with recurrent migration, the phenomenon that
+The consequences of different scalings is taken up in the results and
+discussion throughout the paper where I believe this is relevant.
+In the methods section the relationship between allelic effects and additive
+genetic variance is stated explicitly.
+In the results shown in Fig. 2, the effect of scaling assumptions on
+establishment probability is explicitly shown and described.
+In the section on establishment with recurrent migration, the phenomenon that
 polyploid offspring is more extreme on average when allelic effects are not
 exactly scaled by one half (as in Griswold) is taken up. 
-**TODO** (Discussion) **I now added some further discussion, referring
-explicitly to Griswold and Gallais in the Discussion section. I now also refer
-to Porturas et al.**
+I now added some further discussion, referring explicitly to Griswold's scaling
+in the Discussion section. I now also refer to @porturas2019. Sadly, I
+currently do not have access to Gallais' book.
 
@@
 selection and then migration. It would also help support a self-contained paper
-to give the expression for E[eγ(...) ].
+to give the expression for $E[e\gamma(...) ]$.
 
@@
 
->p. 13-15, suppl.: Some notation seemed to not be defined, e.g. VS(2,2), V0,3, ...
+>p. 13-15, suppl.: Some notation seemed to not be defined, e.g. $V_{S(2,2)}$, ...
 
@@
 approach more clear.
+
+## References

### 2025-03-26 doc/paper/paper.tex
revising

@@
 block', see \cite{ramsey1998,kohler2010,brown2024}).
-This positive frequency dependence effect is commonly referred to as
-\textit{minority cytotype exclusion} (MCE), after \cite{levin1975}.
+This positive frequency dependence effect in mixed-ploidy populations is
+commonly referred to as \textit{minority cytotype exclusion} (MCE), after
+\cite{levin1975}.
 It is well-appreciated that, as a consequence of MCE, the rate of unreduced
 gamete formation needs to be extraordinarily high for tetraploids to establish
-in a large random mating initially diploid population (\cite{felber1997}, see
-also \cref{s-sec:det}).
+a stable subpopulation in a large random mating population initially dominated
+by diploids (\cite{felber1997}, see also \cref{s-sec:det}).
 
@@
 chromosomal rearrangements, where local establishment through genetic drift and
-subsequent spreading in a subdivided population through local extinction and 
-recolonization dynamics has been suggested as a plausible model
-\citep{lande1985}.
-However, MCE is quite strong in randomly mating mixed-ploidy populations, and
-the population size has to be very small for local tetraploid establishment to
-occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
+subsequent spreading in a subdivided population by means of local extinction
+and recolonization has been suggested as a plausible model \citep{lande1985}.
+However, MCE is quite strong under random mating, and the population size has
+to be very small for local tetraploid establishment to occur at an appreciable
+rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
 Secondly, any form of \textit{prezygotic isolation} between cytotypes could
@@
 better adapted to (changing) environmental conditions \citep{vandepeer2021}. 
-
-None of these factors is likely to explain by itself the establishment of
-polyploids, and the consensus in the field appears to be that some mix of the
-above is required to explain the occurrence of mixed-ploidy populations in
+However, none of these factors is likely to explain by itself the establishment
+of polyploids, and the consensus in the field appears to be that some mix of
+the above is required to explain the occurrence of mixed-ploidy populations in
 nature \citep{kolar2017,mortier2024}.
-Many empirical studies show a pattern where polyploids established in
-peripheral habitats at the edge of a species' range (reviewed in
-\cite{griswold2021}), in accordance with large-scale biogeographical patterns
-\citep{rice2019}.
+
+Many empirical studies of mixed-ploidy populations find that polyploids
+established in peripheral habitats at the edge of a species' range (reviewed in
+\cite{griswold2021}), and this is in accord with large scale biogeographical
+patterns \citep{rice2019}.
+This raises the question: which aspects of the process of adaptation to marginal
+habitats could promote the establishment of polyploid populations?
 In a peripheral habitat, a new polyploid population may be more likely to reach
-an appreciable size as it may evade to some extent the negative effects of MCE.
+an appreciable size as it evades to some extent the negative effects of MCE.
 However, at the same time, such peripheral habitats are likely to present
@@
 are to establish before diploids do.
-
 While there have been substantial modeling efforts aimed at understanding
@@
 $\theta$ & trait value beyond which the growth rate becomes positive in the new
-habitat \\ $w_{ij}$ & fitness of parental pair $(i,j)$ \\
-$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
+    habitat \\ 
+$w_{ij}$ & fitness of parental pair $(i,j)$ (expected fitness of
+    offspring of pair $(i,j)$)\\
+$w_{ij}^{kl}$ & expected fitness of offspring of parental pair $(i,j)$ when
 $i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
@@
 has spurred considerable interest among botanists and evolutionary biologists
-\citep{rice2015,kolar2017,vandepeer2021,griswold2021,mortier2024}.
+\citep{kolar2017,rice2019,vandepeer2021,griswold2021,mortier2024}.
 An important question is whether such patterns emerge because polyploids are
@@
 issues with endosperm development \citep{bretagnolle1995}.
+
 Similarly implausible is the assumption of a constant equilibrium genetic
 variance across cytotypes ($\beta_4=\sqrt{1/2}$ in our model), which we used in
-most of our results.
-It would be very interesting to experimentally assess whether typical
-quantitative traits in mixed-ploidy populations follow approximately the
-mixed-ploidy infinitesimal model we outlined in this paper, and to actually
+most of our results (but see \cref{fig:est1}).
+The meta-analysis performed by \cite{porturas2019} suggests that phenotypic
+variation across ploidy levels is indeed fairly constant, so this assumption
+may not yield too bad an approximation.
+It should be noted however that other authors have made different assumptions
+on how allelic effects (and hence genetic variance) scale across ploidy levels
+(in particular \cite{griswold2021}, who made an assumption equivalent to
+$\beta_4 = 1/2$ in our model), and such assumptions evidently impact the
+establishment probability (\cref{fig:est1}).
+Clearly, it would be very interesting to experimentally assess whether
+typical quantitative traits in mixed-ploidy populations follow approximately
+the mixed-ploidy infinitesimal model we outlined in this paper, and to actually
 estimate how the variance scales across ploidy levels.

### 2025-03-27 doc/paper/response.md
good progress

@@
 
-**TODO**
+Whereas I don't think listing potential biological scenario's that can be
+investigated using my modeling approach is very useful, I think the reviewer is
+right in that it may be useful to highlight that the modeling approach can be
+employed to address other questions related to the ecology and evolution of
+mixed-ploidy populations. I have now added a small paragraph at the end that
+does so.
 
@@
 
+>
 - Demonstrating how genetic effects are scaled across triploids and tetraploids
@@
 
-Other informative analyses include recursive or Markov models of null expecta-
-tions for frequencies of ploidies, times to diploid ancestry and effective popula-
-tion size across a population of mixed-ploidy.
+>Other informative analyses include recursive or Markov models of null
+expectations for frequencies of ploidies, times to diploid ancestry and
+effective population size across a population of mixed-ploidy.
 Besides this set of results, the paper then examines conditions under which a
@@
 
+I have added a note in the discussion regarding mutation and why I have ignored it.
+
 >In the intro providing empirical examples of polyploid establishment in
@@
 write:
-"Many empirical studies of mixed-ploidy populations show a pattern where
-polyploids established in peripheral habitats at the edge of a species’ range
-(reviewed in Griswold (2021)), and these observations for specific study
-systems are in accord with large scale biogeographical patterns (Rice et al.,
-2019)."
+"Many empirical studies of mixed-ploidy populations find that polyploids
+established in peripheral habitats at the edge of a species’ range (reviewed in
+Griswold (2021)), and this is in accord with large scale biogeographical
+patterns (Rice et al., 2019)".
 
-With regard to adaptation to peripheral habitats: I refer to an important (if
-somewhat old) review article of Kawecki (2008) and recent theoretical work by
-Sachdeva et al. I think this gives a good sample of where to look for more
-references.
-When it comes to empirical work on the *genetics of adaptation to peripheral
-environments in polyploids*, I am quite frankly not aware of relevant studies...
-Reviewing empirical work on adaptation to peripheral habitats in general is
-somewhat out of scope, as this is a vast field on its own with which I myself
-am not that well acquainted, and the paper is already at the word limit for
-Evolution research articles.
+While I think it is out of scope to review in any depth the empirical
+literature on adaptation and polyploidy, I agree that the paper lacked a clear
+connection to such work, and in particular I noticed that it lacked a clear
+motivation for the interest in *polygenic* adaptation.
+I have now tried to better motivate the likely polygenic nature of local
+adaptation and have added a couple of references to recent work on the genetics
+of adaptation in autopolyploids: 
+"More often than not, local adaptation is polygenic in nature (Pritchard and Di
+Rienzo, 2010; Barghi et al., 2020; Bomblies and Peichel, 2022), involving many
+weakly selected variants across the genome, and adaptation during polyploid
+establishment is unlikely to be an exception. Recent studies on local
+adaptation in autopolyploids indicate a polygenic basis of adaptation
+(Bohutinska et al., 2021; Konecna et al., 2021, 2022), however it is not clear
+how this adaptation relates to their initial establishment."
 
@@
 
-**TODO** (Discussion)
+I agree that @oswald2011 could be a relevant paper to relate my work to,
+however it should be noted that this paper does not deal with migration, nor
+establishment in a new environment (they do study establishment after
+environmental change, but as far as I can tell they assumed a constant
+population size, so this is quite different from establishment in a new
+habitat).
+While the study is clearly related, it is not straightforward to explicitly
+relate the results beyond noting that they also found that assortative mating
+and selfing may promote establishment...
+
+However, I have now added a more explicit reference to @oswald2011 in the concluding
+note, to highlight that it would be possible to do a study that is comparable
+to theirs but assuming infinitesimal quantitative genetics instead of a model
+with a small number of loci.
 

### 2025-03-27 doc/paper/submission1.tex
good progress

--- 2025-03-13 doc/paper/paper.tex
@@
 \usepackage{authblk}
-\usepackage{lineno}
-\linenumbers
 \externaldocument[s-]{supp}
@@
 \begin{abstract}
-We define the infinitesimal model of quantitative genetics for the inheritance
-    of an additive quantitative trait in a mixed-ploidy population consisting
-    of diploid, triploid and autotetraploid individuals producing haploid and
-    diploid gametes.
-    We implement efficient simulation methods and use these to study the
+We define the infinitesimal model of quantitative genetics (\textit{sensu}
+    \cite{barton2017}) for the inheritance of an additive quantitative trait in
+    a mixed-ploidy population consisting of diploid, triploid and
+    autotetraploid individuals producing haploid and diploid gametes.  We
+    implement efficient simulation methods and use these to study the
     quantitative genetics of mixed-ploidy populations and the establishment of
-    autotetraploids in a new habitat.
-    We show that for rare migration from the source population, autotetraploids
-    are more likely to found a successful population than diploids under a very
-    broad range of conditions, but that this is unlikely to sufficiently
-    counter the scarcity of tetraploid founders when the source is
-    predominantly diploid.
-    We assess in more detail how minority cytotype exclusion interacts with
-    migration load in the establishment process and evaluate the impact of
-    additional sources of prezygotic isolation, specifically selfing and
-    assortative mating, on the relative establishment probabilities of the
-    different cytotypes.
+    autotetraploids in a new habitat.  We show that for rare migration from the
+    source population, autotetraploids are more likely to found a successful
+    population than diploids under a very broad range of conditions, but that
+    this is unlikely to sufficiently counter the scarcity of tetraploid
+    founders when the source is predominantly diploid.  We assess in more
+    detail how minority cytotype exclusion interacts with migration load in the
+    establishment process and evaluate the impact of additional sources of
+    prezygotic isolation, specifically selfing and assortative mating, on the
+    relative establishment probabilities of the different cytotypes.
     In the discussion we consider how inbreeding depression may impact our
@@
 \section*{Introduction}
+
+% Should connect more with literature and concepts of isolating mechanisms pre
+% vs. postzygotic isolation, and reinforcement etc. (see e.g. Parvathy &
+% Himani's paper introduction)
 
@@
 phenological differences across cytotypes, or differences in pollinators;
-\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, 
-localized dispersal \citep{baack2005,kolar2017} and asexual reproduction
-\citep{vandrunen2022}.
+\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, and
+localized dispersal \citep{baack2005,kolar2017}.
 Finally, selection may be invoked to explain the establishment of polyploids.
 Tetraploids may have higher relative fitness than their diploid counterparts
-due to reduced inbreeding depression
-\citep{ronfort1999,otto2000,husband2008,clo2022b}, or due to being
+due to reduced inbreeding depression \citep{ronfort1999,otto2000}, or due to being
 better adapted to (changing) environmental conditions \citep{vandepeer2021}.
@@
 the proportion of euploid gametes produced by a triploid individual.
-The rate of unreduced gamete production is hence fixed across individuals, and
-is assumed to be the same in diploids and tetraploids.
 
@@
 This is not likely to reflect biological reality: tetraploids do not tend to
-have, for instance, twice the size of their diploid progenitors on average
-(e.g. \cite{porturas2019}).
+have, for instance, twice the size of their diploid progenitors on average.
 Furthermore, the genetic variance at Hardy-Weinberg and linkage equilibrium
@@
 The expectation on the right hand side can be calculated from \cref{eq:oneline}
-using the moment-generating function of the Gaussian, i.e.
-\begin{equation}
-    \Ex\left[e^{\gamma(Z_{ij}^{kl} - \theta)}\right] 
-    = e^{\gamma \left(\Ex \left[Z_{ij}^{kl}\right] -\theta\right) +
-    \frac{\gamma^2}{2}
-    \Var\left[Z_{ij}^{kl}\right]} 
-\end{equation}
+using the moment-generating function of the Gaussian.
 Having calculated the $w_{ij}$, the number of offspring surviving into the next
@@
 implemented in Julia \citep{julia}.
-Documented code and simulation notebooks are available at \url{[anonymized]}.
-%\url{https://github.com/arzwa/InfGenetics}.
+Documented code and simulation notebooks are available at
+\url{[anonymized]}.
 
@@
 critical branching process with a Poisson offspring distribution, so that the
-probability to reach $N=100$ is $\sim 1/100$ \citep{barton2018}.
+probability to reach $N=100$ is $1/100$ \citep{barton2018}.
 
@@
 to meiotic irregularities associated with multivalent formation
-\citep{bomblies2016,novikova2023}, and triploids may be inviable due to
-issues with endosperm development \citep{bretagnolle1995}.
-Similarly implausible is the assumption of a constant equilibrium genetic
-variance across cytotypes ($\beta_4=\sqrt{1/2}$ in our model), which we used in
-most of our results.
+\citep{bomblies2016,novikova2023}.
+Furthermore, in our simulations, we have focused on the case where the
+equilibrium genetic variance is constant across cytotypes ($\beta_4=\sqrt{1/2}$
+in our model).
 It would be very interesting to experimentally assess whether typical
@@
 higher than expected based on the frequency of tetraploid migrants (i.e. is
-roughly of order $u$ instead of $u^2$) when migration is sufficiently weak and
+roughly $O(u)$ instead of $O(u^2)$) when migration is sufficiently weak and
 maladaptation sufficiently high.
@@
 However, autopolyploidy has important consequences whenever dominance is
-relevant, as in the case of inbreeding depression
-\citep{ronfort1999,husband2008,clo2022b}.
+relevant, as in the case of inbreeding depression \citep{ronfort1999}.
 Indeed, when inbreeding depression is due to recessive deleterious variation,
 it is expected to be less expressed in neotetraploids because homozygous genotypes
-should be much rarer than in their diploid parents (the `masking' effect;
-\cite{otto2000}).
-In the presence of deleterious recessive variation, inbreeding during the
+should be much rarer than in their diploid parents.
+When there is indeed deleterious recessive variation, inbreeding during the
 establishment process should therefore incur a higher fitness cost in diploids
@@
 tetraploid establishment.
-How this plays out depends however on the \textit{rate} at which populations
-become inbred, which will differ between cytotypes and will depend strongly on
-the mating system.
+How this plays out depends on the \textit{rate} of inbreeding, which will
+differ between cytotypes. 
 In outcrossing populations, inbreeding occurs at a slower rate in tetraploids,
@@
 and they will not have accumulated more deleterious mutations than their
-diploid counterparts, so that polyploidy should lead to reduced rather than
-increased inbreeding depression when selfing rates are similar (as discussed
-above).
+diploid counterparts, so that polyploidy leads to reduced rather than increased
+inbreeding depression when selfing rates are similar (as discussed above).
 Interestingly, the interplay between the effects of polyploidy on different
-timescales could yield an equilibrium situation that may characterize many
+timescales could yield an equilibrium situation that may characterize
 mixed-ploidy populations in nature: although sometimes polyploids could enjoy

### 2025-03-27 doc/paper/buildanddiff.sh

@@ -1,0 +1,3 @@
+buildtex.sh paper.tex
+latexdiff submission1.tex paper.tex > diff.tex
+buildtex.sh diff.tex

### 2025-04-03 doc/paper/response.md
update

@@
 suggestions made by the reviewers and, if you disagree with them, that you
-justify and explain your reasons in a response letter.
+justify and explain your reasons in a response letter. (emphasis mine)
 
-I have now revised the Introduction slightly to include more discussion of
-empirical work, although, in the interest of brevity (the manuscript is already
-near the Evolution word limit), I mostly resorted to explicitly referring the
-reader to the article by @griswold2021. In the latter, the author provides a
-nice overview of relevant recent studies.
+Thanks a lot for your positive appraisal.
+
+First of all, I wish to note that there was an important error in my original
+submission, which I outline below. Evidently, I have redone all simulations and
+revised the manuscript accordingly. The error affects both diploid and
+tetraploid establishment similarly, so does not affect the results
+qualitatively (since these are concerned, mostly, with relative establishment
+probabilities), but does affect all results with nonzero selection intensity
+($\gamma > 0$). I outline the details of the error in the next section.
+
+I have now revised the Introduction to refer more to empirical work.
+I have included the references suggested by Reviewer 1.
+Following the suggestions of Reviewer 2, I have tried to refer better to
+empirical work on (1) polyploid establishment at range margins and (2) the
+genetics of adaptation to peripheral environments (in polyploids).
+In the interest of brevity, I do not present an in-depth review of these
+topics. For (1), I refer mostly to @griswold2021, who presented a
+nice overview. For (2), I mainly tried to better motivate my interest in
+*polygenic* adaptation, referring to recent work and reviews that suggest that
+local adaptation is typically polygenic, and work that shows that
+autopolyploids present no exception to this.
 
 Regarding the Results and the Discussion sections: I do not want to do away
-with Discussion section altogether (as suggested by Reviewer 1), given that I
+with Discussion section altogether (as suggested by Reviewer 1), as I
 wish to discuss factors relevant to the problem which are not studied in my
-article (dominance, inbreeding depression, *etc.*) and want to relate my work
-to other recent theoretical work.
-I believe it is better to separate the latter topics from the description *and*
-interpretation of the actual results. 
+article (dominance, inbreeding depression, *etc.*) and hence do not feature in
+the results section. Furthermore, I want to relate my work to other recent
+theoretical work (in particular that of @griswold2021). I believe this warrants
+a dedicated Didcussion section.
 I have sought to move 'more discussive text' to the Discussion section,
 although I found there was fairly limited scope to do so when I seek to keep
-the interpretation of the results more or less self-contained. **TODO!**
+the interpretation of the results more or less self-contained.
 
-I have expanded the discussion slightly to include more of the previous
-theoretical literature on the subject, in particular following the suggestions
-of Reviewer 2. **TODO!**
+I have expanded the discussion slightly to address some of the remarks of
+Reviewer 2 concerning the issues of mutation and the scaling of genetic variance. 
+I have added a concluding paragraph following the suggestion of Reviewer 1 to
+highlight that other eco-evolutionary questions about mixed-ploidy populations
+might be fruitfully addresed using the infinitesimal framework. I took the
+opportunity to allude to the work of @oswald2011 in that section, following up
+on a suggestion of Reviewer 2.
+
+Overall, I think these revisions have contributed considerably to a
+better-motivated article that engages more with relevant empirical work. I
+hence thank the reviewers and the editors for their suggestions. Below I
+include a detailed reply to both reviewers.
+
+## Error in the first submission
+
+In the simulations included in the first submission of my article, I did not
+sample the trait values of the next generation correctly when there was nonzero
+selection $\gamma > 0$ (as in figs. 2-5).
+As outlined in my paper, I sample the number of offspring of a particular
+combination of gametes from a particular parental pair after selection in
+proportion with the expected fitness using eq. 11 (number in new version).
+In the previous version, I then sampled, incorrectly, offspring trait values
+from the Gaussian distribution of offspring implied by the mixed-ploidy
+infinitesimal model (eq. 7). This however fails to account for selection on the
+trait value within each parental pair.
+
+The density of trait values after selection is found easily.
+Writing $f(z)$ for the Gaussian density of trait values with mean $\bar{z}$ and
+variance $V$, and $w(z)$ for the fitness function, the density after selection
+$f^\ast(z)$ is
+\begin{align*}
+f^\ast(z) &= \frac{w(z)f(z)}{\int w(z)f(z)dz} \\
+&\propto \exp\left(\gamma(z - \theta) - \frac{(z - \bar{z})^2}{2V}\right) \\
+&\propto \exp\left(-\frac{1}{2V}(z^2 - 2z\bar{z} - 2V\gamma z)\right) \\
+&= \exp\left(-\frac{1}{2V}\left[z^2 - 2z(\bar{z} + \gamma V) + (\bar{z} + \gamma V)^2 - (\bar{z} + \gamma V)^2\right]\right) \\
+&\propto \exp\left(-\frac{1}{2V}\left[z - (\bar{z} + \gamma V)\right]^2\right)
+\end{align*}
+i.e. $f^\ast(z)$ is a Gaussian density with mean $\bar{z} + \gamma V$ and
+variance $V$. 
+The correct trait value distribution for the offspring in parental pair $(i,j)$
+with gametes of ploidy $k$ and $l$ *after selection* is hence a Gaussian with
+mean $\overline{z_{ij}^{kl}} + \gamma V_{ij}^{kl}$ and variance $V_{ij}^{kl}$.
+
+The incorrect results from the previous submission can be seen as a kind of
+zeroth order approximation to the correct results: i.e. they implement
+selection on the mean trait value across families, but not selection on the
+variance. The main consequence of this error is that adaptation is more
+efficient than in my previous results, and hence establishment is easier. This
+is however the case for both diploids and tetraploids. 
+The effect on establishment is however nonlinear, so although the main qualitative
+results of the paper remain the same, the new results do not simply amount to a
+rescaling of the old ones.
+
+I apologize for this oversight.
 
@@
 
-I appreciate the comment and have reworked the text with this in mind.
+I appreciate the comment and have tried to rework the text with this in mind.
 However, following the suggestion of the associate editor, I went with the
@@
 
->Defining establishment when N=100: Is it a classic way to define establishment
-in such models?  Or does it come from your experience with the model and you
-see that Under N=100 extinction remains likely? More details would be needed
-here.
+>Defining establishment when $N=100$: Is it a classic way to define
+establishment in such models?  Or does it come from your experience with the
+model and you see that Under N=100 extinction remains likely? More details
+would be needed here.
 
-$N=100$ is indeed an arbitrary choice. This threshold was also adopted in
+$N=100$ is indeed a fairly arbitrary choice. This threshold was also adopted in
 @barton2018, so it makes comparisons of establishment probabilities *etc.* with
@@
 approach requires tracking the $N \times N$ matrix of identity coefficients). 
+Also, for the migration rates considered, $N=100$ is sufficiently large that
+migration alone (without population growth due to adaptation) can not lead to
+$N=100$, i.e. the population does have to increase the mean trait value in
+order to reach establishment.
 I did a check by putting the threshold at $N=150$, and the establishment
@@
 and whether mutation may affect results. Is there a relationship between $\mu V_m$
-(Barton et al. 2017) and m and V on establishment?
+(Barton et al. 2017) and $m$ and $V$ on establishment?
 Recognizing the point seems to be important, but its analysis can be for later.
@@
 establishment probability between diploids and tetraploids should be
-negligible.
-This is definitely the case when there is ongoing migration: as long as $m$ is
-sufficiently greater than $\mu$ and $V_m$ is sufficiently smaller than the
-genetic variance in the mainland population, we will have that for every
-mutation contributing a very slight decrease in the loss of segregation
+negligible. In more quantitative terms: we assume $1/\mu \gg \tilde{T}$ where
+$\tilde{T}$ is the time between establishment and the arrival of the migrant
+from which the established population derived. 
+In the presence of migration the variance contributed by mutation is even more
+certainly negligible.
+As long as $m V$ is sufficiently larger than $\mu V_m$, we will have that, for
+every mutation contributing a very slight decrease in the loss of segregation
 variance on the island, there will be many more migrant arrivals, which each
 introduce a completely unrelated genotype.
+In all our simulations with migration we assume $mV$ considerably larger than
+conceivable values of $\mu V_m$.
 
@@
 establishment is unlikely to be an exception. Recent studies on local
-adaptation in autopolyploids indicate a polygenic basis of adaptation
-(Bohutinska et al., 2021; Konecna et al., 2021, 2022), however it is not clear
-how this adaptation relates to their initial establishment."
+adaptation in autopolyploids indeed tend to find a polygenic basis of
+adaptation (Bohutı́nská et al., 2021; Konečná et al., 2021, 2022), however it is
+not clear how observed adaptive differentiation in established tetraploid
+populations relates to adaptation that occurred during initial establishment".
 
->The discussion could relate findings to other theoretical works more. For ex-
-ample, it does not go back and place the results in the context of Oswald and
+>The discussion could relate findings to other theoretical works more. For
+example, it does not go back and place the results in the context of Oswald and
 Nuismer (2011), which assumed a two-locus model and varied dosage effects, as
@@
 I agree that @oswald2011 could be a relevant paper to relate my work to,
-however it should be noted that this paper does not deal with migration, nor
-establishment in a new environment (they do study establishment after
-environmental change, but as far as I can tell they assumed a constant
-population size, so this is quite different from establishment in a new
-habitat).
-While the study is clearly related, it is not straightforward to explicitly
-relate the results beyond noting that they also found that assortative mating
-and selfing may promote establishment...
+however it should be noted that this paper focuses on establishment in
+sympatry, and does not deal with migration, nor establishment in a new
+environment (they do study establishment after environmental change, but as far
+as I can tell they assumed a constant population size, so this is quite
+different from establishment in a new habitat).
+While my study is clearly related to theirs in that it is concerned with
+establishment of autotetraploids in mixed-ploidy populations, it is not
+straightforward to explicitly relate the results beyond noting that (1) they
+investigated more processes (assortative mating by trait value, competition,
+stabilizing selection and inbreeding depression) and (2) also found that
+assortative mating and selfing may promote establishment.  The latter was of
+course already found by many authors before @oswald2011...
 
 However, I have now added a more explicit reference to @oswald2011 in the concluding
-note, to highlight that it would be possible to do a study that is comparable
-to theirs but assuming infinitesimal quantitative genetics instead of a model
-with a small number of loci.
+note (see the associated remark of Reviewer 1), to highlight that it would be
+possible to do a study that is comparable to theirs but assuming infinitesimal
+quantitative genetics instead of a model with a small number of loci.
 
->It would be informative to expand on the the consequences of different scalings
-of genetic effects. This paper made the sensible choice to scale such that segre-
-gation variance is equal between ploidies as a baseline. Griswold (2021) scaled
-genetic effects such that autotetraploid and diploid individuals have equal fit-
-ness as a function of proportional allele count in an individual. A consequence
-of Griswold’s scaling is autotetraploids have lower additive genetic variance at
-HWE (Gallais 2003, p. 185), so are expected to respond more slowly in the
-absence of another process. Gallais (2003, p. 186) mentions some studies that
-compared additive genetic variance between diploids and autotetraploids.
+>It would be informative to expand on the the consequences of different
+scalings of genetic effects. This paper made the sensible choice to scale such
+that segregation variance is equal between ploidies as a baseline. Griswold
+(2021) scaled genetic effects such that autotetraploid and diploid individuals
+have equal fitness as a function of proportional allele count in an individual.
+A consequence of Griswold’s scaling is autotetraploids have lower additive
+genetic variance at HWE (Gallais 2003, p. 185), so are expected to respond more
+slowly in the absence of another process. Gallais (2003, p. 186) mentions some
+studies that compared additive genetic variance between diploids and
+autotetraploids.
 
@@
 I now added some further discussion, referring explicitly to Griswold's scaling
-in the Discussion section. I now also refer to @porturas2019. Sadly, I
-currently do not have access to Gallais' book.
+in the Discussion section. I now also refer to @porturas2019 and @gallais2003.
 
@@
 maintain a locally beneficial allele when migrants introduce the deleterious
-allele, and that when adaptation is due to alleles of very small effect this
-may seem to imply that there is little scope for local adaptation.
+allele. So, when adaptation is due to alleles of very small effect, this seems
+to imply that there is little scope for local adaptation, since locally
+beneficail variants will be swamped by gene flow.
 
-There are multiple reasons why this is not the case. The argument outlined in
-@barton2018 is simply that the classical theory assumes complete divergence
-between mainland and island, but that when adaptation is from standing
-variation instead this doe snot hold, since allele frequency differences
-between mainland and island are only slight. In the latter case, many alleles
-of small effect can be divergently maintained by selection. Another reason why
-this does not hold is linkage disequilibrium between locally beneficial
-alleles, this is explored in detail in @sachdeva2022 and @zwaenepoel2024.
+There are multiple reasons why this is not the case in nature. The argument
+outlined in @barton2018 is simply that the classical theory usually assumes
+complete divergence between mainland and island, but that when adaptation is
+from standing variation instead, this does not hold, since allele frequency
+differences between mainland and island are only slight. In the latter case,
+many alleles of small effect can be divergently maintained by selection.
+Another reason why this does not hold is that selected alleles will 
+be in linkage disequilibrium, so that entire sets of deleterious alleles
+introduced by migrants are eliminated jointly. The latter phenomenon is
+explored in detail in @sachdeva2022 and @zwaenepoel2024.
 
@@
 explain and do not have much to do with the problem studied, i.e. polyploid
-establishment.
+establishment, so in the interest of brevity I decided not to revisit this
+point.
 
@@
 1960's.
-I think the reference to Turelli (2017) which reviews the history of the
-infinitesimal model suffices, and prevents confusion as to what we mean when we
-refer to 'the infinitesimal model'.
+I think the reference to Turelli (2017), which reviews the history of the
+infinitesimal model, is quite adequate, and prevents confusion as to what we
+mean when we refer to 'the infinitesimal model'.
 
@@
 selection and then migration. It would also help support a self-contained paper
-to give the expression for $E[e\gamma(...) ]$.
+to give the expression for $E\left[e^{\gamma(\dots)}\right]$.
 
@@
 I do not think, however, that fig 4 suggests such a transition point though. As
-noted in the main text (**line numbers**), the difference between the $m=0$ and
-$m=0.01$ results in fig. 4 are mostly due to the fact that the latter assumes a
-Gaussian distribution of migrant trait values, whereas the former assumes a
-fixed initial migrant trait value.
+noted in the main text the difference between the $m=0$ and $m=0.01$ results in
+fig. 4 are mostly due to the fact that the latter assumes a Gaussian
+distribution of migrant trait values, whereas the former assumes a fixed
+initial migrant trait value.
+(*"The similar trait means for weak migration suggest
+that establishment depends mostly on the chance pick of a well-adapted migrant
+(note that the $m = 0$ simulations in fig. 4 assume the initial migrant has $z =
+0$, whereas the $m = 0.01$ simulations assume the initial trait value to be
+Gaussian)."*)
 
->p. 12: Add qualifier “as m increases” at end of last sentence in last full para-
-graph.
+>p. 12: Add qualifier “as $m$ increases” at end of last sentence in last full
+paragraph.
 
@@
 
->p. 12, suppl.: add “between centromere and locus” after “. . . recombination
-happens with probability c”
+>p. 12, suppl.: add “between centromere and locus” after “... recombination
+happens with probability $c$”
 

### 2025-04-03 doc/paper/paper.tex
update

@@
         assortative mating, on the relative establishment probabilities of the
-    different cytotypes.
+        different cytotypes in the presence of maladaptive migration.
         In the discussion we consider how inbreeding depression may impact our
@@
     phenological differences across cytotypes, or differences in pollinators;
-\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, 
+    \cite{husband2004,kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, 
     localized dispersal \citep{baack2005,kolar2017} and asexual reproduction
@@
     due to reduced inbreeding depression
-\citep{ronfort1999,otto2000,husband2008,clo2022b}, or due to being
+    \citep{husband1997,ronfort1999,otto2000,husband2004,husband2008,clo2022b}, or due to being
     better adapted to (changing) environmental conditions \citep{vandepeer2021}. 
@@
     In a peripheral habitat, a new polyploid population may be more likely to reach
-an appreciable size as it evades to some extent the negative effects of MCE.
+    an appreciable size as it evades to some extent the negative effects of MCE
+    \citep{levin1975}.
     However, at the same time, such peripheral habitats are likely to present
@@
     inbreeding and maladaptive migration, compared to diploids.
-Indeed, when the source population is dominated by diploids, the probability
-that a migrant individual is tetraploid will be very small ($\sim u^2$, where
-$u$ is the probability that a diploid produces an unreduced gamete in meiosis),
-so that tetraploids need a substantial advantage during colonization if they
-are to establish before diploids do.
+
+    More often than not, local adaptation is polygenic in nature
+    \citep{pritchard2010,barghi2020,bomblies2022}, involving many weakly selected
+    variants across the genome, and adaptation during polyploid establishment
+    in a marginal habitat is unlikely to present an exception.
+    Recent studies on local adaptation in autopolyploids indeed tend to find a
+    polygenic basis of adaptation \citep{bohutinska2021,konevcna2021,konevcna2022},
+    however it is not clear how observed adaptive differentiation in established
+    tetraploid populations relates to adaptation that may have occurred during
+    initial establishment.
+
     While there have been substantial modeling efforts aimed at understanding
@@
     i.e., $Z_{ij}^{kl}$ is a Gaussian random variate with distribution
-\begin{equation}
-  Z_{ij}^{kl} \sim \Normal\left(
-    \beta_{k+l} \left(
+        $Z_{ij}^{kl} \sim \Normal\left(\overline{z_{ij}^{kl}},
+        V_{ij}^{kl}\right)$
+    where
+    \begin{align}
+        \overline{z_{ij}^{kl}} &= \beta_{k+l} \left(
               \frac{k}{c_i}\frac{z_i}{\beta_{c_i}} 
-        + \frac{l}{c_j}\frac{z_j}{\beta_{c_j}}\right), 
-    \beta_{k+l}^2 (V_{i,k} + V_{j,l})\right)
+            + \frac{l}{c_j}\frac{z_j}{\beta_{c_j}}\right) \nonumber \\
+        V_{ij}^{kl} &= \beta_{k+l}^2 (V_{i,k} + V_{j,l})
        \label{eq:oneline}
-\end{equation}
+    \end{align}
 
@@
     \textit{twice} the excess of their parent on average.
-In other words, the mixed-ploidy model is not `coordinate free' as the basic
+    In other words, the mixed-ploidy model is not `coordinate-free' as the basic
     infinitesimal model is: \cref{eq:oneline} only makes sense when the trait
@@
         \Ex\left[e^{\gamma(Z_{ij}^{kl} - \theta)}\right] 
-    = e^{\gamma \left(\Ex \left[Z_{ij}^{kl}\right] -\theta\right) +
+        = e^{\gamma \left(\overline{z_{ij}^{kl}} -\theta\right) +
         \frac{\gamma^2}{2}
-    \Var\left[Z_{ij}^{kl}\right]} 
+        V_{ij}^{kl}} 
+        \label{eq:Ew}
     \end{equation}
@@
     generation is calculated as $N^\ast(t+1) = \sum_{i,j}w_{ij}/N(t)$.
-Next, $N^\ast(t+1)$ offspring individuals are sampled by sampling parental
+    Next, $N^\ast(t+1)$ offspring individuals are obtained by sampling parental
     pairs and gametes proportional to $w_{ij}^{kl}$, and sampling a trait value
-in accordance with \cref{eq:oneline}.
+    from a Gaussian distribution with mean $\overline{z_{ij}^{kl}} + \gamma
+    V_{ij}^{kl}$ and variance $V_{ij}^{kl}$.
 
@@
 \centering
-\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/tetfininf.pdf}
+\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/fig1.pdf}
 \caption{
@@
 inbreeding in exact simulations of the infinitesimal model in autotetraploids
-against individual-based simulations of autotetraploid populations with $L$
+against individual-based simulations of autotetraploid populations with $L$.
 unlinked additive loci determining the quantitative trait. 
@@
 increase in genetic variance (as it increases the segregation variance), but
-leads to accelerated inbreeding, causing faster loss of genetic variance in the
+leads to accelerated inbreeding, causing faster loss of variation in the
 long term (\cref{fig:vztet,s-fig:fininf}).
@@
 diploid individuals (to first order in $u$), highlighting that when diploids
-dominate, polyploids have low fitness and do not contribute to the effective
-population size.
+dominate, polyploids do not contribute to the effective population size.
 
@@
 \centering
-\includegraphics[width=\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/est1b.pdf}
+\includegraphics[width=\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/fig2.pdf}
 \caption{
-(A) Probability of establishment from a single diploid or tetraploid individual
-with trait value $z=0$ for increasing selection intensity $\gamma$.
-We assume $m=0$ and $u=0$, i.e. there is no migration, and no unreduced
-gametes are produced. The trait is scaled in tetraploids so as to yield the
-same genetic variance at HWLE ($\beta_4 = \sqrt{1/2}$)
+    (A) Probability of establishment (defined as reaching $N=100$) from a
+    single diploid or tetraploid individual with trait value $z=0$ for
+    increasing selection intensity $\gamma$, for two different values of
+    $\theta$ (degree of maladaptation).  We assume $m=0$ and $u=0$, i.e. there
+    is no migration, and no unreduced gametes are produced. The trait is scaled
+    in tetraploids so as to yield the same genetic variance at HWLE ($\beta_4^2 =
+    1/2$).
+    Note that when $\gamma = 0$, we obtain a critical branching process with a
+    Poisson offspring distribution, so that the probability to reach $N=100$ is
+    $\sim 1/100$ \citep{barton2018}.  
 (B) Probability of a tetraploid individual with trait value $z = 0$
-successfully founding a population ($P_4$), relative to the probability for a
-diploid individual with the same trait value ($P_2$).
-The vertical dashed line marks $\beta_4=\sqrt{1/2}$, for which the variance at
-HWLE is identical between diploids and autotetraploids.
+    successfully founding a population ($P_4$), relative to the probability for
+    a diploid individual with the same trait value ($P_2$).  The vertical
+    dashed line marks $\beta_4^2=1/2$.
 (C) Probability of tetraploid establishment relative to the probability of
     diploid establishment (on a $\log_{10}$ scale) across a range of values for
-$\gamma$ and $\beta_4$. The vertical line again marks $\beta_4=\sqrt{1/2}$.
-All results are estimated from 500.000 replicate simulations and assume
-$\theta=2$. Establishment is defined as reaching $N=100$.
+    $\gamma$ and $\beta_4$ ($\theta=2.5$). The vertical line again marks
+    $\beta_4^2=1/2$.  All results are estimated from 1.000.000 (A\&B) or
+    500.000 (C) replicate simulations.
 \label{fig:est1}}
@@
 is diploid vs. tetraploid (\cref{fig:est1}).
-As noted by \cite{barton2018}, the establishment probability (for a given
-cytotype) depends essentially on two dimensionless parameters,
-$\gamma\sqrt{2V}$ and $\theta/\sqrt{2V}$, corresponding to the intensity of
-selection and the degree of maladaptation, respectively.
-We shall scale our results accordingly, assuming $2V=1$ throughout, and shall
-assume $\theta=2$, so that the average trait value has to increase by two
-standard deviations to achieve a positive growth rate.
-The \textit{relative} establishment probability for different cytotypes will of
-course also depend on how allelic effects scale across ploidy levels ($\beta_4$).
-
-We find that reduced inbreeding in tetraploids substantially increases the
-establishment probability of tetraploids relative to diploids across a large
-part of the parameter space (\cref{fig:est1}).
-Indeed, in the case where allelic effects are scaled across ploidy levels
-to yield the same equilibrium genetic variance ($\beta_4 = \sqrt{1/2}$, so that
-polyploid migrants are not more likely to have high fitness than diploid
-migrants), the establishment probability for tetraploids can be almost five
-times as high as in diploids depending on the selection gradient
+For a given mixed-ploidy model (characterized by parameters $\alpha, \beta,
+u$ and $v$), the establishment probability depends on $\gamma, \theta$ and $V$
+through two dimensionless parameters, $\gamma\sqrt{2V}$ and $\theta/\sqrt{2V}$
+\citep{barton2018}, corresponding to the intensity of selection and the degree
+of maladaptation, respectively.
+We shall scale our results accordingly, assuming $2V=1$ throughout.
+
+For a fixed degree of maladaptation $\theta$, the probability of establishment
+depends in a complicated way on the strength of selection.
+To see this, note that the expected number of offspring of an initial migrant
+of ploidy level $k$ is $e^{-\gamma \theta + \gamma^2 k\beta_k^2 V/4}$, and the
+expected trait value among its offspring will be $\gamma k \beta_k^2 V/2$.
+A higher intensity of selection ($\gamma$) therefore yields a stronger effect
+of initial maladaptation, but also causes a stronger response in the mean trait
+value.
+If the genetic variance is not constant across cytotypes (i.e. $\beta_4^2 \ne
+1/2$), this response will differ for different ploidy levels.
+Differences across cytotypes in how the genetic variance evolves over time as a
+consequence of inbreeding will then further cause rates of adaptation to
+differ, leading to different establishment probabilities even when
+$\beta_4^2=1/2$.
+
+Indeed, we find that reduced inbreeding in tetraploids substantially increases the
+likelihood of tetraploid establishment relative to diploids across a large
+part of the parameter space (\cref{fig:est1}A).
+For the $\beta_4^2 = 1/2$ case, the establishment probability for
+tetraploids can be more than ten times as high as for diploids depending on the
+selection gradient ($\gamma$) and the degree of maladaptation ($\theta$)
 (\cref{fig:est1}A).
@@
 migrant individual.
-
-While the probability of establishment becomes lower as the strength of
-selection becomes large, the establishment probability does not decrease
-monotonically with $\gamma$, i.e. as selection becomes very weak, the
-establishment probability also decreases.
-This happens because, although the probability of surviving the first couple of
-generations becomes higher when selection is weaker, adaptation (i.e. increase
-in the trait mean) will be slower,
-increasing the risk that genetic variation is exhausted due to inbreeding before the
-population is able to reach a consistently positive growth rate.
-Note that when $\gamma = 0$, each individual has fitness 1 and we have a
-critical branching process with a Poisson offspring distribution, so that the
-probability to reach $N=100$ is $\sim 1/100$ \citep{barton2018}.
-
 Evidently, the scaling of the genetic variance across ploidy levels has a
@@
 tetraploids compared to diploids) is the benefit of the slower rate of
-inbreeding in tetraploids canceled (\cref{fig:est1}B,C).
+inbreeding in tetraploids canceled by the reduction in genetic variance 
+(\cref{fig:est1}B,C).
 
@@
 \centering
-\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/paper/figs/estwmig.pdf}
+\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/fig3.pdf}
 \caption{
@@
     in \cref{fig:est1}, but with unreduced gamete formation included). 
+    The line for the $\theta=3$ case lies around 0.25 and is not shown.
     The dashed horizontal line marks the proportion of tetraploid migrants
@@
 \centering
-\includegraphics[width=\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration1d.pdf}
+\includegraphics[width=\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/fig4.pdf}
 \caption{
@@
     \theta=2, 2V=1$ and $u=v=0.05$.
-    The $m=0$ simulations are seeded with a single diploid or tetraploid
-    individual with trait value 0, and assume $u=v=0$.
+    For the $m=0$ simulations, the trait value of the initial migrant was
+    Gaussian with mean zero and variance $2V$, and $u=v=0$ is assumed.
 \label{fig:estwmig2}}
@@
 This suggests that maladaptive gene flow in diploids can increase the
-probability of tetraploid establishment.
+probability of tetraploid establishment in some cases.
 However, at the same time we find that the tetraploid establishment probability
 declines monotonically with $m$, which highlights the negative effects of MCE.
-Together, these results confirm the view that, while cytotype differences are an
-effective barrier to maladaptive gene flow, this advantage to establishment is
-outweighed by the negative effects of MCE as $m$ increases.
 
@@
 
-Conditional on establishment, tetraploids have a higher trait mean both in the
-absence of migration and for moderate to strong migration, but not when
-migration is weak (\cref{fig:estwmig2}, bottom row).
+Conditional on establishment, tetraploids have a higher trait mean than
+diploids (\cref{fig:estwmig2}, bottom row).
 In the absence of migration, this is a consequence of the reduced rate of
 inbreeding and the resulting increased adaptive potential of tetraploids.
-Another consequence of this is that the average trait value in tetraploids at
-low population size (conditional on eventual establishment) is lower than that
-in diploids.
-This is because, in order to adapt before genetic variation is exhausted,
-tetraploids need not increase the trait mean as much as diploids do in the
-first couple of generations.
-The similar trait means for weak migration suggest that establishment depends
-mostly on the chance pick of a well-adapted migrant (note that the $m=0$
-simulations in \cref{fig:estwmig2} assume the initial migrant has $z=0$,
-whereas the $m=0.01$ simulations assume the initial trait value to be
-Gaussian).
+For weak migration, the difference in trait values between diploids and
+tetraploids, conditional on eventual establishment, is limited.
+This indicates the beneficial effects of migration on establishment in
+diploids: migration introduces new variation on which selection can act,
+counteracting the loss of genetic variance due to inbreeding.
+The genetic variance contributed by migration is however negligible in
+tetraploids.
 When migration is strong, tetraploids have markedly larger trait values
@@
 increases with increasing migration rate.
-In this regime, the increased strength of MCE due to increased migration and
-increased reproductive assurance in diploids is compensated by the stronger
-maladaptive gene flow experienced by diploids.
+In this regime, the effects of migration on MCE and reproductive assurance in
+diploids are compensated by the stronger maladaptive gene flow experienced by
+diploids.
 
@@
 \centering
-\includegraphics[width=0.9\linewidth]{/home/arzwa/dev/InfGenetics/doc/paper/figs/prez.pdf} 
+\includegraphics[width=0.9\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/fig5.pdf} 
     \caption{
@@
     $k$-ploid mother is pollinated by a $k$-ploid father.
-    All results are based on 25.000 replicate simulations.  We assume
-    $\gamma=0.25, \theta=1.5$ and $u=v=0.05$.  \label{fig:selfing}}
+    All results are based on 50.000 replicate simulations.  We assume
+    $\gamma=0.25, \theta=2$ and $u=v=0.05$.  \label{fig:selfing}}
 \end{figure}
@@
 and migration of maladapted individuals from a predominantly diploid source,
-we sought to determine the conditions under which tetraploids are more likely
-to establish a stable population.
-
-Importantly, we assume no intrinsic advantage or disadvantage of polyploids in
-the marginal habitat, i.e. the expected fitness of a migrant individual is the
-same regardless of the ploidy level.
-Differences in establishment probabilities are hence caused solely by aspects
-of autopolyploid genetics and the barrier to gene flow between subpopulations
-of different ploidy levels.
-This is undoubtedly unrealistic.
-For instance, neopolyploids are likely to suffer intrinsic fertility issues due
-to meiotic irregularities associated with multivalent formation
-\citep{bomblies2016,novikova2023}, and triploids may be inviable due to
-issues with endosperm development \citep{bretagnolle1995}.
-
-Similarly implausible is the assumption of a constant equilibrium genetic
-variance across cytotypes ($\beta_4=\sqrt{1/2}$ in our model), which we used in
-most of our results (but see \cref{fig:est1}).
-The meta-analysis performed by \cite{porturas2019} suggests that phenotypic
-variation across ploidy levels is indeed fairly constant, so this assumption
-may not yield too bad an approximation.
-It should be noted however that other authors have made different assumptions
-on how allelic effects (and hence genetic variance) scale across ploidy levels
-(in particular \cite{griswold2021}, who made an assumption equivalent to
-$\beta_4 = 1/2$ in our model), and such assumptions evidently impact the
-establishment probability (\cref{fig:est1}).
-Clearly, it would be very interesting to experimentally assess whether
-typical quantitative traits in mixed-ploidy populations follow approximately
-the mixed-ploidy infinitesimal model we outlined in this paper, and to actually
-estimate how the variance scales across ploidy levels.
-In addition, we have assumed a relatively high and constant rate of unreduced
+we sought to determine under which conditions tetraploids are more likely to
+establish a stable population.
+
+Throughout, we have assumed a relatively high and constant rate of unreduced
 gamete formation $u$ and triploid fertility $v$ in all our simulations (5\%),
@@
 prezygotic isolation affect the tetraploid establishment probability.
+
+Similarly, we have ignored mutation, which would reduce the rate at which
+genetic variation is lost through inbreeding \citep{barton2017}, and would
+likely do so differently across cytotypes (i.e. $\mu V_m$ is expected to differ
+for different ploidy levels, where $\mu$ is the mutation rate and $V_m$ the
+mutational variance).
+The contribution of new mutation to the genetic variance on the timescales we
+consider should however be very limited.
+Indeed, any individual at the time of establishment derives from a completely
+outbred migrant individual a relatively short time in the past, so that the
+opportunity for mutation to contribute to differences in establishment
+probability between diploids and tetraploids is negligible for realistic
+$\mu V_m$.
+
+Importantly, we assumed no intrinsic advantage or disadvantage of polyploids in
+the marginal habitat, i.e. the expected fitness of a migrant individual is the
+same regardless of the ploidy level.
+Differences in the likelihood of polyploid establishment are hence caused
+solely by aspects of autopolyploid genetics and the barrier to gene flow
+between subpopulations of different ploidy levels.
+This is undoubtedly unrealistic, as both trait values and fitness will often
+differ systematically across ploidy levels \citep{porturas2019}.
+For instance, neopolyploids are likely to suffer intrinsic fertility issues due
+to meiotic irregularities associated with multivalent formation
+\citep{bomblies2016,novikova2023}, and triploids may be inviable due to
+issues with endosperm development \citep{bretagnolle1995}.
+
+Similarly implausible is the assumption of a constant equilibrium genetic
+variance across cytotypes ($\beta_4=\sqrt{1/2}$ in our model), which we used in
+most of our results (but see \cref{fig:est1}).
+Empirical data on how the genetic variance scales across ploidy levels is scant
+and suggests that there is no general rule \citep{gallais2003,porturas2019}.
+The meta-analysis performed by \cite{porturas2019} does indicate that trait
+variance across ploidy levels is often fairly constant, so the assumption of
+equal genetic variance is arguably a reasonable default.
+It should be noted however that other authors have made different assumptions
+on how allelic effects (and hence genetic variance) scale across ploidy levels
+(in particular \cite{griswold2021}, who scaled allelic effects in a way that is
+equivalent to $\beta_4 = 1/2$ in our model).
+Such assumptions evidently impact the likelihood of polyploid establishment 
+(\cref{fig:est1}).
+More empirical data on quantitative traits in experimental or natural
+mixed-ploidy populations is needed to assess whether the mixed-ploidy
+infinitesimal model can adequately describe the genetics of quantitative traits
+across cytotypes, and to suggest plausible values for the relevant parameters
+($\alpha, \beta$).
 
@@
 relevant, as in the case of inbreeding depression
-\citep{ronfort1999,husband2008,clo2022b}.
+\citep{ronfort1999,gallais2003,husband2008,clo2022b}.
 Indeed, when inbreeding depression is due to recessive deleterious variation,
@@
 should be much rarer than in their diploid parents (the `masking' effect;
-\cite{otto2000}).
-In the presence of deleterious recessive variation, inbreeding during the
-establishment process should therefore incur a higher fitness cost in diploids
-relative to tetraploids, and hence further increase the probability of
-tetraploid establishment.
+\cite{husband1997,otto2000}).
+Inbreeding during the establishment process should therefore incur a higher
+fitness cost in diploids relative to neotetraploids, and hence further increase
+the probability of tetraploid establishment.
 How this plays out depends however on the \textit{rate} at which populations
@@
 Dominance and inbreeding depression may strongly affect the complicated
-relationship between selfing, migration load and tetraploid establishment
-probability.
+interaction between selfing and migration load in determining tetraploid
+establishment.
 \cite{griswold2021} studied the case where local fitness is determined by a
@@
 
+While in this study we focused on polyploid establishment in a peripheral
+habitat and how this is affected by migration from a diploid source, the
+mixed-ploidy infinitesimal framework could be used to address many other
+eco-evolutionary questions that arise in the study of mixed-ploidy populations.
+For instance, it could be of interest to develop a complicated individual-based
+model along the lines of \cite{oswald2011} to study the effects of selfing,
+assortative mating and competition on establishment and coexistence of
+tetraploids within diploid populations, but where the focal trait that
+determines fitness and assortative mating is not controlled by a few
+large-effect loci, but many loci of small effect.
+Similarly, our model could be straightforwardly extended to include population
+regulation and stabilizing selection, which would allow us to study polyploid
+establishment along an environmental gradient and the potential of
+polyploidization to promote range expansions \citep{polechova2015}.
+
 %\section*{Acknowledgements}

### 2025-04-03 notebooks/_estwmigration.jl

@@ -1,0 +1,643 @@
+using Pkg; Pkg.activate("/home/arzwa/dev/InfGenetics/")
+using Distributed, StatsBase, Plots, Serialization, Distributions, Optim
+addprocs(6)
+@everywhere using InfGenetics
+
+# estimate parameter of a censored geometric distribution by ML
+function logpdf_censored_geometric(p, c, x)
+    x < c ? log(p) + (x-1)*log(1-p) : (c-1)*log(1-p)
+end
+
+function mle_censored_geometric(c, xs)
+    fun = p -> -sum(logpdf_censored_geometric.(Ref(p), Ref(c), xs))
+    optimize(fun, 0.0, 1.0).minimizer
+end
+
+# Example        
+# ========
+function dosim(M=InfDemeMixEst(m=1.0, θ=fill(2.0, 3), U=Umat(0.05,0.05)))
+    while true
+        xs = simest(M)
+        cs = counts(xs[2][end][2].c, 2:4)
+        @info cs; cs[3] > cs[2] && return xs
+    end
+end
+
+ex = dosim()
+
+z_ = map(x->x.z, last.(ex[2]))
+P1 = plot()
+for i=1:length(z_)
+    scatter!(fill(i, length(z_[i])), z_[i], color=:black, ms=2, label="")
+end
+plot!(legend=false, xlabel="generation", ylabel="\$z\$")
+hline!([2])
+P2 = plot(mapreduce(x->log10.(counts(x.c, 2:4)), hcat, last.(ex[2]))',
+    legend=:topleft, marker=true, ms=2, ylabel="\$\\log_{10}N_k\$",
+    xlabel="generation", label=["diploid" "triploid" "tetraploid"])
+plot(P1, P2, size=(260*2,210), margin=3Plots.mm)
+
+
+# Basic model
+# ===========
+nrep = 50_000
+ms   = 10 .^ range(log10(0.01), stop=log10(3), length=10)
+zs   = [2.5, 2.0, 1.5, 1.0]
+γ    = 0.25
+U    = Umat(0.05, 0.05)
+nmax = 10000
+Nest = 100
+    
+# get baseline results without migration (but with unreduced gametes)
+base = map(zs) do z
+    M  = InfDemeMixEst(m=0.0, θ=fill(z, 3), U=U, γ=γ)
+    nr = 1_000_000
+    Nest = 150
[…]

### 2025-04-03 notebooks/equilibrium.jl

@@ -12,13 +12,57 @@ map([2,4]) do c
     mean(pop.z), var(pop.z)
 end
 
-MM = InfDemeMix(U=Umat(0.1, 0.1), V=0.5)
-N  = 1000
-pop = InfPop(z=zeros(N), c=fill(2,N))
-for i=1:100
-    pop = generation(pop, MM)
+
+
+# have a closer look at the trait distributions
+MM = InfDemeMix(U=Umat(0.05, 0.05), V=0.5, β=[1., 1, 1], α=maxα)
+N  = 20000
+pop = InfPop(z=randn(N), c=fill(2,N))
+Zs  = [Float64[] for i=1:3]
+for i=1:200
+    pop = InfGenetics.neutral_noinbreeding_generation(pop, MM)
     pop.F .= 0.0
-    i % 10 != 0 && continue
-    zs = pop.z[pop.c .== 4]
-    @info var(zs), 4*MM.β[3]^2*MM.V
+    (i % 1000 == 0 && @info i)
+    for k=2:4
+        length(Zs[k-1]) > 100000 && continue
+        zs = pop.z[pop.c .== k]
+        Zs[k-1] = [Zs[k-1] ; zs]
+    end
 end
+# seems that the higher ploidy levels converge very slowly to the equilibrium?
+plot()
+for Z in Zs
+    stephist!(Z, bins=(-6:0.2:6),normalize=true)
+end
+plot!()
+
+Vpred_ = MM.β .^ 2 .* (2:4) .* MM.V
+Vpred  = MM.β .^ 2 .* (2:4) .* MM.V
+Vpred[2] *= (1 + (2/3)*MM.α[1])
+Vpred[3] *= (1 + MM.α[1])
+map(1:3) do k
+    stephist(Zs[k], normalize=true, color=:black, 
+        legend=k==3 ? :outertopright : false,
+        title="$(strs[k+1])", label="simulation")
+    plot!(Normal(0,1), fill=true, color=:gray, alpha=0.2, label="Normal(0,2V)")
+    plot!(Normal(0, √(Vpred[k])), label="approximate eq.")
+    plot!(Normal(0, √(Vpred_[k])), label="monocytotypic eq.")
+end |> x->plot(x..., size=(800,180), xlabel="Trait value \$z\$", 
[…]

### 2025-04-03 notebooks/est-details.jl

@@ -1,0 +1,17 @@
+
+M = InfDemeMixEst(
+    U=Umat(0.0,0.0),  # no unreduced gametes 
+    γ=0.25,           # selection strength
+    m=0.0,            # no migration
+    θ=fill(1.5, 3),   # maladaptation
+    V=0.5)            # segregation variance
+
+Random.seed!(446)
+pop = InfPop(z=zeros(10), c=fill(2, 10))
+plot()
+for i=1:15
+    pop = generation(M, pop)
+    @info length(pop.z)
+    scatter!(fill(i, length(pop.z)), pop.z, color=:black, label="", ms=2)
+end
+plot!()

### 2025-04-03 notebooks/establishment.jl

@@ -1,96 +1,167 @@
 # Consider first, as in Barton & Etheridge 2018, for a given source population
 # the probability of a single diploid/tetraploid migrant establishing.
-using Distributed, Plots
-#addprocs(6)
-@everywhere using Parameters, InfGenetics
-@everywhere M = InfDemeMixEst(U=Umat(0.0,0.0), γ=0.25, m=0.0, θ=[2.0,2.0,2.0], V=0.5)
-gs = [0.0 ; 10 .^ range(log10(0.02), stop=log10(2), length=10)]
+using Distributed, Plots, Serialization, Parameters
+addprocs(8)
+@everywhere using InfGenetics, Parameters
+
+# Define the model
+M = InfDemeMixEst(
+    U=Umat(0.0,0.0),  # no unreduced gametes 
+    γ=0.25,           # selection strength
+    m=0.0,            # no migration
+    θ=fill(2.5, 3),   # maladaptation
+    V=0.5)            # segregation variance
+
+ps = pmap(_->simest2(M, x0=InfPop(z=[0.0], c=[2])), 1:1_000_000)
+pp = sum(first.(ps)) / length(ps)
+#pp = 0.002928
+
+
+# 1. Examine the effect of selection strength (γ)
+# -----------------------------------------------
+# Define a range of selection strengths to examine
+gs = [0.0 ; 10 .^ range(log10(0.02), stop=log10(2), length=15)]
 
 res = pmap(gs) do g
-    nr = 500_000
-    x2 = map(_->simest2(reconstruct(M, γ=g), x0=InfPop(z=[0.0], c=[2])), 1:nr)
-    x4 = map(_->simest2(reconstruct(M, γ=g), x0=InfPop(z=[0.0], c=[4])), 1:nr)
-    (sum(first.(x2)) / nr, sum(first.(x4)) / nr)
+    @info g
+    nr = 1_000_000  # number of replicate simulations
+    M = InfDemeMixEst(
+        U=Umat(0.0,0.0),  # no unreduced gametes 
+        γ=g,              # selection strength
+        m=0.0,            # no migration
+        θ=fill(2.5, 3),   # maladaptation
[…]

### 2025-04-03 notebooks/estwmigration.jl

@@ -1,6 +1,7 @@
-using Pkg; Pkg.activate("/home/arthur_z/dev/InfGenetics/")
+using Pkg; Pkg.activate("/home/arzwa/dev/InfGenetics/")
 using Distributed, StatsBase, Plots, Serialization, Distributions, Optim
-addprocs(4)
+using DataFrames, CSV
+addprocs(6)
 @everywhere using InfGenetics
 
 # estimate parameter of a censored geometric distribution by ML
@@ -13,337 +14,149 @@ function mle_censored_geometric(c, xs)
     optimize(fun, 0.0, 1.0).minimizer
 end
 
-# Example        
-# ========
-function dosim(M=InfDemeMixEst(m=1.0, θ=fill(2.0, 3), U=Umat(0.05,0.05)))
-    while true
-        xs = simest(M)
-        cs = counts(xs[2][end][2].c, 2:4)
-        @info cs; cs[3] > cs[2] && return xs
-    end
+function make_a_df(sims, nmax)
+    map(eachcol(sims)) do col
+        map(col) do xs 
+            ps, reps = xs
+            ts = getindex.(reps, 2)
+            tt = 1/mle_censored_geometric(nmax, ts)
+            nsucc = sum(first.(reps))
+            xx = sum(map(x->x[3][3] > x[3][1] && x[1], reps))
+            p = xx/nsucc
+            e1,e2 = jeffreys_interval(xx, nsucc, 0.05)
+            merge(ps, (nsucc=nsucc, n=xx, t=tt, p=p, p1=p-e1, p2=e2-p))
+        end 
+    end |> x->vcat(x...) |> DataFrame
 end
-
-ex = dosim()
-
-z_ = map(x->x.z, last.(ex[2]))
-P1 = plot()
-for i=1:length(z_)
-    scatter!(fill(i, length(z_[i])), z_[i], color=:black, ms=2, label="")
-end
-plot!(legend=false, xlabel="generation", ylabel="\$z\$")
-hline!([2])
-P2 = plot(mapreduce(x->log10.(counts(x.c, 2:4)), hcat, last.(ex[2]))',
-    legend=:topleft, marker=true, ms=2, ylabel="\$\\log_{10}N_k\$",
-    xlabel="generation", label=["diploid" "triploid" "tetraploid"])
[…]

### 2025-04-03 notebooks/figs.jl

@@ -1,0 +1,73 @@
+using CSV, DataFrames, Plots, Serialization, InfGenetics
+plotsdefault()
+
+ms = 10 .^ range(log10(0.01), stop=log10(3), length=15)
+
+df1 = CSV.read("data/sims1b-df.csv", DataFrame)
+df2 = CSV.read("data/sims2b-df.csv", DataFrame)
+sort!(df1, [:z2, :m], rev=true)
+sort!(df2, [:z2, :m], rev=true)
+
+#base = @NamedTuple{z::Float64, p2::Float64, p4::Float64}[(z = 2.0, p2 = 6.46e-5, p4 = 0.0014164), (z = 1.5, p2 = 0.0004414, p4 = 0.0046594), (z = 1.0, p2 = 0.0023506, p4 = 0.0139396)]
+base = deserialize("data/baseline3.jls")
+
+P1 = plot(title="(A) Time to establishment")
+for gdf in groupby(df1, :z2)
+    plot!(gdf[:,:m], gdf[:,:t], marker=true, ms=2, label="\$\\theta=$(gdf[1,:z2])\$")
+end
+P1 = plot(P1, marker=true, ms=2, yscale=:log10, xlabel="\$m\$",
+    ylabel="\$\\overline{T}\$", xscale=:log10, legend=:bottomleft, 
+    ylim=(10,500_000), yticks=[10,100,1000,10000,100000])
+P2 = plot(title="(B) Tetraploid establishment", xscale=:log10)
+for (df, ls, label) in zip([df1, df2], [:solid, :dot], [true, false])
+    for (i,gdf) in enumerate(groupby(df, [:z2]))
+        gdf = sort(gdf, :m)
+        plot!(gdf[:,:m], gdf[:,:p], ribbon=(gdf[:,:p1], gdf[:,:p2]),
+            marker=true, ms=2, ls=ls, color=i,
+            label=label ? "\$\\theta=$(gdf[1,:z2])\$" : "", fillalpha=0.2)
+    end
+end
+P2 = plot(P2, marker=true, ms=2, xlabel="\$m\$",
+    ylabel="\$P_4\$", xscale=:log10, legend=false)
+ppi = InfGenetics.cytotype_equilibrium(Umat(0.05,0.05))
+hline!(ppi[[3]], label="",
+    color=:black, xlabel="\$m\$", ylabel="\$P_4\$", ls=:dash)
+for i=1:4
[…]

### 2025-04-03 notebooks/inbreeding.jl

@@ -1,0 +1,39 @@
+
+nrep = 100
+m    = 1.0
+zs   = 2.0
+#ρs   = [fill(r, 3) for r=0:0.2:1.0]
+σs   = [[0.0, s, s] for s=0:0.2:1.0]
+γ    = 0.25
+U    = Umat(0.05, 0.05)
+nmax = 10000
+Nest = 100
+
+function tetest(x) 
+    y = counts(x.c, 2:4)
+    y[3] > y[1]
+end
+
+sims = pmap(σs) do σ
+    @info σ
+    reps2 = Vector{Float64}[]
+    reps4 = Vector{Float64}[]
+    while length(reps2) < nrep || length(reps4) < nrep 
+        _, sim = simest(
+            InfDemeMixEst(m=m, θ=fill(zs, 3), U=U, σ=σ), nmax=nmax, Nest=Nest)
+        ff = map(x->mean(x[2].F), sim)
+        if tetest(sim[end][2]) 
+            length(reps4) < nrep && push!(reps4, ff) 
+        else
+            length(reps2) < nrep && push!(reps2, ff)
+        end
+    end
+    reps2, reps4
+end
+
+map(sims) do (s2, s4)
+    x2 = avg(s2)
+    x4 = avg(s4)
+    plot(x2, xscale=:log10, ylim=(0,1))
+    plot!(x4)
+end |> x->plot(x..., legend=false, size=(900,300))

### 2025-04-03 notebooks/ri-establishment.jl

@@ -1,0 +1,197 @@
+# Consider for the case of succesful establishment (without/with migration),
+# how relative fitness of migrants evolves
+using Plots
+using Parameters
+using Distributed
+using Serialization
+addprocs(5)
+@everywhere using InfGenetics
+@everywhere using StatsBase
+
+@everywhere const Sims = Vector{InfPop{Float64}}
+
+# collect n2 and n4 establishment replicates for both diploid and tetraploids
+@everywhere function collect_est(f, n2, n4)
+    x2 = Vector{Sims}(undef, n2)
+    x4 = Vector{Sims}(undef, n4)
+    c2 = c4 = 1
+    while c2 <= n2 || c4 <= n4
+        res = simuntil(f)
+        cs = counts(res[end][2].c, 2:4)
+        if cs[3] > cs[1] && c4 <= n4
+            x4[c4] = last.(res)
+            c4 += 1
+            @info c4, c2
+        elseif cs[1] > cs[3] && c2 <= n2
+            x2[c2] = last.(res)
+            c2 += 1
+            @info c4, c2
+        end
+    end
+    return x2, x4
+end
+
+function avg(xs)
+    map(i->mean(filter(
+        y->y>0, [x[i] for x in xs if length(x) >= i])), 
+        1:maximum(length, xs))
+end
+
+# initial migrant z=0
+M = InfDemeMixEst(U=Umat(0.00,0.00), γ=0.25, m=0.0, θ=[2.0,2.0,2.0], V=0.5)
+f2=()->simest(reconstruct(M, γ=0.25), x0=InfPop(z=zeros(1), c=Int64[2]))
+f4=()->simest(reconstruct(M, γ=0.25), x0=InfPop(z=zeros(1), c=Int64[4]))
+x2, _ = collect_est(f2, 1000, 0) 
+_, x4 = collect_est(f4, 0, 1000) 
+
+# initial migrant random
+M = InfDemeMixEst(U=Umat(0.00,0.00), γ=0.25, m=0.0, θ=[2.0,2.0,2.0], V=0.5)
+f2=()->simest(reconstruct(M, γ=0.25), x0=InfPop(z=randn(1), c=Int64[2]))
+f4=()->simest(reconstruct(M, γ=0.25), x0=InfPop(z=randn(1), c=Int64[4]))
+x2b, _ = collect_est(f2, 1000, 0) 
+_, x4b = collect_est(f4, 0, 1000) 
+
[…]

### 2025-04-03 notebooks/traits.jl

@@ -1,11 +1,86 @@
-using InfGenetics, Distributions, Plots
+using InfGenetics, Distributions, Plots, StatsPlots, Parameters
+using InfGenetics: haploidgamete, diploidgamete
+plotsdefault()
 
-N = 2000
-M = InfDemeMix(U=Umat(1.0, 0.0))
-pop = InfPop(z=rand(Normal(-2, √(2M.V)), N), c=fill(2,N))
-gen1 = generation(pop, M)
+N = 1000
 
-stephist(pop.z, fill=true, fillalpha=0.2)
-stephist!(gen1.z, fill=true, fillalpha=0.2)
+xs = map(-2:1.0:2) do z̄
+    # set u=1, all offspring is tetraploid
+    M = InfDemeMix(U=Umat(0.0, 0.0))
+    pop = InfPop(z=rand(Normal(z̄, √(2M.V)), N), c=fill(2,N))
+    pop2 = generation(pop, M)
+    pop4 = generation(pop, reconstruct(M, U=Umat(1.0, 0.0)))
+    z̄, pop2.z, pop4.z
+end
 
+map(xs) do (z̄, z2, z4)
+    bins = -6:0.25:6
+    stephist(z2, fill=true, fillalpha=0.2, label="diploids", normalize=true, bins=bins)
+    stephist!(z4, fill=true, fillalpha=0.2, label="tetraploids", normalize=true, bins=bins)
+    vline!([z̄], label="", legend=:outertopright, size=(400,200)) 
+end |> x->plot(x..., size=(900,300))
+
+function familydist(zi, zj, ki, kj, ci, cj, Fi, Fj, M)
+    @unpack β = M
+    k   = ki + kj
+    yi  = (ki/ci)*(zi/β[ci-1])
+    yj  = (kj/cj)*(zj/β[cj-1])
+    yi, Vi = ki == 1 ? haploidgamete(M, zi, Fi, ci) : diploidgamete(M, zi, Fi, ci)
+    yj, Vj = kj == 1 ? haploidgamete(M, zj, Fj, cj) : diploidgamete(M, zj, Fj, cj)
+    z̄ij = (β[k-1]/β[ci-1])*yi + (β[k-1]/β[cj-1])*yj
+    Vij = β[k-1]^2*(Vi + Vj)
+    (z=z̄ij, V=Vij)
+end
+
[…]

### 2025-04-03 notebooks/unreducedgametes.jl

@@ -1,0 +1,43 @@
+
+using Pkg; Pkg.activate("/home/arzwa/dev/InfGenetics/")
+using Distributed, StatsBase, Plots, Serialization, Distributions, Optim
+addprocs(8)
+@everywhere using InfGenetics
+
+# estimate parameter of a censored geometric distribution by ML
+function logpdf_censored_geometric(p, c, x)
+    x < c ? log(p) + (x-1)*log(1-p) : (c-1)*log(1-p)
+end
+
+function mle_censored_geometric(c, xs)
+    fun = p -> -sum(logpdf_censored_geometric.(Ref(p), Ref(c), xs))
+    optimize(fun, 0.0, 1.0).minimizer
+end
+
+us = 0.02:0.01:0.1
+m  = 0.75
+n  = 20
+γ  = 0.125
+z  = 1.5
+sims = pmap(us) do u
+    U = Umat(u, u)
+    M = InfDemeMixEst(m=m, θ=fill(z, 3), U=U, γ=γ)
+    ks = Int64[]
+    K  = 0
+    while length(ks) < n 
+        est, ngen, c = simest2(M)
+        K += 1
+        if c[end] > c[1]
+            push!(ks, K)
+            K = 0
+        end
+    end
+    @info u, ks
+    u, ks
+end
+
+map(sims) do (u, ks)
+    p = mle_censored_geometric(Inf, ks)
+    u, log(p/(1-p))
+end |> z->plot(z, ms=2, marker=true)
+    

### 2025-04-03 src/InfGenetics.jl

@@ -2,13 +2,14 @@ module InfGenetics
 
 using LinearAlgebra, Distributions, StatsBase, Random, Parameters
 
-export jeffreys_interval, Umat, simest, simest2, simuntil, fiterate
+export jeffreys_interval, Umat, simest, simest2, simuntil, fiterate, maxα, strs
 export InfDemeMixEstPair, InfDemeMixEstSelf, InfDemeMixEst, InfDemeMix, InfDemeHS, InfPop, FinPop, generation
 
 include("infmix.jl")
 include("finmix.jl")
 include("infmixest.jl")
 include("simulation.jl")
+include("rv.jl")
 #include("infmixhs.jl")
 #include("infmixestpair.jl")
 #include("infmixestself.jl")
@@ -33,5 +34,7 @@ end
 
 mpeq(u, v) = (3 - 3u - 6v + √((u + 2v - 1)*(5u + 2v - 1)))/(4 - 2u -8v)
 Umat(u, v) = [1-u u ; v v ; 0.0 1-u]
+const maxα = [1/2, 1/4, 1/6]
+const strs = ["", "diploid", "triploid", "tetraploid"]
 
 end # module

### 2025-04-03 src/infmix.jl

@@ -130,3 +130,30 @@ function diploidgamete(M, z, F, c)
     end
 end
 
+function neutral_noinbreeding_generation(
+        pop::InfPop, M::InfDemeMix)
+    @unpack F, Φ = pop
+    @unpack α, β, w, U = M
+    N = popsize(pop)
+    ws = fitnesses(pop, U, w, z->1.0)
+    ws ./= mean(ws)
+    idx = sample(1:N, Weights(ws), 2N, replace=true) 
+    pop_ = similar(pop)
+    for k=1:N
+        i = idx[k]; j = idx[N+k]
+        zi, Fi, ci = pop[i]
+        zj, Fj, cj = pop[j]
+        gi, yi, Vi = gamete(M, zi, Fi, ci)
+        gj, yj, Vj = gamete(M, zj, Fj, cj)
+        cij = gi + gj
+        βij = β[cij-1]
+        βi  = β[ci-1]
+        βj  = β[cj-1]
+        Yi  = rand(Normal(yi*βij/βi, √(βij^2*Vi))) 
+        Yj  = rand(Normal(yj*βij/βj, √(βij^2*Vj))) 
+        pop_.z[k] = Yi + Yj
+        pop_.c[k] = cij
+    end
+    return pop_
+end
+

### 2025-04-03 src/infmixest.jl

@@ -16,6 +16,21 @@
     c::Vector{T} = cytotype_equilibrium(U)
 end
 
+function parameters(M::InfDemeMixEst)
+    @unpack U, β, α, σ, ρ, γ, m, V, θ, c, w = M 
+    w2, w3, w4 = w
+    b2, b3, b4 = β
+    s2, s3, s4 = σ
+    a2, a3, a4 = α
+    r2, r3, r4 = ρ
+    z2, z3, z4 = θ
+    c2, c3, c4 = c
+    (u21=U[1,1], u22=U[1,2], u31=U[2,1], u32=U[2,2], u41=U[3,1], u42=U[3,2],
+     w2=w2, w3=w3, w4=w4, b2=b2, b3=b3, b4=b4, s2=s2, s3=s3, s4=s4, r2=r2,
+     r3=r3, r4=r4, z2=z2, z3=z3, z4=z4, c2=c2, c3=c3, c4=c4, V=V, g=γ, 
+     a2=a2, a3=a3, a4=a4, m=m)
+end
+
 function cytotype_equilibrium(U)
     u21, u22 = U[1,:]
     u31, u32 = U[2,:]
@@ -140,17 +155,27 @@ function assortative_mating!(W, ρ, c)
 end
 
 function migration(M::InfDemeMixEst, pop::InfPop)
-    @unpack m, θ, V, c, β = M
+    @unpack m, θ, V, c, β, α = M
     @unpack z, c, F, Φ = pop
     nr = popsize(pop)
     nm = rand(Poisson(m))
     cm = rand(Multinomial(nm, M.c))
     cm = vcat([fill(i+1, k) for (i, k) in enumerate(cm)]...)
     # Vzₖ/Vz₂ = (k/2)βₖ² => Vzₖ = kβₖ²V
-    zm = [rand(Normal(0.0, √(m*β[m-1]^2*V))) for m in cm]
+    # But, this is not really the case: this is an equilibrium result for an
+    # isolated monocytotypic population. It does hold for α = 0...
+    # Here we use the small u approximation
+    Vs = (β .^ 2) .* [2,3,4] .* V .* [1, 1+2α[1]/3, 1+α[1]]  
+    zm = [rand(Normal(0.0, √Vs[m-1])) for m in cm]
+    add_unrelated_individuals(pop, zm, cm, zeros(nm))
+end
+
+function add_unrelated_individuals(pop, zm, cm, Fm)
+    nr = popsize(pop)
+    nm = length(zm)
     z  = [pop.z ; zm]
[…]

### 2025-04-03 src/rv.jl

@@ -1,0 +1,43 @@
+# approximate the reproductive value of an individual in a given population
+# assuming the population composition does not change.
+const cs = [2,3,3,4]
+
+# Recursive! Exponential in `n`!
+"""
+    approx_rv
+
+This gives a crude idea of the reproductive value of an individual dropped in a
+population. It calculates the mean relative fitness of the individual, and then
+convolves over its different possible families, computing the fitness for the
+average offspring within the family... (this is not very clear) i.e. it does
+the recursion:
+```   
+RV = W(z₀)*(Σᵢ Pᵢ W(zᵢ))
+```
+where Pᵢ is the probability that, conditional on the individual with trait z₀
+reproducing, it does so in a family `i` with average offspring trait value zᵢ.
+The recursion is broken after `n` iterates. This is not feasible for
+populations of decent size and `n` larger than 4 say.
+"""
+function approx_rv(M, pop, individual, n=3)
+    n == 0 && return 1.0
+    w, W, Z = mean_relative_fitness(M, pop, individual)
+    w == 0.0 && return 0.0
+    P = W ./ sum(W)
+    w_ = 0.0
+    for i=1:size(P,1), j=1:4
+        wij = approx_rv(M, pop, (z=Z[i,j], c=cs[j]), n-1)
+        w_ += P[i,j]*wij
+    end
+    return w*w_
+end
+
+function mean_relative_fitness(M, pop, individual)
+    @unpack z, c = individual
+    pop_ = InfGenetics.add_unrelated_individuals(pop, [z], [c], [0.0])
+    W, Z, _ = InfGenetics.fitnesses(M, pop_) 
+    # ignore the variance (and selection) and go with the mean
+    w = mapreduce(i->vec(mean(W[:,:,i], dims=2)), +, 1:4)
+    w ./= mean(w)
+    w[end], W[end,:,:], Z[end,:,:]
+end

### 2025-04-03 src/simulation.jl

@@ -10,7 +10,7 @@ function simest(M; x0=InfPop(z=Float64[]), nmax=Inf, Nest=100)
     xs = [(1, generation(M, x0))]
     while !condition(last(xs)..., M, nmax, Nest)
         (n, x) = last(xs) 
-        x_ = generation(M, x)
+        x_ = generation(M, x, Nmax=Nest) # no need to simulate > Nest
         if popsize(x_) == 0 
             # extinction -- prune the stored generations
             xs = [(n + 1, x_)]
@@ -25,7 +25,7 @@ end
 function simest2(M; x0=InfPop(z=Float64[]), nmax=Inf, Nest=100)
     x = generation(M, x0); n = 1
     while !condition(n, x, M, nmax, Nest)
-        x = generation(M, x); n += 1
+        x = generation(M, x, Nmax=Nest); n += 1
 	end
     N = popsize(x)
     c = N > 0 ? counts(x.c, 2:4) : [0, 0, 0]

### 2025-04-07 doc/paper/response.md
update

@@
 combination of gametes from a particular parental pair after selection in
-proportion with the expected fitness using eq. 11 (number in new version).
+proportion with the expected fitness using eq. 11 (equation number in new
+version).
 In the previous version, I then sampled, incorrectly, offspring trait values
@@
 
-The density of trait values after selection is found easily.
+The density of trait values *after* selection is however also Gaussian and
+found easily.
 Writing $f(z)$ for the Gaussian density of trait values with mean $\bar{z}$ and
@@
 The incorrect results from the previous submission can be seen as a kind of
-zeroth order approximation to the correct results: i.e. they implement
+'zeroth order' approximation to the correct results: i.e. they implement
 selection on the mean trait value across families, but not selection on the
@@
 mixed-ploidy populations. I have now added a small paragraph at the end that
-does so.
+does so (lines 549-561).
 
@@
 establishment in such models?  Or does it come from your experience with the
-model and you see that Under N=100 extinction remains likely? More details
+model and you see that Under $N=100$ extinction remains likely? More details
 would be needed here.
@@
 refer to his paper instead of paraphrasing his paragraph. Specifically, I now
-write:
+write (lines 63-65):
 "Many empirical studies of mixed-ploidy populations find that polyploids
@@
 adaptation and have added a couple of references to recent work on the genetics
-of adaptation in autopolyploids: 
+of adaptation in autopolyploids (lines 75-82): 
 "More often than not, local adaptation is polygenic in nature (Pritchard and Di
@@
 However, I have now added a more explicit reference to @oswald2011 in the concluding
-note (see the associated remark of Reviewer 1), to highlight that it would be
-possible to do a study that is comparable to theirs but assuming infinitesimal
-quantitative genetics instead of a model with a small number of loci.
+note (lines 549-561, see also the associated remark of Reviewer 1), to
+highlight that it would be possible to do a study that is comparable to theirs
+but assuming infinitesimal quantitative genetics instead of a model with a
+small number of loci.
 
@@
 I now added some further discussion, referring explicitly to Griswold's scaling
-in the Discussion section. I now also refer to @porturas2019 and @gallais2003.
+in the Discussion section (lines 462-475). I now also refer to @porturas2019
+and @gallais2003.
 
@@
 
-As far as I can tell, $N$ is defined when first mentioned. I slightly altered
-the sentence to make this more clear.
+As far as I can tell, $N$ is defined when first mentioned (line 216). I
+slightly altered the sentence to make this more clear.
 
@@
 offspring survives with a probability that is determined by its fitness. This
-is indeed 'selection before migration'. I tried to make this more clear.
+is indeed 'selection before migration'. I tried to make this more clear (lines
+218-219).
 
@@
 
-I do not think, however, that fig 4 suggests such a transition point though. As
-noted in the main text the difference between the $m=0$ and $m=0.01$ results in
-fig. 4 are mostly due to the fact that the latter assumes a Gaussian
-distribution of migrant trait values, whereas the former assumes a fixed
-initial migrant trait value.
-(*"The similar trait means for weak migration suggest
-that establishment depends mostly on the chance pick of a well-adapted migrant
-(note that the $m = 0$ simulations in fig. 4 assume the initial migrant has $z =
-0$, whereas the $m = 0.01$ simulations assume the initial trait value to be
-Gaussian)."*)
+Note that I now also included lines showing the results for the limiting
+establishment probabilities as $m \rightarrow 0$, i.e.
+  $$\lim_{m \rightarrow 0}P_{4} = \frac{\sum_k P_{k,4}\pi_k}{\sum_{k}(P_{k,2} + P_{k,4})\pi_k}$$
+where $\pi_k$ is the equilibrium frequency of ploidy level $k$ on the mainland,
+and $P_{k,j}$ is the probability of $j$-ploid establishment when the initial
+migrant is of ploidy level $k$.
+These are more informative than the horizontal lines in the previous version,
+which were a somewhat *ad hoc* extrapolation of the results from figure 2 but
+not really the proper limiting probabilities.
+
+I do not think, however, that the previous version of fig 4 suggests such a
+transition point though. As noted in the main text (previous version) the
+difference between the $m=0$ and $m=0.01$ results in the previous fig. 4 are
+mostly due to the fact that the latter assumes a Gaussian distribution of
+migrant trait values, whereas the former assumes a fixed initial migrant trait
+value.
+As I no longer consider this case very relevant, I replaced the $m=0$ results
+in fig 4. with results from simulations where the initial migrant has a trait
+value drawn from a Gaussian distribution. This makes the results more
+straightforward to interpret.
 

### 2025-04-07 doc/paper/paper.tex
update

@@
 1/2$), this response will differ for different ploidy levels.
-Differences across cytotypes in how the genetic variance evolves over time as a
-consequence of inbreeding will then further cause rates of adaptation to
-differ, leading to different establishment probabilities even when
-$\beta_4^2=1/2$.
+Different rates of inbreeding due to differences in ploidy level will then
+further cause rates of adaptation to differ, leading to different establishment
+probabilities, even when $\beta_4^2=1/2$.
 
@@
 tetraploids compared to diploids) is the benefit of the slower rate of
-inbreeding in tetraploids canceled by the reduction in genetic variance 
-(\cref{fig:est1}B,C).
+inbreeding in tetraploids canceled by the reduction in the equilibrium genetic
+variance (\cref{fig:est1}B,C).
 
@@
     $\alpha_4=1/6$ (i.e. maximum $\alpha$).
-    The solid horizontal lines mark $\frac{p_4\pi_4}{(p_4\pi_4 + p_2\pi_2)}$,
-    where $p_2$ and $p_4$ are the establishment probabilities for a single
-    diploid and a single tetraploid individual in the absence of migration (as
-    in \cref{fig:est1}, but with unreduced gamete formation included). 
-    The line for the $\theta=3$ case lies around 0.25 and is not shown.
-    The dashed horizontal line marks the proportion of tetraploid migrants
+    The horizontal lines mark the establishment probabilities in the
+    limit as $m\rightarrow 0$ (solid lines: without double reduction; dashed
+    lines: maximum $\alpha$).
+    The black horizontal line marks the proportion of tetraploid migrants
     (i.e. the proportion of tetraploids at equilibrium in the source
     population, $\approx 0.3$\%).
-    The baseline predictions (horizontal lines) are based on $5\times 10^6$
-    simulation replicates.
-    All other results are based on 50.000 replicate simulations.
+    The baseline predictions (horizontal lines) are based on 500.000 simulation
+    replicates.
+    All other results are based on 100.000 replicate simulations.
     We assume $\gamma=0.25$ and $u=v=0.05$.
@@
 Importantly, the tetraploid establishment probability is considerably larger
-than the expected proportion of tetraploid migrants over a large part of the
-parameter range (\cref{fig:estwmig}B, black dashed line).
-Furthermore, we find that for small migration rates, the probability that
-tetraploids establish rather than diploids can be larger than expected based on
-the establishment probabilities in the absence of migration (colored lines in
-\cref{fig:estwmig}), depending on how maladapted migrant individuals are.
-This suggests that maladaptive gene flow in diploids can increase the
-probability of tetraploid establishment in some cases.
-However, at the same time we find that the tetraploid establishment probability
-declines monotonically with $m$, which highlights the negative effects of MCE.
+than the expected proportion of tetraploid migrants over almost the entire
+parameter range examined (\cref{fig:estwmig}B, black dashed line).
+However, the probability of tetraploid establishment does decline monotonically
+with the migration rate, showing that the negative effects of MCE on tetraploid
+establishment outweigh the positive effects of reduced maladaptive gene flow in
+the absence of prezygotic isolation.
 
@@
 and establishment from a single or limited number of outbred individuals, 
-the inbreeding coefficient is expected to increase less rapidly conditional on
-the event of tetraploid establishment.
+the inbreeding coefficient is expected to increase less rapidly.
 On the other hand, migration mostly introduces unrelated diploids, which cross
@@
 strongly from maladaptive gene flow when the population size is low, while
-tetraploids enjoy some reproductive isolation.
+tetraploids are much more reproductively isolated from migrants.
 Furthermore, in these replicates, tetraploids tend to emerge and rise in
@@
 This is undoubtedly unrealistic, as both trait values and fitness will often
-differ systematically across ploidy levels \citep{porturas2019}.
+differ systematically across ploidy levels (see e.g. \cite{porturas2019}).
 For instance, neopolyploids are likely to suffer intrinsic fertility issues due
@@
 Similarly implausible is the assumption of a constant equilibrium genetic
-variance across cytotypes ($\beta_4=\sqrt{1/2}$ in our model), which we used in
+variance across cytotypes ($\beta_4^2=1/2$ in our model), which we used in
 most of our results (but see \cref{fig:est1}).
@@
 determines fitness and assortative mating is not controlled by a few
-large-effect loci, but many loci of small effect.
+large-effect loci (as in \cite{oswald2011}), but many loci of small effect.
 Similarly, our model could be straightforwardly extended to include population

### 2025-04-07 doc/paper/titlepage.tex
update

@@
 
-I thank Christelle Fraïsse for encouragement and feedback. Thanks to Quinten
+I thank two anonymous reviewers and the associate editor for their constructive
+criticism that resulted in substantial improvements of the present article.
+I thank Christelle Fraïsse for encouragement and feedback, as well as Quinten
 Bafort, Felipe Kauai and Frederik Mortier for comments on an early draft

### 2025-04-07 notebooks/estwmigration.jl
update

@@ -2,7 +2,7 @@ using Pkg; Pkg.activate("/home/arzwa/dev/InfGenetics/")
 using Distributed, StatsBase, Plots, Serialization, Distributions, Optim
 using DataFrames, CSV
 addprocs(6)
-@everywhere using InfGenetics
+@everywhere using InfGenetics, Distributions
 
 # estimate parameter of a censored geometric distribution by ML
 function logpdf_censored_geometric(p, c, x)
@@ -29,42 +29,54 @@ function make_a_df(sims, nmax)
     end |> x->vcat(x...) |> DataFrame
 end
 
-# Basic model
-# ===========
-nrep = 50_000
-ms   = 10 .^ range(log10(0.01), stop=log10(3), length=15)
+# Basic model (m -> 0 limit)
+# ==========================
+nrep = 100_000
+Nest = 100
 zs   = [3.0, 2.5, 2.0, 1.5]
 γ    = 0.25
 U    = Umat(0.05, 0.05)
-nmax = 10000
-Nest = 100
-M  = InfDemeMixEst(m=0.0, θ=fill(2.5, 3), U=U, γ=γ)
+M    = InfDemeMixEst(m=0.0, θ=fill(2.5, 3), U=U, γ=γ)
 
 # baseline
 base = pmap(zs) do z
     @info z
     M  = InfDemeMixEst(m=0.0, θ=fill(z, 3), U=U, γ=γ)
-    nr = 2_500_000
-    x2 = map(_->simest2(M, x0=InfPop(z=randn(1), c=[2]), Nest=Nest), 1:nr)
-    x3 = map(_->simest2(M, x0=InfPop(z=randn(1), c=[3]), Nest=Nest), 1:nr)
-    x4 = map(_->simest2(M, x0=InfPop(z=randn(1), c=[4]), Nest=Nest), 1:nr)
-    p22 = sum(first.(x2) .&& map(x->argmax(x[3]) == 1, x2)) / nr
-    p24 = sum(first.(x2) .&& map(x->argmax(x[3]) == 3, x2)) / nr
-    p32 = sum(first.(x2) .&& map(x->argmax(x[3]) == 1, x2)) / nr
-    p34 = sum(first.(x2) .&& map(x->argmax(x[3]) == 3, x2)) / nr
-    p4  = sum(first.(x4) .&& map(x->argmax(x[3]) == 3, x4)) / nr
[…]

### 2025-04-07 notebooks/figs.jl

@@ -3,13 +3,16 @@ plotsdefault()
 
 ms = 10 .^ range(log10(0.01), stop=log10(3), length=15)
 
-df1 = CSV.read("data/sims1b-df.csv", DataFrame)
-df2 = CSV.read("data/sims2b-df.csv", DataFrame)
+df1 = CSV.read("data/sims1c-df.csv", DataFrame)
+df2 = CSV.read("data/sims2c.csv", DataFrame)
 sort!(df1, [:z2, :m], rev=true)
 sort!(df2, [:z2, :m], rev=true)
 
-#base = @NamedTuple{z::Float64, p2::Float64, p4::Float64}[(z = 2.0, p2 = 6.46e-5, p4 = 0.0014164), (z = 1.5, p2 = 0.0004414, p4 = 0.0046594), (z = 1.0, p2 = 0.0023506, p4 = 0.0139396)]
-base = deserialize("data/baseline3.jls")
+m0 = [
+#    (3.0, 0.18622614267240276),
+    (2.5, 0.13471031236560937),
+    (2.0, 0.08502786246117869),
+    (1.5, 0.05699203484708603)]
 
 P1 = plot(title="(A) Time to establishment")
 for gdf in groupby(df1, :z2)
@@ -32,25 +35,24 @@ P2 = plot(P2, marker=true, ms=2, xlabel="\$m\$",
 ppi = InfGenetics.cytotype_equilibrium(Umat(0.05,0.05))
 hline!(ppi[[3]], label="",
     color=:black, xlabel="\$m\$", ylabel="\$P_4\$", ls=:dash)
-for i=1:4
-    px = (ppi[3]*base[i].p4)/(ppi[1]*base[i].p2 + ppi[3]*base[i].p4)
-    @info px
-    hline!([px ], color=i, label="")
+for i=1:3
+    hline!([m0[i][2]], color=i, label="")
 end
-plot!(bg_legend=:transparent, ylim=(0,0.17))
-plot(P1, P2, size=(275*2,230), margin=3Plots.mm, titlefont=9)
-savefig("doc/img_/fig3.pdf")
+plot!(bg_legend=:transparent, ylim=(0,0.162),)
+plot(P1, P2, size=(275*2,230), margin=3Plots.mm, titlefont=9,  xlim=(0.0095,3.2))
+
+savefig("doc/img/fig3.pdf")
 
 
 # Selfing/assortative mating
[…]

### 2025-04-07 src/infmix.jl

@@ -31,7 +31,7 @@ A mixed-ploidy population with trait values `z`, inbreeding coefficients `F`,
 coancestry coefficients `Φ` and ploidy levels `c`.
 """
 @with_kw struct InfPop{T}
-    z::Vector{T}   # trait values (scalars, for now)
+    z::Vector{T}   = Float64[] # trait values (scalars, for now)
     F::Vector{T}   = zeros(length(z)) # inbreeding
     Φ::Matrix{T}   = diagm(fill(0.5, length(z))) # coancestry
     c::Vector{Int} = fill(2, length(z)) # ploidy levels

### 2025-04-07 src/infmixest.jl

@@ -159,6 +159,12 @@ function migration(M::InfDemeMixEst, pop::InfPop)
     @unpack z, c, F, Φ = pop
     nr = popsize(pop)
     nm = rand(Poisson(m))
+    add_n_migrants(M, nm, pop)
+end
+
+function add_n_migrants(M::InfDemeMixEst, nm, pop::InfPop)
+    @unpack m, θ, V, c, β, α = M
+    @unpack z, c, F, Φ = pop
     cm = rand(Multinomial(nm, M.c))
     cm = vcat([fill(i+1, k) for (i, k) in enumerate(cm)]...)
     # Vzₖ/Vz₂ = (k/2)βₖ² => Vzₖ = kβₖ²V
@@ -176,7 +182,8 @@ function add_unrelated_individuals(pop, zm, cm, Fm)
     z  = [pop.z ; zm]
     c  = [pop.c ; cm]
     F  = [pop.F ; Fm]
-    Φ  = [pop.Φ zeros(nr, nm); zeros(nm, nr) zeros(nm, nm)]
+    Φm = diagm(fill(0.5, nm))  # coancestry with self is 0.5!
+    Φ  = [pop.Φ zeros(nr, nm); zeros(nm, nr) Φm]
     InfPop(z=z, c=c, F=F, Φ=Φ)
 end
 
@@ -191,10 +198,13 @@ function generation(M::InfDemeMixEst, pop::InfPop{T}; Nmax=100000) where T
     # _pop is next generation
     pop_ = migration(M, pop)
     N_   = popsize(pop_)
+    N_ == 0 &&  return InfPop()  # XXX popsize after migration = 0 ? Nothing can happen
     W, Z, V = fitnesses(M, pop_)
-    𝔼N = isempty(W) ? 0.0 : sum(W) 
+#   𝔼N = isempty(W) ? 0.0 : sum(W) 
+    𝔼N = sum(W)  
     # number of individuals in the next generation
     N = min(rand(Poisson(𝔼N)), Nmax)
+    N == 0 && return InfPop()    # XXX popsize next generation is zero
     # sample parental pairs x offspring ploidy combinations
     P = zeros(N, N_)
     C = CartesianIndices((1:N_, 1:N_, 1:4))
@@ -234,9 +244,11 @@ function generation(M::InfDemeMixEst, pop::InfPop{T}; Nmax=100000) where T
[…]

### 2025-04-07 src/simulation.jl

@@ -29,7 +29,26 @@ function simest2(M; x0=InfPop(z=Float64[]), nmax=Inf, Nest=100)
 	end
     N = popsize(x)
     c = N > 0 ? counts(x.c, 2:4) : [0, 0, 0]
-    return (popsize(x) > 0, n, c)
+    return (N > 0, n, c)
+end
+
+# weak migration
+function _simest3(M; x=InfPop(z=Float64[]), nmax=Inf, Nest=100)
+    p0 = pdf(Poisson(M.m), 0)
+    M0 = reconstruct(M, m=0)
+    n = 0
+    while !condition(n, x, M, nmax, Nest)
+        if popsize(x) == 0
+            n += rand(Geometric(1-p0))
+            x = add_n_migrants(M, 1, x)  
+            # XXX this should be the distribution conditional on being nonzero,
+            # for small m this is very likely 1
+        end
+        x = generation(M0, x, Nmax=Nest); n+= 1
+	end
+    N = popsize(x)
+    c = N > 0 ? counts(x.c, 2:4) : [0, 0, 0]
+    return (N > 0, n, c)
 end
 
 function simuntil(f) 
@@ -38,3 +57,4 @@ function simuntil(f)
 		x && return xs
 	end
 end
+

### 2025-04-07 test/runtests.jl

@@ -66,3 +66,6 @@ v = v0(x)
 @info "" v v0(x, 0.3) v*1.3
 
 
+model = InfDemeMixEst(m=0.1, θ=fill(2.5, 3), U=Umat(0.05,0.05))
+pop = InfPop(z=randn(20), c=fill(2, 20))
+@btime InfGenetics.generation(model, pop)

### 2025-04-22 doc/paper/paper.tex
update

@@
     \title{Autopolyploid establishment through polygenic adaptation}
-    %\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
-    %\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
-    \author{}
+    \author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
+    \affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
+    %\author{}
     \date{\vspace{-5ex}}
@@
     implemented in Julia \citep{julia}.
-    Documented code and simulation notebooks are available at \url{[anonymized]}.
-%\url{https://github.com/arzwa/InfGenetics}.
+    %Documented code and simulation notebooks are available at \url{[anonymized]}.
+    Documented code and simulation notebooks are available at
+    \url{https://github.com/arzwa/InfGenetics}.
 
@@
 
-%\section*{Acknowledgements}
-%
-%I thank Christelle Fraïsse for encouragement and feedback. Thanks to Quinten
-%Bafort, Felipe Kauai and Frederik Mortier for comments on an early draft
-%version.
-%I acknowledge funding from the Research Foundation -- Flanders (FWO, Junior
-%Postdoctoral Fellowship 1272625N).
+\section*{Funding}
+
+I acknowledge funding from the Research Foundation -- Flanders (FWO, Junior
+Postdoctoral Fellowship 1272625N).
+
+\section*{Conflict of interest}
+
+I have no conflict of interest to declare.
+
+\section*{Acknowledgements}
+
+I thank two anonymous reviewers and the associate editor for their constructive
+criticism that resulted in substantial improvements of the present article.
+I thank Christelle Fraïsse for encouragement and feedback, as well as Quinten
+Bafort, Felipe Kauai and Frederik Mortier for comments on an early draft
+version.
 

### 2025-04-22 notebooks/estwmigration.jl

@@ -31,17 +31,16 @@ end
 
 # Basic model (m -> 0 limit)
 # ==========================
-nrep = 100_000
+nrep = 500_000
 Nest = 100
-zs   = [3.0, 2.5, 2.0, 1.5]
+zs   = [2.5, 2.0, 1.5]
 γ    = 0.25
 U    = Umat(0.05, 0.05)
-M    = InfDemeMixEst(m=0.0, θ=fill(2.5, 3), U=U, γ=γ)
 
 # baseline
 base = pmap(zs) do z
     @info z
-    M  = InfDemeMixEst(m=0.0, θ=fill(z, 3), U=U, γ=γ)
+    M  = InfDemeMixEst(m=0.0, θ=fill(z, 3), α=maxα, U=U, γ=γ)
     zsource = Normal(0,1)
     x2 = map(_->simest2(M, x0=InfPop(z=rand(zsource, 1), c=[2]), Nest=Nest), 1:nrep)
     x3 = map(_->simest2(M, x0=InfPop(z=rand(zsource, 1), c=[3]), Nest=Nest), 1:nrep)
@@ -49,6 +48,7 @@ base = pmap(zs) do z
     (z, x2, x3, x4)
 end
 
+M  = InfDemeMixEst(m=0.0, θ=fill(2.0, 3), U=U, γ=γ)
 m0 = map(base) do y
     p4 = map(y[2:end]) do xs
         length(filter(x->x[1] && (argmax(x[3]) == 3), xs))/nrep
@@ -72,7 +72,7 @@ end
 # ========================
 nrep = 50_000
 ms   = 10 .^ range(log10(0.01), stop=log10(3), length=15)
-zs   = [3.0, 2.5, 2.0, 1.5]
+zs   = [2.5, 2.0, 1.5]
 γ    = 0.25
 U    = Umat(0.05, 0.05)
 nmax = 10_000
@@ -88,7 +88,7 @@ sims1 = pmap(Iterators.product(ms, zs)) do (m, z)
 end
 
 df1 = make_a_df(sims1, nmax)
-#CSV.write("data/sims1b-df-z3.csv", df1)
+CSV.write("data/sims1c_.csv", df1)
 
 
 # With double reduction etc. (maxα)
@@ -103,7 +103,7 @@ sims2 = pmap(Iterators.product(ms, zs)) do (m, z)
 end
 
 df2 = make_a_df(sims2, nmax)
[…]

### 2025-04-22 notebooks/figs.jl

@@ -3,16 +3,34 @@ plotsdefault()
 
 ms = 10 .^ range(log10(0.01), stop=log10(3), length=15)
 
-df1 = CSV.read("data/sims1c-df.csv", DataFrame)
-df2 = CSV.read("data/sims2c.csv", DataFrame)
+df1a = CSV.read("data/sims1c-df.csv", DataFrame)
+df1b = CSV.read("data/sims1c_.csv", DataFrame)
+
+function docombine(df, on)
+    map(collect(groupby(df, on))) do gd 
+        nsucc = sum(gd[:,:nsucc])
+        n     = sum(gd[:,:n])
+        t     = mean(gd[:,:t])
+        p     = n/nsucc
+        e1,e2 = jeffreys_interval(n, nsucc, 0.05) 
+        merge((t=t, nsucc=nsucc, n=n, p=p, p1=p-e1, p2=e2-p), 
+            Dict(k=>gd[1,k] for k in on), )
+    end |> DataFrame
+end
+
+df1 = docombine(vcat(df1a,df1b), [:m,:z2])
+df2a=CSV.read("data/sims2c.csv", DataFrame)
+df2b=CSV.read("data/sims2c_.csv", DataFrame)
+df2 = docombine(vcat(df2a,df2b), [:m,:z2])
 sort!(df1, [:z2, :m], rev=true)
 sort!(df2, [:z2, :m], rev=true)
 
 m0 = [
 #    (3.0, 0.18622614267240276),
-    (2.5, 0.13471031236560937),
-    (2.0, 0.08502786246117869),
-    (1.5, 0.05699203484708603)]
+    (2.5, 0.128),
+    (2.0, 0.088),
+    (1.5, 0.063)]
+m0b = [(2.5, 0.152), (2.0, 0.101), (1.5, 0.066)]
 
 P1 = plot(title="(A) Time to establishment")
 for gdf in groupby(df1, :z2)
@@ -36,7 +54,8 @@ ppi = InfGenetics.cytotype_equilibrium(Umat(0.05,0.05))
 hline!(ppi[[3]], label="",
     color=:black, xlabel="\$m\$", ylabel="\$P_4\$", ls=:dash)
 for i=1:3
-    hline!([m0[i][2]], color=i, label="")
+    hline!([m0[i][2]], color=i, alpha=0.4, lw=2, label="")
[…]

### 2025-04-22 notebooks/_est1.jl
cleaning up some more

@@ -1,259 +1,0 @@
-### A Pluto.jl notebook ###
-# v0.19.42
-
-using Markdown
-using InteractiveUtils
-
-# ╔═╡ baef8376-1514-11ef-1d35-277242ce9bf3
-using Pkg; Pkg.activate("..")
-
-# ╔═╡ 019ac4ed-1679-4683-8d8c-04217475d2f8
-using InfGenetics, StatsBase, Distributions, Parameters, Plots
-
-# ╔═╡ 117ed0b3-7d8a-42f7-92aa-94d59d236f39
-default(size=(280,220), grid=false, legend=:outertopright, framestyle=:box, 
-	fontfamily="Computer modern",  bg_legend=:transparent, titlefont=9, guidefont=9, title_loc=:left, markerstrokewidth=0)
-
-# ╔═╡ cfb8cc81-7801-476a-a2b9-a88286e6c9d7
-md"""
-## Equilibrium variance
-"""
-
-# ╔═╡ 247e5eea-2ba4-4792-80a2-eef8e6d8acdd
-let MM = InfDemeMix(U=[1.0 0.0 ; NaN NaN; 0.0 1.0], V=0.5)
-	pop = InfPop(z=zeros(500), c=fill(4,500))
-	for i=1:20
-		pop = generation(pop, MM)
-		pop.F .= 0.0
-	end
-	mean(pop.z), var(pop.z)
-end
-
-# ╔═╡ 289ed1fc-c191-412c-b045-42d59c636fef
-md"""
-## Establishment of a single migrant individual
-
-This is the first thing Barton & Etheridge examined: finding the probability of non-extinction when starting with a single individual randomly drawn from the source population. This sounds simple, but it does not appear analytically tractable.
-
[…]

### 2025-04-22 notebooks/_est1b.jl

@@ -1,135 +1,0 @@
-### A Pluto.jl notebook ###
-# v0.19.42
-
-using Markdown
-using InteractiveUtils
-
-# ╔═╡ 2bf45042-10ad-464a-9329-4012d1acaadb
-using Pkg; Pkg.activate("..")
-
-# ╔═╡ 604c8252-4f64-43fb-bab8-07d7abd20c40
-using InfGenetics, StatsBase, Distributions, Parameters, Plots
-
-# ╔═╡ 49ab8458-7c71-474e-9077-7800e2d97f44
-default(size=(280,220), grid=false, legend=:outertopright, framestyle=:box, 
-	fontfamily="Computer modern", fg_legend=:transparent, bg_legend=:transparent, titlefont=9, guidefont=9, title_loc=:left, markerstrokewidth=0)
-
-# ╔═╡ 3dad5ec8-c615-4811-9ae5-a10ed7286151
-md"""
-## Simulation functions
-"""
-
-# ╔═╡ b7b2f2cc-f2ba-4ac9-a16f-797eccfea03e
-function simest(M, x=InfPop([rand(Normal(M.θ[1], √(2M.V)))]))
-	xs = [x]
-	while 0 < length(x.z) < 100
-		x = generation(M, x)
-		push!(xs, x)
-	end
-	return (length(x.z) > 0, xs)
-end
-
-# ╔═╡ a34aba59-c35a-4d8c-9480-13b73bcdec38
-function simuntil(f) 
-	while true
-		x, xs = f()
-		x && return xs
-	end
-end
-
-# ╔═╡ 8092b6c6-e903-4915-b293-a0114f411535
-function sim(f, n) 
-	map(1:n) do _
-		x, xs = f()
-	end
-end
-
-# ╔═╡ cf762c4d-c0ac-4667-bd92-327714b6e00c
-md"""
-## Establishment from a single migrant
-"""
-
-# ╔═╡ ac88335c-ef4d-43e5-af0c-33e18095317a
-InfDemeMixEst(U=Umat(0.0, 0.0))
-
-# ╔═╡ b1e578d6-a2cb-44e2-81c0-db50d746e086
-ps = let n=100000, M=InfDemeMixEst(U=Umat(0.0, 0.0), m=0.0, θ=fill(2.0,3))
[…]

### 2025-04-22 notebooks/_est2.jl

@@ -1,136 +1,0 @@
-### A Pluto.jl notebook ###
-# v0.19.42
-
-using Markdown
-using InteractiveUtils
-
-# ╔═╡ 99fd9810-18e3-11ef-3ccf-0f13790a9f6e
-using Pkg; Pkg.activate("..")
-
-# ╔═╡ 65220b6c-c975-4a05-8377-19cfb88f92d8
-using InfGenetics, StatsBase, Distributions, Parameters, Plots
-
-# ╔═╡ 97808c41-877f-454b-a61a-9b3160a176e4
-default(size=(280,220), grid=false, 
-	fg_legend=:transparent, 
-	legend=:outertopright, framestyle=:box, 
-	fontfamily="Computer modern", titlefont=9, 
-	markerstrokewidth=0,
-	guidefont=9, title_loc=:left)
-
-# ╔═╡ a7f8a332-9e11-4025-93ba-d25116994de2
-function simest(M, nmax, Nest=100)
-	i = 0
-	while i < nmax
-		xs = [generation(M, InfPop(z=Float64[]))]; i += 1
-		while 0 < length(xs[end].z) < Nest 
-			push!(xs, generation(M, xs[end])); i += 1
-		end
-		est = length(xs[end].z) > 0
-		if est
-			cyt = counts(xs[end].c, 2:4)
-			return (est, i, cyt, xs)
-		end
-	end
-	return (false, nmax, [0, 0, 0], InfPop{Float64}[])
-end
-
-# ╔═╡ 1e8cd76a-9f61-4da4-8906-9c81beb59924
-let M = InfDemeMixEst(U=Umat(0.05, 0.05), γ=0.25, m=1.0, V=0.5)
-	simest(M, Inf)
-end
-
-# ╔═╡ 424f97c8-7f30-4532-a212-ed353788886a
-mss = 10 .^ range(log10(0.1), stop=log10(3), length=10)
-
-# ╔═╡ 585b30e3-5f6b-4294-8399-4d3c52958a0a
-Xs = map(mss) do m
-	M = InfDemeMixEst(U=Umat(0.05, 0.05), γ=0.25, m=m, V=0.5, θ=fill(-1.0, 3))
-	map(1:250) do _
-		simest(M, Inf)
-	end
-end
-
-# ╔═╡ f90e5d0b-6cb7-41fa-bc1b-b46be2ebde31
-Ys = map(Xs) do xs
[…]

### 2025-04-22 notebooks/_estwmigration.jl

@@ -1,643 +1,0 @@
-using Pkg; Pkg.activate("/home/arzwa/dev/InfGenetics/")
-using Distributed, StatsBase, Plots, Serialization, Distributions, Optim
-addprocs(6)
-@everywhere using InfGenetics
-
-# estimate parameter of a censored geometric distribution by ML
-function logpdf_censored_geometric(p, c, x)
-    x < c ? log(p) + (x-1)*log(1-p) : (c-1)*log(1-p)
-end
-
-function mle_censored_geometric(c, xs)
-    fun = p -> -sum(logpdf_censored_geometric.(Ref(p), Ref(c), xs))
-    optimize(fun, 0.0, 1.0).minimizer
-end
-
-# Example        
-# ========
-function dosim(M=InfDemeMixEst(m=1.0, θ=fill(2.0, 3), U=Umat(0.05,0.05)))
-    while true
-        xs = simest(M)
-        cs = counts(xs[2][end][2].c, 2:4)
-        @info cs; cs[3] > cs[2] && return xs
-    end
-end
-
-ex = dosim()
-
-z_ = map(x->x.z, last.(ex[2]))
-P1 = plot()
-for i=1:length(z_)
-    scatter!(fill(i, length(z_[i])), z_[i], color=:black, ms=2, label="")
-end
-plot!(legend=false, xlabel="generation", ylabel="\$z\$")
-hline!([2])
-P2 = plot(mapreduce(x->log10.(counts(x.c, 2:4)), hcat, last.(ex[2]))',
-    legend=:topleft, marker=true, ms=2, ylabel="\$\\log_{10}N_k\$",
-    xlabel="generation", label=["diploid" "triploid" "tetraploid"])
-plot(P1, P2, size=(260*2,210), margin=3Plots.mm)
-
-
-# Basic model
-# ===========
-nrep = 50_000
-ms   = 10 .^ range(log10(0.01), stop=log10(3), length=10)
-zs   = [2.5, 2.0, 1.5, 1.0]
-γ    = 0.25
-U    = Umat(0.05, 0.05)
-nmax = 10000
-Nest = 100
-    
-# get baseline results without migration (but with unreduced gametes)
-base = map(zs) do z
-    M  = InfDemeMixEst(m=0.0, θ=fill(z, 3), U=U, γ=γ)
-    nr = 1_000_000
-    Nest = 150
[…]

### 2025-04-22 notebooks/autotets.jl

@@ -1,39 +1,0 @@
-
-M = InfDemeMix(α=fill(0.0, 3), U=[0. 0.; 0. 0.; 0. 1.0])
-N = 200
-zinit = rand(Normal(0, √(4M.V)), N)
-pop = InfPop(z=zinit, c=fill(4,N))
-xs = fiterate((pop,i)->generation(pop, M), pop, 1000, callback=x->(z=x[2].z,F=x[2].F))
-
-plot(map(x->var(x.z), xs))
-plot!(t->exp(-t/4N))
-
-α=0.3
-M = InfDemeMix(α=fill(α, 3), U=[0. 0.; 0. 0.; 0. 1.0])
-N = 200
-zinit = rand(Normal(0, √(4M.V)), N)
-pop = InfPop(z=zinit, c=fill(4,N))
-xs = fiterate((pop,i)->generation(pop, M), pop, 500, callback=x->(z=x[2].z,F=x[2].F))
-
-F = map(x->mean(x.F), xs)
-plot(F)
-hline!([α/(2+α)])
-
-
-vz = map(x->var(x.z), xs)
-plot(vz)
-plot!(t->2*(1+2α)*M.V*exp(-t/(4N)))
-
-
-
-α = 0.1
-N = 200
-M = InfDemeMix(α=fill(α, 3), U=[0. 0.; 0. 0.; 0. 1.0])
-pop = InfPop(z=rand(Normal(0, √(4M.V)), N), c=fill(4,N))
-
-for i=1:100
-    pop = generation(pop, M, z->1)
-    pop.F .= 0.0
-    pop.Φ .= 0.0
-    @info var(pop.z), 2*(1+2α)*M.V
-end

### 2025-04-22 notebooks/det.jl

@@ -1,73 +1,0 @@
-### A Pluto.jl notebook ###
-# v0.19.42
-
-using Markdown
-using InteractiveUtils
-
-# ╔═╡ bc36ad7c-18ca-11ef-286f-a523cc8cf5e5
-using Pkg; Pkg.activate("..")
-
-# ╔═╡ 18369b1e-76e0-4ba3-9e6d-4fb2a4099307
-using InfGenetics, StatsBase, Distributions, Parameters, Plots, SymPy
-
-# ╔═╡ 2c9b1135-5bd0-46bf-ba0b-b2629d4ab511
-default(
-	size=(280,220), 
-	grid=false, 
-	fg_legend=:transparent, 
-	legend=:outertopright, 
-	framestyle=:box, 
-	fontfamily="Computer modern", 
-	titlefont=9, 
-	guidefont=9, 
-	title_loc=:left
-)
-
-# ╔═╡ d22a3791-2965-4e29-bdcf-6a17b8e6993e
-u, v, g, w2, w3, w4 = @syms u v g w2 w3 w4
-
-# ╔═╡ 66822d67-3ca4-4427-8863-0fc3b430b2f9
-function nextgen(g1, u, v, w2, w3, w4) 
-	g2 = 1 - g1 
-	x2 = g1^2
-	x3 = 2g1*g2
-	x4 = g2^2
-	w̄  = x2*w2 + x3*w3 + x4*w4
-	x2_ = x2*w2/w̄
-	x3_ = x3*w3/w̄
-	x4_ = x4*w4/w̄
-	g1_ = x2_*(1-u) + x3_*v
-	g2_ = x2_*u + x3_*v + x4_*(1-u)
-	g1__ = g1_/(g1_ + g2_)	
-	g2__ = g2_/(g1_ + g2_)
-	g1__
-end
-
-# ╔═╡ e4478703-873d-4394-8159-46a42516db22
-sols = solve(simplify((nextgen(g, u, v, w2, w3, w4) - g)/g), g)
-
-# ╔═╡ 828aa9c7-038d-4926-a211-83074e4c723a
-ex = simplify(subs(sols[2], w3=>w4, w2=>1))
-
-# ╔═╡ ac0f69fb-c68c-48b9-a5ae-f2c75e148b4b
[…]

### 2025-04-22 notebooks/equilibrium.jl

@@ -1,68 +1,0 @@
-
-using InfGenetics, StatsBase, Distributions, Parameters, Plots
-
-MM = InfDemeMix(U=Umat(0.0, 0.0), V=0.5)
-N  = 1000
-map([2,4]) do c
-    pop = InfPop(z=zeros(N), c=fill(c,N))
-    for i=1:20
-        pop = generation(pop, MM)
-        pop.F .= 0.0
-    end
-    mean(pop.z), var(pop.z)
-end
-
-
-
-# have a closer look at the trait distributions
-MM = InfDemeMix(U=Umat(0.05, 0.05), V=0.5, β=[1., 1, 1], α=maxα)
-N  = 20000
-pop = InfPop(z=randn(N), c=fill(2,N))
-Zs  = [Float64[] for i=1:3]
-for i=1:200
-    pop = InfGenetics.neutral_noinbreeding_generation(pop, MM)
-    pop.F .= 0.0
-    (i % 1000 == 0 && @info i)
-    for k=2:4
-        length(Zs[k-1]) > 100000 && continue
-        zs = pop.z[pop.c .== k]
-        Zs[k-1] = [Zs[k-1] ; zs]
-    end
-end
-# seems that the higher ploidy levels converge very slowly to the equilibrium?
-plot()
-for Z in Zs
-    stephist!(Z, bins=(-6:0.2:6),normalize=true)
-end
-plot!()
-
-Vpred_ = MM.β .^ 2 .* (2:4) .* MM.V
-Vpred  = MM.β .^ 2 .* (2:4) .* MM.V
-Vpred[2] *= (1 + (2/3)*MM.α[1])
-Vpred[3] *= (1 + MM.α[1])
-map(1:3) do k
-    stephist(Zs[k], normalize=true, color=:black, 
-        legend=k==3 ? :outertopright : false,
-        title="$(strs[k+1])", label="simulation")
-    plot!(Normal(0,1), fill=true, color=:gray, alpha=0.2, label="Normal(0,2V)")
-    plot!(Normal(0, √(Vpred[k])), label="approximate eq.")
-    plot!(Normal(0, √(Vpred_[k])), label="monocytotypic eq.")
-end |> x->plot(x..., size=(800,180), xlabel="Trait value \$z\$", 
-    layout=grid(1,3,widths=[0.26,0.26,0.48]), margin=4Plots.mm)
-
[…]

### 2025-04-22 notebooks/est-details.jl

@@ -1,17 +1,0 @@
-
-M = InfDemeMixEst(
-    U=Umat(0.0,0.0),  # no unreduced gametes 
-    γ=0.25,           # selection strength
-    m=0.0,            # no migration
-    θ=fill(1.5, 3),   # maladaptation
-    V=0.5)            # segregation variance
-
-Random.seed!(446)
-pop = InfPop(z=zeros(10), c=fill(2, 10))
-plot()
-for i=1:15
-    pop = generation(M, pop)
-    @info length(pop.z)
-    scatter!(fill(i, length(pop.z)), pop.z, color=:black, label="", ms=2)
-end
-plot!()

### 2025-04-22 notebooks/finvsinf2.jl

@@ -1,113 +1,0 @@
-### A Pluto.jl notebook ###
-# v0.19.42
-
-using Markdown
-using InteractiveUtils
-
-# ╔═╡ 2e34eb4f-d93f-477a-9d6a-a8844dde46f5
-using Pkg; Pkg.activate("..")
-
-# ╔═╡ 85e0ef20-c60c-45f8-82b4-eea93e03a7a6
-using InfGenetics, StatsBase, Distributions, Parameters, Plots
-
-# ╔═╡ 2b4feedf-f7ff-4b15-a665-2dfc98a1030e
-default(size=(280,220), grid=false, fg_legend=:transparent, 
-	legend=:outertopright, framestyle=:box, 
-	fontfamily="Computer modern", titlefont=9, guidefont=9, title_loc=:left)
-
-# ╔═╡ ec095e7e-e18b-43bb-b510-eb6b9ccc9be0
-function smooth(xs::Vector{T}, window, step) where T
-    n = 1
-    ys = Tuple{Float64,T}[]
-    while n+window < length(xs)
-        x = n+window/2
-        push!(ys, (x, mean(filter(!isnan, xs[n:n+window]))))
-        n += step
-    end
-    xl = (n + length(xs))/2 
-    push!(ys, (xl, mean(xs[n:end])))
-end
-
-# ╔═╡ d521aceb-0edf-4227-8008-b0695e03423a
-function cb(x)  # callback function
-    _, pop = x
-    z2 = pop.z[pop.c .== 2]
-    z3 = pop.z[pop.c .== 3]
-    z4 = pop.z[pop.c .== 4]
-    z2, z3, z4
-end
-
-# ╔═╡ 73ce667b-eb15-4d6f-95a3-6ed70bd6e644
-X1 = let N = 250, Ls = [500, 100, 50, 20, 10], ngen=2500
-    M = InfDemeMix(U=[1.0 0.0; 0.0 0.0; 0.0 1.0])
-    p1 = InfPop(z=rand(Normal(0, √(2M.V)), N), c=fill(4,N))
-    xs1 = fiterate((x,i)->generation(x, M, z->1), deepcopy(p1), ngen, callback=cb)
-    xss = map(Ls) do L
-         @info L
-         Gs = [randn(L,4)/√(2L) for i=1:N]
[…]

### 2025-04-22 notebooks/finvsinf3.jl

@@ -1,86 +1,0 @@
-using Distributed, Distributions, StatsBase, Serialization
-addprocs(6)
-@everywhere using InfGenetics, Distributions
-
-nrep = 10
-ngen = 2000
-L    = 50
-N    = 250
-M    = InfDemeMix(U=[1.0 0.0; 0.0 0.0; 0.0 1.0])
-
-@everywhere function cb(x)  # callback function
-    _, pop = x
-    z2 = pop.z[pop.c .== 2]
-    z3 = pop.z[pop.c .== 3]
-    z4 = pop.z[pop.c .== 4]
-    z2, z3, z4
-end
-
-# Tetraploids
-Xi = pmap(1:nrep) do _
-    P = InfPop(z=rand(Normal(0, √(2M.V)), N), c=fill(4,N))
-    X = InfGenetics.fiterate((x,i)->generation(x, M, z->1), deepcopy(P), ngen, callback=cb)
-end
-Xf = pmap(1:nrep) do _
-    G = [randn(L,4)/√(2L) for i=1:N]
-    P = FinPop(sum.(G), G, fill(4, N))
-    X = fiterate((x,i)->generation(x, M, z->1), deepcopy(P), ngen, callback=cb)
-end
-vsi = mapreduce(x->map(var, last.(x)), hcat, Xi) 
-vsf = mapreduce(x->map(var, last.(x)), hcat, Xf) 
-
-P1 = plot(mean(vsi, dims=2), label="\$L \\rightarrow \\infty\$")
-plot!(mean(vsf, dims=2)    , label="\$L = $L\$")
-plot!(t->exp(-t/4N), 
-    xlabel="generation", 
-    ylabel="\$V\$", 
-    color=:black, ls=:dash, label="") 
-
-# Tetraploids with double reduction
-M   = InfDemeMix(U=[1.0 0.0; 0.0 0.0; 0.0 1.0], α=[1/6, 1/6, 1/6])
-Xib = pmap(1:nrep) do _
-    P = InfPop(z=rand(Normal(0, √(2M.V)), N), c=fill(4,N))
-    X = InfGenetics.fiterate((x,i)->generation(x, M, z->1), deepcopy(P), ngen, callback=cb)
-end    
-Xfb = pmap(1:nrep) do _
-    G = [randn(L,4)/√(2L) for i=1:N]
-    P = FinPop(sum.(G), G, fill(4, N))
[…]

### 2025-04-22 notebooks/hardsel.jl

@@ -1,38 +1,0 @@
-using InfGenetics, Plots, PlotThemes, StatsBase, Parameters; theme(:hokusai)
-using Distributions
-
-N = 500
-M = InfDemeHS()
-
-# equilibrate
-xs = let pop = InfPop(z=rand(Normal(0, √(2M.V)), N), c=fill(2,N))
-    map(1:2000) do i
-        pop = generation(pop, M)
-        pop.F .= 0.0
-        pop.Φ .= 0.0
-        pop
-    end
-end
-
-# environmental change
-ys = let M = InfDemeHS(β=ones(3))
-    os = map(x->4*sin(0.05x), 0:251)
-    map(1:100) do j
-        pop = deepcopy(xs[end])
-        res = map(enumerate(os)) do (i,o)
-            M = reconstruct(M, θ=fill(o, 3))
-            pop = generation(pop, M)
-        end
-        cs = counts(pop.c, 1:4)
-        @info j, cs
-        (sum(cs) > 0, cs[4] > cs[2], res)
-    end
-end;
-
-y = ys[10][3]
-P1 = plot(permutedims(hcat(map(x->counts(x.c, 2:4), y)...)), ylabel="\$N\$", label=["diploid" "triploid" "tetraploid"], legend=:topleft)
-P2 = plot(map(x->5*sin(0.05x), 0:251), ylabel="\$z\$", label="\$\\theta\$", color=:gray, legend=:topright)
-plot!(map(x->mean(x.z), y), ribbon=map(x->2std(x.z), y), color=2, label="\$\\bar{z}\$")
-plot(P1, P2, size=(550,200), layout=(1,2), xlabel="generation", margin=3Plots.mm)
-
-

### 2025-04-22 notebooks/inbreeding.jl

@@ -1,39 +1,0 @@
-
-nrep = 100
-m    = 1.0
-zs   = 2.0
-#ρs   = [fill(r, 3) for r=0:0.2:1.0]
-σs   = [[0.0, s, s] for s=0:0.2:1.0]
-γ    = 0.25
-U    = Umat(0.05, 0.05)
-nmax = 10000
-Nest = 100
-
-function tetest(x) 
-    y = counts(x.c, 2:4)
-    y[3] > y[1]
-end
-
-sims = pmap(σs) do σ
-    @info σ
-    reps2 = Vector{Float64}[]
-    reps4 = Vector{Float64}[]
-    while length(reps2) < nrep || length(reps4) < nrep 
-        _, sim = simest(
-            InfDemeMixEst(m=m, θ=fill(zs, 3), U=U, σ=σ), nmax=nmax, Nest=Nest)
-        ff = map(x->mean(x[2].F), sim)
-        if tetest(sim[end][2]) 
-            length(reps4) < nrep && push!(reps4, ff) 
-        else
-            length(reps2) < nrep && push!(reps2, ff)
-        end
-    end
-    reps2, reps4
-end
-
-map(sims) do (s2, s4)
-    x2 = avg(s2)
-    x4 = avg(s4)
-    plot(x2, xscale=:log10, ylim=(0,1))
-    plot!(x4)
-end |> x->plot(x..., legend=false, size=(900,300))

### 2025-04-22 notebooks/markovchain.jl

@@ -1,650 +1,0 @@
-### A Pluto.jl notebook ###
-# v0.19.40
-
-using Markdown
-using InteractiveUtils
-
-# ╔═╡ 79b8767e-1d98-11ef-3714-11ec27859293
-using SymPy, Combinatorics
-
-# ╔═╡ aa153ad4-8db6-4532-a2fb-5cba6623f33b
-using Distributions
-
-# ╔═╡ 70a0a20a-6eff-45da-9b2f-90f99ebf735d
-using LinearAlgebra
-
-# ╔═╡ ce47f179-e473-44f9-8172-03bd6681cd19
-u,v,N,i,j = @syms u v N i j 
-
-# ╔═╡ af107df1-be26-484a-9ea2-8a7fd1527006
-hap = (i/N)*(1-u) + (j/N)*v
-
-# ╔═╡ a2db8450-259a-43d3-a21e-de604406278c
-dip = (i/N)*u + (j/N)*v + ((N-i-j)/N)*(1-u)
-
-# ╔═╡ 875f8d97-7e3e-4314-aa5d-509604e59963
-g1 = simplify(hap/(hap + dip))
-
-# ╔═╡ aa3882da-7d13-4b6e-a721-99b1fc3e0b0f
-dipp = simplify(g1^2)
-
-# ╔═╡ 65341879-b2b6-4388-b1eb-e335369b8e23
-tripp = simplify(2g1*(1-g1))
-
-# ╔═╡ c8d74230-05c8-4044-bedd-de3a8ec3d7d1
-tetp = simplify((1-g1)^2)
-
-# ╔═╡ a6f49211-ddfe-427d-ac3d-e7629cdd0f2a
-p2 = lambdify(dipp, (N, u, v, i, j))
-
-# ╔═╡ 8f3370c5-76ef-4b11-be62-896e22056719
-p3 = lambdify(tripp,(N, u, v, i, j))
-
-# ╔═╡ e1e79282-0adc-4d2b-9ba5-5f005b9fa080
-multinomialpmf(k, l, N, p1, p2) = 
-	pdf(Multinomial(N, [p1, p2, 1-p1-p2]), [k, l, N-k-l])
-
-# ╔═╡ 2bbf6d73-5492-4cec-9c73-447cf8e9f49e
-function transitions(N, u, v)
-	P = zeros((N+1)*(N+1), (N+1)*(N+1))
-	L = LinearIndices((N+1,N+1))
[…]

### 2025-04-22 notebooks/mixedploidy-inf.jl

@@ -1,230 +1,0 @@
-using InfGenetics, Distributions, StatsBase
-
-N = 500
-M = InfDemeMix(α=zeros(3), β=ones(3))
-pop = InfPop(z=rand(Normal(0, √(2M.V)), N), c=fill(2,N))
-
-@btime generation(pop, M, z->1.0)
-
-# In an infinite population (i.e. no inbreeding) we should get to m*M.V
-# quickly, where m is the ploidy level, at least in the absence of double
-# reduction.
-for i=1:100
-    pop = generation(M, z->1, pop)
-    pop.F .= 0.0
-    pop.Φ .= 0.0
-    @info var(pop.z), 2M.V
-end
-
-function gen(pop, M)
-    pop = generation(pop, M)
-    pop.z .-= mean(pop.z)
-    pop.F .= 0.0
-    pop.Φ .= 0.0
-    return pop
-end
-    
-function cb(x)
-    _, pop = x
-    z2 = pop.z[pop.c .== 2]
-    z3 = pop.z[pop.c .== 3]
-    z4 = pop.z[pop.c .== 4]
-    z2, z3, z4
-end
-
-
-bs = [ones(3), [1.0, 0.75, 0.5], [1.0, √(2/3), √(1/2)]]
-ps = map(bs) do b
-    M = InfDemeMix(α=zeros(3), β=b)
-    N = 500
-    zinit = rand(Normal(0, √(2M.V)), N)
-    pop = InfPop(z=zinit, c=fill(2,N))
-    res = fiterate((pop,i)->gen(pop, M), pop, 1000, callback=cb)
-    zs = map(i->vcat(getindex.(res, i)...), 1:3) 
-    plot(title="\$\\beta = $(round.(b, digits=2))\$")
-    map(i->stephist!(zs[i], norm=true, bins=-4:0.3:4,
-        label=["diploid", "triploid", "tetraploid"][i]), 1:3)
-    map(zip(b,2:4)) do (bb, k)
-        plot!(Normal(0, bb*√(k*M.V)), color=k-1, 
-            label="",
-            fill=true, alpha=0.2, linealpha=0.2)
-    end
-    plot!()
-end
-
-plot(ps..., ylabel="density", legend=:topright, layout=(1,3), size=(700,200), xlabel="\$z\$", margin=3Plots.mm)
-
-M = InfDemeMix()
-N = 500
[…]

### 2025-04-22 notebooks/optimumshift.jl

@@ -1,49 +1,0 @@
-# Simulate a shifting optimum in a mixed-ploidy deme
-
-function simshift(M, N, Vs, θ, n1=100, n2=100)
-    pop = InfPop(z=rand(Normal(0, √(2M.V)), N), c=fill(2,N))
-    # equilibrate, assume the population is very large
-    ys = [(z=pop.z, c=counts(pop.c, 2:4))]
-    for _=1:n1
-        pop = generation(pop, M, z->exp(-γ*z^2))
-        pop.F .= 0.0
-        pop.Φ .= 0.0
-        push!(ys, (z=pop.z, c=counts(pop.c, 2:4)))
-    end
-    for _=1:n2
-        pop = generation(pop, M, z->exp(-(z - θ)^2/(2Vs)))
-        push!(ys, (z=pop.z, c=counts(pop.c, 2:4)))
-    end
-    return ys
-end
-
-function tetraploid_est(M, N, Vs, θ, n1=100, n2=100, nmax=100)
-    k = 0
-    while true
-        ys = simshift(M, N, Vs, θ, n1, n2)  
-        (ys[end].c[3] > ys[end].c[1]) && return (k+1, ys)
-        k > nmax && return (k, NaN)
-        k += 1
-    end
-end
-
-# There appears to be a very sharp threshold for establishment that is a
-# function of θ, Vz, Vs, β, u and v (I guess). This is probably something one
-# could try to predict. Note however that as θ/√Vs becomes large enough, we get
-# establishment even when the different ploidy levels are phenotypically
-# indistinguishable. There are only two differences between ploidy levels in
-# that case: the different rate of inbreeding and the different gametic
-# outputs.
-N = 200
-θ = 9
-M = InfDemeMix()
-Vs = 2*2M.V   # assume Vs >> Vz
-
-tw, xs2 = tetraploid_est(M, N, Vs, θ, 50, 20)
-
-xs1 = simshift(M, N, Vs, θ, 100, 100)
-
-[exp(-θ/(2Vs) + (1/(2Vs))^2/2*m*M.β[m-1]^2*M.V) for m=2:4]
-
-
-

### 2025-04-22 notebooks/ri-establishment.jl

@@ -1,197 +1,0 @@
-# Consider for the case of succesful establishment (without/with migration),
-# how relative fitness of migrants evolves
-using Plots
-using Parameters
-using Distributed
-using Serialization
-addprocs(5)
-@everywhere using InfGenetics
-@everywhere using StatsBase
-
-@everywhere const Sims = Vector{InfPop{Float64}}
-
-# collect n2 and n4 establishment replicates for both diploid and tetraploids
-@everywhere function collect_est(f, n2, n4)
-    x2 = Vector{Sims}(undef, n2)
-    x4 = Vector{Sims}(undef, n4)
-    c2 = c4 = 1
-    while c2 <= n2 || c4 <= n4
-        res = simuntil(f)
-        cs = counts(res[end][2].c, 2:4)
-        if cs[3] > cs[1] && c4 <= n4
-            x4[c4] = last.(res)
-            c4 += 1
-            @info c4, c2
-        elseif cs[1] > cs[3] && c2 <= n2
-            x2[c2] = last.(res)
-            c2 += 1
-            @info c4, c2
-        end
-    end
-    return x2, x4
-end
-
-function avg(xs)
-    map(i->mean(filter(
-        y->y>0, [x[i] for x in xs if length(x) >= i])), 
-        1:maximum(length, xs))
-end
-
-# initial migrant z=0
-M = InfDemeMixEst(U=Umat(0.00,0.00), γ=0.25, m=0.0, θ=[2.0,2.0,2.0], V=0.5)
-f2=()->simest(reconstruct(M, γ=0.25), x0=InfPop(z=zeros(1), c=Int64[2]))
-f4=()->simest(reconstruct(M, γ=0.25), x0=InfPop(z=zeros(1), c=Int64[4]))
-x2, _ = collect_est(f2, 1000, 0) 
-_, x4 = collect_est(f4, 0, 1000) 
-
-# initial migrant random
-M = InfDemeMixEst(U=Umat(0.00,0.00), γ=0.25, m=0.0, θ=[2.0,2.0,2.0], V=0.5)
-f2=()->simest(reconstruct(M, γ=0.25), x0=InfPop(z=randn(1), c=Int64[2]))
-f4=()->simest(reconstruct(M, γ=0.25), x0=InfPop(z=randn(1), c=Int64[4]))
-x2b, _ = collect_est(f2, 1000, 0) 
-_, x4b = collect_est(f4, 0, 1000) 
-
[…]

### 2025-04-22 notebooks/segvartests.jl

@@ -1,110 +1,0 @@
-using Random, StatsBase, Distributions
-
-function segtetraploid(X, δs, α, n)
-    @assert sum(δs) == 1
-    hom = sample(1:5, Weights(δs))
-    shuffle!(X)
-    if hom == 2
-        X[1] = X[2]
-    elseif hom == 3
-        X[1] = X[2]
-        X[3] = X[4]
-    elseif hom == 4
-        X[1] = X[2] = X[3]
-    elseif hom ==5
-        X[1] = X[2] = X[3] = X[4]
-    end
-    map(1:n) do _
-        if rand() < α 
-            # double reduction
-            2*rand(X)
-        else
-            X1, X2 = sample(X, 2, replace=false)
-            X1 + X2
-        end
-    end
-end
-
-δs = [0.6, 0.2, 0.1, 0.07, 0.03]
-F = δs[2]/6 + δs[3]/3 + δs[4]/2 + δs[5]
-Vx = 1
-α = 0.2
-V = map(1:10000) do _
-    # Var[Y|X]
-    var(segtetraploid(rand(Normal(0,√Vx), 4), δs, α, 1000))
-end |> mean  # E[Var[Y|X]]
-@info V, (1+2α)*(1-F)*Vx
-
-function segtriploid_diploid(X, δs, ξ, n)
-    @assert sum(δs) == 1
-    hom = sample(1:3, Weights(δs))
-    shuffle!(X)
-    if hom == 2
-        X[1] = X[2]
-    elseif hom == 3
-        X[1] = X[2] = X[3]
-    end
-    map(1:n) do _
-        if rand() < ξ
-            2*rand(X)
-        else
-            X1, X2 = sample(X, 2, replace=false)
-            X1 + X2
-        end
-    end
-end
-
-δs = [0.6, 0.25, 0.15]
-F = δs[2]/3 + δs[3]
-Vx = 1
-ξ = 0.4
-V = map(1:10000) do _
-    # Var[Y|X]
-    var(segtriploid_diploid(rand(Normal(0,√Vx), 3), δs, ξ, 1000))
-end |> mean  # E[Var[Y|X]]
-@info V, (2/3)*(1+3ξ)*(1-F)*Vx
-
-
-function segtriploid_haploid(X, δs, n)
-    @assert sum(δs) == 1
-    hom = sample(1:3, Weights(δs))
[…]

### 2025-04-22 notebooks/selfing.jl

@@ -1,130 +1,0 @@
-using InfGenetics, Plots, PlotThemes, StatsBase, Parameters; theme(:hokusai)
-using Distributions, ThreadTools
-
-function simest(M, minn=100, maxgen=Inf)
-    pops = [InfPop(z=zeros(0))]
-    t = 0
-    while true
-        t += 1
-        pop = generation(pops[end], M)
-        push!(pops, pop)
-        (length(pop.z) > minn || t > maxgen) && (return t, pops)
-    end
-end
-
-function simuntil(simfun, condition)
-    while true
-        x = simfun()
-        condition(x) && return x
-    end
-end
-
-function ρmat(a)
-    ρ = [1.0 a a; a 1.0 a; a a 1.0]
-    ρ ./= mean(ρ)
-end
-
-# As in Abu Awad
-# assume loss of self-incompatibility
-M = InfDemeMixEstSelf(σ=[0.0, 0.5, 0.5])
-t, pops = simest(M); @info t
-plot(permutedims(hcat(map(x->counts(x.c, 2:4), pops[end-30:end])...)))
-
-xs = map(1:1000) do _
-    t, pops = simest(M)
-    x = (t, counts(pops[end].c, 2:4))
-    @info x; x
-end
-
-filter(x->x[3]>x[1], last.(xs))
-
-nrep = 10000
-θs = -2.5:0.5:-1.5
-ss = 0:0.2:1
-res = map(θs) do θ
-    map(ss) do σ
-        @info σ, θ
-        M = InfDemeMixEstSelf(σ=[0.0, σ, σ], θ=fill(θ, 3))
-        map(1:nrep) do _
-            t, pops = simest(M)
-            counts(pops[end].c, 2:4)
-        end
-    end
-end
-
-y = map(res) do x
-    map(xs->length(filter(x->x[3]>x[1], xs))/nrep, x)
-end
-plot(ss, y, 
-    label=reshape(map(x->"\$\\theta=$x\$", -3:0.5:-1.5), 1, 4), 
-    marker=true, xlabel="\$\\sigma\$", ylabel="\$P\$", size=(320,260), 
-    legend=:topleft)
-
-function tetswin(pop)
-    xs = counts(pop.c, 2:4)
-    xs[3] > xs[1]
-end
-
-function extract(pops)
-    (N = map(x->length(x.z), pops),
[…]

### 2025-04-22 notebooks/traits.jl

@@ -1,86 +1,0 @@
-using InfGenetics, Distributions, Plots, StatsPlots, Parameters
-using InfGenetics: haploidgamete, diploidgamete
-plotsdefault()
-
-N = 1000
-
-xs = map(-2:1.0:2) do z̄
-    # set u=1, all offspring is tetraploid
-    M = InfDemeMix(U=Umat(0.0, 0.0))
-    pop = InfPop(z=rand(Normal(z̄, √(2M.V)), N), c=fill(2,N))
-    pop2 = generation(pop, M)
-    pop4 = generation(pop, reconstruct(M, U=Umat(1.0, 0.0)))
-    z̄, pop2.z, pop4.z
-end
-
-map(xs) do (z̄, z2, z4)
-    bins = -6:0.25:6
-    stephist(z2, fill=true, fillalpha=0.2, label="diploids", normalize=true, bins=bins)
-    stephist!(z4, fill=true, fillalpha=0.2, label="tetraploids", normalize=true, bins=bins)
-    vline!([z̄], label="", legend=:outertopright, size=(400,200)) 
-end |> x->plot(x..., size=(900,300))
-
-function familydist(zi, zj, ki, kj, ci, cj, Fi, Fj, M)
-    @unpack β = M
-    k   = ki + kj
-    yi  = (ki/ci)*(zi/β[ci-1])
-    yj  = (kj/cj)*(zj/β[cj-1])
-    yi, Vi = ki == 1 ? haploidgamete(M, zi, Fi, ci) : diploidgamete(M, zi, Fi, ci)
-    yj, Vj = kj == 1 ? haploidgamete(M, zj, Fj, cj) : diploidgamete(M, zj, Fj, cj)
-    z̄ij = (β[k-1]/β[ci-1])*yi + (β[k-1]/β[cj-1])*yj
-    Vij = β[k-1]^2*(Vi + Vj)
-    (z=z̄ij, V=Vij)
-end
-
-M = InfDemeMix(U=Umat(0.05, 0.05), α=[1/6, 1/6, 1/6])
-
-familydist(0., 0., 2, 2, 2, 2, 0.0, 0.0, M)
-
-# trait value of newly formed tetraploid, with diploid parents is Gaussian with
-# mean `z̄ = (2zi + 2zj)/β₄` and variance `β₄²(Vik + Vjl)`.
-
-# Source population (mean zero)
[…]

### 2025-04-22 notebooks/unreducedgametes.jl

@@ -1,43 +1,0 @@
-
-using Pkg; Pkg.activate("/home/arzwa/dev/InfGenetics/")
-using Distributed, StatsBase, Plots, Serialization, Distributions, Optim
-addprocs(8)
-@everywhere using InfGenetics
-
-# estimate parameter of a censored geometric distribution by ML
-function logpdf_censored_geometric(p, c, x)
-    x < c ? log(p) + (x-1)*log(1-p) : (c-1)*log(1-p)
-end
-
-function mle_censored_geometric(c, xs)
-    fun = p -> -sum(logpdf_censored_geometric.(Ref(p), Ref(c), xs))
-    optimize(fun, 0.0, 1.0).minimizer
-end
-
-us = 0.02:0.01:0.1
-m  = 0.75
-n  = 20
-γ  = 0.125
-z  = 1.5
-sims = pmap(us) do u
-    U = Umat(u, u)
-    M = InfDemeMixEst(m=m, θ=fill(z, 3), U=U, γ=γ)
-    ks = Int64[]
-    K  = 0
-    while length(ks) < n 
-        est, ngen, c = simest2(M)
-        K += 1
-        if c[end] > c[1]
-            push!(ks, K)
-            K = 0
-        end
-    end
-    @info u, ks
-    u, ks
-end
-
-map(sims) do (u, ks)
-    p = mle_censored_geometric(Inf, ks)
-    u, log(p/(1-p))
-end |> z->plot(z, ms=2, marker=true)
-    

### 2025-05-13 doc/paper/buildanddiff.sh
submission 3

@@ -1,3 +1,3 @@
 buildtex.sh paper.tex
-latexdiff submission1.tex paper.tex > diff.tex
+latexdiff submission2.tex paper.tex > diff.tex
 buildtex.sh diff.tex

### 2025-05-31 doc/paper/paper.tex
update

@@
     \affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
-    %\author{}
     \date{\vspace{-5ex}}
@@
         habitat \\ 
-    $w_{ij}$ & fitness of parental pair $(i,j)$ (expected fitness of
+    $w_{ij}$ & fitness of parental pair $(i,j)$ (expected zygotic fitness of
         offspring of pair $(i,j)$)\\
-    $w_{ij}^{kl}$ & expected fitness of offspring of parental pair $(i,j)$ when
+    $w_{ij}^{kl}$ & expected fitness of zygote from parental pair $(i,j)$ when
         $i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
+    $w_i$ & expected number of offspring from individual $i$ surviving to
+        reproductive age \\
     $\sigma_k$ & rate of self-fertilization in $k$-ploids \\
@@
 
-    In generation $t$, $M(t)$ migrant individuals arrive on an island (the new
-    habitat) joining $N^\ast(t)$ resident individuals, where $M(t)$ is Poisson
-    distributed with mean $m$, and $N^\ast(0)=0$ unless stated otherwise. 
+    In generation $t$, $M(t) \sim \mathrm{Poisson}(m)$ migrant individuals
+    arrive on an island (the new habitat) joining $N^\ast(t)$ resident
+    individuals.
+    We assume $N^\ast(0)=0$ unless stated otherwise. 
     The migrant individuals are assumed to be unrelated to the resident
     individuals.
-    After migration there are $N(t) = N^\ast(t) + M(t)$ individuals on the island
-    which reproduce sexually, and the offspring thus produced survives until the
-    next generation with a probability determined by their trait value.
-    Note that this corresponds to a life cycle where selection occurs before
-    migration within a generation.
-    In the basic model, random selfing is allowed (but see below for a model with
-    self-incompatibility).
-    We assume the trait is under directional selection, with fitness $w(z) =
-    e^{\gamma(z - \theta)}$, where $\gamma$ is the intensity of directional
-    selection and $\theta$ is the trait value for which the growth rate of the
-    island population becomes positive.
+    After migration there are $N(t) = N^\ast(t) + M(t)$ individuals on the
+    island.
+    These individuals reproduce sexually by random mating (allowing random
+    selfing, see below for modifications involving partial selfing and
+    assortative mating) to produce a large (effectively infinite) number of
+    zygotes, of which $N^\ast(t+1)$ survive to reproductive age with a
+    probability proportional to individual fitness, where $N^\ast(t+1)$ is
+    Poisson distributed with mean $N(t)\overline{W}$ and $\overline{W}$ is the
+    mean fitness among zygotes.
+    We assume the trait to be under directional selection, with the fitness of a
+    zygote with trait value $z$ given by $w(z) = e^{\gamma(z - \theta)}$, where
+    $\gamma$ is the intensity of directional selection and $\theta$ is the
+    trait value for which the growth rate of the island population becomes
+    positive.
     As the population becomes better adapted, eventually, some form of density
@@
     phase, defining establishment as reaching a population of size 100.
+    Note that this model corresponds to a life cycle where selection occurs
+    before migration within a generation.
 
     Again following \cite{barton2018}, we simulate the model by first calculating
-    the fitness of each parental pair $(i,j)$, which is the expected fitness of
-    offspring of this pair
+    the fitness of each parental pair $(i,j)$, which is defined as the expected
+    fitness of a random zygote derived from this pair
     \begin{equation}
@@
         = \sum_{k=1}^2\sum_{l=1}^2 u_{c_{i},k}u_{c_{j},l}
-            \Ex\left[e^{\gamma(Z_{ij}^{kl} - \theta)}\right]
+            \Ex\left[e^{\gamma(Z_{ij}^{kl} - \theta)}\right] \label{eq:wij}
     \end{equation}
-    The expectation on the right hand side can be calculated from \cref{eq:oneline}
-    using the moment-generating function of the Gaussian, i.e.
+    The expected number of offspring individuals of an individual $i$
+    surviving to reproductive age is then $w_i = \frac{1}{N(t)}\sum_j w_{ij}$.
+    The expectation on the right hand side of \cref{eq:wij} can be calculated
+    from \cref{eq:oneline} using the moment-generating function of the
+    Gaussian, i.e.
     \begin{equation}
@@
     Having calculated the $w_{ij}$, the number of offspring surviving into the next
-    generation is calculated as $N^\ast(t+1) = \sum_{i,j}w_{ij}/N(t)$.
-    Next, $N^\ast(t+1)$ offspring individuals are obtained by sampling parental
-    pairs and gametes proportional to $w_{ij}^{kl}$, and sampling a trait value
-    from a Gaussian distribution with mean $\overline{z_{ij}^{kl}} + \gamma
-    V_{ij}^{kl}$ and variance $V_{ij}^{kl}$.
+    generation $N^\ast(t+1)$ is sampled from a Poisson distribution with mean
+    $N(t)\overline{W} = \sum_i w_i = \sum_{i,j}w_{ij}/N(t)$.
+    Offspring cytotypes and trait values are then determined by sampling
+    parental pairs and gametes proportional to $w_{ij}^{kl}$, and sampling a
+    trait value from the offspring trait value distribution conditional on
+    survival to reproductive age, which is Gaussian with mean
+    $\overline{z_{ij}^{kl}} + \gamma V_{ij}^{kl}$ and variance $V_{ij}^{kl}$ .
 
@@
     \begin{equation}
-    \Ex[w_i] = \sigma_{c_i} w_{ii} +
+    w_i = \sigma_{c_i} w_{ii} +
       (1-\sigma_{c_i})\left[\frac{1}{N}\sum_{j=1}^N w_{ij}\right]
@@
     \begin{equation}
-    \Ex[w_i] = \begin{cases}
+    w_i = \begin{cases}
         \frac{1}{N-1}\sum_{j \ne i }w_{ij} & \text{if } N > 1\\ 
@@
     \begin{equation}
-    \Ex[w_i] = \rho_{c_i} \frac{1}{N_{c_{i}}} \sum_{j=1}^N \delta_{c_i,c_j}w_{ij}
+    w_i = \rho_{c_i} \frac{1}{N_{c_{i}}} \sum_{j=1}^N \delta_{c_i,c_j}w_{ij}
      + (1-\rho_{c_i})
@@
     implemented in Julia \citep{julia}.
-    %Documented code and simulation notebooks are available at \url{[anonymized]}.
-    Documented code and simulation notebooks are available at
-    \url{https://github.com/arzwa/InfGenetics}.
+    Documented code and simulation notebooks are available at \url{https://github.com/arzwa/InfGenetics}.
 
@@
 eco-evolutionary questions that arise in the study of mixed-ploidy populations.
-For instance, it could be of interest to develop a complicated individual-based
-model along the lines of \cite{oswald2011} to study the effects of selfing,
-assortative mating and competition on establishment and coexistence of
-tetraploids within diploid populations, but where the focal trait that
-determines fitness and assortative mating is not controlled by a few
+For instance, it could be of interest to develop an individual-based
+model along the lines of \cite{oswald2011} to study competitive exclusion
+versus coexistence of cytotypes in mixed-ploidy populations with partial
+selfing, assortative mating, competition, \textit{etc.}, but where the focal
+trait that determines the various fitness components is not controlled by a few
 large-effect loci (as in \cite{oswald2011}), but many loci of small effect.

### 2025-05-31 doc/paper/supp.tex
update

@@
 \centering
-\includegraphics[width=\textwidth]{/home/arzwa/dev/InfGenetics/doc/img/finsvsinf-polyploids-validation.pdf}
+\includegraphics[width=\textwidth]{/home/arzwa/dev/InfGenetics/doc/supp_img/finsvsinf-polyploids-validation.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=\textwidth]{/home/arzwa/vimwiki/build/img/2023-10-23/mixedploidy-Vz.pdf}
+\includegraphics[width=\textwidth]{/home/arzwa/dev/InfGenetics/doc/supp_img/mixedploidy-Vz.pdf}
 \caption{Comparison of the mixed-ploidy infinitesimal model with the
@@
 \centering
-\includegraphics[width=\linewidth]{/home/arzwa/vimwiki/build/img/2023-10-23/mixedploidy-F.pdf}
+\includegraphics[width=\linewidth]{/home/arzwa/dev/InfGenetics/doc/supp_img/mixedploidy-F.pdf}
 \caption{Average inbreeding coefficient \(\bar{F}\) in each cytotype in
@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration-dr-check.pdf}
+\includegraphics[width=0.9\linewidth]{/home/arzwa/dev/InfGenetics/doc/supp_img/estwmigration-dr-check.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=\linewidth]{/home/arzwa/vimwiki/build/img/2023-10-13/g1diff.pdf}
+\includegraphics[width=\linewidth]{/home/arzwa/dev/InfGenetics/doc/supp_img/g1diff.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=0.5\textwidth]{/home/arzwa/vimwiki/build/img/2023-10-22/timetodiploidancestord.pdf}
+\includegraphics[width=0.5\textwidth]{/home/arzwa/dev/InfGenetics/doc/supp_img/timetodiploidancestord.pdf}
 \caption{Expected time to diploid ancestry. The solid lines show
@@
 \centering
-\includegraphics[width=\textwidth]{/home/arzwa/vimwiki/build/img/2023-10-13/mixedploidy-F-Ne.pdf}
+\includegraphics[width=\textwidth]{/home/arzwa/dev/InfGenetics/doc/supp_img/mixedploidy-F-Ne.pdf}
 \caption{The evolution of \(\bar{F}\) in the mixed-ploidy population and
@@
 \centering
-\includegraphics[width=\textwidth]{/home/arzwa/dev/InfGenetics/doc/img/mixeq.pdf}
+\includegraphics[width=\textwidth]{/home/arzwa/dev/InfGenetics/doc/supp_img/mixeq.pdf}
 \caption{

### 2025-05-31 notebooks/fitness.jl

@@ -1,0 +1,62 @@
+using Distributions, Plots, StatsPlots, StatsBase
+
+V = 1.0   # segregation variance
+N = 100   # population size
+γ = 0.25  # selection strength
+θ = 2.0   # maladaptation
+
+# Current population trait values
+z = rand(Normal(0, √2V), N)
+
+# Number of replicate offspring populations to simulate
+nrep = 2500
+
+# Number of individuals in next generation is NW̄
+sims1 = map(1:nrep) do _
+    Noff = 10000
+    off = map(1:Noff) do _
+        z1 = rand(z)
+        z2 = rand(z)
+        zoff = rand(Normal((z1 + z2)/2, √V))
+        woff = exp(γ*(zoff - θ))
+        woff, zoff
+    end
+    w̄ = mean(first.(off))
+    𝔼N = N*w̄
+    N′ = rand(Poisson(𝔼N))
+    z′ = sample(last.(off), Weights(first.(off)), N′, replace=false)
+end
+
+# Calculate expected contribution of each family, sample families and sample
+# offspring trait values from the Gaussian after selection within each family
+sims2 = map(1:nrep) do _
+    W = [exp(γ*((z[i]+z[j])/2 - θ) + γ^2/2 * V) for i=1:N, j=1:N]
+    𝔼N = sum(W)/N
+    N′ = rand(Poisson(𝔼N))
+    C = CartesianIndices((1:N, 1:N))
+    idx = sample(C, Weights(vec(W)), N′)
+    z′ = map(idx) do ix
+        i, j = Tuple(ix)
+        z̄ = (z[i] + z[j])/2 + γ*V
+        rand(Normal(z̄, √V))
+    end
+end
+
+P1 = density(vcat(sims1...), label="sims1", xlabel="\$z'\$", title="(A)")
+density!(vcat(sims2...),label="sims2")
+z1 = map(mean, sims1)
+z2 = map(mean, sims2)
+P2 = density([z1 z2], xlabel="\$\\overline{z}'\$", legend=false, title="(B)") 
+v1 = map(std, sims1)
+v2 = map(std, sims2)
+P3 = density([v1 v2], xlabel="\$\\sqrt{V′}\$", legend=false, title="(C)")
+P4 = stephist(map(length, sims1), xlabel="\$N'\$", legend=false, title="(D)")
+stephist!(map(length, sims2))
+plot(P1,P2,P3,P4,size=(900,200),bottom_margin=5Plots.mm,layout=(1,4))
+
+
[…]

### 2025-05-31 src/_infdeme.jl

@@ -1,70 +1,0 @@
-
-"""
-    InfDeme
-
-A finite random mating population consisting of a single cytotype with
-individuals characterized by a one-dimensional trait obeying infinitesimal
-quantitative genetics. Defined for haploid, diploid and tetraploid cytotypes.
-"""
-struct InfDeme{K,T}  # K is ploidy level
-    z::Vector{T}  # trait value (scalar, at least for now)
-    F::Vector{T}  # inbreeding coefficients
-    Φ::Matrix{T}  # coancestry coefficients
-    V::T          # segregation variance
-    α::T          # probability of double reduction
-end
-
-# constructor
-InfDeme(z::Vector{T};
-        k = 1,
-        V = 1.,
-        F = zeros(length(z)),
-        Φ = zeros(length(z),length(z)),
-        α = 0. 
-       ) where T = 
-    InfDeme{Val{k},T}(z, F, Φ, V, α)
-
-(d::InfDeme{K,T})(z, F, Φ; V=d.V, α=d.α) where {K,T} = 
-    InfDeme{K,T}(z, F, Φ, V, α)
-
-popsize(d::InfDeme) = length(d.z)
-ploidy(d::InfDeme{K}) where K = K.parameters[1] 
-
-function wfgen(d::InfDeme)
-    z = similar(d.z)
-    N = popsize(d)
-    P = zeros(N,N) 
-    F = zeros(length(z))
-    for k=1:N
-        i = rand(1:N)
-        j = rand(1:N)
-        P[k,i] += 0.5
-        P[k,j] += 0.5
-        z[k] = randoff(d, i, j)
-        F[k] = Fcoeff(d, i, j) 
-    end
-    Φ = P * d.Φ * P'
-    setdiagΦ!(d, Φ, F)
-    return d(z, F, Φ)
-end
-
-function randoff(d, i, j)
-    Cij = segvar(d, i, j)
-    zmp = (d.z[i] + d.z[j]) / 2
-    return rand(Normal(zmp, √Cij))
-end
-
-Fcoeff(d::InfDeme{Val{1}}, i, j) = 1.
[…]

### 2025-05-31 src/_infdememix.jl

@@ -1,267 +1,0 @@
-# Think of an environmental context (natural selection, density regulation, ...)
-# we might explicitly split this up in genetic/ecological component
-abstract type PopulationContext end
-
-# ~ Barton & Etheridge
-struct DirectionalSelection{T} <: PopulationContext
-    θ::T  # selection gradient
-    r::T
-    K::T
-end
-
-# ~ Polechova & Barton
-@with_kw struct StabilizingSelection{T} <: PopulationContext
-    θ::T    # phenotypic optimum
-    ω::T    # variance of stabilizing selection
-    r::T = 1.0   # growth rate
-    K::T = Inf   # carrying capacity (can be Inf when no density dependence)
-    ν::T = 0.0   # migration rate
-    z̄::T = NaN   # mean of migrant
-    V::T = NaN   # variance of migrant
-end
-
-(c::StabilizingSelection)(; kwargs...) = reconstruct(c; kwargs...) 
-
-@with_kw struct InfDemeMix{T}
-    z::Vector{T}   # trait values (scalars, for now)
-    F::Vector{T}   = zeros(length(z)) # inbreeding
-    Φ::Matrix{T}   = diagm(fill(0.5, length(z))) # coancestry
-    m::Vector{Int} = fill(2, length(z)) # ploidy levels
-    U::Matrix{T}   = [0.95 0.05 ; 0.0 0.0 ; 0.0 0.95]  # cytotype × gamete 
-    w::Vector{T}   = [1., 1., 1.]  # cytotype specific viability 
-    β::Vector{T}   = [1., 0.75, 0.5]     # variance scaler, will be in [0,1]
-    γ::Vector{T}   = [1., √(2*β[2]/3), √(β[3]/2)]  # trait scalers
-    V::T = 0.5     # half the diploid segvar
-    α::T = 0.0     # double reduction rate
-    ξ::T = 0.5     # probability of an IBD unreduced gamete
-end
-# U is a matrix with for each cytotype (rows) the probability of generating a
-# particular (euploid) gamete.
-# Note that `U` can be seen as capturing a fertility component of fitness,
-# whereas `w` captures a viability component of fitness
-# Cytotype-specific assortativity is one additional thing we may be interested
[…]

### 2025-05-31 src/_infmixest.jl

@@ -1,84 +1,0 @@
-# Establishment model (migration + directional selection)
-# Note that while it is straightforward to include density regulation, this
-# does not make much sense with directional selection. So density regulation
-# should go hand in hand with stabilizing selection. 
-@with_kw struct InfDemeMixEst{T}
-    U::Matrix{T} = [0.95 0.05 ; 0.05 0.05 ; 0.0 0.95]  # cytotype × gamete 
-    w::Vector{T} = ones(3)              # cytotype specific viability 
-    β::Vector{T} = [1, √(2/3), √(1/2)]  # variance scaler, will be in [0,1]
-    α::Vector{T} = zeros(3)             # double reduction rate
-    γ::T = 0.25                         # selection gradient
-    m::T = 1.0                          # 𝔼 number of migrants/gen                          
-    V::T = 0.5                          # Vx (haplotype variance in diploids)
-    θ::Vector{T} = fill(-3.0, 3)        # trait mean of migrants
-    c::Vector{T} = cytotype_equilibrium(U)
-end
-
-function cytotype_equilibrium(U)
-    u21, u22 = U[1,:]
-    u31, u32 = U[2,:]
-      _, u42 = U[3,:]
-    g1 = (u21 - 4*u31 - 2*u32 + 2*u42)/(
-        2*(u21 + u22 - 2*u31 - 2*u32 + u42)) - sqrt(
-            u21^2 - 4*u21*u32 + 8*u22*u31 - 4*u22*u42 + 4*u32^2)/(
-                2*(-u21 - u22 + 2*u31 + 2*u32 - u42))
-    [g1^2, 2g1*(1-g1), (1-g1)^2] 
-end
-
-function generation(pop::InfPop{T}, M::InfDemeMixEst) where T
-    @unpack θ, α, β, γ, w, U, m, V = M
-    # Poisson number of migrants
-    nr = length(pop.z)
-    nm = rand(Poisson(m))
-    cm = rand(Multinomial(nm, M.c))
-    cm = vcat([fill(i+1, k) for (i, k) in enumerate(cm)]...)
-    zm = [rand(Normal(θ[m-1], β[m-1]*√V)) for m in cm]
-    z  = [pop.z ; zm]
[…]

### 2025-05-31 src/_infmixestpair.jl

@@ -1,152 +1,0 @@
-# Establishment model (migration + directional selection)
-# With fitness determined for pairs of individuals, this allows to better
-# incorporate selfing and assortative mating... (different sexes requires a bit
-# more work...)
-# XXX Should study more carefully how assortative mating/selfing are usually
-# modelled.
-@with_kw struct InfDemeMixEstPair{T}
-    U::Matrix{T} = [0.95 0.05 ; 0.05 0.05 ; 0.0 0.95]  # cytotype × gamete 
-    w::Vector{T} = ones(3)              # cytotype specific viability 
-    β::Vector{T} = [1, √(2/3), √(1/2)]  # variance scaler, will be in [0,1]
-    α::Vector{T} = zeros(3)             # double reduction rate
-    σ::Vector{T} = zeros(3)             # selfing rate by ploidy
-    ρ::Matrix{T} = ones(3,3)            # assortativity by ploidy
-    γ::T = 0.25                         # selection gradient
-    m::T = 1.0                          # 𝔼 number of migrants/gen                          
-    V::T = 0.5                          # Vx (haplotype variance in diploids)
-    θ::Vector{T} = fill(-3.0, 3)        # trait mean of migrants
-    c::Vector{T} = cytotype_equilibrium(U)
-end
-
-# But how exactly do we do this?
-# 1. We let the number of offspring in the next generation be Poisson(NW̄) where
-#    W̄ is the viability component. But note this is not what we've done
-#    earlier, we also included the fertility component in the offspring
-#    number (i.e. `uk1 + uk2`)
-# 2. We take W̄ = E[wᵢⱼ] = ∑∑wᵢⱼ/N², hence NW̄ = ∑∑wᵢⱼ/N, where wᵢⱼ includes both
-#    fertility, viability of offspring/parents, ...
-# Note, B&E use the expected viability fitness of the offspring, but one could
-# also take the viability of the parents? That is somewhat easier, as we don't
-# need to consider the ploidy levels of the offspring etc.
-
[…]

### 2025-05-31 src/_infmixestself.jl

@@ -1,96 +1,0 @@
-# Establishment model (migration + directional selection)
-@with_kw struct InfDemeMixEstSelf{T}
-    U::Matrix{T} = [0.95 0.05 ; 0.05 0.05 ; 0.0 0.95]  # cytotype × gamete 
-    w::Vector{T} = ones(3)              # cytotype specific viability 
-    β::Vector{T} = [1, √(2/3), √(1/2)]  # variance scaler, will be in [0,1]
-    α::Vector{T} = zeros(3)             # double reduction rate
-    σ::Vector{T} = zeros(3)             # selfing rates
-    γ::T = 0.25                         # selection gradient
-    m::T = 1.0                          # 𝔼 number of migrants/gen                          
-    V::T = 0.5                          # Vx (haplotype variance in diploids)
-    θ::Vector{T} = fill(-3.0, 3)        # trait mean of migrants
-    c::Vector{T} = cytotype_equilibrium(U)
-end
-
-function fitnesses(z, c, M::InfDemeMixEstSelf)
-    @unpack γ, U, w = M
-    return [(U[k-1,1] + U[k-1,2])*w[k-1]*exp(γ*zk) for (zk,k) in zip(z,c)]  
-end
-
-# As in Abu Awad et al. 2014
-function selfingp(w, c, σ)
-    N = length(w)
-    x = [(1-σ[ci-1])*wi for (wi, ci) in zip(w, c)]
-    Z = sum(x)
-    [σ[ci-1]*wi/(σ[ci-1]*wi + (Z - xi)/(N-1)) for (wi, xi, ci) in zip(w, x, c)]
-end
-
-function samplediff(ws, i)
-end
-
-function generation(pop::InfPop{T}, M::InfDemeMixEstSelf) where T
-    @unpack θ, α, β, m, V = M
-    # Poisson number of migrants
-    nr = length(pop.z)
-    nm = rand(Poisson(m))
-    cm = rand(Multinomial(nm, M.c))
-    cm = vcat([fill(i+1, k) for (i, k) in enumerate(cm)]...)
-    zm = [rand(Normal(θ[m-1], β[m-1]*√V)) for m in cm]
-    z  = [pop.z ; zm]
-    c  = [pop.c ; cm]
-    F  = [pop.F ; zeros(nm)]
-    Φ  = [pop.Φ zeros(nr, nm); zeros(nm, nr) zeros(nm, nm)]
-    # fitness, this is the contribution to the gamete pool
[…]

### 2025-05-31 src/_infmixhs.jl

@@ -1,69 +1,0 @@
-# Deme with hard selection, stabilizing selection, and density regulation
-@with_kw struct InfDemeHS{T}
-    U::Matrix{T} = [0.95 0.05 ; 0.05 0.05 ; 0.0 0.95]  # cytotype × gamete 
-    w::Vector{T} = ones(3)              # cytotype specific Malthusian growth rate
-    β::Vector{T} = [1, √(2/3), √(1/2)]  # variance scaler, will be in [0,1]
-    α::Vector{T} = zeros(3)             # double reduction rate
-    θ::Vector{T} = zeros(3)             # optimum
-    γ::T = 0.25                         # selection gradient
-    V::T = 0.5                          # Vx (haplotype variance in diploids)
-    K::T = 200.                         # carrying capacity
-    c::Vector{T} = cytotype_equilibrium(U)
-end
-
-function fitnesses(z, c, M::InfDemeHS)
-    @unpack w, U, θ, γ, K = M
-    N = length(z)
-    return [(U[k-1,1] + U[k-1,2])*exp(w[k-1]*(1 - N/K) - γ*(zk - θ[k-1])^2) 
-        for (zk,k) in zip(z,c)]  
-end
-
-function generation(pop::InfPop{T}, M::InfDemeHS) where T
-    @unpack θ, α, β, γ, w, U, K, V = M
-    @unpack z, F, Φ, c = pop
-    N = length(z)
-    # fitness, this is the contribution to the gamete pool
-    ws = fitnesses(z, c, M)
-    w̄  = length(ws) == 0 ? 0.0 : mean(ws)
-    # number of individuals in the next generation
-    N_ = rand(Poisson(w̄*length(z)))   
-    # 2N gametes
-    ix = sample(1:length(ws), Weights(ws), 2N_, replace=true) 
-    P  = zeros(N_,N)
-    pop_ = InfPop(z=zeros(N_), F=zeros(N_), Φ=zeros(N_,N_), c=zeros(Int, N_))
-    for k=1:N_
-        i = ix[k]; j = ix[N_+k]
-        zi, Fi, ci = z[i], F[i], c[i]
-        zj, Fj, cj = z[j], F[j], c[j]
-        gi, yi, Vi = gamete(M, zi, Fi, ci)
[…]

### 2025-05-31 src/_pedigree.jl

@@ -1,51 +1,0 @@
-# Pedigree data structure
-# - each individual has two parents (left, right)
-# - we need to be able to prune the pedigree efficiently (remove all irrelevant
-#   parts based on who is still alive)
-# - we need to be able to efficiently find the generation in which two
-#   individuals share an ancestor
-# - we might wish to include ploidy level rightaway, because in the mixed
-#   ploidy case, a polyploid individual may inherit asymmetrically from its
-#   parents (however, this may be a parametric type thing)
-# - do we want a recursive data structure? or more a la Graphs.jl?
-
-struct PedigreeNode{T} 
-    this::T
-    left::Union{PedigreeNode{T},Nothing}
-    rght::Union{PedigreeNode{T},Nothing}
-end
-
-Base.show(io::IO, n::PedigreeNode) = write(io, "PedigreeNode($(n.this))")
-
-istoplevel(n::PedigreeNode) = isnothing(n.left) && isnothing(n.rght)
-
-# monoidal
-wfgeneration!(pop) = [mate(rand(pop), rand(pop)) for i=1:length(pop)]
-
-function mate(a, b)
-    k1 = length(a.this)
-    k2 = length(b.this)
-    y1 = sample(a.this, k1÷2, replace=false)
-    y2 = sample(b.this, k2÷2, replace=false)
-    PedigreeNode([y1 ; y2], a, b)
-end
-
-homozygosity(n::PedigreeNode) = homozygosity(n.this)
-homozygosity(x::Vector) = sort(collect(values(countmap(hash.(x)))))
-
-function homozygosity_tetpop(pop)
-    hs = Dict([1,1,1,1]=>0., [1,1,2]=>0., [2,2]=>0., [1,3]=>0., [4]=>0.)
-    for (k,v) in proportionmap(map(homozygosity, pop))
-        hs[k] += v
-    end
-    [hs[[4]], hs[[1,3]], hs[[2,2]], hs[[1,1,2]], hs[[1,1,1,1]]]
-end
-
-function simulate(N, n)
-    pop = [PedigreeNode(randn(4), nothing, nothing) for i=1:N]
-    hs = map(1:n) do _
-        pop = wfgeneration!(pop) 
-        homozygosity_tetpop(pop)
-    end
-    return pop, permutedims(hcat(hs...))
-end

### 2025-05-31 src/_rvmatrix.jl

@@ -1,38 +1,0 @@
-
-function cytotype_equilibrium(U)
-    u21, u22 = U[1,:]
-    u31, u32 = U[2,:]
-      _, u42 = U[3,:]
-    g1 = (u21 - 4*u31 - 2*u32 + 2*u42)/(
-        2*(u21 + u22 - 2*u31 - 2*u32 + u42)) - sqrt(
-            u21^2 - 4*u21*u32 + 8*u22*u31 - 4*u22*u42 + 4*u32^2)/(
-                2*(-u21 - u22 + 2*u31 + 2*u32 - u42))
-    [g1^2, 2g1*(1-g1), (1-g1)^2] 
-end
-
-function rvmatrix(U)
-    p = cytotype_equilibrium(U)
-    Z1 = sum(U[:,1] .* p)
-    Z2 = sum(U[:,2] .* p)
-    M = [
-        p[1]*U[1,1]/Z1                          p[2]*U[2,1]/Z1                             0.0              ;
-        p[1]*U[1,1]/(3Z1) + 2p[1]*U[1,2]/(3Z2)  p[2]*U[2,1]/(3Z1) + 2p[2]*U[2,2]/(3Z2)   2*p[3]*U[3,2]/(3Z2);
-        p[1]*U[1,2]/Z2                          p[2]*U[2,2]/Z2                             p[3]*U[3,2]/Z2
-    ]
-    return M
-end
-
-function rvs(U)
-    M = rvmatrix(U)
-    _, evecs = eigen(M')
-    v = evecs[:,3]
-    v ./ sum(v)
-end
-
-function Nerv(U, N)
-    alpha = rvs(U)
-    pis = cytotype_equilibrium(U)
-    Nis = N .* pis
-    1 / sum((alpha .^ 2) ./ Nis)
-end
-

### 2025-05-31 src/_sandbox.jl

@@ -1,118 +1,0 @@
-
-# additive genetic variance in HWLE
-function addvarhw(k, a, p)
-    Ex  = [i*a*binomial(k,i)*p^i*(1-p)^(k-i) for i=0:k]
-    Ex² = [(i*a)^2*binomial(k,i)*p^i*(1-p)^(k-i) for i=0:k]
-    sum(Ex²) - sum(Ex)^2, k*a^2*p*(1-p)
-end
-
-# segregation variance in autotetraploids
-# there are four relevant inbreeding coefficients (5 states of homozygosity)
-
-# Here I estimate Eₓ[Var[Y|X]] by Monte Carlo
-using Random, StatsBase
-for genotype=[1,2,3,4], α=[0.,0.1,0.2]
-    vs = map(1:10000) do _
-        x = randn(4) 
-        (genotype == 2 || genotype == 3) && (x[1] = x[2])
-        genotype == 3 && (x[3] = x[4])
-        genotype == 4 && (x[1] = x[2] = x[3])
-        genotype == 5 && (x[1] = x[2] = x[3] = x[4])
-        ys = map(1:1000) do _
-            if rand() < α
-                2*rand(x)
-            else
-                y = shuffle(x)[1:2]
-                y[1] + y[2]
-            end
-        end
-        var(ys)
-    end
-    r = [0., 1/6, 1/3, 1/2, 1.][genotype]
-    pred = (1-r)*(1+2α)
-    @info (genotype, α) (mean(vs), pred)
-end
-
-using Distributions
-v = 2.5
-map(1:10000) do _
-    x = rand(Normal(0,√v), 4)
-    var([rand(x) for i=1:10000])
-end |> mean
-3/4 * v
-
-
-F(ϕ) = ϕ[2]/6 + ϕ[3]/3 + ϕ[4]/2 + ϕ[5]
-Φ(ϕ) = first([1 5/6 2/3 1/2 0] * ϕ)
-
-# for a given probability distribution over homozygosity states, simulate a
-# random set of unlinked loci, and generate gametes by Mendelian segregation.
-function manyloci(ϕ, nloci, noff, nrep)
-    vs = map(1:nrep) do rep1
-        # draw the homozygosity states for each locus
-        genotypes = map(i->sample(1:5, Weights(ϕ)), 1:nloci)
-        # draw the allelic effects, v0/2 = 1
[…]

### 2025-05-31 src/_unlinkedloci.jl

@@ -1,126 +1,0 @@
-# Simulation methods for a genome of a large but finite number of additive
-# unlinked loci with small effect under Mendelian inheritance.
-#
-# Consider the model which is also discussed in Barton et al. in an
-# example there where we have n loci, each of effect ± 1/√n.
-# generalize to k-ploids -> ± 1/√(kn)
-
-const Genome{T} = Matrix{T}
-const Population{T} = Vector{Genome{T}}
-
-ploidy(g::Genome) = size(g)[2]
-
-# ϕ denotes a vector of homozygosity states in tetraploids
-# ϕ = [abcd, aabc, aabb, aaab, aaaa] in tets
-
-# compute inbreeding coefficient from homozygosity coefficients
-F(ϕ) = ϕ[2]/6 + ϕ[3]/3 + ϕ[4]/2 + ϕ[5]
-
-# compute the variance reduction factor from homozygosity coefficients
-Φ(ϕ) = first([1 5/6 2/3 1/2 0] * ϕ)  # this is 1 - F(ϕ)...
-
-# generate a random genome with a given probability distribution `ϕ` over
-# homozygosity states. `V` is the phenotypic variance of the non-inbred base
-# population, i.e. twice the segregation variance.
-function randgenome(nloci, ploidy, ϕ, V)
-    genotypes = map(i->sample(1:length(ϕ), Weights(ϕ)), 1:nloci)
-    xs = if ploidy == 4
-        mapreduce(i->randlocustet(genotypes[i], V), hcat, 1:nloci)
-    elseif ploidy == 2
-        mapreduce(i->randlocusdip(genotypes[i], V), hcat, 1:nloci)
-    elseif ploidy == 3
-        mapreduce(i->randlocustrip(genotypes[i], V), hcat, 1:nloci)
-    else
-        @error "should implement"
-    end
-    return permutedims(xs) ./ √(nloci*ploidy)
-end
-
-function randlocustet(genotype, V)
-    x = rand(Normal(0, √V), 4)
-    (genotype == 2 || genotype == 3) && (x[1] = x[2])
-    genotype == 3 && (x[3] = x[4])
-    genotype == 4 && (x[1] = x[2] = x[3])
-    genotype == 5 && (x[1] = x[2] = x[3] = x[4])
-    return x
-end
-
-function randlocusdip(genotype, V)
[…]

### 2025-05-31 test/infdememix.jl

@@ -1,94 +1,0 @@
-using Pkg; Pkg.activate(@__DIR__); using InfGenetics
-using InfGenetics: InfDemeMix, offspring_distribution, 𝔼fitness, generation
-using InfGenetics: StabilizingSelection, selection
-using Distributions, StatsBase, LaTeXStrings, Plots, PlotThemes
-theme(:wong2, titlefont=10, title_loc=:left)
-
-# plot a series of populations
-plotpops(pops) = 
-    plot(
-         plot(map(x->mean(x[2].z), pops), title=L"\bar{z}"),
-         plot(map(x->var(x[2].z), pops), title=L"V_z"),
-         plot(permutedims(mapreduce(x->counts(x[2].m, 2:4), hcat, pops)), title=L"N"),
-         plot(map(x->mean(x[2].F), pops), title=L"\bar{F}", ylim=(0,1)),
-         legend=false,
-        )
-
-# simulates an adaptive challenge. We should have a branching process
-# approximation to get a rough probability of establishment. Then check how
-# often tets/dips do establish?
-d = InfDemeMix(z = randn(100), 
-               m = fill(2, 100),
-               U = [0.9 0.1 ; 0.05 0.05; 0.0 0.95], 
-               w = [1.0, 1.0, 1.0],
-               β = [1.0, 1.5, 2.0], 
-               α = 0.2, 
-               ξ = 0.5)
-
-context = StabilizingSelection(θ=0., ω=1.0, r=1.1, K=100., ν=2., z̄=-5., V=4d.V)
-
-# comparing dips/tets
-pops2 = fiterate((d,_)->InfGenetics.generation(d, context), d(U=[1.0 0.;0. 0.; 0. 0.]), 500)
-pops4 = fiterate((d,_)->InfGenetics.generation(d, context), d(m=fill(4,100)), 500)
-ps = plot(map(plotpops, [pops2, pops4])..., size=(800,300))
-
-context = StabilizingSelection(θ=0., ω=1.0, r=1.1, K=500., ν=2., z̄=0., V=2d.V)
-sims = map(1:100) do i
-    pops = fiterate((d,i)->InfGenetics.generation(d, context(θ=i>50 ? 3. : 0.)), d, 100)
-    ms = counts(pops[end][2].m, 1:4)
-    @info ms
-    (argmax(ms), pops)
-end;
-
[…]

### 2025-05-31 test/selfing.jl

@@ -1,7 +1,0 @@
-
-M = InfDemeMixEst(m=1.0, γ=0.0, σ=ones(3), θ=fill(0.0, 3), U=Umat(0.0,0.0))
-x = InfPop(z=zeros(4))
-W, Z, V = InfGenetics.families(M, x)
-InfGenetics.selfing!(W, fill(Inf, 3), x.c)
-sum(W)
-

### 2025-05-31 test/tests.jl

@@ -1,83 +1,0 @@
-using Pkg; Pkg.activate(@__DIR__)
-
-# some checks of infinitesimal predictions against an actual Mendelian system
-using Random, StatsBase, Distributions
-include("src/unlinkedloci.jl")
-
-ϕ = [1., 0., 0., 0., 0.]
-V = 1.2
-v2 = map(1:1000) do _
-    G = randgenome(100, 2, ϕ, V)
-    var(map(_->sum(gamete(G, 0.)), 1:200))
-end |> mean
-v4 = map(1:1000) do _
-    G = randgenome(100, 4, ϕ, V)
-    var(map(_->sum(gamete(G, 0.)), 1:200))
-end |> mean
-v2, V/2/2, v4, V/2/2
-
-# check whether the segregation variance matches predictions
-ϕ = [0.1, 0.54, 0.07, 0.25, 0.04]
-α = 0.3
-V = 3  # phenotypic var, segvar=V/2, gameticvar = V/4
-vs = map(1:1000) do _
-    G = randgenome(100, 4, ϕ, V)
-    var(map(_->sum(gamete(G, α)), 1:100))
-end
-mean(vs), Φ(ϕ) * (V/4) * (1+2α), (1 - F(ϕ)) * (V/4) * (1+2α)
-
-# the segregation variance for unreduced gametes
-ξ = 0.3
-V = 1  # segvar V0 = V/2, gameticvar = V0/2 = V/4
-ϕ = [0.8, 0.2]
-vs = map(1:1000) do _
-    G = randgenome(100, 2, ϕ, V)
-    v1 = var(map(_->sum(ugamete(G, ξ)), 1:100))
-    v2 = var(map(_->sum( gamete(G, 0)), 1:100))
-    v1, v2
-end
-mean(first.(vs)), mean(last.(vs)), 4*(V/4)*ϕ[1]*ξ, ϕ[1]*V/4
-
-# triploids
-ξ = 0.3
-V = 1.2  # segvar V0 = V/2, gameticvar = V/3 
-ϕ = [.5, 0.3, 0.2]
-f = ϕ[3] + ϕ[2]/3
-vs = map(1:1000) do _
-    G = randgenome(100, 3, ϕ, V)
-    v1 = var(map(_->sum(tripdipgamete(G, ξ)), 1:100))
-    v2 = var(map(_->sum(triphapgamete(G)), 1:100))
-    v1, v2
-end
-vs2 = first.(vs)
-vs1 = last.(vs)
[…]
